### 2015-01-08 inst/doc/sparseHessianFD.Rmd
it works, with lots of new documentation

---
title:  Using sparseHessianFD
author:  Michael Braun
date:  "`r Sys.Date()`"
output:
  html_vignette:
    toc:  false
  pdf_vignette:
    toc:  false	
header-includes:
  \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{sparseHessian example}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

The **sparseHessianFD** package provides an interface to ACM TOMS Algorithm
636 for estimating a Hessian that is sparse, in that a large proportion of
the cross-partial derivatives are zero.  The user provides the following:

1.  an R function that returns the value of the objective function
whene valuated at a variable vector $P$;
2.  an R function that returns the gradient of that objective function
at $P$; and
3.  the row and column indices for the non-zero elements in the lower triangle of the Hessian.

The non-zero elements are estimated through finite differencing of
the gradients in a way that exploits the sparsity pattern of the
Hessian.

There are, of course, other methods to estimate sparse Hessians, such
as automatic differentiation (AD).  However, operator-overloading
implementations of AD require the objective function to be written
using specialized libraries.  These libraries are available for C++,
Fortran, Matlab, Python, and possibly other programming languages and
environments.  At the present, there are no AD implementations that
are "native" to R.  Other than deriving the Hessian analytically,
finite differencing remains the best option when the objective and
gradient functions are written purely in R.

## Example function

Before going into the details of how to use the package, let's
consider the following example of an objective function with a sparse Hessian.
 Suppose we have a dataset of $N$ households, each with $T$ opportunities to purchase a particular product.  Let $y_i$ be the number of times household $i$ purchases the product, out of the $T$ purchase opportunities.  Furthermore, let $p_i$ be the probability of purchase; $p_i$ is the same for all $T$ opportunities, so we can treat $y_i$ as a binomial random variable.  The purchase probability $p_i$ is heterogeneous, and depends on both $k$ continuous covariates $x_i$, and a heterogeneous coefficient vector $\beta_i$, such that
$$
  p_i=\frac{\exp(x_i'\beta_i)}{1+\exp(x_i'\beta_i)},~i=1 ... N
$$

The coefficients can be thought of as sensitivities to the covariates, and they are distributed across the population of households following a multivariate normal distribution with mean $\mu$ and covariance $\Sigma$.   We assume that we know $\Sigma$, but we do not know $\mu$.  Instead, we place a multivariate normal prior on $\mu$, with mean $0$ and covariance $\Omega_0$.  Thus, each $\beta_i$, and $\mu$ are $k-$dimensional vectors, and the total number of unknown variables in the model is $(N+1)k$.

The log posterior density, ignoring any normalization constants, is
$$
  \log \pi(\beta_{1:N},\mu|Y, X, \Sigma_0,\Omega_0)=\sum_{i=1}^Np_i^{y_i}(1-p_i)^{T-y_i}
  -\frac{1}{2}\left(\beta_i-\mu\right)'\Sigma^{-1}\left(\beta_i-\mu\right)
-\frac{1}{2}\mu'\Omega_0^{-1}\mu
$$

```{r, echo=FALSE}
N <- 6
k <- 2
nv1 <- (N+1)*k
nels1 <- nv1^2
nnz1 <- (N+1)*k^2 + 2*N*k^2
nnz1LT <- (N+1)*k*(k+1)/2 + N*k^2
Q <- 1000
nv2 <- (Q+1)*k
nels2 <- nv2^2
nnz2 <- (Q+1)*k^2 + 2*Q*k^2
nnz2LT <- (Q+1)*k*(k+1)/2 + Q*k^2
options(scipen=999)
```

Since the $\beta_i$ are drawn iid from a multivariate normal,
$\dfrac{\partial^2\log\pi }{\partial\beta_i\beta_j}=0$ for all $i\neq
j$.  We also know that all of the $\beta_i$ are correlated with
$\mu$.  The structure of the Hessian depends on how the variables are
ordered within the vector. One such ordering is to group all of the
coefficients for each unit together.

$$
\beta_{11},...,\beta_{1k},\beta_{21},...,\beta_{2k},...~,~...~,~\beta_{N1}~,~...~,~\beta_{Nk},\mu_1,...,\mu_k
$$

In this case, the Hessian has a "block-arrow" structure.  For example,
if $N=`r N`$ and $k=`r k`$, then there are `r nv1` total variables, and the Hessian will have the following pattern.

```{r, echo=FALSE}
M <- as(kronecker(diag(N),matrix(1,k,k)),"lMatrix")
M <- rBind(M, Matrix(TRUE,k,N*k))
M <- cBind(M, Matrix(TRUE, k*(N+1), k))
print(M)
```

There are `r nels1` elements in this symmetric matrix, but only  `r nnz1` are
non-zero, and only `r nnz1LT` values are unique.  Although the reduction in
RAM from using a sparse matrix structure for the Hessian may be
modest, consider what would happen if $N=`r Q`$ instead.  In that case,
there are `r nv2` variables in the problem, and more than $`r 
floor(nels2/10^6)`$ million
elements in the Hessian.  However, only $`r nnz2`$ of those elements are
non-zero.  If we work with only the lower triangle of the Hessian we only need to work with
only `r nnz2LT` values.

Another possibility is to group coefficients for each covariate
together.

$$
\beta_{11},...,\beta_{1N},\beta_{21},...,\beta_{2N},...,...,\beta_{k1},...,\beta_{kN},\mu_1,...,\mu_k
$$

Now the Hessian has an "off-diagonal" sparsity pattern.

```{r, echo=FALSE}
M <- as(kronecker(matrix(1,k,k), diag(N)),"lMatrix")
M <- rBind(M, Matrix(TRUE,k,N*k))
M <- cBind(M, Matrix(TRUE, k*(N+1), k))
print(M)
```

In both cases, the number of non-zeros is the same.  The
block-diagonal case may lead to slightly faster initialization,
but repeated computation of the Hessian will take the same time as
with the "off-diagonal" case.

## Using the package

The functions for computing the objective function, gradient and
Hessian for this example are in the R/ex\_funcs.R file.  The package
also includes a sample dataset with simulated data from the binary choice example.

To start, we load the data, set some dimension parameters, set prior
values for $\Sigma^{-1}$ and $\Omega^{-1}$, and simulate a
vector of variables at which to evaluate the function.

```{r}
set.seed(123)
data(binary)
Y <- binary$Y
X <- binary$X
T <- binary$T
N <- length(Y)
k <- NROW(X)
nvars <- as.integer(N*k + k)
P <- rnorm(nvars) ## random starting values
inv.Sigma <- rWishart(1,k+5,diag(k))[,,1]
inv.Omega <- diag(k)
```

```r
options(scipen=-999)
```

This dataset represents the simulated choices for $N= `r N`$ customers
over $T= `r T`$ purchase opportunties, where the probability of purchase
is influenced by $k= `r k`$ covariates.

The objective function for the binary choice example is `binary.f` and the gradient function is
`binary.grad`.  The first argument to both is the variable vector, and
the argument lists must be the same for both.  For this example, we
need to provide the data ($X$ and $Y$) and the prior parameters
($\Sigma^{-1}$ and $\Omega^{-1}$). The functions also have an
`order.row` argument to change the ordering of the variables (and
thus, the sparsity pattern). If `order.row=TRUE`, then the Hessian
will have an off-diagonal pattern.  If `order.row=FALSE`, then the
Hessian will have a block-arrow pattern.

For testing and demonstration purposes, we also have a `binary.hess`
function that returns the Hessian as a sparse `dgCMatrix` object (see
the Matrix package).

```{r, tidy=TRUE}
true.f <- binary.f(P, Y=Y, X=X, T=T, inv.Sigma=inv.Sigma, inv.Omega=inv.Omega, order.row=FALSE)
true.grad <- binary.grad(P, Y=Y, X=X, T=T, inv.Sigma=inv.Sigma, inv.Omega=inv.Omega, order.row=FALSE)
true.hess <- binary.hess(P, Y=Y, X=X, T=T, inv.Sigma=inv.Sigma, inv.Omega=inv.Omega, order.row=FALSE)
```

The sparsity pattern of the Hessian is specified by two **integer**
vectors:  one each for the row and column indices of the non-zero
elements of the ***lower triangular part*** of the Hessian.  If you
happen have have an example of a matrix with the same sparsity pattern
of the Hessian you are trying to compute, you can use the following
convenience function to extract the appropriate index vectors.

```{r}
pattern <- Matrix.to.Coord(tril(true.hess))
str(pattern)	    	    
```

Next, we create a new instance of a sparseHessianFD object with an
"initial variable" $P$, and the row and column indices of the non-zero
elements in the lower triangle of the Hessian.  We also pass in any
other arguments for `binary.f` and `binary.grad`.  We accept the default
values for other arguments to `sparseHessianFD.new`.
```{r}
obj <- sparseHessianFD.new(P, binary.f, binary.grad,
       rows=pattern$iRow, cols=pattern$jCol,
       X=X, Y=Y, T=T, inv.Sigma=inv.Sigma, inv.Omega=inv.Omega,
       order.row=FALSE)
```

Now we can evaluate the function value, gradient and Hessian through
`obj`.

```{r}
f <- obj$fn(P)
gr <- obj$gr(P)
hs <- obj$hessian(P)
```
Note that the member functions in the sparseHessianFD class take only
one argument:  the variable vector.  All of the other arguments are
stored in `obj`.

Do we get the same results that we would get after calling `binary.f`,
`binary.grad` and `binary.hess` directly?  Let's see.
```{r}

all.equal(f, true.f)
all.equal(gr, true.grad)
all.equal(hs, true.hess)	      	     	     
```

If there is any difference, keep in mind that `hs` is a numerical
estimate that is not always exact.  I certainly wouldn't worry about
mean relative differences smaller than, say, $10^{-6}$.

### Speed comparison

Instead of using this package, we could treat the Hessian as dense,
and use the hessian function numDeriv package. The advantage of the
numDeriv package is that it does not require the gradient.  However,
you can see that it takes some time to run.

```{r, tidy=TRUE}
library(numDeriv)
hess.time <- system.time(H1 <- obj$hessian(P))	  
fd.time <- system.time(H2 <- hessian(obj$fn, P))
H2 <- drop0(H2, 1e-7) ## treat values < 1e-8 as zero
print(hess.time)
print(fd.time)	   
```

The sparseHessianFD package can be substantially faster than estimating a dense
Hessian by brute force finite differencing.  The cost of this speed up
is that the user does need to provide the gradient and the sparsity
structure.  As with everything in life, there are trade-offs.

### 2015-01-08 R/ex_funcs.R

@@ -4,125 +4,148 @@
 ## Copyright (C) 2013 Michael Braun
 
 ## Functions to compute objective function, gradient, and Hessian for
-## example.R
+## binary choice example, and some unit tests.
 
 
-log.f <- function(pars, Y, X, inv.Omega, inv.Sigma, ..., order.row=FALSE) {
+#' @name binary
+#' @title Binary choice example
+#' @description Functions for binary choice example in the vignette.
+#' @param P Numeric vector of length (N+1)*k.  First N*k elements are heterogeneous coefficients. The remaining k elements are population parameters.
+#' @param Y,X Data matrices
+#' @param inv.Omega,inv.Sigma Prior parameters
+#' @return Log posterior density, gradient and Hessian.
+#' @details Hessian is sparse, and returned as a dgcMatrix object
+#' @rdname binary
+binary.f <- function(P, Y, X, T, inv.Omega, inv.Sigma, order.row=FALSE) {
 
-  beta <- matrix(pars[1:(N*k)], k, N, byrow=order.row)
-  mu <- pars[(N*k+1):(N*k+k)]
-
-  bx <- colSums(X * beta)
-  
-  log.p <- bx - log1p(exp(bx))
-  log.p1 <- -log1p(exp(bx))
-  
-  data.LL <- sum(Y*log.p + (T-Y)*log.p1)
-  
-  Bmu <- apply(beta, 2, "-", mu)
-  
-  prior <- -0.5 * sum(diag(tcrossprod(Bmu) %*% inv.Sigma))
-  hyp <- -0.5 * t(mu) %*% inv.Omega %*% mu
-  res <- data.LL + prior + hyp
-  return(as.numeric(res))
-  
+    N <- length(Y)
+    k <- NROW(X)
+    beta <- matrix(P[1:(N*k)], k, N, byrow=order.row)
+    mu <- P[(N*k+1):(N*k+k)]
+    
+    bx <- colSums(X * beta)
+    
+    log.p <- bx - log1p(exp(bx))
+    log.p1 <- -log1p(exp(bx))
+    
+    data.LL <- sum(Y*log.p + (T-Y)*log.p1)
+    
+    Bmu <- apply(beta, 2, "-", mu)
+    
[…]

### 2015-01-08 R/matrices.R

@@ -4,6 +4,23 @@
 
 
 
+#' @name Matrix.to.Coord
+#' @title Row and column indices of non-zero elements of a sparse symmetric matrix
+#' @param M A symmwetric Matrix
+#' @description Takes an object of class Matrix and returns the row and column indices of the non-zero elements in the lower triangle
+#' @return A list with two named elements.
+#' \item{iRow} Integer vector containing row indices of non-zero elements in the lower triangle of M
+#' \item{jCol} Integer vector containing column indices of non-zero elements in the lower triangle of M
+#' @export
+Matrix.to.Coord <- function(M) {
+  res <- vector("list",length=2)
+  names(res) <- c("iRow","jCol")
+  M <- tril(as(M,"TsparseMatrix"))
+  res$iRow <- as.integer(M@i + 1) ## return to 1-based indexingk
+  res$jCol <- as.integer(M@j + 1)
+  return(res)
+}
+
 Sym.CSC.to.Matrix <- function(H,nvars) {
 
 ## H is a list of data stored in compressed sparse column (CSC) format
@@ -15,16 +32,7 @@ Sym.CSC.to.Matrix <- function(H,nvars) {
 }
 
 
-Matrix.to.Coord <- function(M) {
 
-  res <- vector("list",length=2)
-  names(res) <- c("iRow","jCol")
-  M <- tril(as(M,"TsparseMatrix"))
-  res$iRow <- as.integer(M@i)+1 ## return to 1-based indexingk
-  res$jCol <- as.integer(M@j)+1
-  return(res)
-
-}
 
 Coord.to.Pattern.Matrix <- function(H,nrows, ncols=nrows) {
 

### 2015-01-08 R/wrappers.R

@@ -1,45 +1,64 @@
 ## wrappers.R --   Part of the sparsehessianFD package for the R programming language.
 ##
-## Copyright (C) 2013 Michael Braun
+## Copyright (C) 2015 Michael Braun
 
 
+#' @name sparseHessianFD.new
+#' @title Create and initialize a new sparseHessianFD object
+#' @param x vector of variables at which to evaluate value, gradient and Hessian during initialization.
+#' @param fn R function that returns function value
+#' @param gr R function that returns the gradient of the function
+#' @param rows Integer vector of row indices of non-zero elements of the lower triangle of the Hessian
+#' @param cols Integer vector of column indices of non-zero elements of the lower triangle of the Hessian
+#' @param direct If TRUE, use direct method.  Otherwise, use indirect/substitution method.
+#' @param eps Difference parameter for finite differencing
+#' @param ... Other parameters to be passed to fn and gr.
+#' @return An object of class sparseHessianFD
+#' @details The function new.sparse.hessian.obj has been deprecated.  Use sparseHessianFD.new instead.
+#' @seealso sparseHessianFD-class
+#' @export
+sparseHessianFD.new <- function(x, fn, gr, rows, cols, direct=FALSE,
+                            eps=sqrt(.Machine$double.eps), ...) {
+    
+    
+    stopifnot(is.function(fn))
+    stopifnot(is.function(gr))
+    
+    fn1 <- function(x) fn(x,...)  ## create closure
+    gr1 <- if (!is.null(gr)) function(x) gr(x,...)
+    
+    ## test fn and gr
 
+    stopifnot(is.finite(fn1(x)))
+    gradient <- gr1(x)
+    stopifnot(all(is.finite(gradient)))
+    stopifnot(length(gradient)==length(x))
+    stopifnot(!is.null(rows))
+    stopifnot(!is.null(cols))
+    
+    obj <- new("sparseHessianFD", length(x), fn1, gr1)
+    rows <- as.integer(rows)
+    cols <- as.integer(cols)
+    stopifnot(length(rows)==length(cols))
+    stopifnot(all(is.finite(rows)) & all(is.finite(cols)))
+    if (any(cols > rows)) stop("sparseHessianFD: Some elements in upper triangle.  Provide lower triangle only.")
[…]

### 2015-01-08 data-raw/binary.R

@@ -7,16 +7,14 @@ if (!require(devtools)) {
 
 set.seed(123)
 
-N <- 100
-k <- 3
+N <- 30
+k <- 2
 T <- 20
 
 x.mean <- rep(0,k)
 x.cov <- diag(k)
 mu <- rnorm(k, 0, 1)
 Omega <- diag(k)
-##inv.Sigma <- rWishart(1,k+5,diag(k))[,,1]
-##inv.Omega <- solve(Omega)
 
 X <- t(rmvnorm(N, mean=x.mean, sigma=x.cov)) ## k x N
 B <- t(rmvnorm(N, mean=mu, sigma=Omega)) ## k x N
@@ -26,6 +24,6 @@ Y <- sapply(log.p, function(q) return(rbinom(1,T,exp(q))))
 
 binary <- list(Y=Y, X=X, T=T)
 
-devtools::use_data(binary)
+devtools::use_data(binary, overwrite=TRUE)
  
 

### 2015-01-08 inst/doc/sparseHessianFD.R

@@ -1,0 +1,74 @@
+## ----, echo=FALSE--------------------------------------------------------
+N <- 6
+k <- 2
+nv1 <- (N+1)*k
+nels1 <- nv1^2
+nnz1 <- (N+1)*k^2 + 2*N*k^2
+nnz1LT <- (N+1)*k*(k+1)/2 + N*k^2
+Q <- 1000
+nv2 <- (Q+1)*k
+nels2 <- nv2^2
+nnz2 <- (Q+1)*k^2 + 2*Q*k^2
+nnz2LT <- (Q+1)*k*(k+1)/2 + Q*k^2
+options(scipen=999)
+
+## ----, echo=FALSE--------------------------------------------------------
+M <- as(kronecker(diag(N),matrix(1,k,k)),"lMatrix")
+M <- rBind(M, Matrix(TRUE,k,N*k))
+M <- cBind(M, Matrix(TRUE, k*(N+1), k))
+print(M)
+
+## ----, echo=FALSE--------------------------------------------------------
+M <- as(kronecker(matrix(1,k,k), diag(N)),"lMatrix")
+M <- rBind(M, Matrix(TRUE,k,N*k))
+M <- cBind(M, Matrix(TRUE, k*(N+1), k))
+print(M)
+
+## ------------------------------------------------------------------------
+set.seed(123)
+data(binary)
+Y <- binary$Y
+X <- binary$X
+T <- binary$T
+N <- length(Y)
+k <- NROW(X)
+nvars <- as.integer(N*k + k)
+P <- rnorm(nvars) ## random starting values
+inv.Sigma <- rWishart(1,k+5,diag(k))[,,1]
+inv.Omega <- diag(k)
+
+## ----, tidy=TRUE---------------------------------------------------------
+true.f <- binary.f(P, Y=Y, X=X, T=T, inv.Sigma=inv.Sigma, inv.Omega=inv.Omega, order.row=FALSE)
+true.grad <- binary.grad(P, Y=Y, X=X, T=T, inv.Sigma=inv.Sigma, inv.Omega=inv.Omega, order.row=FALSE)
+true.hess <- binary.hess(P, Y=Y, X=X, T=T, inv.Sigma=inv.Sigma, inv.Omega=inv.Omega, order.row=FALSE)
+
+## ------------------------------------------------------------------------
+pattern <- Matrix.to.Coord(tril(true.hess))
+str(pattern)	    	    
+
[…]

### 2015-01-08 inst/examples/ex2.R

@@ -43,34 +43,32 @@ Y <- sapply(log.p, function(q) return(rbinom(1,T,exp(q))))
 nvars <- as.integer(N*k + k)
 par <- rnorm(nvars) ## random starting values
 print("Creating true hessian")
-true.hess <- drop0(get.hess(par, Y, X, inv.Omega, inv.Sigma, order.row=order.row))
+true.hess <- drop0(binary.hess(par, Y, X, T, inv.Omega, inv.Sigma, order.row=order.row))
 hess.struct <- Matrix.to.Coord(as(tril(true.hess), "lMatrix"))
 
-make.f <- function(Y, X, inv.Omega, inv.Sigma, order.row=FALSE) {
+make.f <- function(Y, X, T, inv.Omega, inv.Sigma, order.row=FALSE) {
     function(pars) {
-        log.f(pars, Y=Y, X=X, inv.Omega=inv.Omega, inv.Sigma=inv.Sigma, order.row=order.row)
+        binary.f(pars, Y=Y, X=X, T=T, inv.Omega=inv.Omega, inv.Sigma=inv.Sigma, order.row=order.row)
     }
 }
 
-make.df <- function(Y, X, inv.Omega, inv.Sigma, order.row=FALSE) {
+make.df <- function(Y, X, T, inv.Omega, inv.Sigma, order.row=FALSE) {
     function(pars) {
-        get.grad(pars, Y=Y, X=X, inv.Omega=inv.Omega, inv.Sigma=inv.Sigma, order.row=order.row)
+        binary.grad(pars, Y=Y, X=X, T=T, inv.Omega=inv.Omega, inv.Sigma=inv.Sigma, order.row=order.row)
     }
 }
 
-get.f <- make.f(Y, X, inv.Omega, inv.Sigma, order.row)
-get.df <- make.df(Y, X, inv.Omega, inv.Sigma, order.row)
+get.f <- make.f(Y, X, T, inv.Omega, inv.Sigma, order.row)
+get.df <- make.df(Y, X, T, inv.Omega, inv.Sigma, order.row)
 
 
 print("Creating object")
 new.time <- system.time(obj <- new("sparseHessianFD", nvars, get.f, get.df) )
 print(new.time)
 print("Initializing")
-init.time <- system.time(obj$hessian.init(hess.struct$iRow, hess.struct$jCol, 1, 1e-6))
[…]

### 2015-01-08 inst/examples/ex3.R

@@ -1,84 +1,0 @@
-## example.R -- this file is part of sparseHessianFD, a contributed package
-## for the R statistical programming platform.
-##
-## Copyright (C) 2013 Michael Braun.  See LICENSE file for details.
-
-## this file demonstrates how to create a sparseHessianObj object, and use
-## it to compute the Hessian of the function, given the gradient and sparsity
-## structure.
-
-rm(list=ls())
-gc()
-
-
-library(mvtnorm)
-library(Rcpp)
-library(RcppEigen)
-library(numDeriv)
-
-
-
-set.seed(123)
-
-
-data(binary)
-Y <- binary$Y
-X <- binary$X
-T <- binary$T
-N <- length(Y)
-k <- NROW(X)
-
-nvars <- as.integer(N*k + k)
-par <- rnorm(nvars) ## random starting values
-
-Omega <- diag(k)
-inv.Sigma <- rWishart(1,k+5,diag(k))[,,1]
-inv.Omega <- solve(Omega)
-
-
-make.f <- function(Y, X, inv.Omega, inv.Sigma) {
-    function(pars) {
-        log.f(pars, Y=Y, X=X, inv.Omega=inv.Omega, inv.Sigma=inv.Sigma)
-    }
-}
-
-make.df <- function(Y, X, inv.Omega, inv.Sigma) {
-    function(pars) {
-        get.grad(pars, Y=Y, X=X, inv.Omega=inv.Omega, inv.Sigma=inv.Sigma)
-    }
-}
-
-make.hess <- function(Y, X, inv.Omega, inv.Sigma) {
-    function(pars) {
-        get.hess(pars, Y=Y, X=X, inv.Omega=inv.Omega, inv.Sigma=inv.Sigma)
-    }
-}
-
-get.f <- make.f(Y, X, inv.Omega, inv.Sigma)
-get.df <- make.df(Y, X, inv.Omega, inv.Sigma)
-get.d2f <- make.hess(Y, X, inv.Omega, inv.Sigma)
-
-## True values for test
-
-true.f <- get.f(par)
-true.df <- get.df(par)
-true.hess <- get.d2f(par)
-
-
-## Get hessian structure
-hess.struct <- Matrix.to.Coord(as(tril(true.hess), "lMatrix"))
-
-obj <- new("sparseHessianFD", nvars, get.f, get.df) )
[…]

### 2015-01-08 inst/examples/ex_funcs.R

@@ -1,124 +1,0 @@
-## ex_funcs.R -- this file is part of sparseHessianFD, a contributed package
-## for the R statistical programming platform.
-##
-## Copyright (C) 2013 Michael Braun
-
-## Functions to compute objective function, gradient, and Hessian for
-## example.R
-
-
-log.f <- function(pars, Y, X, inv.Omega, inv.Sigma, ...) {
-
-  beta <- matrix(pars[1:(N*k)], k, N, byrow=FALSE)
-  mu <- pars[(N*k+1):(N*k+k)]
-
-  bx <- colSums(X * beta)
-  
-  log.p <- bx - log1p(exp(bx))
-  log.p1 <- -log1p(exp(bx))
-  
-  data.LL <- sum(Y*log.p + (T-Y)*log.p1)
-  
-  Bmu <- apply(beta, 2, "-", mu)
-  
-  prior <- -0.5 * sum(diag(tcrossprod(Bmu) %*% inv.Sigma))
-  hyp <- -0.5 * t(mu) %*% inv.Omega %*% mu
-  res <- data.LL + prior + hyp
-  return(as.numeric(res))
-  
-}
-
-dlog.f.db <- function(pars, Y, X, inv.Omega, inv.Sigma) {
-
-  beta <- matrix(pars[1:(N*k)], k, N, byrow=FALSE)
-  mu <- pars[(N*k+1):length(pars)]
-  bx <- colSums(X * beta)
-
-  p <- exp(bx)/(1+exp(bx))
-
-  tmp <- Y - T*p
-
-  dLL.db <- apply(X,1,"*",tmp)
-    
-  Bmu <- apply(beta, 2, "-", mu)
-  dprior <- -inv.Sigma %*% Bmu
-  
-  res <- t(dLL.db) + dprior
-
-  return(as.vector(res))
- 
-}
-
-dlog.f.dmu <- function(p, Y, X, inv.Omega, inv.Sigma) {
-
-  beta <- matrix(p[1:(N*k)], k, N, byrow=FALSE)
-  mu <- p[(N*k+1):length(p)]
-  Bmu <- apply(beta, 2, "-", mu)
-
-  res <- inv.Sigma %*% (rowSums(Bmu)) -  inv.Omega %*% mu
-  return(res)
-}
-
-
-
[…]

### 2015-01-08 inst/examples/example.R

@@ -1,65 +1,0 @@
-## example.R -- this file is part of sparseHessianFD, a contributed package
-## for the R statistical programming platform.
-##
-## Copyright (C) 2013 Michael Braun.  See LICENSE file for details.
-
-## this file demonstrates how to create a sparseHessianObj object, and use
-## it to compute the Hessian of the function, given the gradient and sparsity
-## structure.
-
-
-library(plyr)
-library(Rcpp)
-library(RcppEigen)
-library(sparseHessianFD)
-
-source("ex_funcs.R")
-
-set.seed(123)
-
-N <- 15
-k <- 3
-T <- 20
-
-
-## Simulate data and set priors
-
-x.mean <- rep(0,k)
-x.cov <- diag(k)
-mu <- rnorm(k,0,10)
-Omega <- diag(k)
-inv.Sigma <- rWishart(1,k+5,diag(k))[,,1]
-inv.Omega <- solve(Omega)
-X <- t(rmvnorm(N, mean=x.mean, sigma=x.cov)) ## k x N
-B <- t(rmvnorm(N, mean=mu, sigma=Omega)) ## k x N
-XB <- colSums(X * B)
-log.p <- XB - log1p(exp(XB))
-Y <- laply(log.p, function(q) return(rbinom(1,T,exp(q))))
-
-nvars <- as.integer(N*k + k)
-par <- rnorm(nvars) ## random starting values
-
-
-hess.struct <- get.hess.struct(N, k)  ## for SparseFD method only
-
-obj <- new.sparse.hessian.obj(par, log.f, get.grad, hess.struct, 
-                              Y=Y, X=X, inv.Omega=inv.Omega, inv.Sigma=inv.Sigma)
-
-fn <- obj$fn(par)
-gr <- obj$gr(par)
-hs <- obj$hessian(par)
-fdf <- obj$fngr(par)
-fdfh <- obj$all(par)
-
-fn1 <- get.fn(par, obj)
-gr1 <- get.gr(par, obj)
-hs1 <- get.hessian(par, obj)
-get.fngr(par, obj)
-
-H2 <- get.hess(par,  Y=Y, X=X, inv.Omega=inv.Omega, inv.Sigma=inv.Sigma)
-
-
-
-
- 
-

### 2015-01-08 tests/testthat/test_binary.R

@@ -19,45 +19,82 @@ test_that("binary_example", {
     inv.Omega <- solve(Omega)
     inv.Sigma <- rWishart(1,k+5,diag(k))[,,1]
 
-    make.funcs <- function(Y, X, inv.Omega, inv.Sigma) {
+    make.funcs <- function(Y, X, T, inv.Omega, inv.Sigma, order.row) {
         res <- vector("list", length=3)
         names(res) <- c("fn", "gr", "hessian")
         res$fn <-  function(pars) {
-            log.f(pars, Y=Y, X=X, inv.Omega=inv.Omega, inv.Sigma=inv.Sigma)
+            binary.f(pars, Y=Y, X=X, T=T, inv.Omega=inv.Omega,
+                     inv.Sigma=inv.Sigma, order.row=order.row)
         }
         res$gr <-  function(pars) {
-            get.grad(pars, Y=Y, X=X, inv.Omega=inv.Omega, inv.Sigma=inv.Sigma)
+            binary.grad(pars, Y=Y, X=X, T=T, inv.Omega=inv.Omega,
+                        inv.Sigma=inv.Sigma, order.row=order.row)
         }
         res$hessian <-  function(pars) {
-            get.hess(pars, Y=Y, X=X, inv.Omega=inv.Omega, inv.Sigma=inv.Sigma)
+            binary.hess(pars, Y=Y, X=X, T=T, inv.Omega=inv.Omega,
+                        inv.Sigma=inv.Sigma, order.row=order.row)
         }
         return(res)
     }
         
-    f <- make.funcs(Y, X, inv.Omega, inv.Sigma)
+    f1 <- make.funcs(Y, X, T, inv.Omega, inv.Sigma, order.row=FALSE) ## block-arrow
+    f2 <- make.funcs(Y, X, T, inv.Omega, inv.Sigma, order.row=TRUE) ## off-diagonals
     
     ## True values for test
     
-    true.val <- f$fn(P)
-    true.grad <- f$gr(P)
-    true.hess <- f$hessian(P)    
-    
-    ## Get hessian structure
-    hess.struct <- Matrix.to.Coord(as(tril(true.hess), "lMatrix"))
-    
-    obj <- new("sparseHessianFD", nvars, f$fn, f$gr)
-    obj$hessian.init(hess.struct$iRow, hess.struct$jCol, 1, 1e-6)
-    
[…]

### 2015-01-09 inst/doc/sparseHessianFD.Rmd
version 0.2.0 probably ready for CRAN

@@
 ```{r, echo=FALSE}
+require(Matrix)
+require(sparseHessianFD)
 N <- 6
@@
 The functions for computing the objective function, gradient and
-Hessian for this example are in the R/ex\_funcs.R file.  The package
+Hessian for this example are in the R/binary.R file.  The package
 also includes a sample dataset with simulated data from the binary choice example.
@@
 data(binary)
-Y <- binary$Y
-X <- binary$X
-T <- binary$T
-N <- length(Y)
-k <- NROW(X)
+str(binary)
+N <- length(binary$Y)
+k <- NROW(binary$X)
 nvars <- as.integer(N*k + k)
 P <- rnorm(nvars) ## random starting values
-inv.Sigma <- rWishart(1,k+5,diag(k))[,,1]
-inv.Omega <- diag(k)
+priors <- list(inv.Sigma = rWishart(1,k+5,diag(k))[,,1],
+               inv.Omega = diag(k))
+
 ```
@@
 the argument lists must be the same for both.  For this example, we
-need to provide the data ($X$ and $Y$) and the prior parameters
+need to provide the data list ($X$, $Y$ and $T$) and the prior
+parameter list
 ($\Sigma^{-1}$ and $\Omega^{-1}$). The functions also have an
@@
 ```{r, tidy=TRUE}
-true.f <- binary.f(P, Y=Y, X=X, T=T, inv.Sigma=inv.Sigma, inv.Omega=inv.Omega, order.row=FALSE)
-true.grad <- binary.grad(P, Y=Y, X=X, T=T, inv.Sigma=inv.Sigma, inv.Omega=inv.Omega, order.row=FALSE)
-true.hess <- binary.hess(P, Y=Y, X=X, T=T, inv.Sigma=inv.Sigma, inv.Omega=inv.Omega, order.row=FALSE)
+true.f <- binary.f(P, binary, priors, order.row=FALSE)
+true.grad <- binary.grad(P, binary, priors, order.row=FALSE)
+true.hess <- binary.hess(P, binary, priors, order.row=FALSE)
 ```
@@
 obj <- sparseHessianFD.new(P, binary.f, binary.grad,
-       rows=pattern$iRow, cols=pattern$jCol,
-       X=X, Y=Y, T=T, inv.Sigma=inv.Sigma, inv.Omega=inv.Omega,
+       rows=pattern$rows, cols=pattern$cols,
+       data=binary, priors=priors,
        order.row=FALSE)
@@
 structure.  As with everything in life, there are trade-offs.
+
+## Quick summary
+In short, to use the package, follow the following steps:
+
+1.  Write R functions to return the value and gradient of the
+objective function.
+2.  Determine the row and column indices of the non-zero elements of
+the lower triangle of the Hessian.
+3.  Pick a variable vector (i.e., a starting value) at which you can
+    initialize the sparseHessian object.  It doesn't really matter
+    what this vector is, as long as the function value and gradient
+    elements are all finite.
+3.  Create a new sparseHessianFD object using the sparseHessianFD.new
+function.  For this example, call that object F.
+4.  Compute Hessian at x by calling F$hessian(x).
+5.  ???
+6.  Profit.

### 2015-01-09 R/binary.R

@@ -1,7 +1,6 @@
-## ex_funcs.R -- this file is part of sparseHessianFD, a contributed package
-## for the R statistical programming platform.
-##
-## Copyright (C) 2013 Michael Braun
+## binary.R -- Part of the sparseHessianFD package 
+## Copyright (C) 2013-2015 Michael Braun
+## See LICENSE file for details.
 
 ## Functions to compute objective function, gradient, and Hessian for
 ## binary choice example, and some unit tests.
@@ -11,37 +10,48 @@
 #' @title Binary choice example
 #' @description Functions for binary choice example in the vignette.
 #' @param P Numeric vector of length (N+1)*k.  First N*k elements are heterogeneous coefficients. The remaining k elements are population parameters.
-#' @param Y,X Data matrices
-#' @param inv.Omega,inv.Sigma Prior parameters
+#' @param data List of data matrices Y and X, and choice count integer T
+#' @param priors List of named matrices inv.Omega and inv.Sigma
+#' @param order.row Determines order of heterogeneous coefficients in
+#' parameter vector. Affects sparsity pattern of Hessian. See vignette. 
 #' @return Log posterior density, gradient and Hessian.
 #' @details Hessian is sparse, and returned as a dgcMatrix object
 #' @rdname binary
-binary.f <- function(P, Y, X, T, inv.Omega, inv.Sigma, order.row=FALSE) {
+#' @export
 
-    N <- length(Y)
-    k <- NROW(X)
+binary.f <- function(P, data, priors, order.row=FALSE) {
+
+    N <- length(data$Y)
+    k <- NROW(data$X)
     beta <- matrix(P[1:(N*k)], k, N, byrow=order.row)
     mu <- P[(N*k+1):(N*k+k)]
     
-    bx <- colSums(X * beta)
+    bx <- colSums(data$X * beta)
     
     log.p <- bx - log1p(exp(bx))
     log.p1 <- -log1p(exp(bx))
     
-    data.LL <- sum(Y*log.p + (T-Y)*log.p1)
+    data.LL <- sum(data$Y*log.p + (data$T-data$Y)*log.p1)
     
     Bmu <- apply(beta, 2, "-", mu)
     
[…]

### 2015-01-09 R/deprecated.R

@@ -1,0 +1,89 @@
+## deprecated.R -- Part of the sparseHessianFD package 
+## Copyright (C) 2013-2015 Michael Braun
+## See LICENSE file for details.
+
+
+
+## Place to hold functions that will not be maintained in the future
+
+
+#' @name sparseHessianFD-deprecated
+#' @aliases Sym.CSC.to.Matrix Coord.to.Sym.Pattern.Matrix
+#' @title Deprecated functions
+#' @description These functions were in earlier versions, but will no
+#' longer be maintained, and are not even guaranteed to work now.
+NULL
+
+#' @title Sym.CSC.to.Matrix
+#' @description Build sparse matrix from data in CSC (column
+#' compressed) format.
+#' @param H a list containing Hessian data.  See details.
+#' @param nvars the number of rows (and columns) in the matrix.
+#' @return An object of Matrix class.
+#' @details Use Matrix::sparseMatrix instead of Sym.CSC.to.Matrix.
+#' @rdname sparseHessianFD-deprecated
+#' @export
+Sym.CSC.to.Matrix <- function(H,nvars) {
+
+    .Deprecated("Matrix::spMatrix")
+    
+  M <- new("dsCMatrix", i = H$indrow, p = H$jpntr, x = H$vals, Dim=c(nvars, nvars),uplo="L")
+  return(M)
+}
+
+
+
+#' @name Coord.to.Sym.Pattern.Matrix
+#' @inheritParams Sym.CSC.to.Matrix
+#' @details
+#' Use Coord.to.Pattern.Matrix with symmetric=TRUE instead of Coord.to.Sym.Pattern.Matrix.
+#' @rdname sparseHessianFD-deprecated
+#' @export
+Coord.to.Sym.Pattern.Matrix <- function(H, nvars) {
+
+    .Deprecated("Coord.to.Pattern.Matrix")
+
+## coords are for lower triangle, but coerces to symmetric pattern matrix
+## H is a list with two integer vectors:  iRow and jCol
+
+  
+    
+    res <- new("nsTMatrix",i=as.integer(H$iRow-1), j=as.integer(H$jCol-1),
+               Dim=c(as.integer(nvars), as.integer(nvars)),uplo="L")
+
+    return(res)
+
+}
+
+#' @name new.sparse.hessian.obj
+#' @title Deprecated constructor
+#' @param x variable vector for initialization
+#' @param fn R function that returns function value
+#' @param gr R function that returns the gradient of the function
[…]

### 2015-01-09 R/matrices.R

@@ -1,82 +1,50 @@
-## matrices.R --   Part of the sparseHessianFD package for the R programming language.
-##
-## Copyright (C) 2013 Michael Braun
+## matrices.R -- Part of the sparseHessianFD package 
+## Copyright (C) 2013-2015 Michael Braun
+## See LICENSE file for details.
 
 #' @name Matrix.to.Coord
-#' @title Row and column indices from sparse symmetric matrix.
+#' @title Row and column indices from sparse matrix.
 #' @description Returns list of row and column indices of the non-zero
-#' elements of the lower triangle of a symmetric matrix.
-#' @param M A symmetric Matrix
-#' @details Takes an object of class Matrix and returns the row and
-#' column indices of the non-zero elements in the lower triangle 
+#' elements of a sparse matrix.
+#' @param M A sparse Matrix, as defined in the Matrix package.
 #' @return A list with two named elements.
 #' \describe{
-#' \item{iRow}{ Integer vector containing row indices of non-zero elements in the lower triangle of M}
-#' \item{jCol}{ Integer vector containing column indices of non-zero
-#' elements in the lower triangle of M}
+#' \item{rows}{ Integer vector containing row indices of non-zero elements}
+#' \item{cols}{ Integer vector containing column indices of non-zero elements}
 #' }
 #' @export
 Matrix.to.Coord <- function(M) {
   res <- vector("list",length=2)
-  names(res) <- c("iRow","jCol")
-  M <- tril(as(M,"TsparseMatrix"))
-  res$iRow <- as.integer(M@i + 1) ## return to 1-based indexing
-  res$jCol <- as.integer(M@j + 1)
+  names(res) <- c("rows","cols")
+  M <- as(M,"TsparseMatrix")
+  res$rows <- as.integer(M@i + 1) ## return to 1-based indexing
+  res$cols <- as.integer(M@j + 1)
   return(res)
-}
-
-
-#' @title Sym.CSC.to.Matrix
-#' @description Build sparse matrix from data in CSC (column
-#' compressed) format.
-#' @param H a list containing Hessian data.  See details.
-#' @param nvars the number of rows (and columns) in the matrix.
-#' @return An object of Matrix class.
-#' @details
-#' H is a list with three elements.
-#' \describe{
-#' \item{iRow}{Row indices for each of the non-zero elements in the lower triangle of H}
[…]

### 2015-01-09 R/sparseHessianFD-class.R

@@ -1,5 +1,8 @@
-## Copyright (C) 2014-15 <package author:  Michael Braun>
-## This file is part of sparseHessianFD.  Terms of use are defined in the LICENSE file.
+## sparseHessianFD-class.R -- Part of the sparseHessianFD package 
+## Copyright (C) 2014-2015 Michael Braun
+## See LICENSE file for details.
+
+
 
 
 ##' @title sparseHessianFD
@@ -8,7 +11,7 @@
 ##' @aliases Rcpp_sparseHessianFD-class
 ##' @docType class
 ##' @details
-##' This class contains methods to estimate the Hessian of an
+##' Methods to estimate the Hessian of an
 ##' objective function using finite ifferencing of gradients. An
 ##' object is created with the names of R functions that return the
 ##' value and the gradient, and initialized with the row and column
@@ -21,7 +24,7 @@
 ##' \item{$fngr(x)}{\code{signature(x="numeric")}: returns list of fn(x) and gr(x)}
 ##' \item{$hessian(x)}{\code{signature(x="numeric")}: returns sparse
 ##' Hessian as dgCMatrix object}
-##' \item{$hessian.init(rows,cols,direct,eps)}{ Used internally to initialize Hessian with sparsity pattern}
+##' \item{$hessian.init(rows, cols, direct, eps)}{ Used internally to initialize Hessian with sparsity pattern}
 ##' \item{$nnz()}{ Number of non-zero elements in lower triangle of Hessian, as provided by sparsity pattern}
 ##' \item{$nvars()}{ Length of parameter vector that was provided to constructor}
 ##' }

### 2015-01-09 R/sparseHessianFD-new.R

@@ -1,6 +1,6 @@
-## wrappers.R --   Part of the sparsehessianFD package for the R programming language.
-##
-## Copyright (C) 2015 Michael Braun
+## matrices.R -- Part of the sparseHessianFD package 
+## Copyright (C) 2013-2015 Michael Braun
+## See LICENSE file for details.
 
 
 #' @name sparseHessianFD.new
@@ -18,10 +18,9 @@
 #' @param eps The perturbation amount for finite differencing of the gradient to compute the Hessian. Defaults to sqrt(.Machine$double.eps).
 #' @param ... Other parameters to be passed to fn and gr.
 #' @return An object of class sparseHessianFD
-#' @details  Indexing starts at 1, and should include the diagonal
-#' elements, even though it is obvious that the diagonal elements are
-#' non-zero.  Entries must be ordered by column, and by row within
-#' columns.  Do not include any entries for the upper triangle.
+#' @details  Indexing starts at 1, and must include the diagonal
+#' elements.  Any upper triangle coordinates will be removed, and
+#' replaced with their lower triangle counterparts. 
 #' The algorithms used for estimating sparse Hessians using finite
 #' differencing are described in the reference below.
 #'
@@ -31,9 +30,28 @@
 #' the partition and permutation:  direct and indirect.  The direct
 #' method tends to require more computation, but may be more
 #' accurate.  We recommend the indirect method to start, so the
-#' default value is 0.
+#' default value is FALSE.
 #'
-#' The function new.sparse.hessian.obj has been deprecated.  Use sparseHessianFD.new instead.
+#' Here is the description of the two methods, as included in the
+#' original ACM TOMS Fortran code:
+#'
+#'   The direct method (method = 1) first determines a partition of
+#'   the columns of symmetric matrix A such that two columns in a
+#'   group have a non-zero element in row K only if column K is in an
+#'   earlier group. Using this partition, the subroutine then computes
+#'   a symmetric permutation of A consistent with the determination of
+#'   A by a lower triangular substitution method. 
+#' 
+#'   The indirect method first computes a symmetric permutation of A
+#'   which minimizes the maximum number of non-zero elements in any 
+#'   row of L, where L is the lower triangular part of the permuted
[…]

### 2015-01-09 inst/doc/sparseHessianFD.R

@@ -1,4 +1,6 @@
 ## ----, echo=FALSE--------------------------------------------------------
+require(Matrix)
+require(sparseHessianFD)
 N <- 6
 k <- 2
 nv1 <- (N+1)*k
@@ -27,20 +29,19 @@ print(M)
 ## ------------------------------------------------------------------------
 set.seed(123)
 data(binary)
-Y <- binary$Y
-X <- binary$X
-T <- binary$T
-N <- length(Y)
-k <- NROW(X)
+str(binary)
+N <- length(binary$Y)
+k <- NROW(binary$X)
 nvars <- as.integer(N*k + k)
 P <- rnorm(nvars) ## random starting values
-inv.Sigma <- rWishart(1,k+5,diag(k))[,,1]
-inv.Omega <- diag(k)
+priors <- list(inv.Sigma = rWishart(1,k+5,diag(k))[,,1],
+               inv.Omega = diag(k))
+
 
 ## ----, tidy=TRUE---------------------------------------------------------
-true.f <- binary.f(P, Y=Y, X=X, T=T, inv.Sigma=inv.Sigma, inv.Omega=inv.Omega, order.row=FALSE)
-true.grad <- binary.grad(P, Y=Y, X=X, T=T, inv.Sigma=inv.Sigma, inv.Omega=inv.Omega, order.row=FALSE)
-true.hess <- binary.hess(P, Y=Y, X=X, T=T, inv.Sigma=inv.Sigma, inv.Omega=inv.Omega, order.row=FALSE)
+true.f <- binary.f(P, binary, priors, order.row=FALSE)
+true.grad <- binary.grad(P, binary, priors, order.row=FALSE)
+true.hess <- binary.hess(P, binary, priors, order.row=FALSE)
 
 ## ------------------------------------------------------------------------
 pattern <- Matrix.to.Coord(tril(true.hess))
@@ -48,8 +49,8 @@ str(pattern)
 
 ## ------------------------------------------------------------------------
 obj <- sparseHessianFD.new(P, binary.f, binary.grad,
-       rows=pattern$iRow, cols=pattern$jCol,
-       X=X, Y=Y, T=T, inv.Sigma=inv.Sigma, inv.Omega=inv.Omega,
+       rows=pattern$rows, cols=pattern$cols,
+       data=binary, priors=priors,
        order.row=FALSE)
 
 ## ------------------------------------------------------------------------

### 2015-01-09 inst/include/func.h

@@ -208,7 +208,6 @@ void Rfunc::hessian_init(const IntegerVector& hess_iRow_,
       pert(i, ngrp(i)-1) = 1.;  // construct perturbation matrix from DSSM results
     }
 
-
     tmp1.setZero(nvars);
     tmp2.setZero(nvars);
     tmp3.setZero(nvars);

### 2015-01-09 tests/testthat/test_binary.R

@@ -1,44 +1,41 @@
+## test_binary.R -- Part of the sparseHessianFD package 
+## Copyright (C) 2013-2015 Michael Braun
+## See LICENSE file for details.
+
 context("binary example")
 
 test_that("binary_example", {
     require(numDeriv)
 
     set.seed(123)
-
-
     data(binary)
-    Y <- binary$Y
-    X <- binary$X
-    T <- binary$T
-    N <- length(Y)
-    k <- NROW(X)
+ 
+    N <- length(binary$Y)
+    k <- NROW(binary$X)
     nvars <- as.integer(N*k + k)
     P <- rnorm(nvars) ## random starting values
 
     Omega <- diag(k)
-    inv.Omega <- solve(Omega)
-    inv.Sigma <- rWishart(1,k+5,diag(k))[,,1]
+    priors <- list(inv.Omega = solve(Omega),
+                   inv.Sigma = rWishart(1,k+5,diag(k))[,,1])
 
-    make.funcs <- function(Y, X, T, inv.Omega, inv.Sigma, order.row) {
+    make.funcs <- function(D, priors, order.row) {
         res <- vector("list", length=3)
         names(res) <- c("fn", "gr", "hessian")
         res$fn <-  function(pars) {
-            binary.f(pars, Y=Y, X=X, T=T, inv.Omega=inv.Omega,
-                     inv.Sigma=inv.Sigma, order.row=order.row)
+            binary.f(pars, data=D, priors=priors, order.row=order.row)
         }
         res$gr <-  function(pars) {
-            binary.grad(pars, Y=Y, X=X, T=T, inv.Omega=inv.Omega,
-                        inv.Sigma=inv.Sigma, order.row=order.row)
+            binary.grad(pars, data=D, priors=priors, order.row=order.row)
         }
         res$hessian <-  function(pars) {
-            binary.hess(pars, Y=Y, X=X, T=T, inv.Omega=inv.Omega,
-                        inv.Sigma=inv.Sigma, order.row=order.row)
+            binary.hess(pars, data=D, priors=priors, order.row=order.row)
         }
         return(res)
     }
         
-    f1 <- make.funcs(Y, X, T, inv.Omega, inv.Sigma, order.row=FALSE) ## block-arrow
[…]

### 2015-01-09 tests/testthat/test_matrices.R

@@ -1,25 +1,41 @@
+## test_matrices.R -- Part of the sparseHessianFD package 
+## Copyright (C) 2013-2015 Michael Braun
+## See LICENSE file for details.
+
+
 context("Matrix utility functions")
 
 test_that("Matrix.to.Coord", {
 
-    nnz <- 6
+    ## LT matrix data
+    nnz <- 7
     k <- 5
-    rows <- c(1:5,5)
-    cols <- c(1:5,2)
-    vals <- c(1,2,3,4,5,6)
+    rows <- c(1,2,5,3,4,5,2)
+    cols <- c(1,2,2,3,4,5,5)
+    vals <- c(1,2,6,3,4,5,6)
 
 
-    
-    ## M <- Matrix(0,k,k)
-    ## for (i in 1:nnz) {
-    ##     M[rows[i], cols[i]] <- vals[i]
-    ## }
-    
+    M1 <- sparseMatrix(i=rows, j=cols, x=vals, dims=c(k,k) )
+    C1 <- Matrix.to.Coord(M1)
+    expect_equal(names(C1), c("rows","cols"))
+    M2 <- sparseMatrix(i=C1$rows, j=C1$cols, dims=c(k,k))
+    M3 <- as(M1, "nMatrix")
+    expect_equal(M2, M3)
 
-    res <- Matrix.to.Coord(M)
+ 
+    P1 <- Coord.to.Pattern.Matrix(C1$rows, C1$cols, c(k,k))
+    expect_is(P1, "ngCMatrix")
+    T1 <- tril(M3)
+    expect_equal(P1, M3)
+    expect_equal(T1, tril(M3))
 
-    print(cbind(res$iRow, res$jCol))
-    expect_equal(names(res), c("iRow","jCol"))
- ##   expect_equal(res$iRow, rows)
- ##   expect_equal(res$jCol, cols)
+    C2 <- Matrix.to.Coord(tril(M1))
+    P2 <- Coord.to.Pattern.Matrix(C2$rows, C2$cols, c(k,k), symmetric=TRUE)
+    expect_that(P2, is_a("nsCMatrix"))
+    expect_equal(forceSymmetric(tril(P1)), P2)
+
+    P3 <- Coord.to.Pattern.Matrix(C2$rows, C2$cols, c(k,k), compressed=FALSE)
+    expect_is(P3, "ngTMatrix")
+    expect_equivalent(P1, P3)
+
 })

### 2015-01-28 README.Rmd
closer to release

---
title:  The sparseHessianFD package
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Why use the sparseHessianFD package?

Suppose you have a scalar-valued objective function of _N_ variables.  You have an **R** function that returns the function value, and another that returns the gradient.  You still need the
Hessian, which would be hard to derive analytically, and translate
into an **R** function.  If _N_ is very large, it can take a long time
to estimate the Hessian numerically, and a lot of memory to store all _N^2_
elements.

This scalability problem comes up in at least two areas of statistics:

1.  using derivative-based algorithms for nonlinear optimization; and
2.  sampling from, and estimating the density of, high-dimensional
multivariate normal distributions.

The *sparseHessianFD* package solves this problem when the Hessian is
sparse, and the sparsity pattern is known in advance.  The sparsity pattern is just the row and column indices of the
non-zero elements in the lower triangle of the Hessian.  Sparse Hessians arise, for example, in hierarchical models in
which parameters are conditionally independent across heterogeneous
units.  In that case, we know which cross-partial derivatives are zero.

See the vignette for more about sparsity patterns in Hessians.

# Getting the package

The package is available on CRAN, so a simple

```
install.packages("sparseHessianFD")
```

should suffice.  The package is being developed on github at
github.com/braunm/sparseHessian, and you can get development branches
there.

# Using the package

To use the package, you need at least the following:

-  an R function that takes a parameter vector as the first argument,
   and returns the value of the function for which you need the
   Hessian;
-  an R function that takes the same arguments as above, and returns
the gradient of the target function.
-  two integer vectors, one for the row indices of the non-zero
   elements of the lower triangle of the Hessian, and one for the
   column indices.

The `sparseHessian.new` function returns an object of class
sparseHessianFD.  This class contains methods to return the function
value, gradient and Hessian.  The Hessian is returned in a sparse,
compressed format as a `dgCMatrix` object (defined in the *Matrix*
package).

A typical usage might look like this.

```{r, eval=FALSE}
P <- rnorm(N) ## random "starting value" for initialization
obj <- sparseHessianFD.new(P, f, df, rows, cols)
X <- rnorm(N) ## another set of variables at which the function is evaluated
val <- obj$fn(X) ## function value
gr <- obj$gr(X) ## gradient
hess <- obj$hessian(X)  ## Hessian, as a sparse dgCMatrix object
```

See the vignette and package manual for more details, options,
convenience functions, etc.

### 2015-04-14 R/matrices.R
more work on vignette

@@ -6,7 +6,7 @@
 #' @title Row and column indices from sparse matrix.
 #' @description Returns list of row and column indices of the non-zero
 #' elements of a sparse matrix.
-#' @param M A sparse Matrix, as defined in the Matrix package.
+#' @param M A matrix that can be coerced into a \code{TsparseMatrix} object, as defined in the Matrix package.  Base R matrices, and most Matrix matrices, are valid.
 #' @param index1 TRUE if the index of the first element should be 1, and FALSE if 0.
 #' @return A list with two named elements.
 #' \describe{
@@ -75,7 +75,7 @@ Coord.to.Pattern.Matrix <- function(rows, cols, dims, compressed=TRUE,
 #' \item{pntr}{ Integer vector containing pointers to elements of idx at which the next row or column begins.}
 #' }
 #' @export
-Matrix.to.Pointers <- function(M, order=c("column", "row", "symmetric"),
+Matrix.to.Pointers <- function(M, order="symmetric",
                                index1=TRUE, out.index1=index1) {
 
 

### 2015-04-14 R/sparseHessianFD-class.R

@@ -36,15 +36,15 @@ sparseHessianFD <-
                     colors = "list",
                     colors_vec = "integer"),
                 methods = list(
-                    initialize = function(x, fn, gr, rows, cols, direct=NULL,
+                    initialize = function(x.init, fn, gr, rows, cols, direct=NULL,
                                           eps=sqrt(.Machine$double.eps),
                                           index1 = TRUE, ...) {
-                        "Initialize object with functions to compute the objective function and gradient (fn and gr), row and column indices of non-zero elements (rows and cols), an initial variable vector x at which fn and gr can be evaluated, a finite differencing parameter eps, flags for 0 or 1-based indexing (index1), whether sparsity pattern is just for the lower triangle (indexLT), and other arguments (...) to be passed to fn and gr."
+                        "Initialize object with functions to compute the objective function and gradient (fn and gr), row and column indices of non-zero elements (rows and cols), an initial variable vector x.init at which fn and gr can be evaluated, a finite differencing parameter eps, flags for 0 or 1-based indexing (index1), whether sparsity pattern is just for the lower triangle (indexLT), and other arguments (...) to be passed to fn and gr."
 
                         if (!is.null(direct)) {
                             warning(" 'direct' argument is ignored. Only indirect method is, and will be, supported.")
                         }
-                        validate(fn, gr, rows, cols, x, eps, index1, ...)
+                        validate(fn, gr, rows, cols, x.init, eps, index1, ...)
                         ww <- which(cols <= rows)
 
                         initFields(fn1 = function(y) fn(y, ...),
@@ -53,7 +53,7 @@ sparseHessianFD <-
                                    jCol = as.integer(cols[ww]),
                                    eps = eps,
                                    index1 = index1,
-                                   nvars = length(x),
+                                   nvars = length(x.init),
                                    nnz = length(iRow),
                                    ready = FALSE)
                         if (any(cols > rows)) {

### 2015-04-14 R/sparseHessianFD-new.R

@@ -60,37 +60,42 @@
 sparseHessianFD.new <- function(x, fn, gr, rows, cols, direct=FALSE,
                             eps=sqrt(.Machine$double.eps), ...) {
 
-
-    stopifnot(is.function(fn))
-    stopifnot(is.function(gr))
-
-    fn1 <- function(x) fn(x,...)  ## create closure
-    gr1 <- if (!is.null(gr)) function(x) gr(x,...)
-
-    ## test fn and gr
-
-    stopifnot(is.finite(fn1(x)))
-    gradient <- gr1(x)
-    stopifnot(all(is.finite(gradient)))
-    stopifnot(length(gradient)==length(x))
-    stopifnot(!is.null(rows))
-    stopifnot(!is.null(cols))
-
-    obj <- new("sparseHessianFD", length(x), fn1, gr1)
-    rows <- as.integer(rows)
-    cols <- as.integer(cols)
-    stopifnot(length(rows)==length(cols))
-    stopifnot(all(is.finite(rows)) & all(is.finite(cols)))
-    if (any(cols > rows)) {
-        warning("sparseHessianFD:  Some elements are in upper triangle and will be deleted.")
-        cat("Provide lower triangle only.\n")
-        ww <- which(cols > rows)
-        rows <- rows[-ww]
-        cols <- cols[-ww]
-    }
-    obj$hessian.init(rows, cols, as.integer(direct), eps)
-
-    return(obj)
-
+    .Deprecated("sparseHessianFD")
+    sparseHessianFD(x, fn=fn, gr=gr,
+                           rows=rows, cols=cols, eps=eps,
+                           index1=TRUE, ...)
 }
+
+ ##    stopifnot(is.function(fn))
+##     stopifnot(is.function(gr))
+
+##     fn1 <- function(x) fn(x,...)  ## create closure
+##     gr1 <- if (!is.null(gr)) function(x) gr(x,...)
+
+##     ## test fn and gr
+
+##     stopifnot(is.finite(fn1(x)))
+##     gradient <- gr1(x)
+##     stopifnot(all(is.finite(gradient)))
+##     stopifnot(length(gradient)==length(x))
+##     stopifnot(!is.null(rows))
+##     stopifnot(!is.null(cols))
+
+##     obj <- new("sparseHessianFD", length(x), fn1, gr1)
+##     rows <- as.integer(rows)
+##     cols <- as.integer(cols)
+##     stopifnot(length(rows)==length(cols))
[…]

### 2015-04-14 tests/testthat/test_binary.R

@@ -70,7 +70,7 @@ test_that("binary_example", {
     print(th1)
     cat("\nBanded:\n")
     print(th2)
-
+browser()
     expect_equal(test.val1, true.val1)
     expect_equal(test.grad1, true.grad1)
     expect_equal(test.hess1, true.hess1, tolerance=5e-8)

### 2015-04-14 R/matrices.R
clean some code and doxygen comments

@@ -34,6 +34,7 @@ Matrix.to.Coord <- function(M, index1=TRUE) {
 #' @param compressed If TRUE, returns a matrix is compressed column (default=TRUE)
 #' @param symmetric If TRUE, matrix will be symmetric, and only the
 #' lower triangular elements need to be provided (default=FALSE)
+#' @param index1 TRUE if input row and col use 1-based indexing, and FALSE for 0-based indexing.
 #' @return A sparse pattern matrix
 #' @details This function is useful to prototype a sparsity pattern.
 #' No assumptions are made about symmetry.
@@ -58,7 +59,7 @@ Coord.to.Pattern.Matrix <- function(rows, cols, dims, compressed=TRUE,
 #' compressed format.
 #' @param M A sparse Matrix, as defined in the Matrix package.
 #' @param order Pointers can be ordered according to a row-oriented or column-oriented compression pattern.  If symmetric is chosen, the compression method does not matter.  The input matrix must be symmetric (although not necessarily stored that way) to get symmetric output.  The only difference between "symmetric," and either "row" or "column," is the names of the elements in the list.
-#' @param index1 TRUE if the index of the first element should be 1, and FALSE if 0.
+#' @param index1 TRUE if return values use 1-based indexing, and FALSE for 0-based indexing.
 #' @return A list with two named elements. If matrix is row-oriented,
 #' \describe{
 #' \item{jCol}{ Integer vector containing column indices of non-zero elements}
@@ -76,7 +77,7 @@ Coord.to.Pattern.Matrix <- function(rows, cols, dims, compressed=TRUE,
 #' }
 #' @export
 Matrix.to.Pointers <- function(M, order="symmetric",
-                               index1=TRUE, out.index1=index1) {
+                               index1=TRUE) {
 
 
     stopifnot(order %in% c("column","row","symmetric"))
@@ -85,19 +86,19 @@ Matrix.to.Pointers <- function(M, order="symmetric",
         stopifnot(Matrix::isSymmetric(M))
         M <- as(M,"dgCMatrix")
         names(res) <- c("idx","pntr")
-        res$idx <- as.integer(M@i) + as.integer(out.index1)
-        res$pntr <- as.integer(M@p) + as.integer(out.index1)
+        res$idx <- as.integer(M@i) + as.integer(index1)
+        res$pntr <- as.integer(M@p) + as.integer(index1)
     } else {
         if (order == "row") {
[…]

### 2015-04-14 inst/doc/sparseHessianFD.R

@@ -1,81 +1,0 @@
-## ----, echo=FALSE--------------------------------------------------------
-require(Matrix)
-require(sparseHessianFD)
-N <- 6
-k <- 2
-nv1 <- (N+1)*k
-nels1 <- nv1^2
-nnz1 <- (N+1)*k^2 + 2*N*k^2
-nnz1LT <- (N+1)*k*(k+1)/2 + N*k^2
-Q <- 1000
-nv2 <- (Q+1)*k
-nels2 <- nv2^2
-nnz2 <- (Q+1)*k^2 + 2*Q*k^2
-nnz2LT <- (Q+1)*k*(k+1)/2 + Q*k^2
-options(scipen=999)
-
-## ----, echo=FALSE--------------------------------------------------------
-M <- as(kronecker(diag(N),matrix(1,k,k)),"lMatrix")
-M <- rBind(M, Matrix(TRUE,k,N*k))
-M <- cBind(M, Matrix(TRUE, k*(N+1), k))
-print(M)
-
-## ----, echo=FALSE--------------------------------------------------------
-M <- as(kronecker(matrix(1,k,k), diag(N)),"lMatrix")
-M <- rBind(M, Matrix(TRUE,k,N*k))
-M <- cBind(M, Matrix(TRUE, k*(N+1), k))
-print(M)
-
-## ----, collapse=TRUE-----------------------------------------------------
-set.seed(123)
-data(binary)
-str(binary)
-N <- length(binary$Y)
-k <- NROW(binary$X)
-nvars <- as.integer(N*k + k)
-P <- rnorm(nvars) ## random starting values
-priors <- list(inv.Sigma = rWishart(1,k+5,diag(k))[,,1],
-               inv.Omega = diag(k))
-
-
-## ----, echo=FALSE--------------------------------------------------------
-options(scipen=-999)
-
-## ----, tidy=TRUE---------------------------------------------------------
-true.f <- binary.f(P, binary, priors, order.row=FALSE)
-true.grad <- binary.grad(P, binary, priors, order.row=FALSE)
-true.hess <- binary.hess(P, binary, priors, order.row=FALSE)
-
-## ----, collapse=TRUE-----------------------------------------------------
-pattern <- Matrix.to.Coord(tril(true.hess))
-str(pattern)	    	    
-
-## ------------------------------------------------------------------------
-obj <- sparseHessianFD.new(P, binary.f, binary.grad,
[…]

### 2015-04-15 R/matrices.R
more vignette words

@@ -127,7 +127,7 @@ Matrix.to.Pointers <- function(M, order="symmetric",
 Coord.to.Pointers <- function(rows, cols, dims,
                               triangle=TRUE, lower=TRUE,
                               order=c("column", "row", "symmetric"),
-                              index1=TRUE, out.index1=index1) {
+                              index1=TRUE) {
 
     stopifnot(is.logical(triangle),
               is.logical(index1),
@@ -151,7 +151,7 @@ Coord.to.Pointers <- function(rows, cols, dims,
     } else {
         A <- R
     }
-    Matrix.to.Pointers(A, order, index1, out.index1)
+    Matrix.to.Pointers(A, order, index1, index1)
 }
 
 

### 2015-04-15 R/sparseHessianFD-class.R

@@ -60,7 +60,7 @@ sparseHessianFD <-
                             warning("Some elements are in upper triangle, and will be ignored.")
                         }
 
-
+print("Break 1")
 
                         ptr <- Coord.to.Pointers(iRow, jCol, c(nvars, nvars),
                                                  triangle=TRUE,
@@ -86,6 +86,7 @@ sparseHessianFD <-
                             colors_vec[colors[[i]]+1] <<- as.integer(i-1)
                         }
 
+
                         coord2vec <- function(j) {
                             z <- rep(0,nvars)
                             z[j+1] <- eps
@@ -100,6 +101,8 @@ sparseHessianFD <-
 
                     partition = function() {
                         "Return the partitioning used to compute finite differences"
+
+                        print("Break 3")
                         if (ready) {
                             res <- colors
                             if (index1) {

### 2015-04-20 src/color.cpp
start writing algorithm details in vignette

@@ -95,7 +95,7 @@ List color_graph(const IntegerVector& pntr, //row/col pointer
     for (auto i : uncolored) {
       for (auto rr : Wk) {
 	P[i].erase(rr);
-	if (LT) {
+	if (LT) { // Can degree calculation go outside loop on Wk?
 	  deg[i] = std::distance(P[i].begin(), P[i].upper_bound(i));
 	} else {
 	  deg[i] = P[i].size();

### 2015-04-23 R/RcppExports.R
vignette is finally taking shape

@@ -5,9 +5,9 @@
 #' @description Generate valid cyclic coloring of variables that is consistent with estimating sparse Hessian with a triangle substitution method.
 #' @param pntr,idx row pointers and column indices (CSC or CSR format; same since Hessian matrix is symmetric). Must use zero-based indexing.
 #' @param nvars Dimension of Hessian (number of variables)
-#' @param LT true (default) if ordering with degree in lower triangle, and false if using full Hessian.  Included only for testing, to see if it matters (I don't think it does).
+#' @param LT true if ordering with degree in lower triangle, and false (default) if using full Hessian.  Included only for testing, to see if it matters (I don't think it does).
 #' @return A list.  Each element of the list represents a color, and contains an integer vector with the indices of the variables with that color.  Indices are zero-based.
-color_graph <- function(pntr, idx, nvars, LT = TRUE) {
+color_graph <- function(pntr, idx, nvars, LT = FALSE) {
     .Call('sparseHessianFD_color_graph', PACKAGE = 'sparseHessianFD', pntr, idx, nvars, LT)
 }
 

### 2015-04-23 src/color.cpp

@@ -37,13 +37,13 @@ void print_vec(const Eigen::MatrixBase<T>& X) {
 //' @description Generate valid cyclic coloring of variables that is consistent with estimating sparse Hessian with a triangle substitution method.
 //' @param pntr,idx row pointers and column indices (CSC or CSR format; same since Hessian matrix is symmetric). Must use zero-based indexing.
 //' @param nvars Dimension of Hessian (number of variables)
-//' @param LT true (default) if ordering with degree in lower triangle, and false if using full Hessian.  Included only for testing, to see if it matters (I don't think it does).
+//' @param LT true if ordering with degree in lower triangle, and false (default) if using full Hessian.  Included only for testing, to see if it matters (I don't think it does).
 //' @return A list.  Each element of the list represents a color, and contains an integer vector with the indices of the variables with that color.  Indices are zero-based.
 //[[Rcpp::export]]
 List color_graph(const IntegerVector& pntr, //row/col pointer
 		 const IntegerVector& idx, // col/row index
 		 const int nvars,
-		 const bool LT = true) {
+		 const bool LT = false) {
   
   // All pointers and indices are ZERO-BASED
   // and for the FULL MATRICES (not just LT)

### 2015-04-28 R/binary-data.R
now back to the vignette

@@ -1,5 +1,5 @@
 #' @name binary-data
-#' @aliases binary_small binary_large
+#' @aliases binary_small binary_large binary_super
 #' @title Sample simulated data for binary choice model in vignette
 #' @description Simulated data.  See vignette.  Generated from data-raw/binary.R
 #' @docType data

### 2015-04-28 R/sparseHessianFD.R

@@ -18,7 +18,7 @@
 #' @references
 #' Coleman, Thomas F, Burton S Garbow, and Jorge J More. 1985. Software
 #' for Estimating Sparse Hessian Matrices. ACM Transaction on
-#' Mathematical Software 11 (4) (December): 363-377. 
+#' Mathematical Software 11 (4) (December): 363-377.
 #'
 #' Coleman, Thomas F, Burton S Garbow, and Jorge J More. 1985. Algorithm
 #' 636:  FORTRAN Subroutines for Estimating Sparse Hessian Matrices. ACM

### 2015-04-29 R/coloring.R
timing code

@@ -7,6 +7,7 @@
 #' @param L  nMatrix with sparsity pattern of lower triangle of Hessian
 #' @return Integer vector of length nvars with color assignments for each variable.
 #' @description cyclic coloring from a lower triangular pattern matrix
+#' @export
 coloring <- function(L) {
 
     stopifnot(is(L,"nMatrix"),

### 2015-04-29 inst/examples/summarize_timings.R

@@ -1,0 +1,43 @@
+library(ggplot2)
+library(plyr)
+library(dplyr)
+library(tidyr)
+library(reshape2)
+
+theme_set(theme_bw())
+
+data.file <- "timings3.Rdata"
+load(data.file)
+
+tab <- mutate(res, ms=bench.time/1000000) %>%
+  dcast(N+k+T+ord+bench.rep+ncolors~bench.expr, value.var="ms")  %>%
+  mutate(nvars=N*k+k, hess_f=hess/f, hess_df=hess/df,
+         hess_k=hess/k, hess_ncol = hess/ncolors, total=setup+hess,
+         total_f=total/f, total_df = total/df, df/nvars)
+
+tab2 <- gather(tab, stat, ms, c(setup:hess,hess_f:total_df)) %>%
+  group_by(N, k, T, ord, stat, nvars) %>%
+  summarize(mean=mean(ms), sd=sd(ms), lower=mean-1.96*sd, upper=mean+1.96*sd)
+
+
+## D2 <- subset(data.frame(tab2), stat %in% c("nvars","setup","colors","f","df","hess",
+##                                "hess_df","total","total_f","total_df"))
+
+D2 <- data.frame(tab2)
+
+P2 <- ggplot(D2, aes(x=N,y=mean, color=as.factor(k))) %>%
+  + geom_line() %>%
+  + scale_x_continuous("Number of heterogeneous units") %>%
+  + scale_y_continuous("Computation time (milliseconds)") %>%
+  + facet_wrap(~stat, scales="free")
+
+
+P3 <- ggplot(filter(D2,k==4), aes(x=N)) %>%
+  + geom_line(aes(y=mean), linetype=1) %>%
+  + geom_line(aes(y=upper), linetype=3) %>%
+  + geom_line(aes(y=lower), linetype=3) %>%
+  + facet_wrap(~stat, scales="free")
+
+
+
+print(P2)

### 2015-04-29 inst/examples/timing.R

@@ -1,110 +1,0 @@
-rm(list=ls())
-gc()
-##library(testthat)
-
-set.seed(1234)
-data(binary_super)
-binary <- binary_super
-
-N <- length(binary$Y)
-k <- NROW(binary$X)
-nvars <- as.integer(N*k + k)
-P <- rnorm(nvars) ## random starting values
-
-Omega <- diag(k)
-priors <- list(inv.Omega = solve(Omega),
-               inv.Sigma = rWishart(1,k+5,diag(k))[,,1])
-
-make.funcs <- function(D, priors, order.row) {
-    res <- vector("list", length=3)
-    names(res) <- c("fn", "gr", "hessian")
-    res$fn <-  function(pars) {
-        binary.f(pars, data=D, priors=priors, order.row=order.row)
-    }
-    res$gr <-  function(pars) {
-        binary.grad(pars, data=D, priors=priors, order.row=order.row)
-    }
-    res$hessian <-  function(pars) {
-        binary.hess(pars, data=D, priors=priors, order.row=order.row)
-    }
-    return(res)
-}
-
-f1 <- make.funcs(D=binary, priors=priors, order.row=FALSE) ## block-arrow
-f2 <- make.funcs(D=binary, priors=priors, order.row=TRUE) ## off-diagonals
-
-## True values for test
-true.hess1 <- drop0(f1$hessian(P))
-true.hess2 <- drop0(f2$hessian(P))
-
-## Get hessian structure2
-pattern1LT <- Matrix.to.Coord(tril(true.hess1))
-pattern2LT <- Matrix.to.Coord(tril(true.hess2))
-pattern1 <- Matrix.to.Coord(true.hess1)
-pattern2 <- Matrix.to.Coord(true.hess2)
-
-time <- array(NA, dim=c(2,2,3,2))
-dimnames(time) <- list(order=c("block","band"),
-                       method=c("direct","indirect"),
-                       stat=c("init","fd","setup"),
-                       version=c("old","new"))
-
-
-rows1 <- pattern1$rows
-rows2 <- pattern2$rows
-cols1 <- pattern1$cols
-cols2 <- pattern2$cols
-
-rows1LT <- pattern1LT$rows
-rows2LT <- pattern2LT$rows
-cols1LT <- pattern1LT$cols
-cols2LT <- pattern2LT$cols
-
-time["block","indirect","init","new"] <- system.time(
[…]

### 2015-04-29 inst/examples/timings.R

@@ -1,0 +1,138 @@
+rm(list=ls())
+gc()
+
+library(sparseHessianFD)
+library(mvtnorm)
+library(plyr)
+library(microbenchmark)
+library(doParallel)
+
+run.par <- TRUE
+if (run.par) registerDoParallel(cores=12) else registerDoParallel(cores=1)
+
+binary_sim <- function(N, k, T) {
+
+    x.mean <- rep(0,k)
+    x.cov <- diag(k)
+    x.cov[1,1] <- .02
+    x.cov[k,k] <- x.cov[1,1]
+    mu <- seq(-2,2,length=k)
+    Omega <- diag(k)
+
+    X <- t(rmvnorm(N, mean=x.mean, sigma=x.cov)) ## k x N
+    B <- t(rmvnorm(N, mean=mu, sigma=Omega)) ## k x N
+    XB <- colSums(X * B)
+    log.p <- XB - log1p(exp(XB))
+    Y <- sapply(log.p, function(q) return(rbinom(1,T,exp(q))))
+
+    list(Y=Y, X=X, T=T)
+}
+
+priors_sim <- function(k) {
+
+    list(inv.Omega = solve(diag(k)),
+         inv.Sigma = rWishart(1,k+5,diag(k))[,,1])
+}
+
+
+make_funcs <- function(D, priors, order.row) {
+    res <- vector("list", length=3)
+    names(res) <- c("fn", "gr", "hessian")
+    res$fn <-  function(pars) {
+        binary.f(pars, data=D, priors=priors, order.row=order.row)
+    }
+    res$gr <-  function(pars) {
+        binary.grad(pars, data=D, priors=priors, order.row=order.row)
+    }
+    res$hessian <-  function(pars) {
+        binary.hess(pars, data=D, priors=priors, order.row=order.row)
+    }
+    return(res)
+}
+
+make_perm <- function(iRow, jCol) {
+    tmp <- sparseMatrix(i=iRow, j=jCol, index1=TRUE, symmetric=TRUE)
+    perm <- order(Matrix::rowSums(tmp), decreasing=TRUE)
+    invperm <- invPerm(perm)
+    L <- tril(tmp[perm,perm])
+    ptr <- Matrix.to.Pointers(L, order="row", index1=TRUE)
+    idx <- ptr[[1]]
+    pntr <- ptr[[2]]
+}
+
+get_id <- function(x) 1:length(x)
+
+run_test <- function(NkTord, reps=50) {
+
+    N <- as.numeric(NkTord["N"])
[…]

### 2015-04-30 R/sparseHessianFD-class.R
the vignette is almost readable

@@ -38,15 +38,15 @@ sparseHessianFD <-
                     perm = "integer",
                     invperm = "integer"),
                 methods = list(
-                    initialize = function(x.init, fn, gr, rows, cols, direct=NULL,
+                    initialize = function(x, fn, gr, rows, cols, direct=NULL,
                                           eps=sqrt(.Machine$double.eps),
                                           index1 = TRUE, ...) {
-                        "Initialize object with functions to compute the objective function and gradient (fn and gr), row and column indices of non-zero elements (rows and cols), an initial variable vector x.init at which fn and gr can be evaluated, a finite differencing parameter eps, flags for 0 or 1-based indexing (index1), whether sparsity pattern is just for the lower triangle (indexLT), and other arguments (...) to be passed to fn and gr."
+                        "Initialize object with functions to compute the objective function and gradient (fn and gr), row and column indices of non-zero elements (rows and cols), an initial variable vector x at which fn and gr can be evaluated, a finite differencing parameter eps, flags for 0 or 1-based indexing (index1), whether sparsity pattern is just for the lower triangle (indexLT), and other arguments (...) to be passed to fn and gr."
 
                         if (!is.null(direct)) {
                             warning(" 'direct' argument is ignored. Only indirect method is, and will be, supported.")
                         }
-                        validate(fn, gr, rows, cols, x.init, eps, index1, ...)
+                        validate(fn, gr, rows, cols, x, eps, index1, ...)
                         ww <- which(cols <= rows)
 
                         initFields(fn1 = function(y) fn(y, ...),
@@ -55,7 +55,7 @@ sparseHessianFD <-
                                    jCol = as.integer(cols[ww]),
                                    eps = eps,
                                    index1 = index1,
-                                   nvars = length(x.init),
+                                   nvars = length(x),
                                    ready = FALSE)
                         if (any(cols > rows)) {
                             warning("Some elements are in upper triangle, and will be ignored.")

### 2015-05-01 R/matrices.R
even mroe vignette

@@ -84,7 +84,6 @@ Matrix.to.Pointers <- function(M, order="symmetric",
     res <- vector("list",length=2)
     if (order == "symmetric") {
         stopifnot(Matrix::isSymmetric(M))
-        ##     M <- as(M,"dgCMatrix")
         M <- as(M,"ngCMatrix")
         names(res) <- c("idx","pntr")
         res$idx <- as.integer(M@i) + as.integer(index1)

### 2015-05-01 inst/examples/summarize_timings.R

@@ -6,7 +6,7 @@ library(reshape2)
 
 theme_set(theme_bw())
 
-data.file <- "timings3.Rdata"
+data.file <- "inst/examples/timings3.Rdata"
 load(data.file)
 
 tab <- mutate(res, ms=bench.time/1000000) %>%

### 2015-05-02 R/binary.R
fixed jss

@@ -161,7 +161,7 @@ binary.hess <- function(P, data, priors, order.row=FALSE) {
 }
 
 .d2.cross <- function(N, inv.Sigma) {
-  res <- kronecker(Matrix(rep(1,N),nrow=1),inv.Sigma)
+  res <- kronecker(matrix(rep(1,N),nrow=1),inv.Sigma)
   return(res)
 }
 

### 2015-05-03 R/binary.R
will this paper ever get done?

@@ -88,9 +88,9 @@ binary.hess <- function(P, data, priors, order.row=FALSE) {
     }
 
     Bmu <- .d2.dmu(N,SX, XO)
-    res <- rBind(cBind(B2, Matrix::t(cross)),cBind(cross, Bmu))
+    res <- rbind(cbind(B2, Matrix::t(cross)),cBind(cross, Bmu))
 
-    return(res)
+    return(drop0(res))
 }
 
 

### 2015-05-03 inst/examples/summarize_timings.R
formatting

@@ -41,10 +41,10 @@ P2 <- ggplot(D2, aes(x=N,y=mean, color=as.factor(k), linetype=as.factor(k))) %>%
   + scale_y_continuous("Computation time (milliseconds)") %>%
   + guides(color=guide_legend("k"), linetype=guide_legend("k")) %>%
   + facet_wrap(~stat, scales="free") %>%
-  + theme(text=element_text(size=8))
+  + theme(text=element_text(size=8), legend.position="right")
 
 
-pdf(file="vignettes/timings.pdf", width=6.5, height=4)
+pdf(file="vignettes/timings.pdf", width=8, height=4.5)
 print(P2)
 dev.off()
 

### 2015-05-03 inst/examples/summarize_timings.R
home stretch

@@ -44,7 +44,7 @@ P2 <- ggplot(D2, aes(x=N,y=mean, color=as.factor(k), linetype=as.factor(k))) %>%
   + theme(text=element_text(size=8), legend.position="right")
 
 
-pdf(file="vignettes/timings.pdf", width=8, height=4.5)
+pdf(file="vignettes/timings.pdf", width=6, height=4)
 print(P2)
 dev.off()
 

### 2015-05-03 R/deprecated.R
deprecating functions

@@ -1,8 +1,5 @@
 ## deprecated.R -- Part of the sparseHessianFD package
 ## Copyright (C) 2013-2015 Michael Braun
-## See LICENSE file for details.
-
-
 
 ## Place to hold functions that will not be maintained in the future
 
@@ -41,7 +38,7 @@ Sym.CSC.to.Matrix <- function(H,nvars) {
 #' @export
 Coord.to.Sym.Pattern.Matrix <- function(H, nvars) {
 
-    .Deprecated("Coord.to.Pattern.Matrix")
+    .Deprecated("Matrix::sparseMatrix")
 
 ## coords are for lower triangle, but coerces to symmetric pattern matrix
 ## H is a list with two integer vectors:  iRow and jCol
@@ -54,6 +51,38 @@ Coord.to.Sym.Pattern.Matrix <- function(H, nvars) {
     return(res)
 
 }
+
+#' @name Coord.to.Pattern.Matrix
+#' @aliases Coord.to.Pattern.Matrix
+#' @title Pattern matrix from row and column indices.
+#' @description Converts row and column indices to a pattern Matrix
+#' object of Matrix class
+#' @param rows,cols row and column indices of non-zero elements
+#' @param dims 2-element vector for number of rows and columns in
+#' matrix
+#' @param compressed If TRUE, returns a matrix is compressed column (default=TRUE)
+#' @param symmetric If TRUE, matrix will be symmetric, and only the
+#' lower triangular elements need to be provided (default=FALSE)
+#' @param index1 TRUE if input row and col use 1-based indexing, and FALSE for 0-based indexing.
+#' @return A sparse pattern matrix
+#' @details This function is useful to prototype a sparsity pattern.
+#' No assumptions are made about symmetry.
+#' @rdname sparseHessianFD-deprecated
+#' @export
+Coord.to.Pattern.Matrix <- function(rows, cols, dims, compressed=TRUE,
+                                    symmetric=FALSE, index1=TRUE) {
+
+    res <- sparseMatrix(i=as.integer(rows),
+                        j=as.integer(cols),
+                        dims=dims,
+                        giveCsparse=compressed,
+                        symmetric=symmetric,
+                        index1=index1)
+    return(res)
+}
+
+
+
 
 #' @name new.sparse.hessian.obj
 #' @title Deprecated constructor
@@ -79,7 +108,7 @@ Coord.to.Sym.Pattern.Matrix <- function(H, nvars) {
 #' @export
[…]

### 2015-05-03 R/matrices.R

@@ -1,6 +1,6 @@
 ## matrices.R -- Part of the sparseHessianFD package
 ## Copyright (C) 2013-2015 Michael Braun
-## See LICENSE file for details.
+
 
 #' @name Matrix.to.Coord
 #' @title Row and column indices from sparse matrix.
@@ -23,33 +23,7 @@ Matrix.to.Coord <- function(M, index1=TRUE) {
   return(res)
 }
 
-#' @name Coord.to.Pattern.Matrix
-#' @aliases Coord.to.Pattern.Matrix
-#' @title Pattern matrix from row and column indices.
-#' @description Converts row and column indices to a pattern Matrix
-#' object of Matrix class
-#' @param rows,cols row and column indices of non-zero elements
-#' @param dims 2-element vector for number of rows and columns in
-#' matrix
-#' @param compressed If TRUE, returns a matrix is compressed column (default=TRUE)
-#' @param symmetric If TRUE, matrix will be symmetric, and only the
-#' lower triangular elements need to be provided (default=FALSE)
-#' @param index1 TRUE if input row and col use 1-based indexing, and FALSE for 0-based indexing.
-#' @return A sparse pattern matrix
-#' @details This function is useful to prototype a sparsity pattern.
-#' No assumptions are made about symmetry.
-#' @export
-Coord.to.Pattern.Matrix <- function(rows, cols, dims, compressed=TRUE,
-                                    symmetric=FALSE, index1=TRUE) {
 
-    res <- sparseMatrix(i=as.integer(rows),
-                        j=as.integer(cols),
-                        dims=dims,
-                        giveCsparse=compressed,
-                        symmetric=symmetric,
-                        index1=index1)
-    return(res)
-}
 
 
 #' @name Matrix.to.Pointers

### 2015-05-03 R/sparseHessianFD-new.R

@@ -1,101 +1,0 @@
-## matrices.R -- Part of the sparseHessianFD package
-## Copyright (C) 2013-2015 Michael Braun
-## See LICENSE file for details.
-
-
-#' @name sparseHessianFD.new
-#' @title Create and initialize a new sparseHessianFD object
-#' @param x A intital vector of variables at which to evaluate value, gradient
-#' and Hessian during initialization.
-#' @param fn R function that returns function value
-#' @param gr R function that returns the gradient of the function
-#' @param rows Integer vector of row indices of non-zero elements of
-#' the lower triangle of the Hessian
-#' @param cols Integer vector of column indices of non-zero elements
-#' of the lower triangle of the Hessian
-#' @param direct If TRUE, use direct method for computatation.  Otherwise, use
-#' indirect/substitution method.  See references.
-#' @param eps The perturbation amount for finite differencing of the gradient to compute the Hessian. Defaults to sqrt(.Machine$double.eps).
-#' @param ... Other parameters to be passed to fn and gr.
-#' @return An object of class sparseHessianFD
-#' @details  Indexing starts at 1, and must include the diagonal
-#' elements.  Any upper triangle coordinates will be removed, and
-#' replaced with their lower triangle counterparts.
-#' The algorithms used for estimating sparse Hessians using finite
-#' differencing are described in the reference below.
-#'
-#' This method involves a partitioning and permutation of the Hessian
-#' matrix to reduce the number of distinct finite differencing
-#' operations of the gradient.  There are two methods for computing
-#' the partition and permutation:  direct and indirect.  The direct
-#' method tends to require more computation, but may be more
-#' accurate.  We recommend the indirect method to start, so the
-#' default value is FALSE.
-#'
-#' Here is the description of the two methods, as included in the
-#' original ACM TOMS Fortran code:
-#'
-#'   The direct method (method = 1) first determines a partition of
-#'   the columns of symmetric matrix A such that two columns in a
-#'   group have a non-zero element in row K only if column K is in an
-#'   earlier group. Using this partition, the subroutine then computes
-#'   a symmetric permutation of A consistent with the determination of
-#'   A by a lower triangular substitution method.
-#'
[…]

### 2015-05-03 R/sparseHessianFD.R

@@ -7,22 +7,29 @@
 #' a function in sparse Matrix format.  The user must supply the
 #' objective function, the gradient, and the row and column indices of
 #' the non-zero elements of the lower triangle of the Hessian (i.e.,
-#' the sparsity structure must be known in advance). Sparse Hessians
-#' occur in many applications, such as log posterior densities of
-#' hierarchical models under conditional independence
-#' assumptions. This package is intended to be useful when optimizing
-#' objective functions with this structure, using optimizers than can
-#' exploit this sparsity, such as the trustOptim package.
-#'
-#' License details are available in the LICENSE file in the source code.
+#' the sparsity structure must be known in advance).
+
 #' @references
-#' Coleman, Thomas F, Burton S Garbow, and Jorge J More. 1985. Software
+#' Coleman, Thomas F, and Jin-Yi Cai. 1986.  The Cyclic Coloring Problem
+#' and Estimation of Sparse Hessian Matrices.  SIAM Journal on Algebraic
+#' Discrete Methods 7 (2): 221-235
+#'
+#' Coleman, Thomas F, Burton S Garbow, and Jorge J Mor\'e. 1985. Software
 #' for Estimating Sparse Hessian Matrices. ACM Transaction on
 #' Mathematical Software 11 (4) (December): 363-377.
 #'
-#' Coleman, Thomas F, Burton S Garbow, and Jorge J More. 1985. Algorithm
+#' Coleman, Thomas F, Burton S Garbow, and Jorge J Mor\'e. 1985. Algorithm
 #' 636:  FORTRAN Subroutines for Estimating Sparse Hessian Matrices. ACM
 #' Transactions on Mathematical Software 11 (4): 378.
+#'
+#' Coleman, Thomas F and Jorge J More\'e.  Estimation of Sparse Hessian
+#' Matrices and Graph Coloring Problems.  Mathematical Programming
+#' 28 (3) (October): 243-270
+#'
+#' Powell, M J D and Ph L Toint. 1979. On the Estimation of Sparse
+#' Hessian Matrices.  SIAM Journal on Numerical Analysis 16 (6)
+#' (December): 1060-1074.
+#'
 #' @keywords package
 #'
 #' @useDynLib sparseHessianFD

### 2015-05-03 tests/testthat/test_matrices.R

@@ -144,31 +144,3 @@ test_that("Coord.to.Pointers", {
     expect_equal(M2.true, P2)
     expect_equal(M2.trueS, P3)
 })
-
-context("Coord.to.Pattern.Matrix")
-test_that("Coord.to.Pattern.Matrix", {
-
-    k <- 5
-    rows <- c(1,2,5,3,4,5,2)
-    cols <- c(1,2,2,3,4,5,5)
-    rowsLT <- c(1,2,5,3,4,4,5)
-    colsLT <- c(1,2,2,3,3,4,5)
-
-    M1.true <- sparseMatrix(i=rows, j=cols, dims=c(k,k))
-    M2.true <- as(tril(sparseMatrix(i=rowsLT, j=colsLT, dims=c(k,k))),"ngCMatrix")
-    M2.trueS <- as(sparseMatrix(i=rowsLT, j=colsLT, dims=c(k,k), symmetric=TRUE),"ngCMatrix")
-
-    P1 <- Coord.to.Pattern.Matrix(rows, cols, dims=c(k,k),
-                                  symmetric=FALSE)
-
-    P2 <- Coord.to.Pattern.Matrix(rowsLT, colsLT, dims=c(k,k),
-                                  symmetric=FALSE)
-
-    P3 <- as(Coord.to.Pattern.Matrix(rowsLT, colsLT, dims=c(k,k),
-                                     symmetric=TRUE), "ngCMatrix")
-
-    expect_equal(M1.true, P1)
-    expect_equal(M2.true, P2)
-    expect_equal(M2.trueS, P3)
-
-})

### 2015-05-04 R/coloring.R
almost there

@@ -17,7 +17,9 @@ coloring <- function(L) {
     nvars <- NROW(L)
     L <- as(L,"nMatrix")
     G <- Matrix::crossprod(L)  # intersection graph
-    ptr <- Matrix.to.Pointers(G, order="symmetric", index1=FALSE)
+ ##   ptr <- Matrix.to.Pointers(G, order="symmetric", index1=FALSE)
+
+    ptr <- Matrix.to.Pointers(G, index1=FALSE)
 
     colors_vec <- get_colors(ptr[[2]], ptr[[1]], nvars)
     return(colors_vec)

### 2015-05-04 R/matrices.R

@@ -6,101 +6,45 @@
 #' @title Row and column indices from sparse matrix.
 #' @description Returns list of row and column indices of the non-zero
 #' elements of a sparse matrix.
-#' @param M A matrix that can be coerced into a \code{TsparseMatrix} object, as defined in the Matrix package.  Base R matrices, and most Matrix matrices, are valid.
+#' @param M A sparse matrix, as defined in the Matrix package.
 #' @param index1 TRUE if the index of the first element should be 1, and FALSE if 0.
 #' @return A list with two named elements.
 #' \describe{
 #' \item{rows}{ Integer vector containing row indices of non-zero elements}
 #' \item{cols}{ Integer vector containing column indices of non-zero elements}
 #' }
+#' @details This function is a wrapper to Matrix.to.Pointers for \code{order='triplet'} and \code{values=FALSE}.  It can be used to extract the row and column indices of a sparsity pattern from a matrix that has that same pattern.
 #' @export
+#'
 Matrix.to.Coord <- function(M, index1=TRUE) {
-  res <- vector("list",length=2)
-  names(res) <- c("rows","cols")
-  M <- as(M,"TsparseMatrix")
-  res$rows <- as.integer(M@i) + as.integer(index1)
-  res$cols <- as.integer(M@j) + as.integer(index1)
-  return(res)
-}
-
-
-
-
-#' @name Matrix.to.Pointers
-#' @title Column (row) indices and row (column) pointers from sparse matrix.
-#' @description Returns list of column (row) indices row (column) pointers
-#' of the non-zero elements of a sparse matrix stored in row (column)-oriented
-#' compressed format.
-#' @param M A sparse Matrix, as defined in the Matrix package.
-#' @param order Pointers can be ordered according to a row-oriented or column-oriented compression pattern.  If symmetric is chosen, the compression method does not matter.  The input matrix must be symmetric (although not necessarily stored that way) to get symmetric output.  The only difference between "symmetric," and either "row" or "column," is the names of the elements in the list.
-#' @param index1 TRUE if return values use 1-based indexing, and FALSE for 0-based indexing.
-#' @return A list with two named elements. If matrix is row-oriented,
-#' \describe{
-#' \item{jCol}{ Integer vector containing column indices of non-zero elements}
-#' \item{ipntr}{ Integer vector containing pointers to elements of jCol at which the next row begins.}
-#' }
-#'  If matrix is column-oriented,
[…]

### 2015-05-04 R/sparseHessianFD-class.R

@@ -62,7 +62,9 @@ sparseHessianFD <-
                         }
                         nnz <<- length(iRow)
 
-                        tmp <- sparseMatrix(i=iRow, j=jCol, index1=index1, symmetric=TRUE)
+                        tmp <- sparseMatrix(i=iRow, j=jCol,
+                                            index1=index1, symmetric=TRUE,
+                                            giveCsparse=FALSE)
                         perm <<- order(Matrix::rowSums(tmp), decreasing=TRUE)
                         invperm <<- invPerm(perm)
 
@@ -113,6 +115,13 @@ sparseHessianFD <-
                                   check.index1.col,
                                   is.finite(val)
                                   )
+
+                        dups <- which(duplicated(cbind(rows, cols)))
+                        if (length(dups)>0) {
+                            cat("Duplicates in sparsity pattern:\n",dups,"\n")
+                            stop("sparseHessianFD does not allow duplicates")
+                        }
+
                     },
 
                     fd = function(d, x, grad.x) {

### 2015-05-04 data-raw/binary.R

@@ -1,29 +1,0 @@
-if (!require(mvtnorm)) {
-    stop("package mvtnorm required")
-}
-if (!require(devtools)) {
-    stop("package devtools required")
-}
-
-set.seed(123)
-
-N <- 30
-k <- 2
-T <- 20
-
-x.mean <- rep(0,k)
-x.cov <- diag(k)
-mu <- rnorm(k, 0, 1)
-Omega <- diag(k)
-
-X <- t(rmvnorm(N, mean=x.mean, sigma=x.cov)) ## k x N
-B <- t(rmvnorm(N, mean=mu, sigma=Omega)) ## k x N
-XB <- colSums(X * B)
-log.p <- XB - log1p(exp(XB))
-Y <- sapply(log.p, function(q) return(rbinom(1,T,exp(q))))
-
-binary <- list(Y=Y, X=X, T=T)
-
-devtools::use_data(binary, overwrite=TRUE)
- 
-

### 2015-05-04 tests/testthat/test_matrices.R

@@ -1,39 +1,5 @@
 ## Part of the sparseHessianFD package
 ## Copyright (C) 2013-2015 Michael Braun
-## See LICENSE file for details.
-
-## context("Matrix utility functions")
-## test_that("Matrix.to.Coord", {
-##     ## LT matrix data
-##     nnz <- 7
-##     k <- 5
-##     rows <- c(1,2,5,3,4,5,2)
-##     cols <- c(1,2,2,3,4,5,5)
-##     vals <- c(1,2,6,3,4,5,6)
-
-##     M1 <- sparseMatrix(i=rows, j=cols, x=vals, dims=c(k,k) )
-##     C1 <- Matrix.to.Coord(M1)
-##     expect_equal(names(C1), c("rows","cols"))
-##     M2 <- sparseMatrix(i=C1$rows, j=C1$cols, dims=c(k,k))
-##     M3 <- as(M1, "nMatrix")
-##     expect_equal(M2, M3)
-
-##     P1 <- Coord.to.Pattern.Matrix(C1$rows, C1$cols, c(k,k))
-##     expect_is(P1, "ngCMatrix")
-##     T1 <- tril(M3)
-##     expect_equal(P1, M3)
-##     expect_equal(T1, tril(M3))
-
-##     C2 <- Matrix.to.Coord(tril(M1))
-##     P2 <- Coord.to.Pattern.Matrix(C2$rows, C2$cols, c(k,k), symmetric=TRUE)
-##     expect_that(P2, is_a("nsCMatrix"))
-##     expect_equal(forceSymmetric(tril(P1)), P2)
-
-##     P3 <- Coord.to.Pattern.Matrix(C2$rows, C2$cols, c(k,k), compressed=FALSE)
-##     expect_is(P3, "ngTMatrix")
-##     expect_equivalent(P1, P3)
-## })
-
 
 context("Matrix.to.Coord")
 test_that("Matrix.to.Coord", {
@@ -64,8 +30,9 @@ test_that("Matrix.to.Pointers", {
     M1.true <- sparseMatrix(i=rows, j=cols, dims=c(k,k) )
     C1r <- Matrix.to.Pointers(M1.true, order="row", index1=TRUE)
     C1c <- Matrix.to.Pointers(M1.true, order="column", index1=TRUE)
-    expect_equal(names(C1r), c("jCol","ipntr"))
-    expect_equal(names(C1c), c("iRow","jpntr"))
+
[…]

### 2015-05-04 README.Rmd
done for the night

@@
 
-The `sparseHessian.new` function returns an object of class
+The `sparseHessianFD` function returns an object of class
 sparseHessianFD.  This class contains methods to return the function
@@
 P <- rnorm(N) ## random "starting value" for initialization
-obj <- sparseHessianFD.new(P, f, df, rows, cols)
+obj <- sparseHessianFD(P, f, df, rows, cols)
 X <- rnorm(N) ## another set of variables at which the function is evaluated

### 2015-05-04 R/RcppExports.R

@@ -1,21 +1,21 @@
 # This file was generated by Rcpp::compileAttributes
 # Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
 
-#' @title Vertex coloring of sparse symmetric matrix
-#' @description Generate proper coloring of a sparse symmetric matrix.
-#' @param pntr,idx row pointers and column indices (CSC or CSR format; same since  matrix is symmetric). Must use zero-based indexing.
-#' @param nvars Dimension of matrix (number of variables)
-#' @return A list.  Each element of the list represents a color, and contains an integer vector with the indices of the variables with that color.  Indices are zero-based.
+#' @title Vertex coloring of a sparse undirected graph
+#' @description Generate proper vertex coloring of a sparse undirected graph.
+#' @param pntr,idx row pointers and column indices of the adjacency matrix, in compressed column-oriented format. Must use zero-based indexing.
+#' @param nvars Number of vertices.
+#' @return An integer vector of length nvars, where each element represents the color of the corresponding vertex. Indices are zero-based.
 get_colors <- function(pntr, idx, nvars) {
     .Call('sparseHessianFD_get_colors', PACKAGE = 'sparseHessianFD', pntr, idx, nvars)
 }
 
 #' @title Estimate sparse Hessian
 #' @description Estimate Hessian using triangular subsitution algorithm
-#' @param Y Matrix of finite differences
+#' @param Y Matrix of finite differences of gradients
 #' @param colors Vector of length nvars that identifies color of each variable
-#' @param jCol,ipntr Column indices and row pointers for non-zero elements of lower triangle of Hessian.
-#' @param delta Perturbation factor used to compute finite differences.
+#' @param jCol,ipntr Column indices and row pointers for non-zero elements of lower triangle of Hessian (row-oriented compressed format).
+#' @param delta Perturbation factor used to compute finite differences of gradients.
 #' @param nvars Dimension of Hessian (number of variables)
 #' @param nnz Number of non-zero elements in the lower triangle of the Hessian.
 #' @return A sparse Hessian of class dgCMatrix.

### 2015-05-04 R/binary-data.R

@@ -1,14 +1,18 @@
 #' @name binary-data
 #' @aliases binary_small binary_large binary_super
-#' @title Sample simulated data for binary choice model in vignette
-#' @description Simulated data.  See vignette.  Generated from data-raw/binary.R
+#' @title Sample simulated data for binary choice example in vignette
+#' @details The package provides four sample datasets for the hierarchical binary choice model described in the vignette. These datasets are:
+#' \describe{
+#' \item{binary}{$N=50$, $k=3$}
+#' \item{binary_small}{$N=20$, $k=2$}
+#' \item{binary_large}{$N=800$, $k=3$}
+#' \item{binary_super}{$N=1200$, $k=3$}
+#' }
+#' The datasets were generated using the code in data-raw/binary_sim.R.
 #' @docType data
 NULL
 
 #' @name binary_small-data
-#' @title Sample simulated data for binary choice model in vignette
-#' @description Simulated data, with $N=20$ and $k=2$.  See vignette.  Generated from
-#' data-raw/binary.R
 #' @docType data
 #' @rdname binary-data
 NULL
@@ -16,17 +20,11 @@ NULL
 
 
 #' @name binary_large-data
-#' @title Sample simulated data for binary choice model in vignette
-#' @description Simulated data, with $N=800$ and $k=3$.  See vignette.  Generated from
-#' data-raw/binary.R
 #' @docType data
 #' @rdname binary-data
 NULL
 
 #' @name binary_super-data
-#' @title Sample simulated data for binary choice model in vignette
-#' @description Simulated data, with $N=1200$ and $k=3$.  See vignette.  Generated from
-#' data-raw/binary.R
 #' @docType data
 #' @rdname binary-data
 NULL

### 2015-05-04 R/coloring.R

@@ -1,10 +1,9 @@
 ## Part of the sparseHessianFD package
 ## Copyright (C) 2013-2015 Michael Braun
-## See LICENSE file for details.
 
 #' @name coloring
 #' @title Triangular partitioning of variables
-#' @param L  nMatrix with sparsity pattern of lower triangle of Hessian
+#' @param L  sparsity pattern of the Hessian as a lower triangular pattern matrix
 #' @return Integer vector of length nvars with color assignments for each variable.
 #' @description cyclic coloring from a lower triangular pattern matrix
 #' @export
@@ -17,8 +16,6 @@ coloring <- function(L) {
     nvars <- NROW(L)
     L <- as(L,"nMatrix")
     G <- Matrix::crossprod(L)  # intersection graph
- ##   ptr <- Matrix.to.Pointers(G, order="symmetric", index1=FALSE)
-
     ptr <- Matrix.to.Pointers(G, index1=FALSE)
 
     colors_vec <- get_colors(ptr[[2]], ptr[[1]], nvars)

### 2015-05-04 R/deprecated.R

@@ -142,6 +142,7 @@ new.sparse.hessian.obj <- function(x, fn, gr, hs, fd.method=0L, eps=sqrt(.Machin
 #' @param eps The perturbation amount for finite differencing of the gradient to compute the Hessian. Defaults to sqrt(.Machine$double.eps).
 #' @param ... Other parameters to be passed to fn and gr.
 #' @return An object of class sparseHessianFD
+#' @details This function is deprecated.  Use \code{sparseHessianFD} instead.
 #' @export
 sparseHessianFD.new <- function(x, fn, gr, rows, cols, direct=FALSE,
                             eps=sqrt(.Machine$double.eps), ...) {

### 2015-05-04 src/color.cpp

@@ -1,46 +1,23 @@
-#include <Rcpp.h>
+// Part of the sparseHessianFD package
+// Copyright (C) 2013-2015 Michael Braun
 
+#include <Rcpp.h>
 
 using Rcpp::List;
 using Rcpp::IntegerVector;
 using Rcpp::Rcout;
 
-
 typedef std::set<int> S;
 
-template<typename T>
-void print_container(const T& X) {
-  int z = 0;
-  for (auto i : X) {
-    Rcout << i << "   ";
-    z++;
-    if (fmod(z,10) == 0) Rcout << "\n";
-  }
-  Rcout << "\n";
-}
-
-static void chkIntFn(void *dummy) {
-  R_CheckUserInterrupt();
-}
-
-void checkInterrupt() {
-   if( ! R_ToplevelExec(chkIntFn, NULL) )
-     Rcpp::stop( "user interuption" );
-}
-
-
-//' @title Vertex coloring of sparse symmetric matrix
-//' @description Generate proper coloring of a sparse symmetric matrix.
-//' @param pntr,idx row pointers and column indices (CSC or CSR format; same since  matrix is symmetric). Must use zero-based indexing.
-//' @param nvars Dimension of matrix (number of variables)
-//' @return A list.  Each element of the list represents a color, and contains an integer vector with the indices of the variables with that color.  Indices are zero-based.
+//' @title Vertex coloring of a sparse undirected graph
+//' @description Generate proper vertex coloring of a sparse undirected graph.
+//' @param pntr,idx row pointers and column indices of the adjacency matrix, in compressed column-oriented format. Must use zero-based indexing.
+//' @param nvars Number of vertices.
+//' @return An integer vector of length nvars, where each element represents the color of the corresponding vertex. Indices are zero-based.
 //[[Rcpp::export]]
 Rcpp::IntegerVector get_colors(const IntegerVector& pntr, //row/col pointer
 			 const IntegerVector& idx, // col/row index
 			 const int nvars) {
-  
-  // All pointers and indices are ZERO-BASED
-  // and for the FULL MATRICES (not just LT)
   
   std::vector<std::set<int> > P(nvars);
   std::vector<std::set<int> > forb(nvars);
@@ -80,5 +57,26 @@ Rcpp::IntegerVector get_colors(const IntegerVector& pntr, //row/col pointer
[…]

### 2015-05-04 src/subst.cpp

@@ -1,17 +1,18 @@
+// Part of the sparseHessianFD package
+// Copyright (C) 2013-2015 Michael Braun
+
 #include <Rcpp.h>
 #include <RcppEigen.h>
 #include <Eigen/Core>
 #include <Eigen/Sparse>
 
 
-
-
 //' @title Estimate sparse Hessian
 //' @description Estimate Hessian using triangular subsitution algorithm
-//' @param Y Matrix of finite differences
+//' @param Y Matrix of finite differences of gradients
 //' @param colors Vector of length nvars that identifies color of each variable
-//' @param jCol,ipntr Column indices and row pointers for non-zero elements of lower triangle of Hessian.
-//' @param delta Perturbation factor used to compute finite differences.
+//' @param jCol,ipntr Column indices and row pointers for non-zero elements of lower triangle of Hessian (row-oriented compressed format).
+//' @param delta Perturbation factor used to compute finite differences of gradients.
 //' @param nvars Dimension of Hessian (number of variables)
 //' @param nnz Number of non-zero elements in the lower triangle of the Hessian.
 //' @return A sparse Hessian of class dgCMatrix.

### 2015-05-04 tests/testthat/test_binary.R

@@ -1,6 +1,5 @@
 ## Part of the sparseHessianFD package
 ## Copyright (C) 2013-2015 Michael Braun
-## See LICENSE file for details.
 
 context("binary example")
 

### 2016-03-08 tests/testthat/test_colors.R
back to the vignette

@@ -13,7 +13,7 @@ test_that("colors", {
     N <- length(binary$Y)
     k <- NROW(binary$X)
     nvars <- as.integer(N*k + k)
-    P <- rnorm(nvars) ## random starting values
+    P <- rnorm(nvars) ## random evaluation values
 
     Omega <- diag(k)
     priors <- list(inv.Omega = solve(Omega),
@@ -44,6 +44,13 @@ test_that("colors", {
 
     g1 <- coloring(as(tril(true.hess1),"nMatrix"))
     g2 <- coloring(as(tril(true.hess2), "nMatrix"))
+
+    pat1 <- Matrix.to.Coord(tril(true.hess1))
+    pat2 <- Matrix.to.Coord(tril(true.hess2))
+
+    obj1 <- sparseHessianFD(P, f1$fn, f1$gr, pat1$rows, pat1$cols)
+    obj2 <- sparseHessianFD(P, f2$fn, f2$gr, pat2$rows, pat2$cols)
+
 
 })
 

### 2016-03-08 R/binary-data.R
done for day: lots of vignette work

@@ -1,6 +1,7 @@
 #' @name binary-data
 #' @aliases binary_small binary_large binary_super
 #' @title Sample simulated data for binary choice example in vignette
+#' @description Sample datasets for vignette
 #' @details The package provides four sample datasets for the hierarchical binary choice model described in the vignette. These datasets are:
 #' \describe{
 #' \item{binary}{$N=50$, $k=3$}

### 2016-03-08 R/deprecated.R

@@ -129,6 +129,8 @@ new.sparse.hessian.obj <- function(x, fn, gr, hs, fd.method=0L, eps=sqrt(.Machin
 
 #' @name sparseHessianFD.new
 #' @title Create and initialize a new sparseHessianFD object
+#' @details This function is deprecated.  Use \code{sparseHessianFD} instead.
+#' @description This function is deprecated.  Use \code{sparseHessianFD} instead.
 #' @param x A intital vector of variables at which to evaluate value, gradient
 #' and Hessian during initialization.
 #' @param fn R function that returns function value
@@ -142,7 +144,6 @@ new.sparse.hessian.obj <- function(x, fn, gr, hs, fd.method=0L, eps=sqrt(.Machin
 #' @param eps The perturbation amount for finite differencing of the gradient to compute the Hessian. Defaults to sqrt(.Machine$double.eps).
 #' @param ... Other parameters to be passed to fn and gr.
 #' @return An object of class sparseHessianFD
-#' @details This function is deprecated.  Use \code{sparseHessianFD} instead.
 #' @export
 sparseHessianFD.new <- function(x, fn, gr, rows, cols, direct=FALSE,
                             eps=sqrt(.Machine$double.eps), ...) {

### 2016-03-08 R/matrices.R

@@ -10,8 +10,8 @@
 #' @param index1 TRUE if the index of the first element should be 1, and FALSE if 0.
 #' @return A list with two named elements.
 #' \describe{
-#' \item{rows}{ Integer vector containing row indices of non-zero elements}
-#' \item{cols}{ Integer vector containing column indices of non-zero elements}
+#' \item{rows}{Integer vector containing row indices of non-zero elements}
+#' \item{cols}{Integer vector containing column indices of non-zero elements}
 #' }
 #' @details This function is a wrapper to Matrix.to.Pointers for \code{order='triplet'} and \code{values=FALSE}.  It can be used to extract the row and column indices of a sparsity pattern from a matrix that has that same pattern.
 #' @export
@@ -37,8 +37,7 @@ Matrix.to.Coord <- function(M, index1=TRUE) {
 #' \describe{
 #' \item{triangle=TRUE}{Input \code{rows} and {cols} represent lower or upper triangle of a matrix. If \code{order="symmetric"}, then the output list will be for a full, symmetric matrix. Otherwise, the output list will be for only the lower or upper triangle.  Any elements outside of the specified triangle will trigger an error.}
 #' \item{triangle=FALSE}{Input \code{rows} and {cols} represent a full matrix. If that matrix is not symmetric, then \code{order=="symmetric"} will trigger an error.}
-#' If \code{symmetric=FALSE} and \code{order='triplet'}, the output list should contain the same row and column indices as the input list.
-#' }
+#' If \code{symmetric=FALSE} and \code{order='triplet'}, the output list should contain the same row and column indices as the input list.}
 #' @return A list.  See Matrix.to.Pointers (no values are included in return list).
 #' @export
 Coord.to.Pointers <- function(rows, cols, dims,

### 2016-03-08 R/sparseHessianFD.R

@@ -3,6 +3,7 @@
 #' @docType package
 #' @title Estimate sparse Hessians using finite differences of
 #' gradients.
+#' @description Estimate sparse Hessian
 #' @details This package contains methods to return the Hessian of
 #' a function in sparse Matrix format.  The user must supply the
 #' objective function, the gradient, and the row and column indices of

### 2016-03-09 R/sparseHessianFD-class.R
more vignette work

@@ -15,7 +15,7 @@
 ##' @field nnz Number of non-zero elements in the lower triangle of the Hessian.
 ##' @field ready TRUE if object has been initialized, and Hessian has been partitioned.
 ##' @field idx,pntr Column indices and row pointers for non-zero elements in lower triangle of the permuted Hessian.  Row-oriented compressed storage.
-##' @field colors A vector representation of the partitioning of the columns.  There are nvars elements, one for each column of the Hessian.  The value corresponds to the "color" for that column.
+##' @field colors A vector representation of the partitioning of the columns.  There are nvars elements, one for each column of the permuted Hessian.  The value corresponds to the "color" for that column.
 ##'@field perm,invperm Permutation vector and its inverse
 ##' @details Do not access any of the fields directly.  Use the initializer instead.
 ##' @export sparseHessianFD

### 2016-03-10 R/binary-data.R
documentation

@@ -18,8 +18,6 @@ NULL
 #' @rdname binary-data
 NULL
 
-
-
 #' @name binary_large-data
 #' @docType data
 #' @rdname binary-data

### 2016-03-10 R/binary.R

@@ -1,6 +1,5 @@
 ## Part of the sparseHessianFD package
-## Copyright (C) 2013-2015 Michael Braun
-## See LICENSE file for details.
+## Copyright (C) 2013-2016 Michael Braun
 
 ## Functions to compute objective function, gradient, and Hessian for
 ## binary choice example, and some unit tests.
@@ -10,12 +9,12 @@
 #' @title Binary choice example
 #' @description Functions for binary choice example in the vignette.
 #' @param P Numeric vector of length (N+1)*k.  First N*k elements are heterogeneous coefficients. The remaining k elements are population parameters.
-#' @param data List of data matrices Y and X, and choice count integer T
-#' @param priors List of named matrices inv.Omega and inv.Sigma
+#' @param data Named list of data matrices Y and X, and choice count integer T
+#' @param priors Named list of matrices inv.Omega and inv.Sigma
 #' @param order.row Determines order of heterogeneous coefficients in
-#' parameter vector. Affects sparsity pattern of Hessian. See vignette.
-#' @return Log posterior density, gradient and Hessian.
-#' @details Hessian is sparse, and returned as a dgcMatrix object
+#' parameter vector. If TRUE, heterogeneous coefficients are ordered by unit.  If FALSE, they are ordered by covariate.
+#' @return Log posterior density, gradient and Hessian. The Hessian is a dgcMatrix object.
+#' @details These functions are used by the heterogeneous binary choice example in the vignette. There are N heterogeneous units, each making T binary choices.  The choice probabilities depend on k covariates.
 #' @rdname binary
 #' @export
 binary.f <- function(P, data, priors, order.row=FALSE) {

### 2016-03-10 R/coloring.R

@@ -1,12 +1,11 @@
 ## Part of the sparseHessianFD package
-## Copyright (C) 2013-2015 Michael Braun
+## Copyright (C) 2013-2016 Michael Braun
 
 #' @name coloring
 #' @title Triangular partitioning of variables
 #' @param L  sparsity pattern of the Hessian as a lower triangular pattern matrix
 #' @return Integer vector of length nvars with color assignments for each variable.
 #' @description cyclic coloring from a lower triangular pattern matrix
-#' @export
 coloring <- function(L) {
 
     stopifnot(is(L,"nMatrix"),

### 2016-03-10 R/deprecated.R

@@ -1,24 +1,24 @@
 ## deprecated.R -- Part of the sparseHessianFD package
-## Copyright (C) 2013-2015 Michael Braun
+## Copyright (C) 2013-2016 Michael Braun
 
 ## Place to hold functions that will not be maintained in the future
 
 
-#' @name sparseHessianFD-deprecated
+#' @name deprecated
 #' @aliases Sym.CSC.to.Matrix Coord.to.Sym.Pattern.Matrix
 #' @title Deprecated functions
 #' @description These functions were in earlier versions, but will no
 #' longer be maintained, and are not even guaranteed to work now.
 NULL
 
-#' @title Sym.CSC.to.Matrix
+## #' @title Sym.CSC.to.Matrix
 #' @description Build sparse matrix from data in CSC (column
 #' compressed) format.
 #' @param H a list containing Hessian data.  See details.
 #' @param nvars the number of rows (and columns) in the matrix.
 #' @return An object of Matrix class.
 #' @details Use Matrix::sparseMatrix instead of Sym.CSC.to.Matrix.
-#' @rdname sparseHessianFD-deprecated
+#' @rdname deprecated
 #' @export
 Sym.CSC.to.Matrix <- function(H,nvars) {
 
@@ -34,7 +34,7 @@ Sym.CSC.to.Matrix <- function(H,nvars) {
 #' @inheritParams Sym.CSC.to.Matrix
 #' @details
 #' Use Coord.to.Pattern.Matrix with symmetric=TRUE instead of Coord.to.Sym.Pattern.Matrix.
-#' @rdname sparseHessianFD-deprecated
+#' @rdname deprecated
 #' @export
 Coord.to.Sym.Pattern.Matrix <- function(H, nvars) {
 
@@ -52,8 +52,8 @@ Coord.to.Sym.Pattern.Matrix <- function(H, nvars) {
 
 }
 
-#' @name Coord.to.Pattern.Matrix
-#' @aliases Coord.to.Pattern.Matrix
+## #' @name Coord.to.Pattern.Matrix
+## #' @aliases Coord.to.Pattern.Matrix
 #' @title Pattern matrix from row and column indices.
 #' @description Converts row and column indices to a pattern Matrix
 #' object of Matrix class
@@ -67,7 +67,7 @@ Coord.to.Sym.Pattern.Matrix <- function(H, nvars) {
 #' @return A sparse pattern matrix
 #' @details This function is useful to prototype a sparsity pattern.
[…]

### 2016-03-10 R/matrices.R

@@ -1,5 +1,5 @@
 ## matrices.R -- Part of the sparseHessianFD package
-## Copyright (C) 2013-2015 Michael Braun
+## Copyright (C) 2013-2016 Michael Braun
 
 
 #' @name Matrix.to.Coord
@@ -13,9 +13,13 @@
 #' \item{rows}{Integer vector containing row indices of non-zero elements}
 #' \item{cols}{Integer vector containing column indices of non-zero elements}
 #' }
-#' @details This function is a wrapper to Matrix.to.Pointers for \code{order='triplet'} and \code{values=FALSE}.  It can be used to extract the row and column indices of a sparsity pattern from a matrix that has that same pattern.
+#' @details A wrapper to Matrix.to.Pointers for \code{order='triplet'} and \code{values=FALSE}, for extracting the row and column indices of a sparsity pattern from a matrix that has that same pattern.
+#' @examples
+#' M1 <- as(kronecker(diag(3), matrix(TRUE,2,2)),"sparseMatrix")
+#' C <- Matrix.to.Coord(Mat)
+#' M2 <- Matrix::sparseMatrix(i=C$rows, j=C$cols)
+#' all.equal(M1,M2)
 #' @export
-#'
 Matrix.to.Coord <- function(M, index1=TRUE) {
     res <- Matrix.to.Pointers(M, values=FALSE, order="triplet", index1=index1)
     list(rows=res$rows, cols=res$cols)
@@ -26,9 +30,8 @@ Matrix.to.Coord <- function(M, index1=TRUE) {
 #' @title Convert a matrix defined by row and column indices to one defined by a row- or column-oriented compression scheme.
 #' @description Returns indices and pointers that define a sparse Hessian in compressed format.  Inputs are the row and column indices.
 #' @param rows,cols row and column indices of non-zero elements
-#' @param dims 2-element vector for number of rows and columns in
-#' matrix
-#' @param order See Matrix.to.Pointers
+#' @param dims 2-element vector for number of rows and columns.
+#' @param order Determines the indexing/compression scheme for the output matrix.  Use 'triplet" to get row and column indices.  Defaults to the same class as M.
 #' @param triangle Is input intended to be a triangular (TRUE) or full (FALSE) matrix.  See Details for how this argument is interpreted for different values of \code{order}.
 #' @param lower If \code{triangular} is true, this argument identifies the input matrix as lower- or upper-triangular.  This argument is ignored if \code{triangle} is FALSE.
[…]

### 2016-03-10 R/sparseHessianFD-class.R

@@ -1,5 +1,5 @@
 ## This file is part of the sparseHessianFD package
-## Copyright (C) 2015 Michael Braun
+## Copyright (C) 2015-2016 Michael Braun
 ##
 ##' @title sparseHessianFD
 ##' @name sparseHessianFD-class
@@ -8,7 +8,7 @@
 ##' @field fn1 A closure for calling fn(x, ...).
 ##' @field gr1 A closure for calling gr(x, ...).
 ##' @field iRow,jCol Numeric vectors with row and column indices of the non-zero elements in the lower triangle (including diagonal) of the Hessian.
-##' @field eps The perturbation amount for finite differencing of the gradient to compute the Hessian. Defaults to sqrt(.Machine$double.eps).
+##' @field delta The perturbation amount for finite differencing of the gradient to compute the Hessian. Defaults to sqrt(.Machine$double.eps).
 ##' @field index1 TRUE if rows and cols use 1-based (R format) indexing (FALSE for 0-based (C format) indexing.
 ##' @field D raw finite differences (internal use only)
 ##' @field nvars Number of variables (length of x)
@@ -17,7 +17,29 @@
 ##' @field idx,pntr Column indices and row pointers for non-zero elements in lower triangle of the permuted Hessian.  Row-oriented compressed storage.
 ##' @field colors A vector representation of the partitioning of the columns.  There are nvars elements, one for each column of the permuted Hessian.  The value corresponds to the "color" for that column.
 ##'@field perm,invperm Permutation vector and its inverse
-##' @details Do not access any of the fields directly.  Use the initializer instead.
+##' @details Do not access any of the fields directly.  Use the initializer instead. The internal structure is subject to change in future versions.
+##' @examples
+##' ## Log posterior density of hierarchical binary choice model. See vignette.
+##' data("binary_small")
+##' N <- length(binary[["Y"]])
+##' k <- NROW(binary[["X"]])
+##' T <- binary[["T"]]
+##' priors <- list(inv.Sigma = rWishart(1,k+5,diag(k))[,,1],
+##'                inv.Omega = diag(k))
+##' true.hess <- binary.hess(P, binary, priors)
+##' pattern <- Matrix.to.Coord(tril(true.hess))
+##' str(pattern)
+##' obj <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
[…]

### 2016-03-10 R/sparseHessianFD.R

@@ -3,9 +3,9 @@
 #' @docType package
 #' @title Estimate sparse Hessians using finite differences of
 #' gradients.
-#' @description Estimate sparse Hessian
-#' @details This package contains methods to return the Hessian of
-#' a function in sparse Matrix format.  The user must supply the
+#' @description Estimate sparse Hessians using finite differences of
+#' gradients.
+#' @details The Hessian is returned as a sparse Matrix.  The user supplies the
 #' objective function, the gradient, and the row and column indices of
 #' the non-zero elements of the lower triangle of the Hessian (i.e.,
 #' the sparsity structure must be known in advance).
@@ -15,15 +15,11 @@
 #' and Estimation of Sparse Hessian Matrices.  SIAM Journal on Algebraic
 #' Discrete Methods 7 (2): 221-235
 #'
-#' Coleman, Thomas F, Burton S Garbow, and Jorge J Mor\'e. 1985. Software
+#' Coleman, Thomas F, Burton S Garbow, and Jorge J Moré. 1985. Software
 #' for Estimating Sparse Hessian Matrices. ACM Transaction on
 #' Mathematical Software 11 (4) (December): 363-377.
 #'
-#' Coleman, Thomas F, Burton S Garbow, and Jorge J Mor\'e. 1985. Algorithm
-#' 636:  FORTRAN Subroutines for Estimating Sparse Hessian Matrices. ACM
-#' Transactions on Mathematical Software 11 (4): 378.
-#'
-#' Coleman, Thomas F and Jorge J More\'e.  Estimation of Sparse Hessian
+#' Coleman, Thomas F and Jorge J Moré. 1979 Estimation of Sparse Hessian
 #' Matrices and Graph Coloring Problems.  Mathematical Programming
 #' 28 (3) (October): 243-270
 #'

### 2016-03-10 inst/examples/repl_table4.R

@@ -1,0 +1,94 @@
+library(sparseHessianFD)
+library(mvtnorm)
+library(plyr)
+library(microbenchmark)
+library(doParallel)
+library(numDeriv)
+
+run.par <- TRUE
+if (run.par) registerDoParallel(cores=12) else registerDoParallel(cores=1)
+
+binary_sim <- function(N, k, T) {
+    x.mean <- rep(0,k)
+    x.cov <- diag(k)
+    x.cov[1,1] <- .02
+    x.cov[k,k] <- x.cov[1,1]
+    mu <- seq(-2,2,length=k)
+    Omega <- diag(k)
+
+    X <- t(rmvnorm(N, mean=x.mean, sigma=x.cov)) ## k x N
+    B <- t(rmvnorm(N, mean=mu, sigma=Omega)) ## k x N
+    XB <- colSums(X * B)
+    log.p <- XB - log1p(exp(XB))
+    Y <- sapply(log.p, function(q) return(rbinom(1,T,exp(q))))
+    list(Y=Y, X=X, T=T)
+}
+
+priors_sim <- function(k) {
+    list(inv.Omega = solve(diag(k)),
+         inv.Sigma = rWishart(1,k+5,diag(k))[,,1])
+}
+
+
+make_funcs <- function(D, priors) {
+    res <- vector("list", length=3)
+    names(res) <- c("fn", "gr", "hessian")
+    res$fn <-  function(pars) binary.f(pars, data=D, priors=priors)
+    res$gr <-  function(pars) binary.grad(pars, data=D, priors=priors)
+    res$hessian <-  function(pars) binary.hess(pars, data=D, priors=priors)
+    return(res)
+}
+
+run_test <- function(Nk, reps=50) {
+    N <- as.numeric(Nk["N"])
+    k <- as.numeric(Nk["k"])
+    data <- binary_sim(N, k, T=20)
+    priors <- priors_sim(k)
+    F <- make_funcs(D=data, priors=priors)
+    nvars <- N*k+k
+    M <- as(Matrix::kronecker(Matrix::Diagonal(N),Matrix(1,k,k)),"nMatrix") %>%
+      rBind(Matrix(TRUE,k,N*k)) %>%
+      cBind(Matrix(TRUE, k*(N+1), k)) %>%
+      as("nMatrix")
+    pat <- Matrix.to.Coord(tril(M))
+    X <- rnorm(nvars)
[…]

### 2016-03-10 inst/examples/summarize_timings.R

@@ -6,7 +6,7 @@ library(reshape2)
 
 theme_set(theme_bw())
 
-data.file <- "inst/examples/timings3.Rdata"
+data.file <- "inst/examples/timings.Rdata"
 load(data.file)
 
 tab <- mutate(res, ms=bench.time/1000000) %>%
@@ -47,14 +47,6 @@ P2 <- ggplot(D2, aes(x=N,y=mean, color=as.factor(k), linetype=as.factor(k))) %>%
 pdf(file="vignettes/timings.pdf", width=6, height=4)
 print(P2)
 dev.off()
-
-
-## P3 <- ggplot(filter(D2,k==4), aes(x=N)) %>%
-##   + geom_line(aes(y=mean), linetype=1) %>%
-##   + geom_line(aes(y=upper), linetype=3) %>%
-##   + geom_line(aes(y=lower), linetype=3) %>%
-##   + facet_wrap(~stat, scales="free", nrow=2, ncol=3)
-
 
 
 

### 2016-03-10 inst/examples/table_numDeriv.R

@@ -3,6 +3,7 @@ library(plyr)
 library(dplyr)
 library(tidyr)
 library(reshape2)
+library(xtable)
 
 theme_set(theme_bw())
 

### 2016-03-10 inst/examples/timings.R

@@ -129,7 +129,7 @@ cases <- expand.grid(k=c(8, 6, 4, 2),
                      ord=c(TRUE))
 
 res <- adply(cases, 1, run_test, reps=200, .parallel=run.par)
-save(cases, res, file="timings3.Rdata")
+save(cases, res, file="timings.Rdata")
 
 
 

### 2016-03-10 src/color.cpp

@@ -14,6 +14,7 @@ typedef std::set<int> S;
 //' @param pntr,idx row pointers and column indices of the adjacency matrix, in compressed column-oriented format. Must use zero-based indexing.
 //' @param nvars Number of vertices.
 //' @return An integer vector of length nvars, where each element represents the color of the corresponding vertex. Indices are zero-based.
+//' @details  For internal use.  You should not have to call this function directly.
 //[[Rcpp::export]]
 Rcpp::IntegerVector get_colors(const IntegerVector& pntr, //row/col pointer
 			 const IntegerVector& idx, // col/row index
@@ -56,27 +57,3 @@ Rcpp::IntegerVector get_colors(const IntegerVector& pntr, //row/col pointer
 
   return(Rcpp::wrap(colors));
 }
-
-
-
-// template<typename T>
-// void print_container(const T& X) {
-//   int z = 0;
-//   for (auto i : X) {
-//     Rcout << i << "   ";
-//     z++;
-//     if (fmod(z,10) == 0) Rcout << "\n";
-//   }
-//   Rcout << "\n";
-// }
-
-// static void chkIntFn(void *dummy) {
-//   R_CheckUserInterrupt();
-// }
-
-// void checkInterrupt() {
-//    if( ! R_ToplevelExec(chkIntFn, NULL) )
-//      Rcpp::stop( "user interuption" );
-// }
-
-

### 2016-03-10 tests/testthat/test_colors.R

@@ -1,60 +1,0 @@
-## Part of the sparseHessianFD package
-## Copyright (C) 2013-2015 Michael Braun
-## See LICENSE file for details.
-
-context("colors")
-
-test_that("colors", {
-
-    set.seed(123)
-    data(binary_small)
-    binary <- binary_small
-
-    N <- length(binary$Y)
-    k <- NROW(binary$X)
-    nvars <- as.integer(N*k + k)
-    P <- rnorm(nvars) ## random evaluation values
-
-    Omega <- diag(k)
-    priors <- list(inv.Omega = solve(Omega),
-                   inv.Sigma = rWishart(1,k+5,diag(k))[,,1])
-
-    make.funcs <- function(D, priors, order.row) {
-        res <- vector("list", length=3)
-        names(res) <- c("fn", "gr", "hessian")
-        res$fn <-  function(pars) {
-            binary.f(pars, data=D, priors=priors, order.row=order.row)
-        }
-        res$gr <-  function(pars) {
-            binary.grad(pars, data=D, priors=priors, order.row=order.row)
-        }
-        res$hessian <-  function(pars) {
-            binary.hess(pars, data=D, priors=priors, order.row=order.row)
-        }
-        return(res)
-    }
-
-    f1 <- make.funcs(D=binary, priors=priors, order.row=FALSE) ## block-arrow
-    f2 <- make.funcs(D=binary, priors=priors, order.row=TRUE) ## off-diagonals
-
-    ## True values for test
-
-    true.hess1 <- drop0(f1$hessian(P))
-    true.hess2 <- drop0(f2$hessian(P))
-
-    g1 <- coloring(as(tril(true.hess1),"nMatrix"))
-    g2 <- coloring(as(tril(true.hess2), "nMatrix"))
-
-    pat1 <- Matrix.to.Coord(tril(true.hess1))
-    pat2 <- Matrix.to.Coord(tril(true.hess2))
-
-    obj1 <- sparseHessianFD(P, f1$fn, f1$gr, pat1$rows, pat1$cols)
-    obj2 <- sparseHessianFD(P, f2$fn, f2$gr, pat2$rows, pat2$cols)
-
-
-})
-
-
-
-
-

### 2016-03-14 nobuild/jss/sparseHessianFD.Rnw
ready for JSS submission

\documentclass[article]{jss}

%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{sparseHessianFD}
%\VignettePackage{sparseHessianFD}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{tabularx}
\usepackage{dcolumn}
\usepackage{booktabs}
\usepackage{etoolbox}
\usepackage{placeins}
\newtoggle{tikz}
\toggletrue{tikz}

\iftoggle{tikz}{
\usepackage{tikz}
}{}

 \newcolumntype{d}[1]{D{.}{.}{#1}}
\newcommand{\func}[1]{\code{#1}}
\newcommand{\class}[1]{\textsl{#1}}
\newcommand{\funcarg}[1]{\code{#1}}
\newcommand{\variable}[1]{\code{#1}}
\newcommand{\method}[1]{\func{#1}}

 \newcommand{\df}[3]{\mathsf{d}^{#1}f(#2;#3)}
 \newcommand{\parD}[3]{\mathsf{D}^{#1}_{#2}#3}
\newcommand{\hess}[2]{\mathsf{H}_{#1}#2}
\newcommand{\hessLT}[2]{\mathsf{L}_{#1}#2}
\newcommand{\Mat}[1]{\mathbf{#1}}

 \author{Michael Braun\\Edwin L. Cox School of Business\\Southern Methodist University}
 \Plainauthor{Michael Braun}
 \title{\pkg{sparseHessianFD}: An R Package for Estimating Sparse Hessian Matrices}
  \Plaintitle{sparseHessianFD: An R Package for Estimating Sparse Hessian Matrices}
 \Shorttitle{sparseHessianFD:  Sparse Hessians}
  \Keywords{sparse Hessians, sparsity, computation of Hessians,
    graph coloring, finite differences, differentiation}
  \Address{
    Michael Braun\\
    Edwin L. Cox School of Business\\
    Southern Methodist University\\
    6212 Bishop Blvd.\\
    Dallas, TX 75275\\
    E-mail:  \email{braunm@smu.edu}\\
    URL:  \url{http://www.smu.edu/Cox/Departments/FacultyDirectory/BraunMichael}
  }

  \Abstract{
    Sparse Hessian matrices occur often
    in statistics, and their fast and accurate estimation can
    improve efficiency of numerical optimization and sampling
    algorithms. By exploiting the known sparsity pattern of a Hessian, methods in the
    sparseHessianFD package require many fewer function or gradient
    evaluations than would be required if the Hessian were treated as
    dense. The package implements established graph coloring and
    linear substitution algorithms that were previously unavailable to
    R users, and is most useful when other numerical,
    symbolic or algorithmic methods are impractical, inefficient or
    unavailable. }

\date{March 9, 2016}

%% need no \usepackage{Sweave.sty}

\newenvironment{knitrout}{\begin{figure}}{\end{figure}}

\begin{document}

<<setup1, echo=FALSE, cache=FALSE>>=
suppressPackageStartupMessages(require(Matrix))
knitr::render_sweave()
knitr::opts_chunk$set(prompt=TRUE, cache=TRUE)
options(replace.assign=TRUE, width=90,prompt="R> ")
@

The Hessian matrix of a log likelihood function or log posterior
density function plays
an important role in statistics.  From a frequentist point of view,
the inverse of the negative Hessian is the asymptotic covariance of the sampling distribution of a maximum
likelihood estimator.  In Bayesian analysis, when evaluated at the
posterior mode, it is the covariance of a Gaussian approximation to
the posterior distribution.  More broadly, many numerical optimization
algorithms require repeated computation, estimation or approximation
of the Hessian or its inverse; see \citet{NocedalWright2006}.

The Hessian of an objective function with $M$ variables has $M^2$
elements, of which $M(M+1)/2$ are unique.  Thus, the storage
requirements of the Hessian, and computational cost of many linear
algebra operations on it, grow quadratically with the number of
decision variables.   For
applications with hundreds of thousands of variables, computing the
Hessian even once might not be practical under time, storage or processor constraints.  Hierarchical models, in
which each additional heterogeneous unit is associated with its own subset of
variables, are particularly vulnerable to this curse of dimensionality

However, for many problems, the Hessian is
\emph{sparse}, meaning that the proportion of non-zero elements in the
Hessian is small.  Consider a log
posterior density in a Bayesian hierarchical
model.  If the outcomes across units are conditionally
independent, the cross-partial derivatives of heterogeneous variables
across units are zero.  As the number of units
increases, the size of the Hessian still grows quadratically, but the number
of \emph{non-zero} elements grows only linearly; the Hessian
becomes increasingly sparse.  The row and column indices of the
non-zero elements comprise the \emph{sparsity pattern} of the Hessian,
and are typically known in advance, before computing the values of
those elements.  \proglang{R} packages such as \pkg{trustOptim}
\citep{R_trustOptim}, \pkg{sparseMVN} \citep{R_sparseMVN} and
\pkg{ipoptr} \citep{R_ipoptr} have the capability to accept Hessians
in a compressed sparse format.

The \pkg{sparseHessianFD} package is a tool for estimating
sparse Hessians using finite differences of gradients.  Section \ref{sec:numdiff} will
cover the specifics, but the basic idea is as follows.  Consider a
function $f(x)$, its gradient $\nabla{f(x)}$, and its Hessian
$\hess{}{f(x)}$, for $x\in\mathbb{R}^M$.  Define the derivative vector
as the transpose of the gradient, and a vector of partial derivatives, so
${\parD{}{}{f(x)}=\nabla f(x)^\top=\left(\parD{}{1}{},\mathellipsis,\parD{}{M}{}\right)}$.   (Throughout the paper, we will try
to reduce notational clutter by referring to the derivative and
Hessian as $\parD{}{}{}$ and $\hess{}{}$, respectively, without the
$f(x)$ symbol).  Let $e_m$ be a vector of zeros, except with a 1 in the $m$th
element, and let $\delta$ be a sufficiently small scalar
constant. A linear approximation to the $m$th column of the Hessian is ${\hess{m}{}\approx\left(\nabla f(x+\delta
  e_m)- \nabla f(x)\right)/\delta}$.  Estimating a dense Hessian in this way involves at least $M+1$ calculations of the
gradient: one for the gradient at $x$, and one after perturbing each
of the $M$ elements of $x$, one at a time.\footnote{More accurate
  approximations require more gradient evaluations.  We will consider
  only forward finite differences.}  In a sparse matrix, most of the
elements are constrained to zero.  Depending
on the sparsity pattern of the Hessian, those constraints may let us recover
the Hessian with fewer gradient evaluations by perturbing multiple
elements of $x$ together.  For some sparsity patterns, estimating a Hessian in
this way can be profoundly efficient.  In fact, for the hierarchical models that we consider in
this paper, the number of gradient evaluations does not increase with
the number of additional heterogeneous units.

The package defines the \class{sparseHessianFD} class, whose initializer
requires the user to provide functions that compute the objective
function and the gradient, as well as the sparsity pattern of the
Hessian.  Some users may find those requirements burdensome, and we
emphasize that there may be some applications for
which \pkg{sparseHessianFD} is not an appropriate package to use. To
extract the maximum benefit from using \pkg{sparseHessianFD}, we need
to accept some conditions and assumptions.

\begin{enumerate}
  \item Preferred alternatives to computing the Hessian are not
  available.  Finite differencing is not generally a ``first choice''
  method.  Deriving a gradient or Hessian symbolically, and writing a
  subroutine to compute it, will give an exact answer, but might be
  tedious or difficult to implement. Algorithmic differentiation (AD) is
  probably the most efficient method, but requires specialized
  libraries that, at this moment, are not yet broadly available in \proglang{R}.
 \pkg{sparseHessianFD} makes the most sense when the gradient is easy to compute, but the Hessian is
  not.
  \item The application can tolerate the approximation error in the Hessian
  that comes with finite differencing methods.
\item The objective function $f(x)$ is twice differentiable, and can be
  computed ``quickly and easily,'' even for a large number of
  variables.  We leave the definition of ``quickly and easily''
  intentionally murky, since no method of differentiation can overcome
  pathologies in a function that itself is hard to compute.
\item The gradient can
  be computed quickly, easily and \emph{exactly} (within machine
  precision).   We do not recommend using finite differenced gradients
  when computing finite differenced Hessians.  The
  approximation errors will be compounded, and the time
  complexity of computing a gradient grows with the number of
  variables.
\item The sparsity pattern is known in advance, and does not depend
    on the values of the variables.
\end{enumerate}

Our perspective on the requirements to use \pkg{sparseHessianFD} is
that deriving a vector of first
derivatives, and writing \proglang{R} functions to compute them, is a
lot easier than doing the same for a matrix of second derivatives, and
more accurate than computing second-order approximations from the
objective function. Even when we have derived the Hessian symbolically, in practice it may still be faster to estimate the Hessian
using \pkg{sparseHessianFD} than coding it directly.   These are the
situations in which \pkg{sparseHessianFD} adds the most value to the
statistician's toolbox.  If AD software is available to compute the gradient, then
it is probably available for sparse Hessians as well, and
\pkg{sparseHessianFD} would not be needed.  Without AD-computed or
symbolically derived Hessians, \pkg{sparseHessianFD} is the best
available option.

The only functions and methods the end user should need to use are the
\class{sparseHessianFD} initializer, methods that return the Hessian
in a sparse compressed format, and perhaps some utility functions that
simplify the construction of the sparsity pattern.  The class also
defines methods that partition
the variables into groups that can be perturbed together in a finite
differencing step, and recovers the elements of the Hessian via linear
substitution.  Those methods perform most of the work, but should be
invisible to the user.

This article proceeds as follows.  First, we present
some background information about numerical differentiation, and
sparse matrices in \proglang{R}, in Section~\ref{sec:background}.  In
Section ~\ref{sec:using}, we explain how to use the package.
Section~\ref{sec:algorithms} explains the underlying algorithms, and
Section~\ref{sec:timing} demonstrates the scalability of those algorithms.

\section[Background]{Background}\label{sec:background}

Before describing how to use the package, we
present two short background notes.  The first note is an informal
mathematical explanation of numerical estimation of the Hessian
matrix, with an illustration of how the number of gradient estimates
can be reduced by exploiting the sparsity pattern and symmetric
structure.  This note borrows heavily from, and use the notation in,
\citet[Chapter 6]{MagnusNeudecker2007}. The
second note is a summary of some of the sparse matrix classes that are
defined in the \pkg{Matrix} package \citep{R_Matrix}, which are used
extensively in \pkg{sparseHessianFD}.

\subsection[Numerical differentiation of sparse Hessians]{Numerical
  differentiation of sparse Hessians}\label{sec:numdiff}

The partial derivative of a scalar-valued function $f(x)$ with respect to $x_j$ (the $j$th
component of $x$) is defined as
\begin{align}
  \label{eq:defParD}
\parD{}{j}{f(x)}=\lim\limits_{\delta\to 0}\frac{f(x+\delta e_j)-f(x)}{\delta}
\end{align}
For a sufficiently small $\delta$, this definition allows for a
linear approximation to $\parD{}{j}{f(x)}$.  The derivative of $f(x)$
is the vector of all $M$ partial derivatives.
\begin{align}
  \parD{}{}{f(x)}=\left(\parD{}{1}{f(x)},\mathellipsis,\parD{}{M}{f(x)}\right)
  \end{align}
 The \emph{gradient} is defined as $\nabla f(x)=\parD{}{}{f(x)}^\top$.

We define the second-order partial derivative as
\begin{align}
  \label{eq:14}
  \parD{2}{jk}{}=\lim\limits_{\delta\to 0}\frac{\parD{}{j}{f(x+\delta e_k)}-\parD{}{j}{f(x)}}{\delta}
\end{align}
and the Hessian as
\begin{align}
  \label{eq:15}
  \hess{}{f(x)}=
  \begin{pmatrix}
    \parD{2}{11}{}&  \parD{2}{12}{}&  \mathellipsis &  \parD{2}{1M}{}\\
    \parD{2}{21}{}&  \parD{2}{22}{}&  \mathellipsis &  \parD{2}{2M}{}\\
    \vdots&\vdots&&\vdots\\
    \parD{2}{M1}{}&  \parD{2}{M2}{}&  \mathellipsis &  \parD{2}{MM}{}
    \end{pmatrix}
\end{align}

The Hessian is symmetric, so $\parD{2}{ij}{}=\parD{2}{ji}{}$.

To estimate the $m$th column of $\hess{}{}$, we again choose a
sufficiently small $\delta$, and compute
\begin{align}
  \label{eq:1}
  \hess{m}{f(x)}&\approx\frac{\parD{}{}{f(x+\delta e_m)}-\parD{}{}{f(x)}}{\delta}
\end{align}

For $M=2$, our estimate of a general $\hess{}{f(x)}$ would be
\begin{align}
  \label{eq:FDhess2}
  \hess{}{f(x)}&=
  \begin{pmatrix}
    \parD{}{1}{f(x_1+\delta, x_2)}-\parD{}{1}{f(x_1,x_2)}& \parD{}{1}{f(x_1,x_2+\delta)}-\parD{}{1}{f(x_1,x_2)}\\
        \parD{}{2}{f(x_1+\delta, x_2)}-\parD{}{2}{f(x_1,x_2)}&  \parD{}{2}{f(x_1,x_2+\delta)}-\parD{}{2}{f(x_1,x_2)}
    \end{pmatrix}/\delta
\end{align}

This estimate requires three evaluations of the gradient to get
${\parD{}{}{f(x_1,x_2)}}$, ${\parD{}{}{f(x_1+\delta,x_2)}}$, and
${\parD{}{}{f(x_1,x_2+\delta)}}$.

Now suppose that the Hessian is sparse, and that the off-diagonal
elements are zero.  That means that
\begin{align}
  \parD{}{1}{f(x_1,x_2+\delta)}-\parD{}{1}{f(x_1,x_2)}&=0\label{eq:I1}\\
    \parD{}{2}{f(x_1+\delta,x_2)}-\parD{}{2}{f(x_1,x_2)}&=0\label{eq:I2}
\end{align}
If the identity in Equation~\ref{eq:I1} holds for $x_1$, it must also
hold for $x_1+\delta$, and if Equation~\ref{eq:I2} holds for $x_2$, it
must also hold for $x_2+\delta$.  Therefore,
\begin{align}
  \label{eq:FDhess2sp}
  \hess{}{f(x)}&=
  \begin{pmatrix}
    \parD{}{1}{f(x_1+\delta, x_2+\delta)}-\parD{}{1}{f(x_1,x_2)}&0\\
        0&  \parD{}{2}{f(x_1+\delta,x_2+\delta)}-\parD{}{2}{f(x_1,x_2)}
    \end{pmatrix}/\delta
\end{align}

Only two gradients,
${\parD{}{}{f(x_1,x_2)}}$ and ${\parD{}{}{f(x_1+\delta,x_2+\delta)}}$,
are needed.
Being able to reduce the number of gradient evaluations from 3 to 2 depends on knowing that
the cross-partial derivatives are zero.

\Citet{CurtisPowellReid1974} describe a method of estimating sparse
Jacobian matrices by perturbing groups of variables together.
\Citet{PowellToint1979} extend this idea to the
general case of sparse Hessians.  This method partitions
the decision variables into $C$ mutually exclusive groups so that the number of gradient
evaluations is reduced.  Let $\Mat{G}$ be a $M\times C$
matrix, where $\Mat{G}_{mc}=\delta$ if variable $m$ belongs to group $c$, and zero
otherwise.  Define $\Mat{G}_c$ as the $c$th column of $\Mat{G}$.

Next, let $\Mat{Y}$ be a $M\times C$ matrix, where each column is a
difference in gradients.
\begin{align}
  \label{eq:Yg}
  \Mat{Y}_c&=\nabla f(x+G_c)-\nabla f(x)
\end{align}

If $C=M$, then $\Mat{G}$ is a diagonal matrix with $\delta$ in each
diagonal element.  The matrix equation $\hess{}{}\Mat{G}=\Mat{Y}$ represents the linear approximation
$\hess{im}{}\delta\approx y_{im}$, and we can solve for all elements of $\hess{}{}$
just by computing $Y$. But if $C<M$,
there must be at least one $\Mat{G}_c$ with $\delta$ in at least two
rows. The corresponding column $\Mat{Y}_c$ is computed by perturbing multiple variables at
once, so we cannot solve for any $\hess{im}{}$
without further constraints.

These constraints come from the sparsity pattern and symmetry
of the Hessian. Consider an example with the following values and
sparsity pattern.
\begin{align}
  \label{eq:7}
 \hess{}{f(x)}= \begin{pmatrix}
    h_{11}&0&h_{31}&0&0\\
    0&h_{22}&0&h_{42}&0\\
    h_{31}&0&h_{33}&0&h_{53}\\
    0&h_{42}&0&h_{44}&0\\
    0&0&h_{53}&0&h_{55}
  \end{pmatrix}
\end{align}

Suppose $C=2$, and define group membership of the five variables through the following $\Mat{G}$ matrix.
\begin{align}
  \label{eq:7}
 \Mat{G}^\top= \begin{pmatrix}
 \delta&\delta&0&0&\delta\\
 0&0&\delta&\delta&0
 \end{pmatrix}
\end{align}
Variables 1, 2 and 5 are in group 1, while variables 3 and 4 are in
group 2.

Next, compute the columns of $\Mat{Y}$ using Equation~\ref{eq:Yg}.  We now
have the following system of linear equations from $\hess{}{}\Mat{G}=\Mat{Y}$.
\begin{align}
  \label{eq:11}
  \begin{aligned}[c]
  h_{11}&=y_{11}\\
  h_{22}&=y_{21}\\
  h_{31}+h_{53}&=y_{31}\\
  h_{42}&=y_{41}\\
  h_{55}&=y_{51}\\
  \end{aligned}
  \qquad
  \begin{aligned}[c]
  h_{31}&=y_{12}\\
  h_{42}&=y_{22}\\
  h_{33}&=y_{32}\\
  h_{44}&=y_{42}\\
  h_{53}&=y_{52}
  \end{aligned}
\end{align}
Note that this system is overdetermined.  Both $h_{31}=y_{12}$ and $h_{53}=y_{52}$
can be determined directly, but $h_{31}+h_{53}=y_{31}$ may not
necessarily hold, and $h_{42}$
could be either $y_{41}$ or $y_{22}$.  \Citet{PowellToint1979} prove
that it is sufficient to solve $\hessLT{}{}\Mat{G}=\Mat{Y}$ instead
via a substitution method, where $\hessLT{}{}$ is the lower
triangular part of $\hess{}{}$.  This has the effect of removing the equations
$h_{42}=y_{22}$ and $h_{31}=y_{12}$ from the system, but retaining
$h_{53}=y_{52}$.  We can then solve for
$h_{31}=y_{31}-y_{52}$. Thus, we have determined a
$5\times 5$ Hessian with only three gradient evaluations, in
contrast with the six that would have been needed had $\hess{}{}$ been treated
as dense.

The \pkg{sparseHessianFD} algorithms assign variables to groups
before computing the values of the Hessian.  This is why the sparsity
pattern needs to be provided in advance.  If a non-zero element is
omitted from the sparsity pattern, the resulting estimate of the
Hessian will be incorrect.  The only problems with erroneously including a zero
element in the sparsity pattern are a possible lack of efficiency
(e.g., an increase in the number of gradient evaluations), and that
the estimated value might be close to, but not exactly, zero. The
algorithms for assigning decision variables to groups, and for
extracting nonzero Hessian elements via substitution, are described in
Section \ref{sec:algorithms}.

\subsection[Sparse matrices in R]{Sparse matrices and the \pkg{Matrix}
  package}\label{sec:Rsparse}

The \pkg{sparseHessianFD} package uses the sparse matrix classes that are defined
in the \pkg{Matrix} package \citep{R_Matrix}.  All of these classes are subclasses of
\class{sparseMatrix}. Only the row and column
indices (or pointers to them), the non-zero values, and some
meta-data, are stored; unreferenced elements are assumed to be
zero. Class names, summarized in Table~\ref{tab:MatrixClasses}, depend on the data type, matrix structure,
and storage format. Values in numeric and logical matrices correspond to the
\proglang{R} data types of the same names.  Pattern matrices contain
row and column information for the non-zero elements, but no
values. The storage format refers to the internal ordering of the
indices and values, and the layout defines a matrix as symmetric (so
duplicated values are stored only once), triangular, or general.  The
levels of these three factors determine the prefix of letters in each
class name. For example, a triangular sparse matrix of numeric (double
precision) data, stored in
column-compressed format, has a class \class{dtCMatrix}.

\begin{table}[tbh]\centering
  \begin{tabular}{ll|ccc}
      \toprule
   Storage&Layout&\multicolumn{3}{c}{Data type}\\
    &&numeric&logical&pattern\\
    \midrule
    Triplet&general&\class{dgTMatrix}&\class{lgTMatrix}&\class{ngTMatrix}\\
    &triangular&\class{dtTMatrix}&\class{ltTMatrix}&\class{ntTMatrix}\\
      &symmetric&\class{dsTMatrix}&\class{lsTMatrix}&\class{nsTMatrix}\\
    \midrule
        Row-compressed&general&\class{dgRMatrix}&\class{lgRMatrix}&\class{ngRMatrix}\\
    &triangular&\class{dtRMatrix}&\class{ltRMatrix}&\class{ntRMatrix}\\
      &symmetric&\class{dsRMatrix}&\class{lsRMatrix}&\class{nsRMatrix}\\
    \midrule
            Column-compressed&general&\class{dgCMatrix}&\class{lgCMatrix}&\class{ngCMatrix}\\
    &triangular&\class{dtCMatrix}&\class{ltCMatrix}&\class{ntCMatrix}\\
      &symmetric&\class{dsCMatrix}&\class{lsCMatrix}&\class{nsCMatrix}\\
\bottomrule
  \end{tabular}
    \caption{Class names for sparse matrices, as defined in the
    Matrix package.}\label{tab:MatrixClasses}
\end{table}

\pkg{Matrix} also defines some other classes of
sparse and dense matrices that we will not discuss here. The
\pkg{Matrix} package uses the \func{as} function to convert sparse
matrices from one format to another, and to convert a \pkg{base}
\proglang{R} matrix to one of the \pkg{Matrix} classes.

The distinction among sparse matrix classes is important because
\pkg{sparseHessianFD}'s \method{hessian} method returns a
\class{dgCMatrix}, even though the Hessian is symmetric. Depending on how the
Hessian is used, it might be useful to coerce the Hessian into a
\class{dsCMatrix} object.  Also, the utility functions in Table~\ref{tab:helpers} expect or return certain classes of matrices, so
some degree of coercion of input and output might be
necessary. Another useful \pkg{Matrix} function is \func{tril}, which
extracts the lower triangle of a general or symmetric matrix.

\section[Using the package]{Using the package}\label{sec:using}

In this section, we demonstrate how to use the \pkg{sparseHessianFD},
using a hierarchical binary choice model as an example.  Then, we
discuss the sparsity pattern of the Hessian, and estimate the Hessian values.

\subsection[Example: hierarchical binary choice]{Example model: hierarchical binary choice}\label{sec:example}

Suppose we have a dataset of $N$ households, each with $T$
 opportunities to purchase a particular product.  Let $y_i$ be the
 number of times household $i$ purchases the product, out of the $T$
 purchase opportunities, and let $p_i$ be the probability of
 purchase.  The heterogeneous parameter $p_i$ is the same for all $T$
 opportunities, so $y_i$ is a binomial random variable.  Define each $p_i$ such that it depends on both $k$ continuous covariates
 $x_i$, and a heterogeneous coefficient vector $\beta_i$.
\begin{align}
  p_i=\frac{\exp(x_i'\beta_i)}{1+\exp(x_i'\beta_i)},~i=1 ... N
\end{align}

The coefficients are distributed across the population of households
following a multivariate normal distribution with mean $\mu$ and
covariance $\Mat{\Sigma}$.   Assume that we know $\Mat{\Sigma}$, but not $\mu$.
Instead, place a multivariate normal prior on $\mu$, with mean $0$ and
covariance $\Mat{\Omega}$.  Thus, each $\beta_i$, and $\mu$ are
$k-$dimensional vectors, and the total number of unknown variables in the model is $(N+1)k$.

The log posterior density, ignoring any normalization constants, is
\begin{align}
  \label{eq:LPD}
  \log \pi(\beta_{1:N},\mu|\Mat{Y}, \Mat{X}, \Mat{\Sigma},\Mat{\Omega})=\sum_{i=1}^N\left(p_i^{y_i}(1-p_i)^{T-y_i}
  -\frac{1}{2}\left(\beta_i-\mu\right)^\top\Mat{\Sigma}^{-1}\left(\beta_i-\mu\right)\right)
-\frac{1}{2}\mu^\top\Mat{\Omega}^{-1}\mu
\end{align}

All of the $\beta_i$ are correlated with
$\mu$.  Thus, $\hess{\beta_{ik},\mu_k}{}\neq 0$ for all $i$. Since the $\beta_i$ are independently distributed, and the $y_i$ are
conditionally independent, the cross-partial derivatives
$\hess{\beta_i,\beta_j}{}=\parD{2}{\beta_i, \beta_j}{}=0$ for all $i\neq
j$.  When $N$ is much greater than $k$, the Hessian of the log
posterior density will be sparse.

\subsection[Sparsity patterns]{Sparsity patterns}\label{sec:sparsity}

<<setup2, echo=FALSE>>=
N <- 5
k <- 2
nv1 <- (N+1)*k
nels1 <- nv1^2
nnz1 <- (N+1)*k^2 + 2*N*k^2
nnz1LT <- (N+1)*k*(k+1)/2 + N*k^2
Q <- 1000
nv2 <- (Q+1)*k
nels2 <- nv2^2
nnz2 <- (Q+1)*k^2 + 2*Q*k^2
nnz2LT <- (Q+1)*k*(k+1)/2 + Q*k^2
options(scipen=999)
@

The sparsity pattern depends on how the variables are
ordered within the vector. One such ordering is to group all of the
coefficients for each unit together.
\begin{align}
\beta_{11},\dotsc,\beta_{1k},\beta_{21},\dotsc,\beta_{2k},~\dotsc~,\beta_{N1},\dotsc,\beta_{Nk},\mu_1,\dotsc,\mu_k
\end{align}
In this case, the Hessian has a ``block-arrow'' structure.  For example,
if $N=\Sexpr{N}$ and $k=\Sexpr{k}$, then there are \Sexpr{nv1} total
variables, and the Hessian will have the pattern in Figure~\ref{fig:blockarrow}.

Another possibility is to group coefficients for each covariate
together.
\begin{align}
\beta_{11},\dotsc,\beta_{N1},\beta_{12},\dotsc,\beta_{N2},~\dotsc~,\beta_{1k},\dotsc,\beta_{Nk},\mu_1,\dotsc,\mu_k
\end{align}
Now the Hessian has an "banded" sparsity pattern, as in Figure~\ref{fig:banded}.
\begin{figure}[tbp]
  \begin{subfigure}[b]{.5\textwidth}
<<blockarrow, echo=FALSE>>=
Mat <- as(kronecker(diag(N),matrix(1,k,k)),"sparseMatrix")
Mat <- rBind(Mat, Matrix(1,k,N*k))
Mat <- cBind(Mat, Matrix(1, k*(N+1), k))
printSpMatrix(as(Mat,"nMatrix"))
@
\caption{A ``block-arrow'' sparsity pattern.}\label{fig:blockarrow}
\end{subfigure}
\begin{subfigure}[b]{.5\textwidth}
<<banded, echo=FALSE>>=
Mat <- kronecker(Matrix(1,k,k), diag(N))
Mat <- rBind(Mat, Matrix(1,k,N*k))
Mat <- cBind(Mat, Matrix(1, k*(N+1), k))
printSpMatrix(as(Mat,"nMatrix"))
@
\caption{A ``banded'' sparsity pattern.}\label{fig:banded}
\end{subfigure}
\caption{Two examples of sparsity patterns for a hierarchical model.}\label{fig:patterns}
\end{figure}

In both cases, the number of non-zeros is the same.   There are
\Sexpr{nels1} elements in this symmetric matrix, but only \Sexpr{nnz1} are
non-zero, and only \Sexpr{nnz1LT} values are unique.  Although the reduction in
RAM from using a sparse matrix structure for the Hessian may be
modest, consider what would happen if $N=\Sexpr{scales::comma(Q)}$ instead.  In that case,
there are \Sexpr{scales::comma(nv2)} variables in the problem, and more than $\Sexpr{scales::comma(floor(nels2/10^6))}$ million
elements in the Hessian.  However, only $\Sexpr{scales::comma(nnz2)}$ of those elements are
non-zero.  If we work with only the lower triangle of the Hessian we only need to work with
only \Sexpr{scales::comma(nnz2LT)} values.

The sparsity pattern required by \pkg{sparseHessianFD} consists of the row and column
indices of the non-zero elements in the \emph{lower triangle} the
Hessian, and it is the responsibility of the user to ensure that
the pattern is correct.  In practice, rather than trying to keep track of the row and column indices
directly, it might be easier to construct a pattern matrix first,
check visually that the matrix has the right pattern, and then extract
the indices.

The package defines utility functions (Table~\ref{tab:helpers}) to convert
between sparse matrices, and the vectors of row and column indices
required by the \class{sparseHessianFD} initializer.

\begin{table}[ptb]
  \centering
  \setlength\extrarowheight{.25em}
  \begin{tabularx}{1.0\linewidth}{lX}
    \toprule
   \func{Matrix.to.Coord} & Returns a list of vectors containing row
                            and column indices of the non-zero
                            elements of a matrix.\\
    \func{Matrix.to.Pointers}&Returns indices and pointers from a
                               sparse matrix.\\
    \func{Coord.to.Pointers}&Converts list of row and column indices
                              (triplet format) to a list of indices
                              and pointers (compressed format).\\
  \bottomrule
  \end{tabularx}
  \caption{sparseHessianFD matrix conversion functions.}
  \label{tab:helpers}
\end{table}

The \func{Matrix.to.Coord} function extracts
row and column indices from a sparse matrix. The following code
constructs a logical block diagonal matrix, converts it to a sparse
matrix, and prints the sparsity pattern of its lower triangle.

\begin{minipage}{\columnwidth}
<<>>=
library("sparseHessianFD")
bd <- kronecker(diag(3), matrix(TRUE,2,2))
Mat <- as(bd, "nMatrix")
printSpMatrix(tril(Mat))
mc <- Matrix.to.Coord(tril(Mat))
mc
@
\end{minipage}

\vspace{1em}
To check that a proposed sparsity pattern represents the
intended matrix visually, use the \pkg{Matrix} \func{sparseMatrix} constructor.

<<>>=
pattern <- sparseMatrix(i=mc$rows, j=mc$cols)
printSpMatrix(pattern)
@

\subsection[The sparseHessianFD class]{The \class{sparseHessianFD} class}

The function \func{sparseHessianFD} is an initializer that returns a reference to a
\class{sparseHessianFD} object.  The initializer determines an
appropriate permutation and partitioning
of the variables, and performs some additional validation tests.  The arguments to the initializer are
described in Table~\ref{tab:init}.

{\setlength{\extrarowheight}{.25em}
\begin{table}[ptb]
\begin{tabularx}{\linewidth}{>{\bfseries}r X}
  \toprule
x& A numeric vector, with length $M$ at which the object will be
  initialized and tested.\\
fn,gr& \proglang{R} Functions that return the value of the
  objective function, and its gradient. The first argument is the numeric
  variable vector.  Other named arguments can be passed to \func{fn}
  and {gr} as well (see the \funcarg{...} argument below).\\
rows, cols& Sparsity pattern: integer vectors of the row and column indices of
  the non-zero elements in the \emph{lower triangle} of the Hessian.\\
direct& This argument is deprecated, and is included only for
  backwards compatibility with earlier versions.\\
delta& The perturbation amount for finite differencing of the
  gradient to compute the Hessian (the $\delta$ in Section \ref{sec:numdiff}).  Defaults to
  \code{sqrt(.Machine\$double.eps)}.\\
index1& If \variable{TRUE} (the default), \funcarg{row} and \funcarg{col} use one-based
  indexing.  If \variable{FALSE}, zero-based indexing is used.\\
...& Additional arguments to be passed to \func{fn} and \func{gr}.\\
 \bottomrule
\end{tabularx}
\caption{Arguments to the sparseHessianFD initializer.}\label{tab:init}
\end{table}
}

To create a \class{sparseHessianFD} object, just call
\func{sparseHessianFD}.  Applying the default values for the optional
arguments,
<<eval=FALSE>>=
obj <- sparseHessianFD(x, fn, gr, rows, cols, ...)
@
where \funcarg{...} represents all other named arguments that are
passed to \funcarg{fn} and \funcarg{gr}.

The \func{fn}, \func{gr} and \func{hessian} methods respectively evaluate the
function, gradient and Hessian at a variable vector $x$.  The
\func{fngr} method returns the function and gradient as a list.  The
\func{fngrhs} method includes the Hessian as well.

<<eval=FALSE>>=
f <- obj$fn(x)          ## returns numeric
df <- obj$gr(x)         ## returns numeric vector
hess <- obj$hessian(x)  ## returns dgCMatrix
fngr <- obj$fngr(x)     ## returns list
fngrhs <- obj$fngrhs(x) ## returns list
@

\subsection[An example]{An example}

Now we can estimate the Hessian for the
log posterior density of the model from Section \ref{sec:example}.
For demonstration purposes, \pkg{sparseHessianFD} includes functions that compute the value
(\func{binary.f}), the gradient (\func{binary.grad}) and the Hessian
(\func{binary.hess}) of this model. We will treat the result from \func{binary.hess} as a
  ``true'' value against which we will compare the numerical
  estimates.

To start, we load the data, set some dimension parameters, set prior
values for $\Mat{\Sigma}^{-1}$ and $\Mat{\Omega}^{-1}$, and simulate a
vector of variables at which to evaluate the function. The \func{binary.f} and \func{binary.grad} functions take the data and
priors as lists.  The \func{data(binary)} call adds the appropriate data
list to the environment, but we need to construct the prior list
ourselves.

<<binaryInit>>=
set.seed(123)
data("binary")
str(binary)
N <- length(binary[["Y"]])
k <- NROW(binary[["X"]])
T <- binary[["T"]]
nvars <- as.integer(N*k + k)
priors <- list(inv.Sigma = rWishart(1,k+5,diag(k))[,,1],
               inv.Omega = diag(k))
@

This dataset represents the simulated choices for $N= \Sexpr{N}$ customers
over $T= \Sexpr{T}$ purchase opportunties, where the probability of purchase
is influenced by $k= \Sexpr{k}$ covariates.

The next code chunk evaluates the ``true'' value, gradient and
Hessian.  The \funcarg{order.row} argument tells the function whether
the variables are ordered by household (\variable{TRUE}) or by covariate
(\variable{FALSE}). If \funcarg{order.row} is \variable{TRUE}, then the Hessian
will have a banded pattern.  If \funcarg{order.row} is \variable{FALSE}, then the
Hessian will have a block-arrow pattern.

<<trueValues>>=
P <- rnorm(nvars)
order.row <- FALSE
true.f <- binary.f(P, binary, priors, order.row=order.row)
true.grad <- binary.grad(P, binary, priors, order.row=order.row)
true.hess <- binary.hess(P, binary, priors, order.row=order.row)
@

The sparsity pattern of the Hessian is specified by two integer
vectors:  one each for the row and column indices of the non-zero
elements of the lower triangule of the Hessian.  For this example, we happen to have a
matrix with the same sparsity pattern
of the Hessian we are trying to compute, so we can use the
\func{Matrix.to.Coord} function to extract the appropriate index
vectors.  In practice, it is more likely that we would need to determine the row
and column indices directly, through our knowledge of the structure of
the problem.  For a hierarchical model, we can create a block-arrow
pattern matrix using either the \func{Matrix::bdiag} or \func{kronecker} functions to
create a block diagonal matrix, and concatenate dense rows and columns to the margins.

<<binaryRowsCols>>=
pattern <- Matrix.to.Coord(tril(true.hess))
str(pattern)
@

Finally, we create an instance of a \class{sparseHessianFD} object,
evaluate the function, gradient and Hessian, and compare the output to
the true values.

<<usingSparseHessianFD>>=
obj <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
       rows=pattern[["rows"]], cols=pattern[["cols"]],
       data=binary, priors=priors, order.row=order.row)
f <- obj$fn(P)
all.equal(f, true.f)
gr <- obj$gr(P)
all.equal(gr, true.grad)
hs <- obj$hessian(P)
all.equal(hs, true.hess)
@

\section[Algorithms]{Algorithms}\label{sec:algorithms}

In this section, we explain how \pkg{sparseHessianFD} works. The
algorithms are adapted from \citet{ColemanGarbow1985}, who provided
Fortran implementations as \citet{ColemanGarbow1985b}.  Earlier
versions of \pkg{sparseHessianFD} included licensed copies of the
\citet{ColemanGarbow1985b} code, on which the current version no
longer depends. Although newer partitioning algorithms have been proposed (e.g.,
\citealp{GebremedhinManne2005,GebremedhinTarafdar2009}), mainly in the
context of automatic differentiation, we have chosen to implement
established algorithms that are known to work well, and are likely
optimal for the hierarchical models that many statisticians will encounter.

\subsection[Partitioning the variables]{Partitioning the variables}\label{sec:coloring}

Finding consistent, efficient partitions can be
characterized as a vertex coloring problem from graph theory \citep{ColemanMore1984}.  In this
sense, each variable is a vertex in an undirected graph, and an edge connects two
vertices $i$ and $j$ if and only if ${\hess{ij}{f(x)}\neq 0}$.  The
sparsity pattern of the Hessian is the adjacency matrix of the graph.
By ``color,'' we mean nothing more than group assignment; if a
variable is in a group, then its vertex has the color associated with
that group.  A ``proper'' coloring of a graph is one in which two
vertices with a common edge do not have the same
color. \Citet{ColemanMore1984} define a ``triangular coloring'' as
a proper coloring with the additional condition that common
neighbors of a vertex do not have the same color.  A triangular
coloring is a special case of an ``cyclic coloring,'' in which any cycle in the graph
uses at least three colors \citep{GebremedhinTarafdar2007}.

An ``intersection set'' contains characteristics that are common to two vertices, and
an ``intersection graph'' connects vertices whose intersection set is
not empty.  In our context, the set in question is the row indices of
the non-zero elements in each column of $\hessLT{}{}$.  In the
intersection graph, two vertices are connected if the corresponding
columns in $\hessLT{}{}$ have at least one non-zero element in a common row.

\Citet{PowellToint1979} write that a partitioning is consistent with a
substitution method if and only if no columns
of the of lower triangle of the Hessian that are
in the same group have a non-zero element in the same
row.  An equivalent statement is that no two adjacent vertices in the
intersection graph can have the same color.  Thus, we can partition
the variables by creating a proper coloring of the intersection graph
of $\hessLT{}{}$.

This intersection graph, and the number of colors needed to
color it, are not invariant to
permutation of the rows and columns of $\hess{}{}$.  Let $\pi$
represent such a permutation, and let $\hessLT{\pi}{}$ be the lower
triangle of $\pi\hess{}{}{\pi^\top}$.    \Citet[Theorem
6.1]{ColemanMore1984} show that a coloring is triangular if and only
if it is also a proper coloring of the intersection
graph of $\hessLT{\pi}{}$.  Furthermore,
\citet{ColemanCai1986} prove that a partitioning is consistent with a
substitution method if and only if it is an acyclic coloring of the
graph of the sparsity pattern of the Hessian.   Therefore,
finding an optimal partitioning of the variables involves finding an
optimal combination of a permutation $\pi$, and coloring algorithm
for the intersection graph of $\hessLT{\pi}{}$.

These ideas are illustrated in Figures~\ref{fig:graph1} and
\ref{fig:graph2}.  Figure~\ref{fig:graph1adj} shows the sparsity pattern of the lower triangle of a
Hessian as an adjacency matrix, and Figure~\ref{fig:graph1pic} is the
associated graph with a proper vertex coloring.  Every column (and
thus, every pair of columns) in Figure~\ref{fig:graph1adj} has a non-zero element
in row 7, so there are no non-empty intersection sets across the
columns.  All vertices are connected to each other in the
intersection graph (Figure~\ref{fig:graph1int}), which requires seven colors for a proper coloring.  Estimating a
sparse Hessian with this partitioning scheme would be no more
efficient than treating the Hessian as if it were dense.
\begin{figure}[ptb]
  \begin{subfigure}[b]{.32\textwidth}\centering
  \begin{tabular}{r|ccccccc}
   & 1&2&3&4&5&6&7\\
    \hline
    1& 1&&&&&&\\
    2&1&1&&&&&\\
    3&0&0&1&&&&\\
    4&0&0&1&1&&&\\
    5&0&0&0&0&1&&\\
    6&1&0&0&0&1&1&\\
    7&1&1&1&1&1&1&1
  \end{tabular}
  \caption{Adjacency matrix.}\label{fig:graph1adj}
\end{subfigure}
\begin{subfigure}[b]{.32\textwidth}\centering
\iftoggle{tikz}{
\begin{tikzpicture}
\tikzstyle{every node}=[draw, shape=circle]
\node at (0,0) [fill=green!20] (v7) {7};
\node at (0:1) [fill=blue!20] (v1) {1};
\node at (60:1) [fill=red!20] (v2) {2};
\node at (120:1) [fill=blue!20] (v3) {3};
\node at (180:1) [fill=red!20] (v4) {4};
\node at (240:1) [fill=blue!20] (v5) {5};
\node at (300:1) [fill=red!20] (v6) {6};
\draw
(v1) -- (v7)  (v2) -- (v7)
(v3) -- (v7)  (v4) -- (v7)
(v5) -- (v7)  (v6) -- (v7)
(v1) -- (v2)  (v3) -- (v4)
(v5) -- (v6);
\end{tikzpicture}
}{} %% end toggle
\caption{Sparsity graph.}\label{fig:graph1pic}
\end{subfigure}
\begin{subfigure}[b]{.32\textwidth}\centering
\iftoggle{tikz}{
\begin{tikzpicture}
\tikzstyle{every node}=[draw, shape=circle]
\node at (0,0) [fill=green!20] (v7) {7};
\node at (0:1) [fill=blue!20] (v1) {1};
\node at (60:1) [fill=red!20] (v2) {2};
\node at (120:1) [fill=yellow!20] (v3) {3};
\node at (180:1) [fill=purple!20] (v4) {4};
\node at (240:1) [fill=brown!20] (v5) {5};
\node at (300:1) [fill=white!20] (v6) {6};
\draw
(v1) -- (v7)  (v2) -- (v7)  (v3) -- (v7)
(v4) -- (v7)  (v5) -- (v7)  (v6) -- (v7)
(v1) -- (v6)  (v2) -- (v6)  (v3) -- (v6)
(v4) -- (v6)  (v5) -- (v6)
(v1) -- (v5)  (v2) -- (v5)  (v3) -- (v5)
(v4) -- (v5)
(v1) -- (v4)  (v2) -- (v4)  (v3) -- (v4)
(v1) -- (v3)  (v2) -- (v3)
(v1) -- (v2);
\end{tikzpicture}
}{} %% end toggle
\caption{Intersection graph.}\label{fig:graph1int}
\end{subfigure}
\caption{Unpermuted matrix.}\label{fig:graph1}
\end{figure}

Now suppose we were to rearrange $\hess{}{}$ so the last row and
and column were moved to the front.  In Figure~\ref{fig:graph2adj}, all columns share at least one non-zero row with
the column for variable 7, but variable groups $\{2,4,6\}$ and
$\{1,3,5\}$ have empty intersection sets.  The intersection graph in
Figure~\ref{fig:graph2pic} has fewer edges, and can be colored
with only three colors.
\begin{figure}[ptb]
  \begin{subfigure}[m]{.32\textwidth}\centering
  \begin{tabular}{r|ccccccc}
   & 7&1&2&3&4&5&6\\
    \hline
    7& 1&&&&&&\\
    1&1&1&&&&&\\
    2&1&1&1&&&&\\
    3&1&0&0&1&&&\\
    4&1&0&0&1&1&&\\
    5&1&0&0&0&0&1&\\
    6&1&0&0&0&0&1&1
  \end{tabular}
  \caption{Adjacency matrix.}\label{fig:graph2adj}
\end{subfigure}
\begin{subfigure}[m]{.32\textwidth}\centering
\iftoggle{tikz}{
\begin{tikzpicture}
\tikzstyle{every node}=[draw, shape=circle]
\node at (0,0) [fill=green!20] (v7) {7};
\node at (0:1) [fill=blue!20] (v1) {1};
\node at (60:1) [fill=red!20] (v2) {2};
\node at (120:1) [fill=blue!20] (v3) {3};
\node at (180:1) [fill=red!20] (v4) {4};
\node at (240:1) [fill=blue!20] (v5) {5};
\node at (300:1) [fill=red!20] (v6) {6};
\draw
(v1) -- (v7)  (v2) -- (v7)  (v3) -- (v7)
(v4) -- (v7)  (v5) -- (v7)  (v6) -- (v7)
(v1) -- (v2)  (v3) -- (v4)  (v5) -- (v6);
\end{tikzpicture}
}{} %% end toggle
\caption{Sparsity graph.}\label{fig:graph2pic}
\end{subfigure}
\begin{subfigure}[m]{.32\textwidth}\centering
\iftoggle{tikz}{
\begin{tikzpicture}
\tikzstyle{every node}=[draw, shape=circle]
\node at (0,0) [fill=green!20] (v7) {7};
\node at (0:1) [fill=blue!20] (v1) {1};
\node at (60:1) [fill=red!20] (v2) {2};
\node at (120:1) [fill=blue!20] (v3) {3};
\node at (180:1) [fill=red!20] (v4) {4};
\node at (240:1) [fill=blue!20] (v5) {5};
\node at (300:1) [fill=red!20] (v6) {6};
\draw
(v1) -- (v7)  (v2) -- (v7)  (v3) -- (v7)
(v4) -- (v7)  (v5) -- (v7)  (v6) -- (v7)
(v1) -- (v2)  (v3) -- (v4)  (v5) -- (v6);
\end{tikzpicture}
}{} %% end toggle
\caption{Intersection graph.}\label{fig:graph2int}
\end{subfigure}
\caption{Permuted matrix.}\label{fig:graph2}
\end{figure}

The practical implication of all of this is that by permuting the
rows and columns of the Hessian, we may be able to reduce the
number of colors needed for a cyclic coloring of the graph of the
sparsity pattern.  Fewer colors means fewer partitions of the
variables, and that means fewer gradient evaluations to
estimate the Hessian.

The \class{sparseHessianFD}
class finds a permutation, and partitions the variables, when it is
initialized. The problem of finding a cyclic
coloring of the graph of the sparsity pattern is NP-complete
\citep{ColemanCai1986}, so the partitioning may not be truly optimal.
Fortunately, we just need the partitioning to be reasonably good, to
make the effort worth our
while.  A plethora of vertex coloring heuristics have been proposed,
and we make no claims that any of the algorithms in \pkg{sparseHessianFD} are even
``best available'' for all situations.

The first step is to permute the rows and columns of the Hessian.   A
reasonable choice is the ``smallest-last'' ordering that
sorts the rows and columns in decreasing order of the number of
elements \citep[Theorem 6.2]{ColemanMore1984}. To justify this permutation, suppose
non-zeros within a row are randomly distributed across
columns.  If the row is near the top of the matrix, there is a higher
probability that any non-zero element is in the upper triangle, not in
the lower.  By putting sparser rows near the bottom, we do not change
the number of non-zeros in the lower triangle, but we should come
close to minimizing the number of non-zeros in each row.  Thus, we
would expect the number of columns with non-zero elements in common
rows to be smaller, and the intersection graph to be sparser \citep{GebremedhinTarafdar2007}.

The adjacency matrix of the intersection graph of the permuted matrix
is the Boolean crossproduct, $\hessLT{\pi}{}^\top \hessLT{\pi}{}$.
Algorithm \ref{alg:coloring} is a ``greedy'' vertex coloring
algorithm, in which vertices are colored sequentially.  The result is
a cyclic coloring on the sparsity graph, which in turn is a consistent
partitioning of the variables.

\begin{algorithm}
  \begin{algorithmic}
    \REQUIRE{$P[i],i=1,\mathellipsis,M$: sets of column indices  of
      non-zero elements in row $i$.}
    \REQUIRE{$F[i],i=1,\mathellipsis,M$:  sets of ``forbidden'' colors for
         vertex $i$ (initially empty)}
    \REQUIRE{$U$:  set of used colors (initially empty)}
    \REQUIRE{$C[i], i=1,\mathellipsis,M$: vector to store output of
      assigned colors (initially all zero)}.
    \STATE{$k\leftarrow 0$} \COMMENT{Largest color index used}
    \STATE{Insert $0$ in $U$}
    \FOR{$i=1$ \TO $M$}
       \IF{$F[i]$ is empty (no forbidden colors)}
           \STATE{$C[i]\leftarrow 0$}
        \ELSE
           \STATE{$V\leftarrow U - F[i]$}\COMMENT{Used colors that are
             not forbidden}
           \IF{$V$ is empty}
              \STATE{$k\leftarrow k+1$}
              \STATE{Insert $k$ into $U$}
              \STATE{$C[i]\leftarrow k$}
        \ELSE
             \STATE{$C[i]\leftarrow\min(V)$}
\COMMENT{Assign smallest existing non-forbidden color to $i$}
             \ENDIF
    \ENDIF
    \ENDFOR
    \FOR{$j$ in $P[i]$}
    \STATE{Insert $C[i]$ into $F[j]$}
\COMMENT{Make $i$'s color forbidden to all of
      its uncolored neighbors}
    \ENDFOR
    \RETURN{$C$}
    \end{algorithmic}
    \caption{Consistent partitioning of variables for a triangular
      substitution method.}\label{alg:coloring}
    \end{algorithm}

\subsection[Computing the Hessian by substitution]{Computing the Hessian by substitution}

The cycling coloring of the sparsity graph defines the $\Mat{G}$
matrix from Section \ref{sec:numdiff}.  We then estimate $\Mat{Y}$
using Equation~\ref{eq:Yg}.   Let $C_m$ be the assigned color to
variable $m$.  The substitution method is defined in
\citet[Equation 6.1]{ColemanMore1984}.
\begin{align}
  \label{eq:subst}
  \hess{ij}{f(x)}&=\Mat{Y}_{i,C_j}/\delta - \sum_{l>i,l\in C_j}\hess{li}{f(x)}
\end{align}

We implement the substitution method using Algorithm \ref{alg:subst}.
This algorithm completes the bottom row of the lower triangle, copies
values to the corresponding column in the upper triangle, and advances upwards.

\begin{algorithm}
  \begin{algorithmic}
    \REQUIRE{$P[i],i=1,\mathellipsis,M$: sets of column indices  of
      non-zero elements in row $i$.}
    \REQUIRE{$C[i], i=1,\mathellipsis,M$: vector of
      assigned colors}
    \REQUIRE{$\Mat{H}$, an $M\times M$ Hessian (initialized to zero)}
    \REQUIRE{$\Mat{B}$, a $\max(C)\times M$ matrix (initialized to zero)}
    \REQUIRE{$\Mat{Y}$, a matrix of finite differences}
    \REQUIRE{$\delta$, the small constant used to estimate $\Mat{Y}$}
    \FOR{$i$ = $M$ \TO $1$}
    \FOR{All $j$ in $P_i$}
    \STATE{$z\leftarrow Y[i,C[j]]/\delta - B[C[j], i]$}
    \STATE{$\Mat{B}[C[i], j]\leftarrow \Mat{B}[C[i], j] + z$}
    \STATE{$H[i,j]\leftarrow z$}
    \STATE{$H[j,i]\leftarrow H[i,j]$}
    \ENDFOR
    \ENDFOR
  \end{algorithmic}
  \caption{Triangular substitution method.}\label{alg:subst}
\end{algorithm}

\subsection[Software libraries]{Software libraries}
The coloring and substitution algorithms use the \pkg{Eigen} numerical
library \citep{Eigen}, and the \pkg{Rcpp} \citep{R_Rcpp} and
\pkg{RcppEigen} \citep{R_RcppEigen} \proglang{R} packages.  The
\pkg{testthat} \citep{R_testthat}, \pkg{scales} \citep{R_scales} and
\pkg{knitr} \citep{R_knitr} packages were used for testing, and to prepare this article.

\section[Speed and scalability]{Speed and scalability}\label{sec:timing}

As far as we know, \pkg{numDeriv} \citep{R_numDeriv} is the only other \proglang{R} package
that computes numerical approximations to derivatives.  It differs from
\pkg{sparseHessianFD} in some important ways.

\begin{enumerate}
\item It treats all Hessians as dense;
  \item It computes each element of the Hessian using a second-order
    finite differencing approximation that does not require the user
    to supply the gradient; and
    \item It implements iterative algorithms to improve accuracy, at
      the expense of speed.
\end{enumerate}

Nevertheless, \pkg{numDeriv} is an easy-to-use tool for numerical differentiation,
so it is worthwhile to compare its performance to that of
\pkg{sparseHessianFD}.  To prepare Table~\ref{tab:numDeriv}, we
estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
units ($N$) and within-unit parameters ($k$).  As in the previous
section, the total number of
variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
and standard deviations (across 20 replications) for the time (in milliseconds) to
to compute a Hessian using each package.  The difference in run times is dramatic,
especially because the computation time for \pkg{numDeriv} grows
quadratically in the number of variables.  The setup time for
\pkg{sparseHessianFD} was about 7 milliseconds for all cases.  Times
were collected using the \pkg{microbenchmark} package
\citep{R_microbenchmark}, using a 2013-edition Mac Pro with 12
processing cores and 64GB of RAM.

\begin{table}[ptb]\centering
  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }cccp{2pt}rrrrp{2pt}rrrr}
  \toprule
&&&&\multicolumn{4}{c}{\bfseries{Hessian
    time}}&&\multicolumn{4}{c}{\bfseries{Hessian/Gradient ratio}}\\
&&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
$N$&$k$&$M$&&mean&sd&mean&sd&&mean&sd&mean&sd\\
  \cmidrule{1-3}\cmidrule{5-8}\cmidrule{10-13}
   9 &    2 &   20 &&  363 & 6.4 &  2.8 & 2.2 &&  871 & 63 & 6.5 & 4.2 \\
    12 &    2 &   26 &&  629 &  8.9 &  2.9 & 2.1 && 1469 &  124 & 6.5 & 3.3 \\
     9 &    3 &   30 &&  819 & 27.7 &  3.6 & 2.2 && 1974 &  161 & 8.4 & 3.7 \\
    15 &    2 &   32 &&  973 & 25.8 &    3.0 & 2.2 && 2191 &  209 &  6.5 & 3.4 \\
    12 &    3 &   39 && 1407 & 39.3 & 3.5 & 2.0 && 3291 &  285 &    8.0 & 3.5 \\
     9 &    4 &   40 && 1445 & 28.8 & 4.3 & 2.1 && 3453 &  292 & 9.9 & 3.1 \\
    15 &    3 &   48 && 2138 & 92.7 & 3.5 & 2.0 && 4861 &  341 & 7.8 & 3.4 \\
    12 &    4 &   52 && 2400 & 98.6 & 4.3 & 2.1 && 5679 &  503 & 10.1 & 3.7 \\
    15 &    4 &   64 && 3643 &  209.0 & 4.2 & 2.2 && 8945 &  828 & 10.0 &  2.8 \\
   \bottomrule
\end{tabularx}
\caption{Computation times (milliseconds) for computing Hessians using the
  \pkg{numDeriv} and \pkg{sparseHessianFD} packages, across 200 replications. Rows are ordered by the
  number of variables.}\label{tab:numDeriv}
\end{table}

Because \pkg{numDeriv} does not scale, we cannot use it to benchmark \pkg{sparseHessianFD} for
larger sparse Hessians.  To help us understand just how scalable
\pkg{sparseHessianFD} is, we ran another set of simulations, for the
same hierarchical model, for different values of $N$ and $k$.  We then
recorded the run times for different steps in the sparse Hessian
estimation, across 200 replications.  The steps are summarized in Table~\ref{tab:timeMeasures}.

\begin{table}[ptb]\centering
  \begin{tabularx}{1.0\linewidth}{>{\bfseries}l X}
    \toprule
    Measure&Description\\
    \midrule
Function&estimating the objective function\\
Gradient&estimating the gradient\\
Hessian&computing the Hessian (not including initialization or
  partitioning time)\\
Partitioning&finding a consistent partitioning of the
  variables (the vertex coloring problem)\\
Initialization&total setup time (including the partitioning time)\\
Hessian/Gradient&ratio of the Hessian time to the
    gradient time\\
    \bottomrule
  \end{tabularx}
  \caption{Summary of timing tests (see Figure~\ref{fig:timing}).}\label{tab:timeMeasures}
\end{table}

In the plots in Figure~\ref{fig:timing}, the number of heterogeneous units ($N$) is
on the x-axis, and mean run time, in milliseconds, is on the y-axis.  Each panel shows the
relationship between $N$ and run time for a different measure of time,
and each curve in a panel represents a different number of within-unit
parameters ($k$).

\begin{figure}[tbp]
  \centering
\includegraphics{fig_timings}
  \caption{Run times for sparse Hessian computation.}
  \label{fig:timing}
\end{figure}

It should not be surprising that run times
for the function and gradient, as well as the setup and
partitioning times for the \class{sparseHessianFD} object, grow
linearly with the number of heterogenous units.  The run time
for the Hessian grows linearly as well, and that might be partially
surprising.  We saw in Section \ref{sec:coloring} that adding
additional heterogeneous units in a hierarchical model does not
increase the number of required gradient evaluations.  So we might
think that the time
to compute a Hessian should not increase with $N$ \emph{at all}.  The
reason it does is that each gradient evaluation takes longer.  The
plot of the ratio of Hessian-to-gradient run times is relatively flat as the size
of the dataset increases.  We can conclude that the
\pkg{sparseHessianFD} algorithms are quite efficient and scalable for
hierarchical models.

\FloatBarrier
\bibliography{./sparseHessianFD}

\end{document}

### 2016-03-14 nobuild/jss/sparseHessianFD.tex
ready for JSS submission

--- 2016-03-14 nobuild/jss/sparseHessianFD.Rnw
@@
-\documentclass[article]{jss}
+\documentclass[article]{jss}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\usepackage{Sweave}
 
@@
    \newcolumntype{d}[1]{D{.}{.}{#1}}
+%%\newcommand{\pkg}[1]{\emph{#1}}
+%%\newcommand{\proglang}[1]{\textsf{#1}}
+%%\newcommand{\filename}[1]{\textit{#1}}
+%%\newcommand{\code}[1]{\texttt{#1}}
 \newcommand{\func}[1]{\code{#1}}
@@
 \newenvironment{knitrout}{\begin{figure}}{\end{figure}}
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
-
-<<setup1, echo=FALSE, cache=FALSE>>=
-suppressPackageStartupMessages(require(Matrix))
-knitr::render_sweave()
-knitr::opts_chunk$set(prompt=TRUE, cache=TRUE)
-options(replace.assign=TRUE, width=90,prompt="R> ")
-@
 
@@
 
-<<setup2, echo=FALSE>>=
-N <- 5
-k <- 2
-nv1 <- (N+1)*k
-nels1 <- nv1^2
-nnz1 <- (N+1)*k^2 + 2*N*k^2
-nnz1LT <- (N+1)*k*(k+1)/2 + N*k^2
-Q <- 1000
-nv2 <- (Q+1)*k
-nels2 <- nv2^2
-nnz2 <- (Q+1)*k^2 + 2*Q*k^2
-nnz2LT <- (Q+1)*k*(k+1)/2 + Q*k^2
-options(scipen=999)
-@
-
 The sparsity pattern depends on how the variables are
@@
 In this case, the Hessian has a ``block-arrow'' structure.  For example,
-if $N=\Sexpr{N}$ and $k=\Sexpr{k}$, then there are \Sexpr{nv1} total
+if $N=5$ and $k=2$, then there are 12 total
 variables, and the Hessian will have the pattern in Figure~\ref{fig:blockarrow}.
@@
   \begin{subfigure}[b]{.5\textwidth}
-<<blockarrow, echo=FALSE>>=
-Mat <- as(kronecker(diag(N),matrix(1,k,k)),"sparseMatrix")
-Mat <- rBind(Mat, Matrix(1,k,N*k))
-Mat <- cBind(Mat, Matrix(1, k*(N+1), k))
-printSpMatrix(as(Mat,"nMatrix"))
-@
+\begin{Schunk}
+\begin{Soutput}
+
+ [1,] | | . . . . . . . . | |
+ [2,] | | . . . . . . . . | |
+ [3,] . . | | . . . . . . | |
+ [4,] . . | | . . . . . . | |
+ [5,] . . . . | | . . . . | |
+ [6,] . . . . | | . . . . | |
+ [7,] . . . . . . | | . . | |
+ [8,] . . . . . . | | . . | |
+ [9,] . . . . . . . . | | | |
+[10,] . . . . . . . . | | | |
+[11,] | | | | | | | | | | | |
+[12,] | | | | | | | | | | | |
+\end{Soutput}
+\end{Schunk}
 \caption{A ``block-arrow'' sparsity pattern.}\label{fig:blockarrow}
@@
 \begin{subfigure}[b]{.5\textwidth}
-<<banded, echo=FALSE>>=
-Mat <- kronecker(Matrix(1,k,k), diag(N))
-Mat <- rBind(Mat, Matrix(1,k,N*k))
-Mat <- cBind(Mat, Matrix(1, k*(N+1), k))
-printSpMatrix(as(Mat,"nMatrix"))
-@
+\begin{Schunk}
+\begin{Soutput}
+
+ [1,] | . . . . | . . . . | |
+ [2,] . | . . . . | . . . | |
+ [3,] . . | . . . . | . . | |
+ [4,] . . . | . . . . | . | |
+ [5,] . . . . | . . . . | | |
+ [6,] | . . . . | . . . . | |
+ [7,] . | . . . . | . . . | |
+ [8,] . . | . . . . | . . | |
+ [9,] . . . | . . . . | . | |
+[10,] . . . . | . . . . | | |
+[11,] | | | | | | | | | | | |
+[12,] | | | | | | | | | | | |
+\end{Soutput}
+\end{Schunk}
 \caption{A ``banded'' sparsity pattern.}\label{fig:banded}
@@
 In both cases, the number of non-zeros is the same.   There are
-\Sexpr{nels1} elements in this symmetric matrix, but only \Sexpr{nnz1} are
-non-zero, and only \Sexpr{nnz1LT} values are unique.  Although the reduction in
+144 elements in this symmetric matrix, but only 64 are
+non-zero, and only 38 values are unique.  Although the reduction in
 RAM from using a sparse matrix structure for the Hessian may be
-modest, consider what would happen if $N=\Sexpr{scales::comma(Q)}$ instead.  In that case,
-there are \Sexpr{scales::comma(nv2)} variables in the problem, and more than $\Sexpr{scales::comma(floor(nels2/10^6))}$ million
-elements in the Hessian.  However, only $\Sexpr{scales::comma(nnz2)}$ of those elements are
+modest, consider what would happen if $N=1,000$ instead.  In that case,
+there are 2,002 variables in the problem, and more than $4$ million
+elements in the Hessian.  However, only $12,004$ of those elements are
 non-zero.  If we work with only the lower triangle of the Hessian we only need to work with
-only \Sexpr{scales::comma(nnz2LT)} values.
+only 7,003 values.
 
@@
 \begin{minipage}{\columnwidth}
-<<>>=
-library("sparseHessianFD")
-bd <- kronecker(diag(3), matrix(TRUE,2,2))
-Mat <- as(bd, "nMatrix")
-printSpMatrix(tril(Mat))
-mc <- Matrix.to.Coord(tril(Mat))
-mc
-@
+\begin{Schunk}
+\begin{Sinput}
+R> library("sparseHessianFD")
+R> bd <- kronecker(diag(3), matrix(TRUE,2,2))
+R> Mat <- as(bd, "nMatrix")
+R> printSpMatrix(tril(Mat))
+\end{Sinput}
+\begin{Soutput}
+
+[1,] | . . . . .
+[2,] | | . . . .
+[3,] . . | . . .
+[4,] . . | | . .
+[5,] . . . . | .
+[6,] . . . . | |
+\end{Soutput}
+\begin{Sinput}
+R> mc <- Matrix.to.Coord(tril(Mat))
+R> mc
+\end{Sinput}
+\begin{Soutput}
+$rows
+[1] 1 2 2 3 4 4 5 6 6
+
+$cols
+[1] 1 1 2 3 3 4 5 5 6
+\end{Soutput}
+\end{Schunk}
 \end{minipage}
@@
 
-<<>>=
-pattern <- sparseMatrix(i=mc$rows, j=mc$cols)
-printSpMatrix(pattern)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> pattern <- sparseMatrix(i=mc$rows, j=mc$cols)
+R> printSpMatrix(pattern)
+\end{Sinput}
+\begin{Soutput}
+
+[1,] | . . . . .
+[2,] | | . . . .
+[3,] . . | . . .
+[4,] . . | | . .
+[5,] . . . . | .
+[6,] . . . . | |
+\end{Soutput}
+\end{Schunk}
 
@@
 arguments,
-<<eval=FALSE>>=
-obj <- sparseHessianFD(x, fn, gr, rows, cols, ...)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> obj <- sparseHessianFD(x, fn, gr, rows, cols, ...)
+\end{Sinput}
+\end{Schunk}
 where \funcarg{...} represents all other named arguments that are
@@
 
-<<eval=FALSE>>=
-f <- obj$fn(x)          ## returns numeric
-df <- obj$gr(x)         ## returns numeric vector
-hess <- obj$hessian(x)  ## returns dgCMatrix
-fngr <- obj$fngr(x)     ## returns list
-fngrhs <- obj$fngrhs(x) ## returns list
-@
+\begin{Schunk}
+\begin{Sinput}
+R> f <- obj$fn(x)          ## returns numeric
+R> df <- obj$gr(x)         ## returns numeric vector
+R> hess <- obj$hessian(x)  ## returns dgCMatrix
+R> fngr <- obj$fngr(x)     ## returns list
+R> fngrhs <- obj$fngrhs(x) ## returns list
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<binaryInit>>=
-set.seed(123)
-data("binary")
-str(binary)
-N <- length(binary[["Y"]])
-k <- NROW(binary[["X"]])
-T <- binary[["T"]]
-nvars <- as.integer(N*k + k)
-priors <- list(inv.Sigma = rWishart(1,k+5,diag(k))[,,1],
-               inv.Omega = diag(k))
-@
-
-This dataset represents the simulated choices for $N= \Sexpr{N}$ customers
-over $T= \Sexpr{T}$ purchase opportunties, where the probability of purchase
-is influenced by $k= \Sexpr{k}$ covariates.
+\begin{Schunk}
+\begin{Sinput}
+R> set.seed(123)
+R> data("binary")
+R> str(binary)
+\end{Sinput}
+\begin{Soutput}
+List of 3
+ $ Y: int [1:50] 13 1 18 18 19 6 16 6 5 8 ...
+ $ X: num [1:4, 1:50] -0.07926 -0.23018 1.55871 0.00997 0.01828 ...
+ $ T: num 20
+\end{Soutput}
+\begin{Sinput}
+R> N <- length(binary[["Y"]])
+R> k <- NROW(binary[["X"]])
+R> T <- binary[["T"]]
+R> nvars <- as.integer(N*k + k)
+R> priors <- list(inv.Sigma = rWishart(1,k+5,diag(k))[,,1],
++                inv.Omega = diag(k))
+\end{Sinput}
+\end{Schunk}
+
+This dataset represents the simulated choices for $N= 50$ customers
+over $T= 20$ purchase opportunties, where the probability of purchase
+is influenced by $k= 4$ covariates.
 
@@
 
-<<trueValues>>=
-P <- rnorm(nvars)
-order.row <- FALSE
-true.f <- binary.f(P, binary, priors, order.row=order.row)
-true.grad <- binary.grad(P, binary, priors, order.row=order.row)
-true.hess <- binary.hess(P, binary, priors, order.row=order.row)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> P <- rnorm(nvars)
+R> order.row <- FALSE
+R> true.f <- binary.f(P, binary, priors, order.row=order.row)
+R> true.grad <- binary.grad(P, binary, priors, order.row=order.row)
+R> true.hess <- binary.hess(P, binary, priors, order.row=order.row)
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<binaryRowsCols>>=
-pattern <- Matrix.to.Coord(tril(true.hess))
-str(pattern)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> pattern <- Matrix.to.Coord(tril(true.hess))
+R> str(pattern)
+\end{Sinput}
+\begin{Soutput}
+List of 2
+ $ rows: int [1:1310] 1 2 3 4 201 202 203 204 2 3 ...
+ $ cols: int [1:1310] 1 1 1 1 1 1 1 1 2 2 ...
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<usingSparseHessianFD>>=
-obj <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
-       rows=pattern[["rows"]], cols=pattern[["cols"]],
-       data=binary, priors=priors, order.row=order.row)
-f <- obj$fn(P)
-all.equal(f, true.f)
-gr <- obj$gr(P)
-all.equal(gr, true.grad)
-hs <- obj$hessian(P)
-all.equal(hs, true.hess)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> obj <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
++        rows=pattern[["rows"]], cols=pattern[["cols"]],
++        data=binary, priors=priors, order.row=order.row)
+R> f <- obj$fn(P)
+R> all.equal(f, true.f)
+\end{Sinput}
+\begin{Soutput}
+[1] TRUE
+\end{Soutput}
+\begin{Sinput}
+R> gr <- obj$gr(P)
+R> all.equal(gr, true.grad)
+\end{Sinput}
+\begin{Soutput}
+[1] TRUE
+\end{Soutput}
+\begin{Sinput}
+R> hs <- obj$hessian(P)
+R> all.equal(hs, true.hess)
+\end{Sinput}
+\begin{Soutput}
+[1] TRUE
+\end{Soutput}
+\end{Schunk}
 

### 2016-03-14 R/coloring.R

@@ -6,6 +6,7 @@
 #' @param L  sparsity pattern of the Hessian as a lower triangular pattern matrix
 #' @return Integer vector of length nvars with color assignments for each variable.
 #' @description cyclic coloring from a lower triangular pattern matrix
+#' @export
 coloring <- function(L) {
 
     stopifnot(is(L,"nMatrix"),

### 2016-03-14 nobuild/jss/repl_fig4.R

@@ -1,17 +1,17 @@
-##library(sparseHessianFD)
+## Code to generate Table 4.  For the manuscript, results
+## were subsequently rounded and formatted. A small amount of random-like
+## variation should be expected in the run times.
+
+library(sparseHessianFD)
 library(Matrix)
 library(mvtnorm)
 library(plyr)
 library(microbenchmark)
-library(doParallel)
 library(dplyr)
 library(tidyr)
 library(reshape2)
 library(ggplot2)
-
-
-run.par <- TRUE
-if (run.par) registerDoParallel(cores=12) else registerDoParallel(cores=1)
+theme_set(theme_bw())
 
 binary_sim <- function(N, k, T) {
     x.mean <- rep(0,k)
@@ -111,10 +111,7 @@ cases <- expand.grid(k=c(8, 6, 4, 2),
                      T=20
                      )
 
-res <- adply(cases, 1, run_test, reps=200, .parallel=run.par)
-
-
-##load("inst/examples/timings.Rdata")
+res <- adply(cases, 1, run_test, reps=200)
 
 tab <- mutate(res, ms=bench.time/1000000) %>%
   dcast(N+k+T+bench.rep+ncolors~bench.expr, value.var="ms")  %>%
@@ -135,7 +132,7 @@ D2$stat <- factor(D2$stat, levels=c("Function","Gradient","Hessian",
                                     "Partitioning","Initialization",
                                     "Hessian/Gradient"))
 
-theme_set(theme_bw())
+
 P2 <- ggplot(D2, aes(x=N,y=mean, color=as.factor(k), linetype=as.factor(k))) %>%
   + geom_line(size=.4) %>%
   + scale_x_continuous("Number of heterogeneous units") %>%
@@ -143,13 +140,7 @@ P2 <- ggplot(D2, aes(x=N,y=mean, color=as.factor(k), linetype=as.factor(k))) %>%
   + guides(color=guide_legend("k"), linetype=guide_legend("k")) %>%
   + facet_wrap(~stat, scales="free") %>%
   + theme(text=element_text(size=8), legend.position="right")
-
-
-pdf(file="vignettes/fig_timings.pdf", width=6, height=4)
 print(P2)
-dev.off()
-
-
 
 
 

### 2016-03-14 nobuild/jss/repl_table4.R

@@ -1,12 +1,18 @@
+## Code to generate Table 4.  For the manuscript, results
+## were subsequently rounded and formatted. A small amount of random-like
+## variation should be expected in the run times.
+
+
 library(sparseHessianFD)
+library(Matrix)
 library(mvtnorm)
 library(plyr)
+library(dplyr)
+library(tidyr)
+library(reshape2)
 library(microbenchmark)
-library(doParallel)
 library(numDeriv)
 
-run.par <- TRUE
-if (run.par) registerDoParallel(cores=12) else registerDoParallel(cores=1)
 
 binary_sim <- function(N, k, T) {
     x.mean <- rep(0,k)
@@ -71,8 +77,7 @@ run_test <- function(Nk, reps=50) {
 
 cases <- expand.grid(k=c(2,3,4),
                      N=c(9, 12, 15))
-
-runs <- adply(cases, 1, run_test, reps=20, .parallel=run.par)
+runs <- adply(cases, 1, run_test, reps=20)
 
 tab4 <-  mutate(runs, ms=bench.time/1000000) %>%
   select(-bench.time) %>%
@@ -86,6 +91,7 @@ tab4 <-  mutate(runs, ms=bench.time/1000000) %>%
   dcast(N+k+M~stat+method+stat2,value.var="value") %>%
   arrange(M)
 
+print(tab4)
 
 
 

### 2016-03-16 nobuild/storageFormats.tex

\subsubsection[Storage formats]{Storage formats}

Matrices with a ``triplet'' format store each non-zero element as a
(row, column, value) set.  The values are either numeric or logical,
for \class{d*TMatrix} or \class{l*TMatrix} matrices, respectively.
Pattern matrices of class \class{n*TMatrix} store only row and column
indices, but no values at all.

Row-compressed (CSR) and column-compressed (CSC) formats are more
efficient alternatives.  Instead of storing the row and column indices, CSC
matrices store only the row indices, and ``pointers''
to the values in the row index and value vectors that start
each column\footnote{For the CSR format, switch references to columns and rows
in the preceding sentence.}.

For example, let's construct a numeric matrix $\Mat{A}$ in triplet form using
the \func{sparseMatrix} function. The arguments \funcarg{i},
\funcarg{j} and \funcarg{x} are the row and column indices, and
values, respectively, of the non-zero elements.  By default, \func{sparseMatrix} expects the input vectors \funcarg{i} and \funcarg{j} to
have one-based indexing, and stores $\Mat{A}$ in CSC format.  We can then
look at the values directly, and the pointers (using the familiar
one-based indexing scheme) using \func{Matrix.to.Pointers}.

<<trip>>=
library("sparseHessianFD")
A <- sparseMatrix(i=c(1,4,2,3,3,4), j=c(1,1,2,2,3,4), x=c(6,7,8,9,10,11))
printSpMatrix(A)
Matrix.to.Pointers(A, values=TRUE, order="column", index1=TRUE)
@

The slot \funcarg{x} contains the matrix values exactly as we entered
them.  The column indices are not stored explicitly, but the pointers in \variable{jpntr}
determine where in the sequence of values and row indices (\variable{iRow}) the next
column will start. The first pointer is always 1, so the first element in column 1 is in
 row \code{iRow[1]=1}, and has value \code{x[1]=6}.  Since \code{jpntr[2]=3}, \code{x[2]=7} is still in
the first column, in row \code{iRow[2]=4}, so $\Mat{A}_{41}=7$.  Elements 3 and 4 in
\code{x} and \code{iRow} refer to values in column 2, since
\code{jpntr[3]=5} means that column 3 does not begin until the element
5. Hence, $\Mat{A}_{22}=x_3=8$ and $\Mat{A}_{32}=x_4=9$.
Column 3 begins with $\Mat{A}_{33}=x_5=10$ (\code{iRow[5]=3} gives us the
row index).  Continuing,
$\Mat{A}_{44}=x_6=11$.  The final value in
\variable{jpntr} is always the number of non-zeros in the matrix.  The differences between
adjacent elements in \variable{jpntr} are the number of non-zeros in each column of $\Mat{A}$.

There are two important advantages to storing sparse matrices in a
compressed format.  First, the storage requirements are usually smaller.  A
matrix with $M$ columns and $N$ non-zeros requires $3N$ elements in a
triplet format, but only $2N+M+1$ elements in a compressed format.
Second, computation on the matrix is more efficient, since all elements in a
single column (CSC) or row (CSR) are stored in contiguous memory, in
order.  In contrast, matrix triplets could be entered or stored in any
order.

### 2016-06-09 nobuild/jss/sparseHessianFD.Rnw
jss

@@
-\documentclass[article]{jss}
+\documentclass[jss]{jss}
 
@@
     symbolic or algorithmic methods are impractical, inefficient or
-    unavailable. }
+    unavailable.}
 
@@
 knitr::render_sweave()
-knitr::opts_chunk$set(prompt=TRUE, cache=TRUE)
+knitr::opts_chunk$set(prompt=TRUE, cache=FALSE)
 options(replace.assign=TRUE, width=90,prompt="R> ")
@@
 can be reduced by exploiting the sparsity pattern and symmetric
-structure.  This note borrows heavily from, and use the notation in,
+structure.  This note borrows heavily from, and uses the notation in,
 \citet[Chapter 6]{MagnusNeudecker2007}. The
@@
 
-The distinction among sparse matrix classes is important because
+The distinctions among sparse matrix classes is important because
 \pkg{sparseHessianFD}'s \method{hessian} method returns a
@@
 elements in the Hessian.  However, only $\Sexpr{scales::comma(nnz2)}$ of those elements are
-non-zero.  If we work with only the lower triangle of the Hessian we only need to work with
+non-zero.  If we work with only the lower triangle of the Hessian, then we need to work with
 only \Sexpr{scales::comma(nnz2LT)} values.
@@
 \pkg{sparseHessianFD} was about 7 milliseconds for all cases.  Times
-were collected using the \pkg{microbenchmark} package
-\citep{R_microbenchmark}, using a 2013-edition Mac Pro with 12
-processing cores and 64GB of RAM.
+were collected using the \pkg{microbenchmark} package \citep{R_microbenchmark}.
 
@@
 
-It should not be surprising that run times
+Computation times
 for the function and gradient, as well as the setup and
 partitioning times for the \class{sparseHessianFD} object, grow
-linearly with the number of heterogenous units.  The run time
+linearly with the number of heterogenous units.  The time
 for the Hessian grows linearly as well, and that might be partially
@@
 \FloatBarrier
-\bibliography{./sparseHessianFD}
+\bibliography{sparseHessianFD}
 

### 2016-06-09 nobuild/jss/repl_fig4.R

@@ -1,4 +1,4 @@
-##library(sparseHessianFD)
+library(sparseHessianFD)
 library(Matrix)
 library(mvtnorm)
 library(plyr)
@@ -114,7 +114,6 @@ cases <- expand.grid(k=c(8, 6, 4, 2),
 res <- adply(cases, 1, run_test, reps=200, .parallel=run.par)
 
 
-##load("inst/examples/timings.Rdata")
 
 tab <- mutate(res, ms=bench.time/1000000) %>%
   dcast(N+k+T+bench.rep+ncolors~bench.expr, value.var="ms")  %>%
@@ -136,7 +135,7 @@ D2$stat <- factor(D2$stat, levels=c("Function","Gradient","Hessian",
                                     "Hessian/Gradient"))
 
 theme_set(theme_bw())
-P2 <- ggplot(D2, aes(x=N,y=mean, color=as.factor(k), linetype=as.factor(k))) %>%
+fig4 <- ggplot(D2, aes(x=N,y=mean, color=as.factor(k), linetype=as.factor(k))) %>%
   + geom_line(size=.4) %>%
   + scale_x_continuous("Number of heterogeneous units") %>%
   + scale_y_continuous("Computation time (milliseconds)") %>%
@@ -145,10 +144,7 @@ P2 <- ggplot(D2, aes(x=N,y=mean, color=as.factor(k), linetype=as.factor(k))) %>%
   + theme(text=element_text(size=8), legend.position="right")
 
 
-pdf(file="vignettes/fig_timings.pdf", width=6, height=4)
-print(P2)
-dev.off()
-
+print(fig4)
 
 
 

### 2016-06-09 nobuild/jss/repl_table4.R

@@ -1,9 +1,15 @@
 library(sparseHessianFD)
+library(Matrix)
 library(mvtnorm)
 library(plyr)
+library(dplyr)
+library(tidyr)
+library(reshape2)
 library(microbenchmark)
 library(doParallel)
 library(numDeriv)
+
+
 
 run.par <- TRUE
 if (run.par) registerDoParallel(cores=12) else registerDoParallel(cores=1)

### 2016-12-19 nobuild/jss_revision/response.tex
copied review comments into response.tex

\documentclass{article}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{tabularx}
\usepackage{dcolumn}
\usepackage{booktabs}
\usepackage{etoolbox}
\usepackage{placeins}
\newtoggle{tikz}
\toggletrue{tikz}

\iftoggle{tikz}{
\usepackage{tikz}
}{}

\newcolumntype{d}[1]{D{.}{.}{#1}}
\newcommand{\pkg}[1]{\emph{#1}}
\newcommand{\proglang}[1]{\textsf{#1}}
\newcommand{\filename}[1]{\textit{#1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\func}[1]{\code{#1}}
\newcommand{\class}[1]{\textsl{#1}}
\newcommand{\funcarg}[1]{\code{#1}}
\newcommand{\variable}[1]{\code{#1}}
\newcommand{\method}[1]{\func{#1}}

 \newcommand{\df}[3]{\mathsf{d}^{#1}f(#2;#3)}
 \newcommand{\parD}[3]{\mathsf{D}^{#1}_{#2}#3}
\newcommand{\hess}[2]{\mathsf{H}_{#1}#2}
\newcommand{\hessLT}[2]{\mathsf{L}_{#1}#2}
\newcommand{\Mat}[1]{\mathbf{#1}}

\newenvironment{revQuote}{\itshape}{\vspace{\baselineskip}}
\newenvironment{response}{\normalfont}{\vspace{\baselineskip}}

\usepackage{geometry}
\geometry{margin=1.5in}

\begin{document}

{\bfseries
\begin{center}
  Manuscript JSS2717\\
 ``sparseHessianFD: An R Package for Estimating Sparse
 Hessian Matrices''\\
 \vspace{\baselineskip}
 Response to reviewers
\end{center}
}
\vspace{\baselineskip}

I would like to express my appreciation for the time and effort that
the reviewers invested in their reviews. The comments and suggestions
are all well-taken, and I have addressed all of the items, point by
point, in this note. The reviewers’ comments are in italics, and my
responses follow.
\vspace{\baselineskip}

\begin{enumerate}
 \setcounter{enumi}{-1}

\item \begin{revQuote}
I had a minor problem checking the package tarball:

{\normalfont (error messages related to algorithm.sty removed).}

I don't remember if there is an easy way to distribute ``special'' sty files,
or if you must force the user to install them. If the sty is not really
critical it might be better to omit it.

\end{revQuote}

\begin{response}
I suspect that the error is a result of the \pkg{algorithm} package
not being present in your LaTeX installation.  I would like to continue to use it,
because it does allow for nicely formatted algorithms.  I believe that this is
a relatively common package, and one that should be easily installed
via your closest CTAN mirror.  However, if there is a strong
objection, I may be able to find another way to format the algorithms.
\end{response}

\item \begin{revQuote}
    I am not familiar with reference classes, but for most users I think it
should not be important that the package uses them. However, as a user, I
find it disconcerting that sparseHessianFD() is used as a (constructor)
function (eg. p12, and in the example in ?sparseHessianFD) but is not
documented in the help with 'usage' and 'arguments' as is usual for
functions. There seems to be little guidance in the usual R places about how
reference classes should be documented. Perhaps you could seek some guidance
from the community on this point.
\end{revQuote}

\begin{response}

\end{response}

\item \begin{revQuote}
It is pointed out (p2) that requirements are burdensome and emphasized
that the package is not appropriate for all uses. Also, the conditions 1-5
are fairly stringent. The reader is left with the impression that there may
be few applications. I think the value of the package could be promoted
better. Perhaps  a small list of different types or applications of
hierarchical problems could be given, or an indication of other classes of
problems where the conditions would apply.

  \end{revQuote}

\begin{response}

\end{response}

\item \begin{revQuote}
Condition 5 might be relaxed to a local condition in a neighborhood of
the evaluation point, rather than a global condition, but I do not know if
that would have any practical value.
  \end{revQuote}

\begin{response}

\end{response}

\item \begin{revQuote}
Section 2.1 p7 I think could be made more easily readable by adding a few
more hints about dimension, and more care about use of the term 
"coefficients vector", for example:

  - 'continuous covariates $x_i$,' -> 'continuous covariates $x_i \in
\\R^k$,'

  - 'heterogeneous coefficient vector $\beta_i$' -> 'heterogeneous
coefficient vector $\beta_i \in \\R^k$'

  - 'The coefficients are distributed' -> 'The coefficient vectors
$\beta_i$ are distributed'

  \end{revQuote}

\begin{response}

\end{response}

\item \begin{revQuote}
p7 following "the cross-partial derivatives"
$\hess{\beta_i,\beta_j}{}=\parD{2}{\beta_i, \beta_j}{}=0$ for all $i\neq
j$.

  Is this a definition of $\hess{\beta_i,\beta_j}{}$ in terms of  $\parD{2}{}{}$
or a statement of equality?

  Also, I am confused by the single subscript $\beta_i$ to \\hess here, but a
double subscript just above in "Thus, $\hess{\beta_{ik},\mu_k}{}\neq 0$".
  \end{revQuote}

\begin{response}

\end{response}

\item \begin{revQuote}
 I think possibly (16) is "banded" and (17) "block arrow" rather than the
reverse which is indicated.
  \end{revQuote}

\begin{response}

\end{response}

\item \begin{revQuote}
Code in the paper prior to table 4 needs to be copied from the
vinettes/sparseHessianFD.Rnw file. For those trying to reproduce results it
would be nice if this where mentioned in the file replication.R.
  \end{revQuote}

\begin{response}

\end{response}

\item \begin{revQuote}
p9, last line. The R> at the beginning of

   R> obj <- sparseHessianFD(x, fn, gr, rows, cols, ...)

suggests that this is code that can be entire at the command line, but ...
causes an error when entered (and the arguments have not been defined at
this point in the vignette. Instead it should be indicated as the usage
syntax.
  \end{revQuote}

\begin{response}

\end{response}

\item \begin{revQuote}
p10. "where ... represents all other named arguments"
   I think the usual usage is the ... represents the arguments other than
the "named" ones, so it is probably better to just say "other arguments".
  \end{revQuote}

\begin{response}

\end{response}

\item \begin{revQuote}
On p12, the function calls all.equal(f, true.f) and all.equal(gr,
true.grad) are comparing f and gr calculated with the exact calculation, so
the difference is zero:

\begin{verbatim}
print(obj$fn(P) - binary.f(P, 
      data=binary, priors=priors, order.row=order.row), digits=20)
[1] 0

print(max(abs(obj$gr(P) - binary.grad(P, 
         data=binary, priors=priors,
order.row=order.row))),digits=20)
[1] 0

\end{verbatim}

It is not clear to me whether obj$fn() and obj$gr() use code as in the true
functions or a modified version using sparse techniques. Some further
clarification would be helpful.

On the other hand, all.equal(hs, true.hess) is comparing a true analytic
calculation with a first order simple difference aproximation using the true
gradient function:
\begin{verbatim}
max(abs(  obj$hessian(P)
     - binary.hess(P, data=binary, priors=priors, order.row=order.row)))
[1] 2.786891e-06
\end{verbatim}

which might also be mentioned in the text. (Really just for exposition
purposes, after all, it is almost the main purpose of the package.)
p10. "where ... represents all other named arguments"
   I think the usual usage is the ... represents the arguments other than
the "named" ones, so it is probably better to just say "other arguments".
  \end{revQuote}

\begin{response}

\end{response}

\item \begin{revQuote}
I think it would be instructive to add some of the following comparisons
with the above on p12.  The package numDeriv function hessian by default
does a second order Richardson approximation using the true function value
approximation. This involves a very large number of function evaluations in
an attempt to obtain some accuracy, but the accuracy is limited by being an
approximation of a second difference:
\begin{verbatim}
max(abs(
     numDeriv::hessian( binary.f, P, method="Richardson", 
                    data=binary, priors=priors,
order.row=order.row) 
   - binary.hess(P, data=binary, priors=priors, order.row=order.row)))
[1] 0.0001610595

\end{verbatim}

Since the hessian is the first difference of the gradient, which is the
calculation used by obj\$hessian() in sparseHessianFD, one could also use the
function numDeriv::jacobian:

\begin{verbatim}
max(abs( 
    numDeriv::jacobian( binary.grad, P, method="Richardson", 
                   data=binary, priors=priors,
order.row=order.row)
  - binary.hess(P, data=binary, priors=priors, order.row=order.row)))
[1] 3.268224e-07
\end{verbatim}

This is still doing the calculation intensive Richardson approximation. The
calculation which would seem to most closely resemble what is done by
obj\$hessian() is

\begin{verbatim}
max(abs( 
    numDeriv::jacobian( binary.grad, P, method="simple", 
                   data=binary, priors=priors,
order.row=order.row)
  - binary.hess(P, data=binary, priors=priors, order.row=order.row)))

[1] 0.0008255852

\end{verbatim}

Another very interesting comparison is

\begin{verbatim}
max(abs( 
  numDeriv::jacobian( binary.grad, P, method="complex", 
                  data=binary, priors=priors,
order.row=order.row)
 - binary.hess(P, data=binary, priors=priors, order.row=order.row)))

[1] 7.105427e-15
\end{verbatim}

The complex step derivative provides extremely accurate approximations with
a  number of function evaluation similar to the simple method. (This does
not seem to be anticipated by footnote 1 in the paper.) However, the method
imposes some serious requirements on the function. (Something like complex
analytic even though the user may only be interested in the real part.) The
code also has to accept complex arguments and return the complex result.
Fortunately most R primitive work with complex numbers so the code
requirement may happen accidentally, which can be partly verified by

binary.grad(P + 0+1i, data=binary, priors=priors, order.row=order.row)

returning a complex result. (This does not rule out all possible problems.)

As I recall, sums, multiplication, and exponentiation are all complex
analytic, so it would not be too surprising if the example in the paper is
too, but I have not analyzed that. However, based on the result being very
good, it seems highly likely.
  \end{revQuote}

\begin{response}

\end{response}

\item \begin{revQuote}
A possible extension to the package would be to implement the complex
method in the sparse code. The function numDeriv:::jacobian.default
implements both simple and complex, so provides a good comparison of the
necessary (non-sparse) computation.

  \end{revQuote}

\begin{response}

\end{response}

\item \begin{revQuote}
 p.13  l. -7  Figure 3b  should be Figure 3c
  \end{revQuote}

\begin{response}

\end{response}

\item \begin{revQuote}
 The file replication.R does not set the RNG seed. This may not be too
important if only times are generated, but will be if resulting values are
included.

  \end{revQuote}

\begin{response}

\end{response}

\item \begin{revQuote}
Table 4 and 5. Some OS, processor, and memory details are helpful to put
timing results in context.
  \end{revQuote}

\begin{response}

\end{response}

\item\begin{revQuote}
Table 4 is really not the proper comparison. I think a comparison with
       numDeriv::jacobian( binary.grad, P, method="simple", 
                   data=binary, priors=priors,
order.row=order.row)

really serves to highlight the improvement of the sparse calculation because
it is a valid comparison. Even though the results are not as exaggerated,
they are still important:

(My laptop is a  Intel(R) Core(TM) i5-3337U CPU @ 1.80GHz,  4GB RAM, SSD
swap, running Mint variant of Ubuntu 14.04.2 LTS.)

\begin{verbatim}
run.par <- FALSE

(Note this is using obj$gr() rather than binary.grad(). I am not sure if
that makes a difference.)

run_test_tab4b <- function(Nk, reps=50) {
    ## Replication function like Table 4  with jacobian of grad and
"simple"
    N <- as.numeric(Nk["N"])
    k <- as.numeric(Nk["k"])
    data <- binary_sim(N, k, T=20)
    priors <- priors_sim(k)
    F <- make_funcs(D=data, priors=priors)
    nvars <- N*k+k
    M <-
as(Matrix::kronecker(Matrix::Diagonal(N),Matrix(1,k,k)),"nMatrix") %>%
      rBind(Matrix(TRUE,k,N*k)) %>%
      cBind(Matrix(TRUE, k*(N+1), k)) %>%
      as("nMatrix")
    pat <- Matrix.to.Coord(tril(M))
    X <- rnorm(nvars)
    obj <- sparseHessianFD(X, F$fn, F$gr, pat$rows, pat$cols)

    bench <- microbenchmark(
        numDeriv = numDeriv::jacobian(obj$gr, X, method="simple"), 
        df = obj$gr(X),
        sparse = obj$hessian(X))
    vals <- plyr::ddply(data.frame(bench), "expr",
                  function(x)
return(data.frame(expr=x$expr,

time=x$time,

rep=1:length(x$expr))))
    res <- data.frame(N=N, k=k,
                      bench=vals)
    cat("Completed N = ",N,"\tk = " , k ,"\n")
    return(res)
}

cases_tab4b <- expand.grid(k=c(2,3,4),
                     N=c(9, 12, 15))
runs_tab4b <- plyr::adply(cases_tab4b, 1, run_test_tab4b, reps=20,
.parallel=run.par)

tab4b <-  mutate(runs_tab4b, ms=bench.time/1000000) %>%
  select(-bench.time) %>%
  spread(bench.expr, ms) %>%
  gather(method, hessian, c(numDeriv, sparse)) %>%
  mutate(M=N*k+k, hessian.df=hessian/df) %>%
  gather(stat, time, c(hessian, hessian.df)) %>%
  group_by(N, k, method, M, stat)  %>%
  summarize(mean=mean(time), sd=sd(time)) %>%
  gather(stat2, value, mean:sd) %>%
  dcast(N+k+M~stat+method+stat2,value.var="value") %>%
  arrange(M)

tab4b 
   N k  M hessian_numDeriv_mean hessian_numDeriv_sd hessian_sparse_mean
1  9 2 20              5.731754          
0.6759421            1.713736
2 12 2 26              7.577777          
0.9481714            1.756165
3  9 3 30              8.763990          
1.0358481            2.348153
4 15 2 32              9.553452          
1.1875816            1.844192
5 12 3 39             11.583360          
1.3371329            2.402663
6  9 4 40             11.584188          
1.2045539            2.978028
7 15 3 48             14.638763          
1.5469615            2.547145
8 12 4 52             15.612216          
2.5401431            3.086010
9 15 4 64             19.672496          
2.7303771            3.055931
  hessian_sparse_sd hessian.df_numDeriv_mean hessian.df_numDeriv_sd
1        0.07532715                
19.02359               3.353629
2        0.07522725                
24.31841               4.715941
3        0.15790507                
28.47628               4.366788
4        0.30441636                
30.52029               4.533416
5        0.08570303                
35.98930               6.033750
6        0.55389180                
36.30998               6.007112
7        0.61666543                
44.46145               6.206399
8        0.69277761                
48.73135               9.180011
9        0.12617835                
60.27624               9.530503
  hessian.df_sparse_mean hessian.df_sparse_sd
1               5.687322            0.7812959
2               5.619639            0.7445722
3               7.615163            0.7901088
4               5.884283            1.0339672
5               7.455115            0.8976474
6               9.329878            2.0531265
7               7.734390            2.0592415
8               9.637077            2.4535258
9               9.356261            0.6168655

\end{verbatim}

  \end{revQuote}

  \begin{response}

  \end{response}

\item\begin{revQuote}
It is possible to do a larger example with this comparison:

(On my laptop the next took about 30 hrs of which 24 was for the last,
N=2500, k=8 comparison.)
\begin{verbatim}
cases_tab4b5 <- expand.grid(k=c(2,4,8),
                     N=c(10, 100, 1000, 2500))
runs_tab4b5 <- plyr::adply(cases_tab4b5, 1, run_test_tab4b, reps=20,
.parallel=run.par)

tab4b5 <-  mutate(runs_tab4b5, ms=bench.time/1000000) %>%
  select(-bench.time) %>%
  spread(bench.expr, ms) %>%
  gather(method, hessian, c(numDeriv, sparse)) %>%
  mutate(M=N*k+k, hessian.df=hessian/df) %>%
  gather(stat, time, c(hessian, hessian.df)) %>%
  group_by(N, k, method, M, stat)  %>%
  summarize(mean=mean(time), sd=sd(time)) %>%
  gather(stat2, value, mean:sd) %>%
  dcast(N+k+M~stat+method+stat2,value.var="value") %>%
  arrange(M)

tab4b5
tab4b5
      N k     M hessian_numDeriv_mean hessian_numDeriv_sd
hessian_sparse_mean
1    10 2    22          6.363689e+00       
6.585771e-01            1.795344
2    10 4    44          1.273715e+01       
1.254662e+00            2.934588
3    10 8    88          2.608751e+01       
1.803375e+00            5.659668
4   100 2   202          1.021658e+02       
3.439251e+00            2.938642
5   100 4   404          2.162431e+02       
1.289940e+01            5.622877
6   100 8   808          4.846450e+02       
1.946854e+01           12.922397
7  1000 2  2002          5.511146e+03       
1.918534e+02           14.694924
8  1000 4  4004          1.194815e+04       
1.499518e+02           29.332463
9  2500 2  5002          3.132878e+04       
4.934896e+02           33.393927
10 1000 8  8008          2.835954e+04       
6.959400e+02           76.881070
11 2500 4 10004          6.874300e+04       
2.197321e+02           68.318820
12 2500 8 20008          8.615465e+05       
9.316236e+04         1806.766254
   hessian_sparse_sd hessian.df_numDeriv_mean hessian.df_numDeriv_sd
1       6.056685e-01                
21.19760               2.433792
2       5.508396e-01                
41.16514               5.648203
3       1.265859e+00                
79.97503               8.748075
4       7.409767e-02               
193.27447              11.790089
5       1.210575e+00               
388.74262              31.184602
6       1.168307e+01               
758.39623             121.225037
7       9.293258e-01              
2042.30264             168.538055
8       1.343366e+00              
4099.05785             221.910206
9       1.377541e+00              
5081.52982             473.962083
10      2.408524e+01              
8323.91140             516.709041
11      2.646135e+00             
10315.16422             832.305746
12      2.272944e+03             
77200.22397           42790.360882
   hessian.df_sparse_mean hessian.df_sparse_sd
1                5.984750            2.1093324
2                9.479244            1.9844323
3               17.337516            4.1347369
4                5.556561            0.2851630
5               10.115250            2.3189536
6               20.160163           17.8116929
7                5.448765            0.5700265
8               10.061368            0.6629470
9                5.416287            0.5421972
10              22.563990            7.1995778
11              10.249990            0.8971753
12             151.538778          217.2031623

I think the complex step takes a similar amount of time, but produces a more
accurate result.

\end{verbatim}

  \end{revQuote}

  \begin{response}

  \end{response}

\item\begin{revQuote}

 p.16  l. -7 "to to compute" -> "to compute"

  \end{revQuote}

  \begin{response}

  \end{response}

\item\begin{revQuote}

While that package is useful, and reasonably demonstrated in the paper,
I think it would be nice to expand the paper in some ways that might be
deemed more "original research". Some possibilities are:

 - Explain and try to assess how much of the speedup is due to simple
sparseness and how much is due to the "sparse patern" (p2) allowing for
perturbing multiple elements together. (I think these are related but
slightly different?)

 - Try to assess how much of the speedup is due to reduce computing demand
and how much is due to different memory demand. (I had the impression in the
larger problems with numDeriv that my computer started to use swap space,
which resulted in a big slow down.)

 - Consider implementing a complex step method, and do a comparison.

 - Assess the difference when multiple CPUs are used. (run.par == FALSE vs
run.par == TRUE)

  \end{revQuote}

  \begin{response}

  \end{response}

\item\begin{revQuote}

  \end{revQuote}

  \begin{response}

  \end{response}

\item\begin{revQuote}

  \end{revQuote}

  \begin{response}

  \end{response}

\item\begin{revQuote}

  \end{revQuote}

  \begin{response}

  \end{response}

\item\begin{revQuote}

  \end{revQuote}

  \begin{response}

  \end{response}

\item\begin{revQuote}

  \end{revQuote}

  \begin{response}

  \end{response}

\item\begin{revQuote}

  \end{revQuote}

  \begin{response}

  \end{response}

\end{enumerate}

\end{document}

### 2016-12-19 nobuild/jss_revision/sparseHessianFD.Rnw
copied review comments into response.tex

--- 2016-06-09 nobuild/jss/sparseHessianFD.Rnw
@@
 
-\date{March 9, 2016}
+\date{July 6, 2016}
 
@@
 knitr::opts_chunk$set(prompt=TRUE, cache=FALSE)
-options(replace.assign=TRUE, width=90,prompt="R> ")
+options(replace.assign=TRUE, width=77, prompt="R> ")
 @
@@
 an important role in statistics.  From a frequentist point of view,
-the inverse of the negative Hessian is the asymptotic covariance of the sampling distribution of a maximum
+the inverse of the negative Hessian is the asymptotic covariance of
+the sampling distribution of a maximum
 likelihood estimator.  In Bayesian analysis, when evaluated at the
@@
 more accurate than computing second-order approximations from the
-objective function. Even when we have derived the Hessian symbolically, in practice it may still be faster to estimate the Hessian
+objective function. Even when we have derived the Hessian
+symbolically, in practice it may still be faster to estimate the Hessian
 using \pkg{sparseHessianFD} than coding it directly.   These are the
@@
 k <- 2
-nv1 <- (N+1)*k
+nv1 <- (N + 1) * k
 nels1 <- nv1^2
-nnz1 <- (N+1)*k^2 + 2*N*k^2
-nnz1LT <- (N+1)*k*(k+1)/2 + N*k^2
+nnz1 <- (N + 1) * k^2 + 2 * N * k^2
+nnz1LT <- (N + 1) * k * (k+1)/2 + N * k^2
 Q <- 1000
-nv2 <- (Q+1)*k
+nv2 <- (Q+1) * k
 nels2 <- nv2^2
-nnz2 <- (Q+1)*k^2 + 2*Q*k^2
-nnz2LT <- (Q+1)*k*(k+1)/2 + Q*k^2
+nnz2 <- (Q+1) * k^2 + 2 * Q * k^2
+nnz2LT <- (Q+1) * k * (k+1)/2 + Q * k^2
 options(scipen=999)
@@
 <<blockarrow, echo=FALSE>>=
-Mat <- as(kronecker(diag(N),matrix(1,k,k)),"sparseMatrix")
-Mat <- rBind(Mat, Matrix(1,k,N*k))
-Mat <- cBind(Mat, Matrix(1, k*(N+1), k))
+Mat <- as(kronecker(diag(N), matrix(1, k, k)),"sparseMatrix")
+Mat <- rBind(Mat, Matrix(1, k, N*k))
+Mat <- cBind(Mat, Matrix(1, k * (N+1), k))
 printSpMatrix(as(Mat,"nMatrix"))
@@
 <<banded, echo=FALSE>>=
-Mat <- kronecker(Matrix(1,k,k), diag(N))
-Mat <- rBind(Mat, Matrix(1,k,N*k))
-Mat <- cBind(Mat, Matrix(1, k*(N+1), k))
+Mat <- kronecker(Matrix(1, k, k), diag(N))
+Mat <- rBind(Mat, Matrix(1, k, N * k))
+Mat <- cBind(Mat, Matrix(1, k * (N + 1), k))
 printSpMatrix(as(Mat,"nMatrix"))
@@
 library("sparseHessianFD")
-bd <- kronecker(diag(3), matrix(TRUE,2,2))
+bd <- kronecker(diag(3), matrix(TRUE, 2, 2))
 Mat <- as(bd, "nMatrix")
@@
 <<eval=FALSE>>=
-f <- obj$fn(x)          ## returns numeric
-df <- obj$gr(x)         ## returns numeric vector
-hess <- obj$hessian(x)  ## returns dgCMatrix
-fngr <- obj$fngr(x)     ## returns list
-fngrhs <- obj$fngrhs(x) ## returns list
+f <- obj$fn(x)
+df <- obj$gr(x)
+hess <- obj$hessian(x)
+fngr <- obj$fngr(x)
+fngrhs <- obj$fngrhs(x)
 @
@@
 T <- binary[["T"]]
-nvars <- as.integer(N*k + k)
-priors <- list(inv.Sigma = rWishart(1,k+5,diag(k))[,,1],
+nvars <- as.integer(N * k + k)
+priors <- list(inv.Sigma = rWishart(1, k + 5, diag(k))[,,1],
   inv.Omega = diag(k))

### 2016-12-19 nobuild/jss_revision/sparseHessianFD.tex
copied review comments into response.tex

--- 2016-03-14 nobuild/jss/sparseHessianFD.tex
@@
-\documentclass[article]{jss}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass[jss]{jss}\usepackage[]{graphicx}\usepackage[]{color}
 %% maxwidth is the original width if it is less than linewidth
@@
   symbolic or algorithmic methods are impractical, inefficient or
-    unavailable. }
-
-\date{March 9, 2016}
+  unavailable.}
+
+\date{July 6, 2016}
 
@@
 elements in the Hessian.  However, only $12,004$ of those elements are
-non-zero.  If we work with only the lower triangle of the Hessian we only need to work with
+non-zero.  If we work with only the lower triangle of the Hessian, then we need to work with
 only 7,003 values.
@@
 R> library("sparseHessianFD")
-R> bd <- kronecker(diag(3), matrix(TRUE,2,2))
+R> bd <- kronecker(diag(3), matrix(TRUE, 2, 2))
 R> Mat <- as(bd, "nMatrix")
@@
 \begin{Sinput}
-R> f <- obj$fn(x)          ## returns numeric
-R> df <- obj$gr(x)         ## returns numeric vector
-R> hess <- obj$hessian(x)  ## returns dgCMatrix
-R> fngr <- obj$fngr(x)     ## returns list
-R> fngrhs <- obj$fngrhs(x) ## returns list
+R> f <- obj$fn(x)
+R> df <- obj$gr(x)
+R> hess <- obj$hessian(x)
+R> fngr <- obj$fngr(x)
+R> fngrhs <- obj$fngrhs(x)
 \end{Sinput}
@@
 R> T <- binary[["T"]]
-R> nvars <- as.integer(N*k + k)
-R> priors <- list(inv.Sigma = rWishart(1,k+5,diag(k))[,,1],
+R> nvars <- as.integer(N * k + k)
+R> priors <- list(inv.Sigma = rWishart(1, k + 5, diag(k))[,,1],
 +   inv.Omega = diag(k))

### 2016-12-19 nobuild/jss_revision/replication.R

@@ -1,0 +1,193 @@
+library("sparseHessianFD")
+library("Matrix")
+library("mvtnorm")
+library("microbenchmark")
+library("doParallel")
+library("dplyr")
+library("tidyr")
+library("numDeriv")
+library("reshape2")
+library("ggplot2")
+
+
+run.par <- TRUE
+if (run.par) registerDoParallel(cores=12) else registerDoParallel(cores=1)
+
+binary_sim <- function(N, k, T) {
+  x.mean <- rep(0,k)
+  x.cov <- diag(k)
+  x.cov[1,1] <- .02
+  x.cov[k,k] <- x.cov[1,1]
+  mu <- seq(-2,2,length=k)
+  Omega <- diag(k)
+  X <- t(rmvnorm(N, mean=x.mean, sigma=x.cov)) ## k x N
+  B <- t(rmvnorm(N, mean=mu, sigma=Omega)) ## k x N
+  XB <- colSums(X * B)
+  log.p <- XB - log1p(exp(XB))
+  Y <- sapply(log.p, function(q) return(rbinom(1,T,exp(q))))
+  list(Y=Y, X=X, T=T)
+}
+
+priors_sim <- function(k) {
+  list(inv.Omega = solve(diag(k)),
+       inv.Sigma = rWishart(1,k+5,diag(k))[,,1])
+}
+
+
+make_funcs <- function(D, priors) {
+  res <- vector("list", length=3)
+  names(res) <- c("fn", "gr", "hessian")
+  res$fn <-  function(pars) binary.f(pars, data=D, priors=priors, order.row=TRUE)
+  res$gr <-  function(pars) binary.grad(pars, data=D, priors=priors, order.row=TRUE)
+  res$hessian <-  function(pars) binary.hess(pars, data=D, priors=priors, order.row=TRUE)
+  return(res)
+}
+
+make_perm <- function(iRow, jCol) {
+  tmp <- sparseMatrix(i=iRow, j=jCol, index1=TRUE, symmetric=TRUE)
+  perm <- order(Matrix::rowSums(tmp), decreasing=TRUE)
+  invperm <- invPerm(perm)
+  L <- tril(tmp[perm,perm])
+  ptr <- Matrix.to.Pointers(L, order="row", index1=TRUE)
+  idx <- ptr[[1]]
+  pntr <- ptr[[2]]
+}
+
+get_id <- function(x) 1:length(x)
+
[…]

### 2016-12-19 nobuild/jss_revision/response.tex
working on response letter and revision

@@
 \documentclass{article}
-
+\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
 \usepackage{amsmath}
@@
 \usepackage{placeins}
-\newtoggle{tikz}
-\toggletrue{tikz}
-
-\iftoggle{tikz}{
-\usepackage{tikz}
-}{}
 
@@
 \newcommand{\Mat}[1]{\mathbf{#1}}
+\newcommand{\Real}[1]{\mathbb{R}^{#1}}
 
@@
 
-I would like to express my appreciation for the time and effort that
-the reviewers invested in their reviews. The comments and suggestions
-are all well-taken, and I have addressed all of the items, point by
-point, in this note. The reviewers’ comments are in italics, and my
-responses follow.
+I would like to express my sincere appreciation for the time and effort that
+the reviewer invested in the review.  It is obvious to me that the
+reviewer went through the paper and code in great detail.  In
+particular, I want to thank the reviewer for bringing the complex step
+method to my attention.  I \emph{never} use complex numbers or
+functions in my own work, so I typically ignore anything with the word
+"complex" in it.  But this review prompted me to learn something new.
+That is always fun.
+
+I am happy to write that I was able to incorporate a large majority of
+the reviewer's suggestions in the revised paper and package. For the
+others, I hope that the editor and reviewer will accept my
+explanations and excuses.  Of course, if there are any strong
+objections to how I responded to the reviewer's concerns, I am happy
+to make further changes in advance of publication.
+
 \vspace{\baselineskip}
@@
 I suspect that the error is a result of the \pkg{algorithm} package
-not being present in your LaTeX installation.  I would like to continue to use it,
+not being present in the reviewer's LaTeX installation.  I would like
+to continue to use this package,
 because it does allow for nicely formatted algorithms.  I believe that this is
 a relatively common package, and one that should be easily installed
-via your closest CTAN mirror.  However, if there is a strong
+through CTAN.  However, if there is a strong
 objection, I may be able to find another way to format the algorithms.
@@
 
+  Note to me:  issue here is that there is no documentation for
+  \func{sparseHessianFD()} as a function.  Good idea to create a
+  documentation page for that.
+
 \end{response}
@@
 
+Note to me:  reword last paragraph on p2 to make the requirements
+sound a lot less burdensome.  Spin these to show that these are
+easy,but acknowledge that supplying a gradient might be difficult.
+
+Then, follow reviewer's advice and explain situations in which the
+requirements might not be met.
+
+Note that there is really no disadvantage to defining the sparsity pattern as
+less sparse than it is (other than suboptimal computation). So, if the
+hierarchical structure is known, but there are potentially independent
+components in the margins, there is nothing wrong with assuming dense margins.
+
 \end{response}
@@
 
+  In order to gain any kind of computational advantage, the sparsity
+  pattern has to be the same for the entire domain of the function. I
+  think that qualifies as a global condition in the reviewer's
+  terminology.
+
+  There is a distinction between a structural zero and a numerical
+  zero.  A structural zero is an element of the Hessian that is zero
+  for all possible variables.  Because that element is always zero, we
+  do not have to reserve memory for it in the sparse matrix object.
+  A Hessian element might be a numerical zero when evaluated at some
+  values, but not others.
+
+  Now, suppose we allowed the sparsity pattern to change when a zero
+  element became non-zero.  At the moment, not only would we need to
+  allocate memory for the new non-zero element, but we would need to
+  rearrange all of the other elements to put the new element in the
+  right place.  The efficiency of sparse matrix operations comes from
+  how the non-zero elements are ordered and indexed.
+
+Note to me:  somewhere in the paper, clarify the difference between
+structural and numerical zeros, and explain why that matters.
+
 \end{response}
@@
   - 'continuous covariates $x_i$,' -> 'continuous covariates $x_i \in
-\\R^k$,'
+\Real{k}$,'
 
   - 'heterogeneous coefficient vector $\beta_i$' -> 'heterogeneous
-coefficient vector $\beta_i \in \\R^k$'
+coefficient vector $\beta_i \in \Real{k}$'
 
@@
 
-It is not clear to me whether obj$fn() and obj$gr() use code as in the true
+It is not clear to me whether obj\$fn() and obj\$gr() use code as in the true
 functions or a modified version using sparse techniques. Some further
@@
 
-\item\begin{revQuote}
-
-  \end{revQuote}
-
-  \begin{response}
-
-  \end{response}
-
-\item\begin{revQuote}
-
-  \end{revQuote}
-
-  \begin{response}
-
-  \end{response}
-
-\item\begin{revQuote}
-
-  \end{revQuote}
-
-  \begin{response}
-
-  \end{response}
-
-\item\begin{revQuote}
-
-  \end{revQuote}
-
-  \begin{response}
-
-  \end{response}
-
-\item\begin{revQuote}
-
-  \end{revQuote}
-
-  \begin{response}
-
-  \end{response}
-
-\item\begin{revQuote}
-
-  \end{revQuote}
-
-  \begin{response}
-
-  \end{response}
-
 \end{enumerate}

### 2016-12-20 nobuild/jss_revision/response.tex
revision

@@
 
-Note to me:  reword last paragraph on p2 to make the requirements
-sound a lot less burdensome.  Spin these to show that these are
-easy,but acknowledge that supplying a gradient might be difficult.
-
-Then, follow reviewer's advice and explain situations in which the
-requirements might not be met.
-
-Note that there is really no disadvantage to defining the sparsity pattern as
-less sparse than it is (other than suboptimal computation). So, if the
-hierarchical structure is known, but there are potentially independent
-components in the margins, there is nothing wrong with assuming dense margins.
+I included this section primarily as a way to avoid overselling the
+method, but the point that the reader might be scared away by this list of
+restrictions is well-taken.  In the revision, I removed the list of
+conditions, and wove them into the text.  The result takes a more
+positive tone that in the original submission.
 
@@
 
-  In order to gain any kind of computational advantage, the sparsity
-  pattern has to be the same for the entire domain of the function. I
-  think that qualifies as a global condition in the reviewer's
-  terminology.
+  The list of conditions is no longer in the paper, but to this point,
+  I added a defintion of a ``structural zero'' in the third
+  paragraph. The sparsity pattern represents the structure of the
+  Hessian, not its value, so any element that \emph{could} be zero,
+  but does not have to be zero, is not a structural zero.  Thus, the
+  condition is global and not local.
 
-  There is a distinction between a structural zero and a numerical
-  zero.  A structural zero is an element of the Hessian that is zero
-  for all possible variables.  Because that element is always zero, we
-  do not have to reserve memory for it in the sparse matrix object.
-  A Hessian element might be a numerical zero when evaluated at some
-  values, but not others.
-
-  Now, suppose we allowed the sparsity pattern to change when a zero
-  element became non-zero.  At the moment, not only would we need to
-  allocate memory for the new non-zero element, but we would need to
-  rearrange all of the other elements to put the new element in the
-  right place.  The efficiency of sparse matrix operations comes from
-  how the non-zero elements are ordered and indexed.
-
-Note to me:  somewhere in the paper, clarify the difference between
-structural and numerical zeros, and explain why that matters.
-
+  Note that other than
+  a potential loss of efficiency, there is no disadvantage to erring
+  on the side of an element being non-zero.  
 \end{response}

### 2016-12-20 nobuild/jss_revision/sparseHessianFD.Rnw
revision

@@
 knitr::render_sweave()
-knitr::opts_chunk$set(prompt=TRUE, cache=FALSE)
+knitr::opts_chunk$set(prompt=TRUE, cache=FALSE) %$
 options(replace.assign=TRUE, width=77, prompt="R> ")
@@
 applications with hundreds of thousands of variables, computing the
-Hessian even once might not be practical under time, storage or processor constraints.  Hierarchical models, in
+Hessian even once might not be practical under time, storage or
+processor constraints.  Hierarchical models, in
 which each additional heterogeneous unit is associated with its own subset of
@@
 However, for many problems, the Hessian is
-\emph{sparse}, meaning that the proportion of non-zero elements in the
-Hessian is small.  Consider a log
+\emph{sparse}, meaning that the proportion of  ``structural''  zeros
+(Hessian matrix elements that are always zero, regardless of the value at which
+function is estimated) is high.  Consider a log
 posterior density in a Bayesian hierarchical
@@
 The package defines the \class{sparseHessianFD} class, whose initializer
-requires the user to provide functions that compute the objective
-function and the gradient, as well as the sparsity pattern of the
-Hessian.  Some users may find those requirements burdensome, and we
-emphasize that there may be some applications for
-which \pkg{sparseHessianFD} is not an appropriate package to use. To
-extract the maximum benefit from using \pkg{sparseHessianFD}, we need
-to accept some conditions and assumptions.
-
-\begin{enumerate}
-  \item Preferred alternatives to computing the Hessian are not
-  available.  Finite differencing is not generally a ``first choice''
-  method.  Deriving a gradient or Hessian symbolically, and writing a
-  subroutine to compute it, will give an exact answer, but might be
-  tedious or difficult to implement. Algorithmic differentiation (AD) is
-  probably the most efficient method, but requires specialized
-  libraries that, at this moment, are not yet broadly available in \proglang{R}.
- \pkg{sparseHessianFD} makes the most sense when the gradient is easy to compute, but the Hessian is
-  not.
-  \item The application can tolerate the approximation error in the Hessian
-  that comes with finite differencing methods.
-\item The objective function $f(x)$ is twice differentiable, and can be
-  computed ``quickly and easily,'' even for a large number of
-  variables.  We leave the definition of ``quickly and easily''
-  intentionally murky, since no method of differentiation can overcome
-  pathologies in a function that itself is hard to compute.
-\item The gradient can
-  be computed quickly, easily and \emph{exactly} (within machine
-  precision).   We do not recommend using finite differenced gradients
-  when computing finite differenced Hessians.  The
-  approximation errors will be compounded, and the time
-  complexity of computing a gradient grows with the number of
-  variables.
-\item The sparsity pattern is known in advance, and does not depend
-    on the values of the variables.
-\end{enumerate}
-
-Our perspective on the requirements to use \pkg{sparseHessianFD} is
-that deriving a vector of first
+requires the user to provide functions that compute an objective
+function, its gradient, and the sparsity pattern of its Hessian
+matrix. The gradient should be computed exactly (within machine
+precision), as opposed to being numerically estimated (e.g., by finite
+differencing). The sparsity pattern (e.g., location of structural
+zeros) must be known in advance, and cannot vary across the domain of the
+objective function. The only functions and methods of the class that
+the end user should need to use are the
+initializer, methods that return the Hessian
+in a sparse compressed format, and perhaps some utility functions that
+simplify the construction of the sparsity pattern.  The class also
+defines methods that partition the variables into groups that can be perturbed together in a finite
+differencing step, and recovers the elements of the Hessian via linear
+substitution.  Those methods perform most of the work, but should be
+invisible to the user.
+
+As with any computing method or algorithm, there are boundaries around
+the space of applications for which \pkg{sparseHessianFD} is the right
+tool for the job. In general, finite differencing is not a ``first
+choice'' method because the result is not exact, so
+\pkg{sparseHessianFD} should not be used when the application cannot
+tolerate any error, no matter how small.  Also, we admit that some
+users might balk at having to provide an exact gradient, even though
+the Hessian will be estimated numerically. However, deriving a vector of first
 derivatives, and writing \proglang{R} functions to compute them, is a
@@
 situations in which \pkg{sparseHessianFD} adds the most value to the
-statistician's toolbox.  If AD software is available to compute the gradient, then
-it is probably available for sparse Hessians as well, and
-\pkg{sparseHessianFD} would not be needed.  Without AD-computed or
-symbolically derived Hessians, \pkg{sparseHessianFD} is the best
-available option.
-
-The only functions and methods the end user should need to use are the
-\class{sparseHessianFD} initializer, methods that return the Hessian
-in a sparse compressed format, and perhaps some utility functions that
-simplify the construction of the sparsity pattern.  The class also
-defines methods that partition
-the variables into groups that can be perturbed together in a finite
-differencing step, and recovers the elements of the Hessian via linear
-substitution.  Those methods perform most of the work, but should be
-invisible to the user.
+statistician's toolbox.
 

### 2016-12-21 nobuild/jss_revision/sparseHessianFD.Rnw
more on revision

@@
 
-\date{July 6, 2016}
+\date{December 20, 2016}
 
@@
 
-\newenvironment{knitrout}{\begin{figure}}{\end{figure}}
+%%\newenvironment{knitrout}{\begin{figure}}{\end{figure}}
 
@@
 knitr::render_sweave()
-knitr::opts_chunk$set(prompt=TRUE, cache=FALSE) %$
+knitr::opts_chunk$set(prompt=TRUE, cache=FALSE) #$
 options(replace.assign=TRUE, width=77, prompt="R> ")
@@
 the decision variables into $C$ mutually exclusive groups so that the number of gradient
-evaluations is reduced.  Let $\Mat{G}$ be a $M\times C$
-matrix, where $\Mat{G}_{mc}=\delta$ if variable $m$ belongs to group $c$, and zero
-otherwise.  Define $\Mat{G}_c$ as the $c$th column of $\Mat{G}$.
-
-Next, let $\Mat{Y}$ be a $M\times C$ matrix, where each column is a
+evaluations is reduced.  Define $\Mat{G}\in\Real{M\times C}$ where $\Mat{G}_{mc}=\delta$ if variable $m$ belongs to group $c$, and zero
+otherwise.  Define $\Mat{G}_c\in\Real{M}$ as the $c$th column of $\Mat{G}$.
+
+Next, define $\Mat{Y}\in\Real{M\times C}$ such that each column is a
 difference in gradients.
@@
  purchase.  The heterogeneous parameter $p_i$ is the same for all $T$
- opportunities, so $y_i$ is a binomial random variable.  Define each $p_i$ such that it depends on both $k$ continuous covariates
- $x_i$, and a heterogeneous coefficient vector $\beta_i$.
+ opportunities, so $y_i$ is a binomial random variable.  Define each
+ $p_i$ such that it depends on a covariate vector
+ $x_i\in\Real{k}$, and a heterogeneous coefficient vector $\beta_i\in\Real{k}$.
 \begin{align}
@@
 
-The coefficients are distributed across the population of households
-following a multivariate normal distribution with mean $\mu$ and
-covariance $\Mat{\Sigma}$.   Assume that we know $\Mat{\Sigma}$, but not $\mu$.
+The coefficient vectors $\beta_i$ are distributed across the population of households
+following a multivariate normal distribution with mean $\mu\in\Real{k}$ and
+covariance $\Mat{\Sigma}\in\Real{k\times k}$.   Assume that we know $\Mat{\Sigma}$, but not $\mu$.
 Instead, place a multivariate normal prior on $\mu$, with mean $0$ and
-covariance $\Mat{\Omega}$.  Thus, each $\beta_i$, and $\mu$ are
+covariance $\Mat{\Omega}\in\Real{k\times k}$.  Thus, each $\beta_i$, and $\mu$ are
 $k-$dimensional vectors, and the total number of unknown variables in the model is $(N+1)k$.
@@
 ordered within the vector. One such ordering is to group all of the
-coefficients for each unit together.
+coefficients in the $\beta_i$ for each unit together.
 \begin{align}

### 2016-12-21 nobuild/jss_revision/sparseHessianFD.tex
more on revision

@@
 
-\newenvironment{knitrout}{\begin{figure}}{\end{figure}}
+%%\newenvironment{knitrout}{\begin{figure}}{\end{figure}}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}

### 2016-12-21 src/RcppExports.cpp

@@ -1,4 +1,4 @@
-// Generated by using Rcpp::compileAttributes() -> do not edit by hand
+// This file was generated by Rcpp::compileAttributes
 // Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
 
 #include <RcppEigen.h>
@@ -10,21 +10,21 @@ using namespace Rcpp;
 Rcpp::IntegerVector get_colors(const IntegerVector& pntr, const IntegerVector& idx, const int nvars);
 RcppExport SEXP sparseHessianFD_get_colors(SEXP pntrSEXP, SEXP idxSEXP, SEXP nvarsSEXP) {
 BEGIN_RCPP
-    Rcpp::RObject rcpp_result_gen;
-    Rcpp::RNGScope rcpp_rngScope_gen;
+    Rcpp::RObject __result;
+    Rcpp::RNGScope __rngScope;
     Rcpp::traits::input_parameter< const IntegerVector& >::type pntr(pntrSEXP);
     Rcpp::traits::input_parameter< const IntegerVector& >::type idx(idxSEXP);
     Rcpp::traits::input_parameter< const int >::type nvars(nvarsSEXP);
-    rcpp_result_gen = Rcpp::wrap(get_colors(pntr, idx, nvars));
-    return rcpp_result_gen;
+    __result = Rcpp::wrap(get_colors(pntr, idx, nvars));
+    return __result;
 END_RCPP
 }
 // subst
 Rcpp::S4 subst(const Rcpp::NumericMatrix& Y, const Rcpp::IntegerVector& colors, const Rcpp::IntegerVector& jCol, const Rcpp::IntegerVector& ipntr, const double& delta, const int& nvars, const int& nnz);
 RcppExport SEXP sparseHessianFD_subst(SEXP YSEXP, SEXP colorsSEXP, SEXP jColSEXP, SEXP ipntrSEXP, SEXP deltaSEXP, SEXP nvarsSEXP, SEXP nnzSEXP) {
 BEGIN_RCPP
-    Rcpp::RObject rcpp_result_gen;
-    Rcpp::RNGScope rcpp_rngScope_gen;
+    Rcpp::RObject __result;
+    Rcpp::RNGScope __rngScope;
     Rcpp::traits::input_parameter< const Rcpp::NumericMatrix& >::type Y(YSEXP);
     Rcpp::traits::input_parameter< const Rcpp::IntegerVector& >::type colors(colorsSEXP);
[…]

### 2016-12-21 nobuild/jss_revision/response.tex
more on revision
added hessian_complex functions for complex step method:

@@
 \begin{response}
-
+  Thank you for the suggestion.  Changes have been made throughout the paper.
 \end{response}
@@
 \begin{response}
-
+  Based on this comment, I rewrote Section 2.2 to distinguish the
+  cross-partial derivatives $\parD{2}{}$ from the corresponding
+  elements of the Hessian $\hess{}{}$.  I did this by defining $x_i$ as
+  a subset of the elements of $x$, and Ind$(x_i)$ as an 
+  ``indexing'' function.  The revised version is slightly more formal
+  and precise, and I hope that this is the kind of change you had in
+  mind.  That said, I do not think it makes the paper easier to read,
+  and, with your permission, I would prefer to revert to the simpler
+  language in the original manuscript.
 \end{response}
@@
 \begin{response}
+No, the descriptions of the two sparsity patterns are correct. The
+first index in the subscript of $\beta$ refers to a household, and the
+second index refers to a covariate. Thus,
+$\parD{2}{\beta_{11},\beta_{12}}\neq 0$, but
+$\parD{2}{\beta_{11},\beta_{21}}=0$.  When the second index changes
+before the first, the pattern is ``block-arrow.''  When the first
+index changes before the second, the pattern is ``banded.''
 
@@
 \begin{response}
-
+  I removed the \func{R>} prompt and now refer to that code as a usage
+  syntax.  
 \end{response}
@@
 \begin{response}
-
+  Done.
 \end{response}
@@
 purposes, after all, it is almost the main purpose of the package.)
-p10. "where ... represents all other named arguments"
-   I think the usual usage is the ... represents the arguments other than
-the "named" ones, so it is probably better to just say "other arguments".
   \end{revQuote}
@@
 \begin{response}
-
+  Good point.  I revised that section to explain that evaluations of
+  the function and gradient have to be identical to the true values,
+  and I even replaced \func{all.equal} with \func{identical} to
+  highlight that fact. I then explained that the
+  \class{sparseHessianFD} calculation of the Hessian will not be
+  identical to the true value, and report the mean relative
+  difference.  I chose the mean relative difference over maximum
+  absolute difference because the former is what the \func{all.equal}
+  function uses to test if two matrices are equal within numeric tolerance.
 \end{response}
@@
 \begin{response}
-
+  Fixed.  Thanks.
 \end{response}
@@
   \begin{response}
-
+    Fixed.  Thanks.
   \end{response}

### 2016-12-21 nobuild/jss_revision/sparseHessianFD.Rnw
added hessian_complex functions for complex step method:

@@
  purchase.  The heterogeneous parameter $p_i$ is the same for all $T$
- opportunities, so $y_i$ is a binomial random variable.  Define each
- $p_i$ such that it depends on a covariate vector
- $x_i\in\Real{k}$, and a heterogeneous coefficient vector $\beta_i\in\Real{k}$.
-\begin{align}
-  p_i=\frac{\exp(x_i'\beta_i)}{1+\exp(x_i'\beta_i)},~i=1 ... N
+ opportunities, so $y_i$ is a binomial random variable.
+
+ Let $\beta_i\in\Real{k}$ be a heterogeneous coefficient vector that
+ is specific to household $i$, such that
+ $\beta_i=(\beta_{i1},\dotsc,\beta_{ik})$. Similarly,
+ $z_i\in\Real{k}$ is a vector of household-specific covariates. Define each
+ $p_i$ such that the log odds of $p_i$ is a linear function of
+ $\beta_i$ and $z_i$, but does not depend directly on $\beta_j$ and $z_j$ for
+ another household $j\neq i$.
+\begin{align}
+  p_i=\frac{\exp(z_i'\beta_i)}{1+\exp(z_i'\beta_i)},~i=1 ... N
 \end{align}
@@
 following a multivariate normal distribution with mean $\mu\in\Real{k}$ and
-covariance $\Mat{\Sigma}\in\Real{k\times k}$.   Assume that we know $\Mat{\Sigma}$, but not $\mu$.
-Instead, place a multivariate normal prior on $\mu$, with mean $0$ and
-covariance $\Mat{\Omega}\in\Real{k\times k}$.  Thus, each $\beta_i$, and $\mu$ are
-$k-$dimensional vectors, and the total number of unknown variables in the model is $(N+1)k$.
+covariance $\Mat{\Sigma}\in\Real{k\times k}$.   Assume that we know
+$\Mat{\Sigma}$, but not $\mu$, so we place a multivariate normal prior
+on $\mu$, with mean $0$ and
+covariance $\Mat{\Omega}\in\Real{k\times k}$.  Thus, the parameter
+vector $x\in\Real{(N+1)k}$ consists of the $Nk$ elements in the $N$ $\beta_i$ vectors,
+and the $k$ elements in $\mu$.
 
@@
   \label{eq:LPD}
-  \log \pi(\beta_{1:N},\mu|\Mat{Y}, \Mat{X}, \Mat{\Sigma},\Mat{\Omega})=\sum_{i=1}^N\left(p_i^{y_i}(1-p_i)^{T-y_i}
+  \log \pi(\beta_{1:N},\mu|\Mat{Y}, \Mat{Z}, \Mat{\Sigma},\Mat{\Omega})=\sum_{i=1}^N\left(p_i^{y_i}(1-p_i)^{T-y_i}
   -\frac{1}{2}\left(\beta_i-\mu\right)^\top\Mat{\Sigma}^{-1}\left(\beta_i-\mu\right)\right)
@@
 \end{align}
-
-All of the $\beta_i$ are correlated with
-$\mu$.  Thus, $\hess{\beta_{ik},\mu_k}{}\neq 0$ for all $i$. Since the $\beta_i$ are independently distributed, and the $y_i$ are
-conditionally independent, the cross-partial derivatives
-$\hess{\beta_i,\beta_j}{}=\parD{2}{\beta_i, \beta_j}{}=0$ for all $i\neq
-j$.  When $N$ is much greater than $k$, the Hessian of the log
-posterior density will be sparse.
 
@@
 
-The sparsity pattern depends on how the variables are
-ordered within the vector. One such ordering is to group all of the
+Let $x_1$ and $x_2$ be two subsets of elements of $x$.  
+Define $\parD{2}{x_1,x_2}$ as the product set of cross-partial
+derivatives between all elements in $x_1$ and all elements in $x_2$.
+Furthermore, define an indexing function Ind$(x_i)$ that returns the
+indices in the parameter vector
+$x$ that correspond to the elements in $x_i$, and $\hess{x_1,x_2}{}$
+as the Hessian elements in rows Ind$(x_1)$ and columns Ind$(x_2)$, and
+in rows Ind$(x_2)$ and columns Ind$(x_1)$ (from the symmetry of the
+Hessian).
+
+From the log posterior density in Equation \ref{eq:LPD}, we can see
+that $\parD{2}{\beta_i, \beta_i}\neq 0$ (one element of $\beta_i$ could be
+correlated with another element of $\beta_i$), and that, for all $i$,
+$\parD{2}{\beta_i, \mu}=0$ (because $\mu$ is the prior mean of each
+$\beta_i$).  However, since the $\beta_i$ and $\beta_j$ are independently distributed, and the $y_i$ are
+conditionally independent, the cross-partial derivatives
+$\parD{2}{\beta_i, \beta_j}=0$ for all $i\neq j$. When $N$ is much
+greater than $k$, there will be many more zero cross-partial
+derivatives than non-zero.  Regardless of how each $\parD{2}{}$ is
+mapped to $\hess{}{}$ through the indexing function, the Hessian of the log
+posterior density will be sparse.
+
+The sparsity pattern depends on the indexing function; that is, on how the variables are
+ordered in $x$. One such ordering is to group all of the
 coefficients in the $\beta_i$ for each unit together.
@@
 \func{sparseHessianFD}.  Applying the default values for the optional
-arguments,
-<<eval=FALSE>>=
+arguments, the usage syntax to create a \class{sparseHessianFD} object is
+<<eval=FALSE, prompt=FALSE>>=
 obj <- sparseHessianFD(x, fn, gr, rows, cols, ...)
 @
-where \funcarg{...} represents all other named arguments that are
+where \funcarg{...} represents all other arguments that are
 passed to \funcarg{fn} and \funcarg{gr}.
@@
 
-Finally, we create an instance of a \class{sparseHessianFD} object,
-evaluate the function, gradient and Hessian, and compare the output to
-the true values.
-
-<<usingSparseHessianFD>>=
+Finally, we create an instance of a \class{sparseHessianFD} object.
+Evaluations of the function and gradient will generate the same
+results as the true values because they are computed using the same function.
+
+<<usingSparseHessianFD1>>=
 obj <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
@@
 f <- obj$fn(P)
-all.equal(f, true.f)
+identical(f, true.f)
 gr <- obj$gr(P)
-all.equal(gr, true.grad)
+identical(gr, true.grad)
+@
+
+The evaluation of the Hessian is a finite differenced approximation,
+so it is very close to, but not identical to, the true value, in terms
+of mean relative difference.
+
+<<echo=FALSE>>=
+options(scipen=0)
+@
+
+<<usingSparseHessianFD2>>=
 hs <- obj$hessian(P)
-all.equal(hs, true.hess)
-@
+mean(abs(hs-true.hess))/mean(abs(hs))
+@
+
+<<echo=FALSE>>=
+options(scipen=999)
+@
+
+Note that variables \variable{f} and \variable{true.f} were computed
+using the same exact calculation, as were variables
+\variable{gr} and \variable{true.grad}.In contrast,
+\variable{true.hess} was computed using an exact calculation, while
+\variable{hs} is the sparse finite difference approximation.
+
+were computed by the same functions
 
@@
 $\{1,3,5\}$ have empty intersection sets.  The intersection graph in
-Figure~\ref{fig:graph2pic} has fewer edges, and can be colored
+Figure~\ref{fig:graph2int} has fewer edges than Figure~\ref{fig:graph1int}, and can be colored
 with only three colors.
@@
 variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
-and standard deviations (across 20 replications) for the time (in milliseconds) to
-to compute a Hessian using each package.  The difference in run times is dramatic,
+and standard deviations (across 20 replications) for the time (in
+milliseconds) to compute a Hessian using each package.  The difference in run times is dramatic,
 especially because the computation time for \pkg{numDeriv} grows

### 2016-12-21 nobuild/jss_revision/sparseHessianFD.tex
added hessian_complex functions for complex step method:

@@
 
-All of the $\beta_i$ are correlated with
-$\mu$.  Thus, $\hess{\beta_{ik},\mu_k}{}\neq 0$ for all $i$. Since the $\beta_i$ are independently distributed, and the $y_i$ are
+\subsection[Sparsity patterns]{Sparsity patterns}\label{sec:sparsity}
+
+Let $x_1$ and $x_2$ be two subsets of elements of $x$.  
+Define $\parD{2}{x_1,x_2}$ as the product set of cross-partial
+derivatives between all elements in $x_1$ and all elements in $x_2$.
+Furthermore, define an indexing function Ind$(x_i)$ that returns the
+indices in the parameter vector
+$x$ that correspond to the elements in $x_i$, and $\hess{x_1,x_2}{}$
+as the Hessian elements in rows Ind$(x_1)$ and columns Ind$(x_2)$, and
+in rows Ind$(x_2)$ and columns Ind$(x_1)$ (from the symmetry of the
+Hessian).
+
+From the log posterior density in Equation \ref{eq:LPD}, we can see
+that $\parD{2}{\beta_i, \beta_i}\neq 0$ (one element of $\beta_i$ could be
+correlated with another element of $\beta_i$), and that, for all $i$,
+$\parD{2}{\beta_i, \mu}=0$ (because $\mu$ is the prior mean of each
+$\beta_i$).  However, since the $\beta_i$ and $\beta_j$ are independently distributed, and the $y_i$ are
 conditionally independent, the cross-partial derivatives
-$\hess{\beta_i,\beta_j}{}=\parD{2}{\beta_i, \beta_j}{}=0$ for all $i\neq
-j$.  When $N$ is much greater than $k$, the Hessian of the log
+$\parD{2}{\beta_i, \beta_j}=0$ for all $i\neq j$. When $N$ is much
+greater than $k$, there will be many more zero cross-partial
+derivatives than non-zero.  Regardless of how each $\parD{2}{}$ is
+mapped to $\hess{}{}$ through the indexing function, the Hessian of the log
 posterior density will be sparse.
 
-\subsection[Sparsity patterns]{Sparsity patterns}\label{sec:sparsity}
-
-The sparsity pattern depends on how the variables are
-ordered within the vector. One such ordering is to group all of the
+The sparsity pattern depends on the indexing function; that is, on how the variables are
+ordered in $x$. One such ordering is to group all of the
 coefficients in the $\beta_i$ for each unit together.
@@
 \func{sparseHessianFD}.  Applying the default values for the optional
-arguments,
+arguments, the usage syntax to create a \class{sparseHessianFD} object is
 \begin{Schunk}
 \begin{Sinput}
-R> obj <- sparseHessianFD(x, fn, gr, rows, cols, ...)
+obj <- sparseHessianFD(x, fn, gr, rows, cols, ...)
 \end{Sinput}
 \end{Schunk}
-where \funcarg{...} represents all other named arguments that are
+where \funcarg{...} represents all other arguments that are
 passed to \funcarg{fn} and \funcarg{gr}.
@@
 
-Finally, we create an instance of a \class{sparseHessianFD} object,
-evaluate the function, gradient and Hessian, and compare the output to
-the true values.
+Finally, we create an instance of a \class{sparseHessianFD} object.
+Evaluations of the function and gradient will generate the same
+results as the true values because they are computed using the same function.
 
@@
 R> f <- obj$fn(P)
-R> all.equal(f, true.f)
+R> identical(f, true.f)
 \end{Sinput}
@@
 R> gr <- obj$gr(P)
-R> all.equal(gr, true.grad)
-\end{Sinput}
-\begin{Soutput}
-[1] TRUE
-\end{Soutput}
-\begin{Sinput}
-R> hs <- obj$hessian(P)
-R> all.equal(hs, true.hess)
+R> identical(gr, true.grad)
 \end{Sinput}
@@
 \end{Schunk}
+
+The evaluation of the Hessian is a finite differenced approximation,
+so it is very close to, but not identical to, the true value, in terms
+of mean relative difference.
+
+\begin{Schunk}
+\begin{Sinput}
+R> hs <- obj$hessian(P)
+R> mean(abs(hs-true.hess))/mean(abs(hs))
+\end{Sinput}
+\begin{Soutput}
+[1] 1.1217e-08
+\end{Soutput}
+\end{Schunk}
+
+Note that variables \variable{f} and \variable{true.f} were computed
+using the same exact calculation, as were variables
+\variable{gr} and \variable{true.grad}.In contrast,
+\variable{true.hess} was computed using an exact calculation, while
+\variable{hs} is the sparse finite difference approximation.
+
+were computed by the same functions
 

### 2016-12-21 R/sparseHessianFD-class.R

@@ -166,12 +166,27 @@ sparseHessianFD <-
                         gr1(x+d) - grad.x
                     },
 
+                    fd_complex = function(d, x) {
+                        Im(gr1(x + 1i * d))
+                    },
+
                     hessian = function(x) {
                         "Return sparse Hessian, evaluated at x, as a dgCMatrix object."
                         usingMethods(fd)
                         stopifnot(ready)
                         grad.x <- gr1(x)
                         Y2 <- apply(D, 2, fd, x = x, grad.x = grad.x)
+                        Y <- Y2[perm,]
+                        res <- subst(Y, colors,
+                                     idx-index1, pntr-index1, delta, nvars, nnz)
+                        return(res[invperm,invperm])
+                    },
+
+                    hessian_complex = function(x) {
+                        "Return sparse Hessian, evaluated at x, as a dgCMatrix object, using complex step method"
+                        usingMethods(fd_complex)
+                        stopifnot(ready)             
+                        Y2 <- apply(D, 2, fd_complex, x = x)
                         Y <- Y2[perm,]
                         res <- subst(Y, colors,
                                      idx-index1, pntr-index1, delta, nvars, nnz)

### 2016-12-22 nobuild/jss_revision/response.tex
complex step method works

@@
 \begin{response}
-  I removed the \func{R>} prompt and now refer to that code as a usage
+  I removed the \func{R>} prompt, and now refer to that code as a usage
   syntax.  
@@
 \begin{response}
-
+  The complex step function is now implemented.  It appears to work on
+  the example function, which I had to modify because \proglang{R}
+  does not have a complex implementation of the \func{log1p}.  I added
+  a couple of paragraphs about this method in the Algorithms section,
+  but I did not want to oversell it for a few reasons.  First, I am
+  still generally unfamiliar with this method, so I do not want to
+  push it too hard.  Second, I think that the requirements of
+  holomorphism, and of being able to accept and return complex values,
+  does significantly limit the utility of the method.
 \end{response}

### 2016-12-22 nobuild/jss_revision/sparseHessianFD.Rnw
complex step method works

@@
 \newcommand{\Real}[1]{\mathbb{R}^{#1}}
+\DeclareMathOperator{\Imag}{Im}
 
@@
 knitr::render_sweave()
-knitr::opts_chunk$set(prompt=TRUE, cache=FALSE) #$
+knitr::opts_chunk$set(prompt=TRUE, cache=FALSE,error=FALSE) #$
 options(replace.assign=TRUE, width=77, prompt="R> ")
@@
   indexing.  If \variable{FALSE}, zero-based indexing is used.\\
+  complex&If \variable{TRUE}, the complex step method is used.
+                If \variable{FALSE} (the default), a simple finite
+                differencing of gradients is used.\\
 ...& Additional arguments to be passed to \func{fn} and \func{gr}.\\
@@
 
+\subsection[Complex step method]{Complex step method}
+
+The \pkg{sparseHessianFD} package also allows for
+estimation of the Hessian using the complex step method
+\citep{SquireTrapp1998} as an alternative to finite differencing. In
+this method, the definition of $\Mat{Y}_c$ in Equation \ref{eq:Yg} is replaced with
+\begin{align}
+  \label{eq:complexYg}
+  \Mat{Y}_c&=\Imag(\nabla f(x+iG_c))
+\end{align}
+where $i=\sqrt{-1}$ and $\Imag$ returns the imaginary part of a complex
+number. The advantage of this method is that there is no longer a
+subtraction operation to generate truncation error. Thus, $\delta$ can
+be made arbitrarily small, and the resulting estimate of the Hessian
+is more accurate.
+
+To use the complex step method, include the argument
+\funcarg{complex=TRUE} in the call to the initializer of the \class{sparseHessianFD}
+object. The functions passed as \funcarg{fn} and \funcarg{gr} must both
+accept a complex argument, and return a complex
+result, even though we are differentiating a real-valued function. Although \pkg{base} \proglang{R} supports complex arguments
+for basic arithmetic, trigonometric, logarithmic and exponential
+functions, many other common functions (e.g.,  \func{gamma}, \func{log1p},
+\func{expm1}, and the probability distribution functions) do not have
+complex implementations. Furthermore, the complex step method is valid
+only if the function is holomorphic
+(complex differentiable). The methods in \class{sparseHessianFD} do
+\emph{not} check that this is the case for the function at hand. We convey the
+following warning from the documentation of the \pkg{numDeriv} package
+\citep{R_numDeriv}, which also implements the complex step method:
+``Avoid this method if you do not know that your function is
+suitable. Your mistake may not be caught and the results will be
+spurious.''
+
+In short, the complex step method is available to users who want
+improved accuracy and speed, and know what they are doing.
+
 \subsection[Software libraries]{Software libraries}

### 2016-12-22 R/sparseHessianFD-class.R

@@ -1,5 +1,5 @@
 ## This file is part of the sparseHessianFD package
-## Copyright (C) 2015-2016 Michael Braun
+## Copyright (C) 2015-2017 Michael Braun
 ##
 ##' @title sparseHessianFD
 ##' @name sparseHessianFD-class
@@ -14,6 +14,7 @@
 ##the gradient to compute the Hessian. Defaults to sqrt(.Machine$double.eps).
 ##' @field index1 TRUE if rows and cols use 1-based (R format)
 ##indexing (FALSE for 0-based (C format) indexing.
+##' @field complex TRUE if Hessian will be computed using the complex step method, and FALSE (default) if using finite differences.
 ##' @field D raw finite differences (internal use only)
 ##' @field nvars Number of variables (length of x)
 ##' @field nnz Number of non-zero elements in the lower triangle of the Hessian.
@@ -61,6 +62,7 @@ sparseHessianFD <-
                     jCol = "numeric",
                     delta = "numeric",
                     index1 = "logical",
+                    complex = "logical",
                     D = "matrix",
                     nvars = "integer",
                     nnz = "integer",
@@ -73,18 +75,15 @@ sparseHessianFD <-
                 methods = list(
                     initialize = function(x, fn, gr, rows, cols, direct=NULL,
                                           delta=sqrt(.Machine$double.eps),
-                                          index1 = TRUE, eps=NULL,...) {
-                        "Initialize object with functions to compute the objective function and gradient (fn and gr), row and column indices of non-zero elements (rows and cols), an initial variable vector x at which fn and gr can be evaluated, a finite differencing parameter delta, flags for 0 or 1-based indexing (index1), whether sparsity pattern is just for the lower triangle (indexLT), and other arguments (...) to be passed to fn and gr."
+                                          index1 = TRUE, eps=NULL, complex=FALSE, ...) {
+                        "Initialize object with functions to compute the objective function and gradient (fn and gr), row and column indices of non-zero elements (rows and cols), an initial variable vector x at which fn and gr can be evaluated, a finite differencing parameter delta, flags for 0 or 1-based indexing (index1), whether the complex step method will be used, and other arguments (...) to be passed to fn and gr."
 
                         if (!is.null(direct)) {
                             warning(" 'direct' argument is ignored. Only indirect method is, and will be, supported.")
-                        }
-                        ## if (!is.null(eps)) {
-                        ##     warning(" 'eps' argument is deprecated and ignored. Use delta instead.")
[…]

### 2016-12-22 nobuild/jss_revision/response.tex
more work on revision and response

--- 2016-12-21 nobuild/jss_revision/response.tex
@@
 \usepackage{geometry}
-\geometry{margin=1.5in}
+\geometry{margin=1.2in}
 
@@
 \begin{verbatim}
-print(obj$fn(P) - binary.f(P, 
-      data=binary, priors=priors, order.row=order.row), digits=20)
-[1] 0
-
-print(max(abs(obj$gr(P) - binary.grad(P, 
-         data=binary, priors=priors,
-order.row=order.row))),digits=20)
-[1] 0
-
+Code removed
 \end{verbatim}
@@
 
-\item \begin{revQuote}
+\item \label{it:complex1}\begin{revQuote}
 I think it would be instructive to add some of the following comparisons
@@
 \begin{verbatim}
-max(abs(
-     numDeriv::hessian( binary.f, P, method="Richardson", 
-                    data=binary, priors=priors,
-order.row=order.row) 
-   - binary.hess(P, data=binary, priors=priors, order.row=order.row)))
-[1] 0.0001610595
+Hessian Richardson
 
@@
 \begin{verbatim}
-max(abs( 
-    numDeriv::jacobian( binary.grad, P, method="Richardson", 
-                   data=binary, priors=priors,
-order.row=order.row)
-  - binary.hess(P, data=binary, priors=priors, order.row=order.row)))
-[1] 3.268224e-07
+Jacobian Richardson
 \end{verbatim}
@@
 \begin{verbatim}
-max(abs( 
-    numDeriv::jacobian( binary.grad, P, method="simple", 
-                   data=binary, priors=priors,
-order.row=order.row)
-  - binary.hess(P, data=binary, priors=priors, order.row=order.row)))
-
-[1] 0.0008255852
+Jacobian simple
 
@@
 \begin{verbatim}
-max(abs( 
-  numDeriv::jacobian( binary.grad, P, method="complex", 
-                  data=binary, priors=priors,
-order.row=order.row)
- - binary.hess(P, data=binary, priors=priors, order.row=order.row)))
-
-[1] 7.105427e-15
+Jacobian complex
 \end{verbatim}
@@
   \begin{response}
-
+    See response to Item \ref{it:complex2}
   \end{response}
 
-\item \begin{revQuote}
+\item\label{it:complex2} \begin{revQuote}
 A possible extension to the package would be to implement the complex
@@
 \begin{response}
+
+Here I am combining my responses to items \ref{it:complex1} and
+\ref{it:complex2} together.
+
+  The complex step function is now implemented in the package, and is
+  introduced in Section 2.5 of the paper.  You
+  are correct that ``most R primitives work with complex numbers,''
+  but there are many commonly used functions that do not, including 
+  \func{log1p}, \func{gamma}, and the probability distribution
+  functions. This limitation on the kinds of functions that can be
+  implemented easily in R, combined with the requirement that the function be
+  holomorphic, and my own general unfamiliarity with the method, make
+  me hesitant to promote it too much in the paper. But it is a nifty
+  trick, and including it in the package was a good idea (thanks!).
+
+In terms of accuracy, the paper includes comparisons between the true
+Hessian and the two \class{sparseHessianFD} methods (finite differences
+and complex step). I am not comparing accuracy with \pkg{numDeriv},
+because that really is not the point of this paper.  However, for the
+timing comparisons in Table 4, I did
+change the baseline \pkg{numDeriv} method from \func{hessian} with the
+\method{Richardson} method, to \func{jacobian} with the
+\method{complex} method. Removing the second-order approximation saves
+a lot of time, and lets me compute Table 4 without resorting to
+parallel computation. That will simplify attempts at replication, and
+removes another variable that might influence results.
 
@@
 \begin{response}
-
+  I added a \func{set.seed} statement near the top of \filename{replication.R}.
 \end{response}
@@
 \begin{verbatim}
-run.par <- FALSE
-
-(Note this is using obj$gr() rather than binary.grad(). I am not sure if
-that makes a difference.)
-
-run_test_tab4b <- function(Nk, reps=50) {
-    ## Replication function like Table 4  with jacobian of grad and
-"simple"
-    N <- as.numeric(Nk["N"])
-    k <- as.numeric(Nk["k"])
-    data <- binary_sim(N, k, T=20)
-    priors <- priors_sim(k)
-    F <- make_funcs(D=data, priors=priors)
-    nvars <- N*k+k
-    M <-
-as(Matrix::kronecker(Matrix::Diagonal(N),Matrix(1,k,k)),"nMatrix") %>%
-      rBind(Matrix(TRUE,k,N*k)) %>%
-      cBind(Matrix(TRUE, k*(N+1), k)) %>%
-      as("nMatrix")
-    pat <- Matrix.to.Coord(tril(M))
-    X <- rnorm(nvars)
-    obj <- sparseHessianFD(X, F$fn, F$gr, pat$rows, pat$cols)
-
-    bench <- microbenchmark(
-        numDeriv = numDeriv::jacobian(obj$gr, X, method="simple"), 
-        df = obj$gr(X),
-        sparse = obj$hessian(X))
-    vals <- plyr::ddply(data.frame(bench), "expr",
-                  function(x)
-return(data.frame(expr=x$expr,
-
-time=x$time,
-
-rep=1:length(x$expr))))
-    res <- data.frame(N=N, k=k,
-                      bench=vals)
-    cat("Completed N = ",N,"\tk = " , k ,"\n")
-    return(res)
-}
-
-cases_tab4b <- expand.grid(k=c(2,3,4),
-                     N=c(9, 12, 15))
-runs_tab4b <- plyr::adply(cases_tab4b, 1, run_test_tab4b, reps=20,
-.parallel=run.par)
-
-tab4b <-  mutate(runs_tab4b, ms=bench.time/1000000) %>%
-  select(-bench.time) %>%
-  spread(bench.expr, ms) %>%
-  gather(method, hessian, c(numDeriv, sparse)) %>%
-  mutate(M=N*k+k, hessian.df=hessian/df) %>%
-  gather(stat, time, c(hessian, hessian.df)) %>%
-  group_by(N, k, method, M, stat)  %>%
-  summarize(mean=mean(time), sd=sd(time)) %>%
-  gather(stat2, value, mean:sd) %>%
-  dcast(N+k+M~stat+method+stat2,value.var="value") %>%
-  arrange(M)
-
-tab4b 
-   N k  M hessian_numDeriv_mean hessian_numDeriv_sd hessian_sparse_mean
-1  9 2 20              5.731754          
-0.6759421            1.713736
-2 12 2 26              7.577777          
-0.9481714            1.756165
-3  9 3 30              8.763990          
-1.0358481            2.348153
-4 15 2 32              9.553452          
-1.1875816            1.844192
-5 12 3 39             11.583360          
-1.3371329            2.402663
-6  9 4 40             11.584188          
-1.2045539            2.978028
-7 15 3 48             14.638763          
-1.5469615            2.547145
-8 12 4 52             15.612216          
-2.5401431            3.086010
-9 15 4 64             19.672496          
-2.7303771            3.055931
-  hessian_sparse_sd hessian.df_numDeriv_mean hessian.df_numDeriv_sd
-1        0.07532715                
-19.02359               3.353629
-2        0.07522725                
-24.31841               4.715941
-3        0.15790507                
-28.47628               4.366788
-4        0.30441636                
-30.52029               4.533416
-5        0.08570303                
-35.98930               6.033750
-6        0.55389180                
-36.30998               6.007112
-7        0.61666543                
-44.46145               6.206399
-8        0.69277761                
-48.73135               9.180011
-9        0.12617835                
-60.27624               9.530503
-  hessian.df_sparse_mean hessian.df_sparse_sd
-1               5.687322            0.7812959
-2               5.619639            0.7445722
-3               7.615163            0.7901088
-4               5.884283            1.0339672
-5               7.455115            0.8976474
-6               9.329878            2.0531265
-7               7.734390            2.0592415
-8               9.637077            2.4535258
-9               9.356261            0.6168655
-
+Code and output removed
 \end{verbatim}
@@
   \begin{response}
-
+    I agree.  See my response to Item \ref{it:complex2}.  I include the
+    times from the 
+    \method{complex} method instead of \method{simple} because of the
+    improved accuracy.  I considered including both times in the
+    table, but that would have cluttered the table, there was very
+    little difference in the times between them, and the comparison
+    between two \pkg{numDeriv} methods is not the point of this
+    paper. But I completely agree that either \func{jacobian} method
+    is a more valid comparison than using \func{hessian}.
   \end{response}
@@
 
-tab4b5
-tab4b5
-      N k     M hessian_numDeriv_mean hessian_numDeriv_sd
-hessian_sparse_mean
-1    10 2    22          6.363689e+00       
-6.585771e-01            1.795344
-2    10 4    44          1.273715e+01       
-1.254662e+00            2.934588
-3    10 8    88          2.608751e+01       
-1.803375e+00            5.659668
-4   100 2   202          1.021658e+02       
-3.439251e+00            2.938642
-5   100 4   404          2.162431e+02       
-1.289940e+01            5.622877
-6   100 8   808          4.846450e+02       
-1.946854e+01           12.922397
-7  1000 2  2002          5.511146e+03       
-1.918534e+02           14.694924
-8  1000 4  4004          1.194815e+04       
-1.499518e+02           29.332463
-9  2500 2  5002          3.132878e+04       
-4.934896e+02           33.393927
-10 1000 8  8008          2.835954e+04       
-6.959400e+02           76.881070
-11 2500 4 10004          6.874300e+04       
-2.197321e+02           68.318820
-12 2500 8 20008          8.615465e+05       
-9.316236e+04         1806.766254
-   hessian_sparse_sd hessian.df_numDeriv_mean hessian.df_numDeriv_sd
-1       6.056685e-01                
-21.19760               2.433792
-2       5.508396e-01                
-41.16514               5.648203
-3       1.265859e+00                
-79.97503               8.748075
-4       7.409767e-02               
-193.27447              11.790089
-5       1.210575e+00               
-388.74262              31.184602
-6       1.168307e+01               
-758.39623             121.225037
-7       9.293258e-01              
-2042.30264             168.538055
-8       1.343366e+00              
-4099.05785             221.910206
-9       1.377541e+00              
-5081.52982             473.962083
-10      2.408524e+01              
-8323.91140             516.709041
-11      2.646135e+00             
-10315.16422             832.305746
-12      2.272944e+03             
-77200.22397           42790.360882
-   hessian.df_sparse_mean hessian.df_sparse_sd
-1                5.984750            2.1093324
-2                9.479244            1.9844323
-3               17.337516            4.1347369
-4                5.556561            0.2851630
-5               10.115250            2.3189536
-6               20.160163           17.8116929
-7                5.448765            0.5700265
-8               10.061368            0.6629470
-9                5.416287            0.5421972
-10              22.563990            7.1995778
-11              10.249990            0.8971753
-12             151.538778          217.2031623
+Output removed
 
@@
   \begin{response}
+These are all good suggestions.  Let me address them point by point.
+    \begin{itemize}
+    \item It really is all about the sparsity pattern, and
+      specifically, the how few partitions are needed to group the
+      variables.  Of course, it will be hard to keep the number of
+      partitions small if there are too many non-zero elements, so the
+      sparsity itself has some effect.  The arrangement of the non-zeros is
+      related to the ordering of the variables, but the permutation
+      handles that during the initialization of the
+      \class{sparseHessianFD} argument.  For example, variables
+      arranged in a way that the Hessian has a block-arrow pattern are
+      permuted to get a Hessian with a banded pattern instead.  This
+      is all discussed in the ``Partitioning the
+      variables'' section of the paper.
+\item I have no idea how to do this in a generalizable way.  I think
+  it has a lot to do with the architecture of the computer
+  itself. Figure 4 does break down the components of the computation
+  into the time it takes to reserve memory (initialization), partition
+  the variables, and compute gradients themselves.
+
+I do not want to ignore this suggestion, but I am simply not clear on
+what is being asked.  I am open to more specific suggestions.
+
+\item The complex step method is now supported, and comparisons of
+  accuracy and computation time are now included in the paper.
+
+\item Parallel computing on a shared memory architecture really
+  muddles up timing comparisons.  The \emph{only} reason I generated
+  Table 4 in parallel was because of the \pkg{numDeriv} \func{hessian}
+  function. Now that \func{hessian} was replaced with \func{jacobian},
+  the baseline measures from \pkg{numDeriv} run fast enough that I can
+  generate Table 4 in serial.
+
+Your point is well taken that timing comparisons can be affected by
+whether computation is being run on multiple processors, but the
+``serialization'' of Table 4 makes the issue moot.
+
+    \end{itemize}
 

### 2016-12-22 nobuild/jss_revision/sparseHessianFD.Rnw
more work on revision and response

@@
 
-\subsection[An example]{An example}
+\subsection[An example]{An example}\label{sec:exHess}
 
@@
 
-<<echo=FALSE>>=
-options(scipen=999)
-@
-
 Note that variables \variable{f} and \variable{true.f} were computed
@@
 
-were computed by the same functions
+\subsection[Complex step method]{Complex step method}
+
+The \pkg{sparseHessianFD} package also allows for
+estimation of the Hessian using the complex step method
+\citep{SquireTrapp1998} as an alternative to finite differencing. In
+this method, the definition of $\Mat{Y}_c$ in Equation \ref{eq:Yg} is replaced with
+\begin{align}
+  \label{eq:complexYg}
+  \Mat{Y}_c&=\Imag(\nabla f(x+iG_c))
+\end{align}
+where $i=\sqrt{-1}$ and $\Imag$ returns the imaginary part of a complex
+number. The advantage of this method is that there is no longer a
+subtraction operation to generate truncation error. Thus, $\delta$ can
+be made arbitrarily small, and the resulting estimate of the Hessian
+is more accurate.
+
+To use the complex step method, include the argument
+\funcarg{complex=TRUE} in the call to the initializer of the \class{sparseHessianFD}
+object. The functions passed as \funcarg{fn} and \funcarg{gr} must both
+accept a complex argument, and return a complex
+result, even though we are differentiating a real-valued function. Although \pkg{base} \proglang{R} supports complex arguments
+for basic arithmetic, trigonometric, logarithmic and exponential
+functions, many other common functions (e.g.,  \func{gamma}, \func{log1p},
+\func{expm1}, and the probability distribution functions) do not have
+complex implementations. Furthermore, the complex step method is valid
+only if the function is holomorphic
+(complex differentiable). The methods in \class{sparseHessianFD} do
+\emph{not} check that this is the case for the function at hand. We convey the
+following warning from the documentation of the \pkg{numDeriv} package
+\citep{R_numDeriv}, which also implements the complex step method:
+``Avoid this method if you do not know that your function is
+suitable. Your mistake may not be caught and the results will be
+spurious.''
+
+The log posterior density in Equation \ref{eq:LPD} is holomorphic, so we
+can estimate its Hessian using the complex step method, and compute
+the mean relative difference from the true Hessian.
+
+<<>>=
+obj2 <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
+  rows=pattern[["rows"]], cols=pattern[["cols"]],
+  complex=TRUE,
+  data=binary, priors=priors, order.row=order.row)
+hs2 <- obj2$hessian(P)
+mean(abs(hs2-true.hess))/mean(abs(hs2))
+@
+
+<<echo=FALSE>>=
+options(scipen=999)
+@
+
+In short, the complex step method can be much more accurate than
+finite differencing, but it comes with theoretical and implementation
+restrictions that may limit its universality.
 
@@
 
-\subsection[Complex step method]{Complex step method}
-
-The \pkg{sparseHessianFD} package also allows for
-estimation of the Hessian using the complex step method
-\citep{SquireTrapp1998} as an alternative to finite differencing. In
-this method, the definition of $\Mat{Y}_c$ in Equation \ref{eq:Yg} is replaced with
-\begin{align}
-  \label{eq:complexYg}
-  \Mat{Y}_c&=\Imag(\nabla f(x+iG_c))
-\end{align}
-where $i=\sqrt{-1}$ and $\Imag$ returns the imaginary part of a complex
-number. The advantage of this method is that there is no longer a
-subtraction operation to generate truncation error. Thus, $\delta$ can
-be made arbitrarily small, and the resulting estimate of the Hessian
-is more accurate.
-
-To use the complex step method, include the argument
-\funcarg{complex=TRUE} in the call to the initializer of the \class{sparseHessianFD}
-object. The functions passed as \funcarg{fn} and \funcarg{gr} must both
-accept a complex argument, and return a complex
-result, even though we are differentiating a real-valued function. Although \pkg{base} \proglang{R} supports complex arguments
-for basic arithmetic, trigonometric, logarithmic and exponential
-functions, many other common functions (e.g.,  \func{gamma}, \func{log1p},
-\func{expm1}, and the probability distribution functions) do not have
-complex implementations. Furthermore, the complex step method is valid
-only if the function is holomorphic
-(complex differentiable). The methods in \class{sparseHessianFD} do
-\emph{not} check that this is the case for the function at hand. We convey the
-following warning from the documentation of the \pkg{numDeriv} package
-\citep{R_numDeriv}, which also implements the complex step method:
-``Avoid this method if you do not know that your function is
-suitable. Your mistake may not be caught and the results will be
-spurious.''
-
-In short, the complex step method is available to users who want
-improved accuracy and speed, and know what they are doing.
-
 \subsection[Software libraries]{Software libraries}

### 2016-12-22 nobuild/jss_revision/sparseHessianFD.tex
more work on revision and response

@@
 \begin{Soutput}
-[1] 1.1217e-08
+[1] 1.12e-08
 \end{Soutput}
@@
 
-were computed by the same functions
+\subsection[Complex step method]{Complex step method}
+
+The \pkg{sparseHessianFD} package also allows for
+estimation of the Hessian using the complex step method
+\citep{SquireTrapp1998} as an alternative to finite differencing. In
+this method, the definition of $\Mat{Y}_c$ in Equation \ref{eq:Yg} is replaced with
+\begin{align}
+  \label{eq:complexYg}
+  \Mat{Y}_c&=\Imag(\nabla f(x+iG_c))
+\end{align}
+where $i=\sqrt{-1}$ and $\Imag$ returns the imaginary part of a complex
+number. The advantage of this method is that there is no longer a
+subtraction operation to generate truncation error. Thus, $\delta$ can
+be made arbitrarily small, and the resulting estimate of the Hessian
+is more accurate.
+
+To use the complex step method, include the argument
+\funcarg{complex=TRUE} in the call to the initializer of the \class{sparseHessianFD}
+object. The functions passed as \funcarg{fn} and \funcarg{gr} must both
+accept a complex argument, and return a complex
+result, even though we are differentiating a real-valued function. Although \pkg{base} \proglang{R} supports complex arguments
+for basic arithmetic, trigonometric, logarithmic and exponential
+functions, many other common functions (e.g.,  \func{gamma}, \func{log1p},
+\func{expm1}, and the probability distribution functions) do not have
+complex implementations. Furthermore, the complex step method is valid
+only if the function is holomorphic
+(complex differentiable). The methods in \class{sparseHessianFD} do
+\emph{not} check that this is the case for the function at hand. We convey the
+following warning from the documentation of the \pkg{numDeriv} package
+\citep{R_numDeriv}, which also implements the complex step method:
+``Avoid this method if you do not know that your function is
+suitable. Your mistake may not be caught and the results will be
+spurious.''
+
+The log posterior density in Equation \ref{eq:LPD} is holomorphic, so we
+can estimate its Hessian using the complex step method, and compute
+the mean relative difference from the true Hessian.
+
+\begin{Schunk}
+\begin{Sinput}
+R> obj2 <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
++   rows=pattern[["rows"]], cols=pattern[["cols"]],
++   complex=TRUE,
++   data=binary, priors=priors, order.row=order.row)
+R> hs2 <- obj2$hessian(P)
+R> mean(abs(hs2-true.hess))/mean(abs(hs2))
+\end{Sinput}
+\begin{Soutput}
+[1] 6.75e-18
+\end{Soutput}
+\end{Schunk}
+
+In short, the complex step method can be much more accurate than
+finite differencing, but it comes with theoretical and implementation
+restrictions that may limit its universality.
 

### 2016-12-22 nobuild/jss_revision/replication.R

@@ -9,9 +9,11 @@ library("numDeriv")
 library("reshape2")
 library("ggplot2")
 
+set.seed(1234)
+
 
 run.par <- TRUE
-if (run.par) registerDoParallel(cores=12) else registerDoParallel(cores=1)
+if (run.par) registerDoParallel(cores=2) else registerDoParallel(cores=1)
 
 binary_sim <- function(N, k, T) {
   x.mean <- rep(0,k)
@@ -34,12 +36,12 @@ priors_sim <- function(k) {
 }
 
 
-make_funcs <- function(D, priors) {
+make_funcs <- function(D, priors, order.row=FALSE) {
   res <- vector("list", length=3)
   names(res) <- c("fn", "gr", "hessian")
-  res$fn <-  function(pars) binary.f(pars, data=D, priors=priors, order.row=TRUE)
-  res$gr <-  function(pars) binary.grad(pars, data=D, priors=priors, order.row=TRUE)
-  res$hessian <-  function(pars) binary.hess(pars, data=D, priors=priors, order.row=TRUE)
+  res$fn <-  function(pars) binary.f(pars, data=D, priors=priors, order.row=order.row)
+  res$gr <-  function(pars) binary.grad(pars, data=D, priors=priors, order.row=order.row)
+  res$hessian <-  function(pars) binary.hess(pars, data=D, priors=priors, order.row=order.row)
   return(res)
 }
 
@@ -55,25 +57,30 @@ make_perm <- function(iRow, jCol) {
 
 get_id <- function(x) 1:length(x)
 
-run_test_fig4 <- function(NkT, reps=50) {
+run_test_fig4 <- function(NkT, reps=50, order.row=FALSE) {
   ## Replication function for Figure 4
   N <- as.numeric(NkT["N"])
   k <- as.numeric(NkT["k"])
   T <- as.numeric(NkT["T"])
   data <- binary_sim(N, k, T)
   priors <- priors_sim(k)
-  F <- make_funcs(D=data, priors=priors)
+  F <- make_funcs(D=data, priors=priors, order.row=order.row)
   nvars <- N*k+k
 
-  M <- as(Matrix::kronecker(Matrix(1,k,k), Matrix::Diagonal(N)),"nMatrix") %>%
-    rBind(Matrix(TRUE,k,N*k)) %>%
-    cBind(Matrix(TRUE, k*(N+1), k)) %>%
[…]

### 2016-12-22 nobuild/jss_revision/response.tex
more on revision

@@
 \usepackage{placeins}
+\usepackage{enumitem}
 
@@
 \usepackage{geometry}
-\geometry{margin=1.2in}
+\geometry{margin=1.1in}
+
+\setlength{\parskip}{1ex}
+\setlength{\parindent}{0ex}
 
@@
  \vspace{\baselineskip}
- Response to reviewers
+ Response to reviewer
 \end{center}
@@
 I would like to express my sincere appreciation for the time and effort that
-the reviewer invested in the review.  It is obvious to me that the
-reviewer went through the paper and code in great detail.  In
-particular, I want to thank the reviewer for bringing the complex step
+you invested in the review.  It is obvious to me that you went through
+the paper and code in great detail.  In
+particular, thank you for bringing the complex step
 method to my attention.  I \emph{never} use complex numbers or
 functions in my own work, so I typically ignore anything with the word
-"complex" in it.  But this review prompted me to learn something new.
-That is always fun.
+``complex'' in it.  But this review prompted me to learn something
+new, and that is always fun.
 
 I am happy to write that I was able to incorporate a large majority of
-the reviewer's suggestions in the revised paper and package. For the
-others, I hope that the editor and reviewer will accept my
+your suggestions into the the revised paper and package. For the
+others, I hope that you, and the editor, will accept my
 explanations and excuses.  Of course, if there are any strong
-objections to how I responded to the reviewer's concerns, I am happy
+objections to how I responded to any of the comments, I am happy
 to make further changes in advance of publication.
+
+Below are my responses to each item, in the same order as in the
+review.  The original comment is in italics, and is followed by my
+response in normal font.  To save space, I removed the code and output
+that were included in the review.
 
@@
 
-\begin{enumerate}
+\begin{enumerate}[align=left]
  \setcounter{enumi}{-1}
@@
 I suspect that the error is a result of the \pkg{algorithm} package
-not being present in the reviewer's LaTeX installation.  I would like
-to continue to use this package,
-because it does allow for nicely formatted algorithms.  I believe that this is
-a relatively common package, and one that should be easily installed
-through CTAN.  However, if there is a strong
-objection, I may be able to find another way to format the algorithms.
+not being present in your LaTeX installation.  It is a standard
+package in the MacTeX distribution, but I cannot speak with authority about other
+platforms. It should be easily installed through CTAN.
+
+I would prefer to continue to use the \pkg{algorithm} package, but I
+could try to find another way to format the algorithms if there is a
+strong objection.
 \end{response}
@@
   \begin{response}
-
 I included this section primarily as a way to avoid overselling the
 method, but the point that the reader might be scared away by this list of
-restrictions is well-taken.  In the revision, I removed the list of
-conditions, and wove them into the text.  The result takes a more
+restrictions is well-taken.  In the revision, I removed the list, and
+wove the restrictions into the text.  The revision projects a more
 positive tone that in the original submission.
-
 \end{response}
@@
 \begin{response}
-
-  The list of conditions is no longer in the paper, but to this point,
-  I added a defintion of a ``structural zero'' in the third
+  The list of conditions is no longer in the paper, but to this point, the
+  condition is global and not local.  I added a defintion of a ``structural zero'' in the third
   paragraph. The sparsity pattern represents the structure of the
   Hessian, not its value, so any element that \emph{could} be zero,
-  but does not have to be zero, is not a structural zero.  Thus, the
-  condition is global and not local.
+  but does not have to be zero, is not a structural zero.
 
-  Note that other than
-  a potential loss of efficiency, there is no disadvantage to erring
-  on the side of an element being non-zero.  
+   At the end of Section 2.2, I added a paragraph that explains that
+   if there is uncertainty about whether an element is a structural
+   zero or not, it is better to include it in the sparsity pattern.
 \end{response}
@@
 "coefficients vector", for example:
-
-  - 'continuous covariates $x_i$,' -> 'continuous covariates $x_i \in
+\begin{itemize}
+  \item 'continuous covariates $x_i$,' -> 'continuous covariates $x_i \in
 \Real{k}$,'
 
-  - 'heterogeneous coefficient vector $\beta_i$' -> 'heterogeneous
+  \item  'heterogeneous coefficient vector $\beta_i$' -> 'heterogeneous
 coefficient vector $\beta_i \in \Real{k}$'
 
-  - 'The coefficients are distributed' -> 'The coefficient vectors
+  \item 'The coefficients are distributed' -> 'The coefficient vectors
 $\beta_i$ are distributed'
-
+\end{itemize}
   \end{revQuote}
@@
 \begin{response}
-  Thank you for the suggestion.  Changes have been made throughout the paper.
+  Thank you for the suggestions.  Changes have been made throughout the paper.
 \end{response}
@@
   Based on this comment, I rewrote Section 2.2 to distinguish the
-  cross-partial derivatives $\parD{2}{}$ from the corresponding
+  cross-partial derivatives $\parD{2}{}{}$ from the corresponding
   elements of the Hessian $\hess{}{}$.  I did this by defining $x_i$ as
   a subset of the elements of $x$, and Ind$(x_i)$ as an 
-  ``indexing'' function.  The revised version is slightly more formal
+  ``indexing'' function.  This change required me to redefine the
+  covariate matrix in the example from $\Mat{X}$ to $\Mat{Z}$.
+
+  The revised version is slightly more formal
   and precise, and I hope that this is the kind of change you had in
-  mind.  That said, I do not think it makes the paper easier to read,
-  and, with your permission, I would prefer to revert to the simpler
-  language in the original manuscript.
+  mind.
 \end{response}
@@
 \item \begin{revQuote}
- I think possibly (16) is "banded" and (17) "block arrow" rather than the
+ I think possibly (16) is ``banded'' and (17) ``block arrow'' rather than the
 reverse which is indicated.
@@
 \begin{response}
-No, the descriptions of the two sparsity patterns are correct. The
+The descriptions of the two sparsity patterns are correct. The
 first index in the subscript of $\beta$ refers to a household, and the
 second index refers to a covariate. Thus,
-$\parD{2}{\beta_{11},\beta_{12}}\neq 0$, but
-$\parD{2}{\beta_{11},\beta_{21}}=0$.  When the second index changes
-before the first, the pattern is ``block-arrow.''  When the first
-index changes before the second, the pattern is ``banded.''
+$\parD{2}{\beta_{11},\beta_{12}}{}\neq 0$, but
+$\parD{2}{\beta_{11},\beta_{21}}{}=0$.  When the second index changes
+faster than the first, the pattern is ``block-arrow.''  When the first
+index changes faster than the second, the pattern is ``banded.''
 
@@
 \begin{response}
-
+  Note to me:  I think what this means is that results for Table 4 and
+  Figure 4 are "hard coded.''  For the jss article, the replication
+  file is provided, but this file is not part of the package.  It
+  should be, so non-jss vignette readers know how to replicate the results. 
 \end{response}
@@
   I removed the \func{R>} prompt, and now refer to that code as a usage
-  syntax.  
+  syntax.  Thank you for the suggestion.
 \end{response}
@@
 
-\begin{verbatim}
-Code removed
-\end{verbatim}
+(Code removed)
 
@@
 gradient function:
-\begin{verbatim}
-max(abs(  obj$hessian(P)
-     - binary.hess(P, data=binary, priors=priors, order.row=order.row)))
-[1] 2.786891e-06
-\end{verbatim}
+
+(Code removed)
 
@@
   Good point.  I revised that section to explain that evaluations of
-  the function and gradient have to be identical to the true values,
-  and I even replaced \func{all.equal} with \func{identical} to
-  highlight that fact. I then explained that the
+  the function and gradient, as called from the
+  \class{sparseHessianFD} object, have to be identical to the true values.
+  I replaced \func{all.equal} with \func{identical} to
+  highlight that fact. I then explain that the
   \class{sparseHessianFD} calculation of the Hessian will not be
@@
   difference.  I chose the mean relative difference over maximum
-  absolute difference because the former is what the \func{all.equal}
-  function uses to test if two matrices are equal within numeric tolerance.
+  absolute difference to be consistent with what \func{all.equal}
+  uses to test if two matrices are equal within numeric tolerance.
 \end{response}
@@
 approximation of a second difference:
-\begin{verbatim}
-Hessian Richardson
 
-\end{verbatim}
+(Code removed)
 
@@
 
-\begin{verbatim}
-Jacobian Richardson
-\end{verbatim}
+(Code removed)
 
@@
 
-\begin{verbatim}
-Jacobian simple
-
-\end{verbatim}
+(Code removed)
 
@@
 
-\begin{verbatim}
-Jacobian complex
-\end{verbatim}
+(Code removed)
 
@@
   \func{log1p}, \func{gamma}, and the probability distribution
-  functions. This limitation on the kinds of functions that can be
-  implemented easily in R, combined with the requirement that the function be
-  holomorphic, and my own general unfamiliarity with the method, make
-  me hesitant to promote it too much in the paper. But it is a nifty
-  trick, and including it in the package was a good idea (thanks!).
+  functions. This limitation, combined with the requirement that the function be
+  holomorphic, and my own inexperience with the complex step method, make
+  me hesitant to promote it too heavily in the paper. But it is a nifty
+  trick, and I am happy that it is now included in the package.
 
-In terms of accuracy, the paper includes comparisons between the true
+In terms of accuracy, the paper now includes comparisons between the true
 Hessian and the two \class{sparseHessianFD} methods (finite differences
@@
 removes another variable that might influence results.
+
+I did collect timings using the \pkg{numDeriv} \method{simple} method,
+but did not include those in Table 4 to save space.  There was not an
+appreciable difference in computation time between the \method{simple}
+and \method{complex} methods.
+
+Footnote 1 was expanded to acknowledge the complex step method.
 
@@
 \begin{response}
-
+  This information is now included in the first paragraph of Section 4.
 \end{response}
@@
 
-\begin{verbatim}
-Code and output removed
-\end{verbatim}
-
+(Code and output removed)
   \end{revQuote}
@@
     I agree.  See my response to Item \ref{it:complex2}.  I include the
-    times from the 
+    times from the \pkg{numDeriv}
     \method{complex} method instead of \method{simple} because of the
-    improved accuracy.  I considered including both times in the
-    table, but that would have cluttered the table, there was very
+    improved accuracy.  I considered including results from both methods,
+    but that would have cluttered the table, there was very
     little difference in the times between them, and the comparison
@@
 N=2500, k=8 comparison.)
-\begin{verbatim}
-cases_tab4b5 <- expand.grid(k=c(2,4,8),
-                     N=c(10, 100, 1000, 2500))
-runs_tab4b5 <- plyr::adply(cases_tab4b5, 1, run_test_tab4b, reps=20,
-.parallel=run.par)
 
-tab4b5 <-  mutate(runs_tab4b5, ms=bench.time/1000000) %>%
-  select(-bench.time) %>%
-  spread(bench.expr, ms) %>%
-  gather(method, hessian, c(numDeriv, sparse)) %>%
-  mutate(M=N*k+k, hessian.df=hessian/df) %>%
-  gather(stat, time, c(hessian, hessian.df)) %>%
-  group_by(N, k, method, M, stat)  %>%
-  summarize(mean=mean(time), sd=sd(time)) %>%
-  gather(stat2, value, mean:sd) %>%
-  dcast(N+k+M~stat+method+stat2,value.var="value") %>%
-  arrange(M)
-
-Output removed
-
-I think the complex step takes a similar amount of time, but produces a more
-accurate result.
-
-\end{verbatim}
+(Code and output removed)
 
@@
   \begin{response}
-
+Because the \func{numDeriv::jacobian} function runs so much faster
+than  \func{numDeriv::hessian}, I was able to increase the size of the
+examples in Table 4. Please forgive me for not going as high as
+$N=2500$, $k=8$. One objective of Table 4 is to compare the relative
+increases in computation time for different $N$ and $k$ between
+\pkg{numDeriv} and \pkg{sparseHessianFD}, and I think that objective
+is met with $N=(15, 50, 100, 500)$.  For the reader who wants to
+replicate the results, a run time of more than a day is a lot to ask.
   \end{response}
@@
 deemed more "original research". Some possibilities are:
-
- - Explain and try to assess how much of the speedup is due to simple
+\begin{itemize}
+ \item Explain and try to assess how much of the speedup is due to simple
 sparseness and how much is due to the "sparse patern" (p2) allowing for
@@
 
- - Try to assess how much of the speedup is due to reduce computing demand
+ \item Try to assess how much of the speedup is due to reduce computing demand
 and how much is due to different memory demand. (I had the impression in the
@@
 
- - Consider implementing a complex step method, and do a comparison.
+ \item Consider implementing a complex step method, and do a comparison.
 
- - Assess the difference when multiple CPUs are used. (run.par == FALSE vs
+ \item  Assess the difference when multiple CPUs are used. (run.par == FALSE vs
 run.par == TRUE)
-
+\end{itemize}
   \end{revQuote}
@@
   \begin{response}
-These are all good suggestions.  Let me address them point by point.
+These are all reasonable suggestions.  Let me address them point by point.
     \begin{itemize}
@@
       permuted to get a Hessian with a banded pattern instead.  This
-      is all discussed in the ``Partitioning the
-      variables'' section of the paper.
+      is all discussed in Section 3.1.
+
 \item I have no idea how to do this in a generalizable way.  I think
   it has a lot to do with the architecture of the computer
-  itself. Figure 4 does break down the components of the computation
+  itself.  For example, my computer has 64 GB of RAM, so memory usage
+  is not a binding constraint, and swap space never comes into play.
+
+  Figure 4 does break down the components of the computation
   into the time it takes to reserve memory (initialization), partition
-  the variables, and compute gradients themselves.
+  the variables, and compute gradients themselves.  Perhaps that is
+  the kind of comparison you had in mind.
 
@@
 
-\item The complex step method is now supported, and comparisons of
-  accuracy and computation time are now included in the paper.
+\item The package now implements the complex step method, and comparisons of
+  accuracy and computation time are included in the paper.
 
@@
 
-Your point is well taken that timing comparisons can be affected by
-whether computation is being run on multiple processors, but the
+Your point that timing comparisons can be affected by
+whether computation is being run on multiple processors is well taken, but the
 ``serialization'' of Table 4 makes the issue moot.
+
+  Data for Figure 4 were generated in parallel because of the large
+  number of required simulations.
 

### 2016-12-22 nobuild/jss_revision/sparseHessianFD.Rnw
more on revision

@@
 
-\date{December 20, 2016}
+\date{December 22, 2016}
 
@@
 of the $M$ elements of $x$, one at a time.\footnote{More accurate
-  approximations require more gradient evaluations.  We will consider
-  only forward finite differences.}  In a sparse matrix, most of the
+  approximations using finite differences require more gradient evaluations.  We will consider
+  only forward finite differences. In Section \ref{sec:complex} we
+  introduce the complex step method that, under certain conditions,
+  can generate more accurate estimates with the same number of
+  gradient evaluations using forward finite differences.} In a sparse matrix, most of the
 elements are constrained to zero.  Depending
@@
 correlated with another element of $\beta_i$), and that, for all $i$,
-$\parD{2}{\beta_i, \mu}=0$ (because $\mu$ is the prior mean of each
+$\parD{2}{\beta_i, \mu}\neq 0$ (because $\mu$ is the prior mean of each
 $\beta_i$).  However, since the $\beta_i$ and $\beta_j$ are independently distributed, and the $y_i$ are
@@
 check visually that the matrix has the right pattern, and then extract
-the indices.
-
-The package defines utility functions (Table~\ref{tab:helpers}) to convert
+the indices.  The package defines utility functions (Table~\ref{tab:helpers}) to convert
 between sparse matrices, and the vectors of row and column indices
@@
 @
+
+If there is uncertainty about whether an element is a structural
+   zero or not, one should err on the side of it being non-zero, and
+   include that element in the sparsity pattern.
+   There might be a loss of efficiency if the element really is a
+   structural zero, but the result will still be correct.  All that
+   would happen is that the numerical estimate for that element would be zero
+   (within machine precision). On the other hand, excluding a non-zero
+   element from the sparsity pattern will likely lead to an incorrect
+   estimate of the Hessian.
 
@@
 Finally, we create an instance of a \class{sparseHessianFD} object.
-Evaluations of the function and gradient will generate the same
-results as the true values because they are computed using the same function.
+Evaluations of the function and gradient using the \method{fn} and
+\method{gr} methods will always give the same
+results as the true values because they are computed using the same functions.
 
@@
 
-Note that variables \variable{f} and \variable{true.f} were computed
-using the same exact calculation, as were variables
-\variable{gr} and \variable{true.grad}.In contrast,
-\variable{true.hess} was computed using an exact calculation, while
-\variable{hs} is the sparse finite difference approximation.
-
-\subsection[Complex step method]{Complex step method}
+\subsection[Complex step method]{Complex step method}\label{sec:complex}
 
@@
 As far as we know, \pkg{numDeriv} \citep{R_numDeriv} is the only other \proglang{R} package
-that computes numerical approximations to derivatives.  It differs from
-\pkg{sparseHessianFD} in some important ways.
-
-\begin{enumerate}
-\item It treats all Hessians as dense;
-  \item It computes each element of the Hessian using a second-order
-    finite differencing approximation that does not require the user
-    to supply the gradient; and
-    \item It implements iterative algorithms to improve accuracy, at
-      the expense of speed.
-\end{enumerate}
-
-Nevertheless, \pkg{numDeriv} is an easy-to-use tool for numerical differentiation,
-so it is worthwhile to compare its performance to that of
-\pkg{sparseHessianFD}.  To prepare Table~\ref{tab:numDeriv}, we
+that computes numerical approximations to derivatives.  Like
+\pkg{sparseHessianFD} it includes functions to compute Hessians from
+user-supplied gradients (through the \func{jacobian} function), and
+implements both the finite differencing and complex step methods.  Its
+most important distinction from \pkg{sparseHessianFD} is that it treats all Hessians as dense.  Thus, we
+will use  \pkg{numDeriv} as the baseline against which we can compare the performance of
+\pkg{sparseHessianFD}.  All of the subsequent speed and scalability
+results were generated on a 2013-vintage Apple Mac Pro with a 2.7 GHz
+12-core Intel Xeon E5 processor and 64 GB of RAM.
+
+To prepare Table~\ref{tab:numDeriv}, we
 estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
@@
 
-Because \pkg{numDeriv} does not scale, we cannot use it to benchmark \pkg{sparseHessianFD} for
-larger sparse Hessians.  To help us understand just how scalable
+To help us understand just how scalable
 \pkg{sparseHessianFD} is, we ran another set of simulations, for the

### 2016-12-22 nobuild/jss_revision/sparseHessianFD.tex
more on revision

@@
 \end{Schunk}
+
+If there is uncertainty about whether an element is a structural
+   zero or not, one should err on the side of it being non-zero, and
+   include that element in the sparsity pattern.
+   There might be a loss of efficiency if the element really is a
+   structural zero, but the result will still be correct.  All that
+   would happen is that the numerical estimate for that element would be zero
+   (within machine precision). On the other hand, excluding a non-zero
+   element from the sparsity pattern will likely lead to an incorrect
+   estimate of the Hessian.
 
@@
 \begin{Soutput}
-[1] 1.12e-08
+[1] 1.1217e-08
 \end{Soutput}
@@
 \begin{Soutput}
-[1] 6.75e-18
+[1] 6.7543e-18
 \end{Soutput}

### 2016-12-22 nobuild/jss_revision/response.tex
almost done with revision

@@
   \begin{response}
+Assuming that the R package builder works the way I think it does,
+the \filename{replication.R} file, and the data for Table 4 (\filename{vignette\_tab4.Rdata}), will be
+copied to the \filename{doc\textbackslash} directory of the package
+once it is installed.  It will also be available as an online supplement to
+this paper.
+\begin{itemize}
+\item 
+\end{itemize}
+
+Meeting the documentation and replication needs for both the journal,
+and CRAN, can be tricky, but I think I have a good solution
+
   Note to me:  I think what this means is that results for Table 4 and
@@
 Hessian and the two \class{sparseHessianFD} methods (finite differences
-and complex step). I am not comparing accuracy with \pkg{numDeriv},
-because that really is not the point of this paper.  However, for the
+and complex step). To maintain the focus of the paper on
+\pkg{sparseHessianFD}, I am not including an assessment of the accuracy of
+\pkg{numDeriv}.  However, for the
 timing comparisons in Table 4, I did
 change the baseline \pkg{numDeriv} method from \func{hessian} with the
-\method{Richardson} method, to \func{jacobian} with the
-\method{complex} method. Removing the second-order approximation saves
+\method{Richardson} method, to \func{jacobian} with both the
+\method{simple} and 
+\method{complex} methods. Removing the second-order approximation saves
 a lot of time, and lets me compute Table 4 without resorting to
@@
 removes another variable that might influence results.
-
-I did collect timings using the \pkg{numDeriv} \method{simple} method,
-but did not include those in Table 4 to save space.  There was not an
-appreciable difference in computation time between the \method{simple}
-and \method{complex} methods.
 
@@
   \begin{response}
-    I agree.  See my response to Item \ref{it:complex2}.  I include the
-    times from the \pkg{numDeriv}
-    \method{complex} method instead of \method{simple} because of the
-    improved accuracy.  I considered including results from both methods,
-    but that would have cluttered the table, there was very
-    little difference in the times between them, and the comparison
-    between two \pkg{numDeriv} methods is not the point of this
-    paper. But I completely agree that either \func{jacobian} method
-    is a more valid comparison than using \func{hessian}.
+    I agree.  See my response to Item \ref{it:complex2}.  Table 4
+    includes times from the \pkg{numDeriv}
+    \method{complex} and \method{simple} methods.
+
+    To reduce clutter in Table 4, I removed comparisons of the
+    hessian-to-gradient computation time ratio.  That was not a
+    particularly interesting metric.
+
   \end{response}

### 2016-12-22 nobuild/jss_revision/sparseHessianFD.tex
almost done with revision

@@
 especially because the computation time for \pkg{numDeriv} grows
-quadratically in the number of variables.  The setup time for
-\pkg{sparseHessianFD} was about 7 milliseconds for all cases.  Times
-were collected using the \pkg{microbenchmark} package \citep{R_microbenchmark}.
+quadratically in the number of variables. Times
+were collected using the \pkg{microbenchmark} package
+\citep{R_microbenchmark}. Code to replicate Table~\ref{tab:numDeriv}
+is in the file \filename{replication.R}, which is available as an
+online supplement to this paper.
+
+\begin{Schunk}
+\begin{Sinput}
+R> library(xtable)
+R> library(dplyr)
+R> load("vignette_tab4.Rdata")
+\end{Sinput}
+\end{Schunk}
 
 \begin{table}[ptb]\centering
-  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }cccp{2pt}rrrrp{2pt}rrrr}
+  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }rrr|rrrr|rrrr}
   \toprule
-&&&&\multicolumn{4}{c}{\bfseries{Hessian
-    time}}&&\multicolumn{4}{c}{\bfseries{Hessian/Gradient ratio}}\\
-&&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
-$N$&$k$&$M$&&mean&sd&mean&sd&&mean&sd&mean&sd\\
-  \cmidrule{1-3}\cmidrule{5-8}\cmidrule{10-13}
-   9 &    2 &   20 &&  363 & 6.4 &  2.8 & 2.2 &&  871 & 63 & 6.5 & 4.2 \\
-    12 &    2 &   26 &&  629 &  8.9 &  2.9 & 2.1 && 1469 &  124 & 6.5 & 3.3 \\
-     9 &    3 &   30 &&  819 & 27.7 &  3.6 & 2.2 && 1974 &  161 & 8.4 & 3.7 \\
-    15 &    2 &   32 &&  973 & 25.8 &    3.0 & 2.2 && 2191 &  209 &  6.5 & 3.4 \\
-    12 &    3 &   39 && 1407 & 39.3 & 3.5 & 2.0 && 3291 &  285 &    8.0 & 3.5 \\
-     9 &    4 &   40 && 1445 & 28.8 & 4.3 & 2.1 && 3453 &  292 & 9.9 & 3.1 \\
-    15 &    3 &   48 && 2138 & 92.7 & 3.5 & 2.0 && 4861 &  341 & 7.8 & 3.4 \\
-    12 &    4 &   52 && 2400 & 98.6 & 4.3 & 2.1 && 5679 &  503 & 10.1 & 3.7 \\
-    15 &    4 &   64 && 3643 &  209.0 & 4.2 & 2.2 && 8945 &  828 & 10.0 &  2.8 \\
+&&&\multicolumn{4}{c|}{\bfseries{Finite differencing}}&\multicolumn{4}{c}{\bfseries{Complex step}}\\
+&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c|}{\pkg{sparseHessianFD}}&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
+$N$&$k$&$M$&mean&sd&mean&sd&mean&sd&mean&sd\\
+%%\midrule
+  \midrule
+15 & 2 & 32 & 12.4 & 1.2 & 2.3 & 0.5 & 12.8 & 1.2 & 2.1 & 0.5 \\ 
+  15 & 4 & 64 & 26.2 & 12.4 & 4.0 & 0.9 & 26.1 & 2.0 & 3.8 & 1.1 \\ 
+  50 & 2 & 102 & 50.5 & 2.4 & 3.0 & 0.9 & 54.8 & 3.5 & 2.9 & 2.1 \\ 
+  15 & 8 & 128 & 50.4 & 2.2 & 8.5 & 12.0 & 55.0 & 2.7 & 7.7 & 1.5 \\ 
+  100 & 2 & 202 & 134.1 & 15.7 & 3.7 & 0.5 & 144.0 & 11.6 & 3.3 & 0.6 \\ 
+  50 & 4 & 204 & 104.0 & 2.1 & 6.3 & 11.6 & 117.2 & 17.6 & 5.1 & 1.0 \\ 
+  100 & 4 & 404 & 284.3 & 34.3 & 8.0 & 10.9 & 322.5 & 40.8 & 8.2 & 11.1 \\ 
+  50 & 8 & 408 & 248.5 & 48.7 & 10.9 & 1.4 & 271.7 & 43.7 & 11.5 & 1.4 \\ 
+  100 & 8 & 808 & 624.5 & 54.0 & 14.4 & 1.8 & 762.6 & 76.9 & 15.7 & 1.7 \\ 
+  500 & 2 & 1002 & 1758.6 & 86.7 & 10.1 & 10.0 & 1990.0 & 90.6 & 9.3 & 9.8 \\ 
+  500 & 4 & 2004 & 3832.5 & 84.9 & 18.4 & 2.3 & 4460.7 & 82.8 & 19.4 & 2.5 \\ 
+  500 & 8 & 4008 & 8846.5 & 82.1 & 45.2 & 6.2 & 11058.4 & 101.3 & 52.5 & 6.3 \\ 
    \bottomrule
+
 \end{tabularx}
 \caption{Computation times (milliseconds) for computing Hessians using the
-  \pkg{numDeriv} and \pkg{sparseHessianFD} packages, across 200 replications. Rows are ordered by the
+  \pkg{numDeriv} and \pkg{sparseHessianFD} packages, and the finite
+ difference and complex step methods, across 200 replications. Rows are ordered by the
   number of variables.}\label{tab:numDeriv}
@@
 Initialization&total setup time (including the partitioning time)\\
-Hessian/Gradient&ratio of the Hessian time to the
-    gradient time\\
     \bottomrule
@@
 In the plots in Figure~\ref{fig:timing}, the number of heterogeneous units ($N$) is
-on the x-axis, and mean run time, in milliseconds, is on the y-axis.  Each panel shows the
+on the x-axis, and median run time, in milliseconds, is on the y-axis.  Each panel shows the
 relationship between $N$ and run time for a different measure of time,
@@
 to compute a Hessian should not increase with $N$ \emph{at all}.  The
-reason it does is that each gradient evaluation takes longer.  The
-plot of the ratio of Hessian-to-gradient run times is relatively flat as the size
-of the dataset increases.  We can conclude that the
+reason it does is that each gradient evaluation takes longer.
+Nevertheless, we can conclude that the
 \pkg{sparseHessianFD} algorithms are quite efficient and scalable for

### 2016-12-22 nobuild/jss_revision/replication.R

@@ -11,7 +11,7 @@ library("ggplot2")
 
 set.seed(1234)
 
-registerDoParallel(cores=10)
+registerDoParallel(cores=11)
 
 binary_sim <- function(N, k, T) {
   x.mean <- rep(0,k)
@@ -160,42 +160,45 @@ run_test_tab4 <- function(Nk, reps=50, order.row=TRUE) {
 
 ## Replicate Figure 4
 
-cases_fig4 <- expand.grid(k=c(8, 6, 4, 2),
-                          N=c(25, 50, 75, seq(100,2000, by=100)),
+cases_fig4 <- expand.grid(k=c(8, 5, 2),
+                          N=c(25, 50, 75, seq(100,2500, by=200)),
                           T=20
 )
-reps_fig4 <- 100
+reps_fig4 <- 200
 
 runs_fig4 <- plyr::adply(cases_fig4, 1, run_test_fig4, reps=reps_fig4,
                          order.row=TRUE, .parallel=TRUE)
 
 tab_fig4 <- mutate(runs_fig4, ms=bench.time/1000000) %>%
   dcast(N+k+T+bench.rep+ncolors~bench.expr, value.var="ms")  %>%
-  mutate(nvars=N*k+k, hess_df=hess/df) %>%
-  gather(stat, ms, c(f,df,hess,colors,setup,hess_df)) %>%
+  mutate(nvars=N*k+k) %>%
+  gather(stat, ms, c(f,df,hess,colors,setup)) %>%
   group_by(N, k, T, stat, nvars) %>%
-  summarize(mean=mean(ms))
-
-
+  summarize(median=median(ms))
 D2 <- filter(data.frame(tab_fig4), stat %in% c("f", "df", "hess",
-                                               "colors", "setup","hess_df"))
+                                               "colors", "setup"))
 D2$stat <- plyr::revalue(D2$stat, c("f"="Function", "df"="Gradient", "hess"="Hessian",
                               "colors"="Partitioning",
-                              "setup"="Initialization",
-                              "hess_df"="Hessian/Gradient"))
-
+                              "setup"="Initialization"))
 D2$stat <- factor(D2$stat, levels=c("Function","Gradient","Hessian",
-                                    "Partitioning","Initialization",
-                                    "Hessian/Gradient"))
-
[…]

### 2016-12-23 nobuild/jss_revision/sparseHessianFD.Rnw
still working on jss revision

@@
 The \pkg{sparseHessianFD} package is a tool for estimating
-sparse Hessians using finite differences of gradients.  Section \ref{sec:numdiff} will
+sparse Hessians using finite differences of gradients.  Section~\ref{sec:numdiff} will
 cover the specifics, but the basic idea is as follows.  Consider a
@@
   approximations using finite differences require more gradient evaluations.  We will consider
-  only forward finite differences. In Section \ref{sec:complex} we
+  only forward finite differences. In Section~\ref{sec:complex} we
   introduce the complex step method that, under certain conditions,
@@
 sparse matrices in \proglang{R}, in Section~\ref{sec:background}.  In
-Section ~\ref{sec:using}, we explain how to use the package.
+Section~\ref{sec:using}, we explain how to use the package.
 Section~\ref{sec:algorithms} explains the underlying algorithms, and
@@
 extracting nonzero Hessian elements via substitution, are described in
-Section \ref{sec:algorithms}.
+Section~\ref{sec:algorithms}.
 
@@
 
-From the log posterior density in Equation \ref{eq:LPD}, we can see
+From the log posterior density in Equation~\ref{eq:LPD}, we can see
 that $\parD{2}{\beta_i, \beta_i}\neq 0$ (one element of $\beta_i$ could be
@@
 delta& The perturbation amount for finite differencing of the
-  gradient to compute the Hessian (the $\delta$ in Section \ref{sec:numdiff}).  Defaults to
+  gradient to compute the Hessian (the $\delta$ in Section~\ref{sec:numdiff}).  Defaults to
   \code{sqrt(.Machine\$double.eps)}.\\
@@
 Now we can estimate the Hessian for the
-log posterior density of the model from Section \ref{sec:example}.
+log posterior density of the model from Section~\ref{sec:example}.
 For demonstration purposes, \pkg{sparseHessianFD} includes functions that compute the value
@@
 \citep{SquireTrapp1998} as an alternative to finite differencing. In
-this method, the definition of $\Mat{Y}_c$ in Equation \ref{eq:Yg} is replaced with
+this method, the definition of $\Mat{Y}_c$ in Equation~\ref{eq:Yg} is replaced with
 \begin{align}
@@
 
-To use the complex step method, include the argument
-\funcarg{complex=TRUE} in the call to the initializer of the \class{sparseHessianFD}
-object. The functions passed as \funcarg{fn} and \funcarg{gr} must both
+A call to the \class{sparseHessianFD} \method{hessian} method will
+apply the complex step method if the 
+\funcarg{complex=TRUE} was set in the initializer. The functions
+passed as \funcarg{fn} and \funcarg{gr} must both
 accept a complex argument, and return a complex
 result, even though we are differentiating a real-valued function. Although \pkg{base} \proglang{R} supports complex arguments
-for basic arithmetic, trigonometric, logarithmic and exponential
-functions, many other common functions (e.g.,  \func{gamma}, \func{log1p},
+for most basic mathematical functions, many common functions (e.g.,  \func{gamma}, \func{log1p},
 \func{expm1}, and the probability distribution functions) do not have
@@
 
-The log posterior density in Equation \ref{eq:LPD} is holomorphic, so we
+Fortunately for demonstration purposes, the log posterior density in Equation~\ref{eq:LPD} is holomorphic, so we
 can estimate its Hessian using the complex step method, and compute
@@
 \pkg{sparseHessianFD}.  All of the subsequent speed and scalability
-results were generated on a 2013-vintage Apple Mac Pro with a 2.7 GHz
-12-core Intel Xeon E5 processor and 64 GB of RAM.
+results were generated on an Apple Mac Pro with a 2.7 GHz
+12-core Intel Xeon E5 processor and 64 GB of RAM, running MacOS
+version 10.12.2, and \proglang{R} version 3.3.2.
 
@@
 especially because the computation time for \pkg{numDeriv} grows
-quadratically in the number of variables.  The setup time for
-\pkg{sparseHessianFD} was about 7 milliseconds for all cases.  Times
-were collected using the \pkg{microbenchmark} package \citep{R_microbenchmark}.
+quadratically in the number of variables. Times
+were collected using the \pkg{microbenchmark} package
+\citep{R_microbenchmark}. Code to replicate Table~\ref{tab:numDeriv}
+is available as an online supplement to this paper, and in the
+\func{doc/} directory of the installed package.
+
+<<echo=FALSE>>=
+suppressMessages(library(xtable))
+suppressMessages(library(dplyr))
+load("vignette_tab4.Rdata")
+@
 
 \begin{table}[ptb]\centering
-  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }cccp{2pt}rrrrp{2pt}rrrr}
+  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }rrr|rrrr|rrrr}
   \toprule
-&&&&\multicolumn{4}{c}{\bfseries{Hessian
-    time}}&&\multicolumn{4}{c}{\bfseries{Hessian/Gradient ratio}}\\
-&&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
-$N$&$k$&$M$&&mean&sd&mean&sd&&mean&sd&mean&sd\\
-  \cmidrule{1-3}\cmidrule{5-8}\cmidrule{10-13}
-   9 &    2 &   20 &&  363 & 6.4 &  2.8 & 2.2 &&  871 & 63 & 6.5 & 4.2 \\
-    12 &    2 &   26 &&  629 &  8.9 &  2.9 & 2.1 && 1469 &  124 & 6.5 & 3.3 \\
-     9 &    3 &   30 &&  819 & 27.7 &  3.6 & 2.2 && 1974 &  161 & 8.4 & 3.7 \\
-    15 &    2 &   32 &&  973 & 25.8 &    3.0 & 2.2 && 2191 &  209 &  6.5 & 3.4 \\
-    12 &    3 &   39 && 1407 & 39.3 & 3.5 & 2.0 && 3291 &  285 &    8.0 & 3.5 \\
-     9 &    4 &   40 && 1445 & 28.8 & 4.3 & 2.1 && 3453 &  292 & 9.9 & 3.1 \\
-    15 &    3 &   48 && 2138 & 92.7 & 3.5 & 2.0 && 4861 &  341 & 7.8 & 3.4 \\
-    12 &    4 &   52 && 2400 & 98.6 & 4.3 & 2.1 && 5679 &  503 & 10.1 & 3.7 \\
-    15 &    4 &   64 && 3643 &  209.0 & 4.2 & 2.2 && 8945 &  828 & 10.0 &  2.8 \\
-   \bottomrule
+&&&\multicolumn{4}{c|}{\bfseries{Finite differencing}}&\multicolumn{4}{c}{\bfseries{Complex step}}\\
+&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c|}{\pkg{sparseHessianFD}}&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
+$N$&$k$&$M$&mean&sd&mean&sd&mean&sd&mean&sd\\
+%%\midrule
+<<results='asis',echo=FALSE>>=
+arrange(tab4, M) %>%
+xtable(digits=c(rep(0,4),rep(1,8))) %>%
+    print(only.contents=TRUE, comment=FALSE, booktabs=TRUE,
+      include.rownames=FALSE, include.colnames=FALSE)
+@
 \end{tabularx}
 \caption{Computation times (milliseconds) for computing Hessians using the
-  \pkg{numDeriv} and \pkg{sparseHessianFD} packages, across 200 replications. Rows are ordered by the
+  \pkg{numDeriv} and \pkg{sparseHessianFD} packages, and the finite
+ difference and complex step methods, across 200 replications. Rows are ordered by the
   number of variables.}\label{tab:numDeriv}
@@
 for the Hessian grows linearly as well, and that might be partially
-surprising.  We saw in Section \ref{sec:coloring} that adding
+surprising.  We saw in Section~\ref{sec:coloring} that adding
 additional heterogeneous units in a hierarchical model does not

### 2016-12-23 nobuild/jss_revision/sparseHessianFD.tex
still working on jss revision

@@
 \citep{R_microbenchmark}. Code to replicate Table~\ref{tab:numDeriv}
-is in the file \filename{replication.R}, which is available as an
-online supplement to this paper.
-
-\begin{Schunk}
-\begin{Sinput}
-R> library(xtable)
-R> library(dplyr)
-R> load("vignette_tab4.Rdata")
-\end{Sinput}
-\end{Schunk}
+is available as an online supplement to this paper, and in the
+\func{doc/} directory of the installed package.
 

### 2016-12-24 nobuild/jss_revision/response.tex
just about done with revision

@@
 \newcommand{\proglang}[1]{\textsf{#1}}
-\newcommand{\filename}[1]{\textit{#1}}
+\newcommand{\filename}[1]{\texttt{#1}}
 \newcommand{\code}[1]{\texttt{#1}}
@@
 explanations and excuses.  Of course, if there are any strong
-objections to how I responded to any of the comments, I am happy
-to make further changes in advance of publication.
+objections to any of my responses, I am happy
+to consider further changes in advance of publication.
 
@@
 
-{\normalfont (error messages related to algorithm.sty removed).}
+(error messages related to algorithm.sty removed).
 
@@
 \begin{response}
-I suspect that the error is a result of the \pkg{algorithm} package
+Because \pkg{sparseHessianFD} passes CRAN check, I suspect that the
+error is a result of the \pkg{algorithm} package
 not being present in your LaTeX installation.  It is a standard
@@
 \begin{response}
-
-  Note to me:  issue here is that there is no documentation for
-  \func{sparseHessianFD()} as a function.  Good idea to create a
-  documentation page for that.
-
+The help page for \pkg{sparseHessianFD} now includes explicit
+instructions for the initializer.  You are correct that there is
+little guidance in how to document reference classes, and I agree that
+there should be a way for the
+existing mechanism to include documentation for the initializer to
+an RC.  But it's hard to do it using Roxygen2 in way that also passes
+CRAN checks. So I just ``hard coded'' a description of the initializer
+into the details section of the sparseHessianFD-class documentation. I
+could not find a more elegant way of doing it.  
 \end{response}
@@
 copied to the \filename{doc\textbackslash} directory of the package
-once it is installed.  It will also be available as an online supplement to
-this paper.
-\begin{itemize}
-\item 
-\end{itemize}
-
-Meeting the documentation and replication needs for both the journal,
-and CRAN, can be tricky, but I think I have a good solution
-
-  Note to me:  I think what this means is that results for Table 4 and
-  Figure 4 are "hard coded.''  For the jss article, the replication
-  file is provided, but this file is not part of the package.  It
-  should be, so non-jss vignette readers know how to replicate the results. 
+once it is installed.  At the end of the second paragraph of Section
+4, I now write that ``code to replicate Table 4 is available as an
+online supplement to this paper.''  I cannot be more specific about
+the name of the replication file, because I believe it depends on the
+volume and issue number of the published version.
 \end{response}
@@
 \begin{response}
-  I added a \func{set.seed} statement near the top of \filename{replication.R}.
+  I added a \func{set.seed} statement near the top of
+  \filename{replication.R}, but you are correct that because only
+  times are reported, this does not matter much.
 \end{response}
@@
 than  \func{numDeriv::hessian}, I was able to increase the size of the
-examples in Table 4. Please forgive me for not going as high as
-$N=2500$, $k=8$. One objective of Table 4 is to compare the relative
+examples in Table 4. One objective of Table 4 is to compare the relative
 increases in computation time for different $N$ and $k$ between
 \pkg{numDeriv} and \pkg{sparseHessianFD}, and I think that objective
-is met with $N=(15, 50, 100, 500)$.  For the reader who wants to
+is met with $N=(15, 50, 100, 500)$.  Please forgive me for not going as high as
+$N=2500$, $k=8$. For the reader who wants to
 replicate the results, a run time of more than a day is a lot to ask.
@@
 
-  I do not want to ignore this suggestion, but I am simply not clear on
-  what is being asked.  I am open to more specific suggestions.
+  I do not want to ignore or dismiss this suggestion, but I am simply not clear on
+  what is being asked.  I am open to more specific recommendations.
 
 \item The package now implements the complex step method, and comparisons of
-  accuracy and computation time are included in the paper.
+  accuracy and computation time are included in the paper.  Thank you
+  for this suggestion.
 
@@
   muddles up timing comparisons.  The \emph{only} reason I generated
-  Table 4 in parallel was because of the \pkg{numDeriv} \func{hessian}
+  Table 4 in parallel was because of the run time of the \pkg{numDeriv} \func{hessian}
   function. Now that \func{hessian} was replaced with \func{jacobian},
@@
   number of required simulations.
-
     \end{itemize}

### 2016-12-24 R/sparseHessianFD-class.R

@@ -2,33 +2,70 @@
 ## Copyright (C) 2015-2017 Michael Braun
 ##
 ##' @title sparseHessianFD
-##' @name sparseHessianFD-class
+##' @name sparseHessianFD
 ##' @description A reference class for computing sparse Hessians
 ##' @docType class
 ##' @field fn1 A closure for calling fn(x, ...).
 ##' @field gr1 A closure for calling gr(x, ...).
 ##' @field iRow,jCol Numeric vectors with row and column indices of
-##the non-zero elements in the lower triangle (including diagonal) of
-##the Hessian.
+##' the non-zero elements in the lower triangle (including diagonal) of
+##' the Hessian.
 ##' @field delta The perturbation amount for finite differencing of
-##the gradient to compute the Hessian. Defaults to sqrt(.Machine$double.eps).
+##' the gradient to compute the Hessian. Defaults to 1e-07.
 ##' @field index1 TRUE if rows and cols use 1-based (R format)
-##indexing (FALSE for 0-based (C format) indexing.
-##' @field complex TRUE if Hessian will be computed using the complex step method, and FALSE (default) if using finite differences.
+##' indexing (FALSE for 0-based (C format) indexing.
+##' @field complex TRUE if Hessian will be computed using the complex
+##' step method, and FALSE (default) if using finite differences.
 ##' @field D raw finite differences (internal use only)
 ##' @field nvars Number of variables (length of x)
 ##' @field nnz Number of non-zero elements in the lower triangle of the Hessian.
 ##' @field ready TRUE if object has been initialized, and Hessian has
-##been partitioned.
+##' been partitioned.
 ##' @field idx,pntr Column indices and row pointers for non-zero
-##elements in lower triangle of the permuted Hessian.  Row-oriented
-##compressed storage.
+##' elements in lower triangle of the permuted Hessian.  Row-oriented
+##' compressed storage.
 ##' @field colors A vector representation of the partitioning of the columns.
 ##' There are nvars elements, one for each column of the permuted
-##Hessian.  The value corresponds to the "color" for that column.
-##'@field perm,invperm Permutation vector and its inverse
-##' @details Do not access any of the fields directly.  Use the initializer
-##'instead. The internal structure is subject to change in future versions.
+##' Hessian.  The value corresponds to the "color" for that column.
+##' @field perm,invperm Permutation vector and its inverse
+##' @details The sparseHessianFD function calls the initializer for the
+##' sparseHessianFD class, and returns a sparseHessianFD object.
[…]

### 2016-12-24 nobuild/jss_revision/sparseHessianFD.Rnw
just about done with revision

@@
 \Keywords{sparse Hessians, sparsity, computation of Hessians,
-  graph coloring, finite differences, differentiation}
+  graph coloring, finite differences, differentiation, complex step}
 \Address{
@@
 %%\newenvironment{knitrout}{\begin{figure}}{\end{figure}}
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
-
-<<setup1, echo=FALSE, cache=FALSE>>=
-suppressPackageStartupMessages(require(Matrix))
-knitr::render_sweave()
-knitr::opts_chunk$set(prompt=TRUE, cache=FALSE,error=FALSE) #$
-options(replace.assign=TRUE, width=77, prompt="R> ")
-@
 
@@
 of the $M$ elements of $x$, one at a time.\footnote{More accurate
-  approximations using finite differences require more gradient evaluations.  We will consider
+  approximations using finite differences require more gradient
+  evaluations.  In this section, we will consider
   only forward finite differences. In Section~\ref{sec:complex} we
@@
 
-<<setup2, echo=FALSE>>=
-N <- 5
-k <- 2
-nv1 <- (N + 1) * k
-nels1 <- nv1^2
-nnz1 <- (N + 1) * k^2 + 2 * N * k^2
-nnz1LT <- (N + 1) * k * (k+1)/2 + N * k^2
-Q <- 1000
-nv2 <- (Q+1) * k
-nels2 <- nv2^2
-nnz2 <- (Q+1) * k^2 + 2 * Q * k^2
-nnz2LT <- (Q+1) * k * (k+1)/2 + Q * k^2
-options(scipen=999)
-@
-
 Let $x_1$ and $x_2$ be two subsets of elements of $x$.
@@
   \begin{subfigure}[b]{.5\textwidth}
-<<blockarrow, echo=FALSE>>=
-Mat <- as(kronecker(diag(N), matrix(1, k, k)),"sparseMatrix")
-Mat <- rBind(Mat, Matrix(1, k, N*k))
-Mat <- cBind(Mat, Matrix(1, k * (N+1), k))
-printSpMatrix(as(Mat,"nMatrix"))
-@
+\begin{Schunk}
+\begin{Soutput}
+
+ [1,] | | . . . . . . . . | |
+ [2,] | | . . . . . . . . | |
+ [3,] . . | | . . . . . . | |
+ [4,] . . | | . . . . . . | |
+ [5,] . . . . | | . . . . | |
+ [6,] . . . . | | . . . . | |
+ [7,] . . . . . . | | . . | |
+ [8,] . . . . . . | | . . | |
+ [9,] . . . . . . . . | | | |
+[10,] . . . . . . . . | | | |
+[11,] | | | | | | | | | | | |
+[12,] | | | | | | | | | | | |
+\end{Soutput}
+\end{Schunk}
 \caption{A ``block-arrow'' sparsity pattern.}\label{fig:blockarrow}
@@
 \begin{subfigure}[b]{.5\textwidth}
-<<banded, echo=FALSE>>=
-Mat <- kronecker(Matrix(1, k, k), diag(N))
-Mat <- rBind(Mat, Matrix(1, k, N * k))
-Mat <- cBind(Mat, Matrix(1, k * (N + 1), k))
-printSpMatrix(as(Mat,"nMatrix"))
-@
+\begin{Schunk}
+\begin{Soutput}
+
+ [1,] | . . . . | . . . . | |
+ [2,] . | . . . . | . . . | |
+ [3,] . . | . . . . | . . | |
+ [4,] . . . | . . . . | . | |
+ [5,] . . . . | . . . . | | |
+ [6,] | . . . . | . . . . | |
+ [7,] . | . . . . | . . . | |
+ [8,] . . | . . . . | . . | |
+ [9,] . . . | . . . . | . | |
+[10,] . . . . | . . . . | | |
+[11,] | | | | | | | | | | | |
+[12,] | | | | | | | | | | | |
+\end{Soutput}
+\end{Schunk}
 \caption{A ``banded'' sparsity pattern.}\label{fig:banded}
@@
 In both cases, the number of non-zeros is the same.   There are
-\Sexpr{nels1} elements in this symmetric matrix, but only \Sexpr{nnz1} are
-non-zero, and only \Sexpr{nnz1LT} values are unique.  Although the reduction in
+144 elements in this symmetric matrix, but only 64 are
+non-zero, and only 38 values are unique.  Although the reduction in
 RAM from using a sparse matrix structure for the Hessian may be
-modest, consider what would happen if $N=\Sexpr{scales::comma(Q)}$ instead.  In that case,
-there are \Sexpr{scales::comma(nv2)} variables in the problem, and more than $\Sexpr{scales::comma(floor(nels2/10^6))}$ million
-elements in the Hessian.  However, only $\Sexpr{scales::comma(nnz2)}$ of those elements are
+modest, consider what would happen if $N=1,000$ instead.  In that case,
+there are 2,002 variables in the problem, and more than $4$ million
+elements in the Hessian.  However, only $12,004$ of those elements are
 non-zero.  If we work with only the lower triangle of the Hessian, then we need to work with
-only \Sexpr{scales::comma(nnz2LT)} values.
+only 7,003 values.
 
@@
 \begin{minipage}{\columnwidth}
-<<>>=
-library("sparseHessianFD")
-bd <- kronecker(diag(3), matrix(TRUE, 2, 2))
-Mat <- as(bd, "nMatrix")
-printSpMatrix(tril(Mat))
-mc <- Matrix.to.Coord(tril(Mat))
-mc
-@
+\begin{Schunk}
+\begin{Sinput}
+R> library("sparseHessianFD")
+R> bd <- kronecker(diag(3), matrix(TRUE, 2, 2))
+R> Mat <- as(bd, "nMatrix")
+R> printSpMatrix(tril(Mat))
+\end{Sinput}
+\begin{Soutput}
+
+[1,] | . . . . .
+[2,] | | . . . .
+[3,] . . | . . .
+[4,] . . | | . .
+[5,] . . . . | .
+[6,] . . . . | |
+\end{Soutput}
+\begin{Sinput}
+R> mc <- Matrix.to.Coord(tril(Mat))
+R> mc
+\end{Sinput}
+\begin{Soutput}
+$rows
+[1] 1 2 2 3 4 4 5 6 6
+
+$cols
+[1] 1 1 2 3 3 4 5 5 6
+\end{Soutput}
+\end{Schunk}
 \end{minipage}
@@
   the non-zero elements in the \emph{lower triangle} of the Hessian.\\
-direct& This argument is deprecated, and is included only for
-  backwards compatibility with earlier versions.\\
 delta& The perturbation amount for finite differencing of the
@@
 arguments, the usage syntax to create a \class{sparseHessianFD} object is
-<<eval=FALSE, prompt=FALSE>>=
+\begin{Schunk}
+\begin{Sinput}
 obj <- sparseHessianFD(x, fn, gr, rows, cols, ...)
-@
+\end{Sinput}
+\end{Schunk}
 where \funcarg{...} represents all other arguments that are
@@
 
-<<eval=FALSE>>=
-f <- obj$fn(x)
-df <- obj$gr(x)
-hess <- obj$hessian(x)
-fngr <- obj$fngr(x)
-fngrhs <- obj$fngrhs(x)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> f <- obj$fn(x)
+R> df <- obj$gr(x)
+R> hess <- obj$hessian(x)
+R> fngr <- obj$fngr(x)
+R> fngrhs <- obj$fngrhs(x)
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<binaryInit>>=
-set.seed(123)
-data("binary")
-str(binary)
-N <- length(binary[["Y"]])
-k <- NROW(binary[["X"]])
-T <- binary[["T"]]
-nvars <- as.integer(N * k + k)
-priors <- list(inv.Sigma = rWishart(1, k + 5, diag(k))[,,1],
-  inv.Omega = diag(k))
-@
-
-This dataset represents the simulated choices for $N= \Sexpr{N}$ customers
-over $T= \Sexpr{T}$ purchase opportunties, where the probability of purchase
-is influenced by $k= \Sexpr{k}$ covariates.
+\begin{Schunk}
+\begin{Sinput}
+R> set.seed(123)
+R> data("binary")
+R> str(binary)
+\end{Sinput}
+\begin{Soutput}
+List of 3
+ $ Y: int [1:50] 13 1 18 18 19 6 16 6 5 8 ...
+ $ X: num [1:4, 1:50] -0.07926 -0.23018 1.55871 0.00997 0.01828 ...
+ $ T: num 20
+\end{Soutput}
+\begin{Sinput}
+R> N <- length(binary[["Y"]])
+R> k <- NROW(binary[["X"]])
+R> T <- binary[["T"]]
+R> nvars <- as.integer(N * k + k)
+R> priors <- list(inv.Sigma = rWishart(1, k + 5, diag(k))[,,1],
++   inv.Omega = diag(k))
+\end{Sinput}
+\end{Schunk}
+
+This dataset represents the simulated choices for $N= 50$ customers
+over $T= 20$ purchase opportunties, where the probability of purchase
+is influenced by $k= 4$ covariates.
 
@@
 
-<<usingSparseHessianFD1>>=
-obj <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
-  rows=pattern[["rows"]], cols=pattern[["cols"]],
-  data=binary, priors=priors, order.row=order.row)
-f <- obj$fn(P)
-identical(f, true.f)
-gr <- obj$gr(P)
-identical(gr, true.grad)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> obj <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
++   rows=pattern[["rows"]], cols=pattern[["cols"]],
++   data=binary, priors=priors, order.row=order.row)
+R> f <- obj$fn(P)
+R> identical(f, true.f)
+\end{Sinput}
+\begin{Soutput}
+[1] TRUE
+\end{Soutput}
+\begin{Sinput}
+R> gr <- obj$gr(P)
+R> identical(gr, true.grad)
+\end{Sinput}
+\begin{Soutput}
+[1] TRUE
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<echo=FALSE>>=
-options(scipen=0)
-@
-
-<<usingSparseHessianFD2>>=
-hs <- obj$hessian(P)
-mean(abs(hs-true.hess))/mean(abs(hs))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> hs <- obj$hessian(P)
+R> mean(abs(hs-true.hess))/mean(abs(hs))
+\end{Sinput}
+\begin{Soutput}
+[1] 2.3357e-09
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<>>=
-obj2 <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
-  rows=pattern[["rows"]], cols=pattern[["cols"]],
-  complex=TRUE,
-  data=binary, priors=priors, order.row=order.row)
-hs2 <- obj2$hessian(P)
-mean(abs(hs2-true.hess))/mean(abs(hs2))
-@
-
-<<echo=FALSE>>=
-options(scipen=999)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> obj2 <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
++   rows=pattern[["rows"]], cols=pattern[["cols"]],
++   complex=TRUE,
++   data=binary, priors=priors, order.row=order.row)
+R> hs2 <- obj2$hessian(P)
+R> mean(abs(hs2-true.hess))/mean(abs(hs2))
+\end{Sinput}
+\begin{Soutput}
+[1] 8.0555e-17
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<echo=FALSE>>=
-suppressMessages(library(xtable))
-suppressMessages(library(dplyr))
-load("vignette_tab4.Rdata")
-@
-
 \begin{table}[ptb]\centering
@@
 %%\midrule
-<<results='asis',echo=FALSE>>=
-arrange(tab4, M) %>%
-xtable(digits=c(rep(0,4),rep(1,8))) %>%
-    print(only.contents=TRUE, comment=FALSE, booktabs=TRUE,
-      include.rownames=FALSE, include.colnames=FALSE)
-@
+  \midrule
+15 & 2 & 32 & 12.4 & 1.2 & 2.3 & 0.5 & 12.8 & 1.2 & 2.1 & 0.5 \\ 
+  15 & 4 & 64 & 26.2 & 12.4 & 4.0 & 0.9 & 26.1 & 2.0 & 3.8 & 1.1 \\ 
+  50 & 2 & 102 & 50.5 & 2.4 & 3.0 & 0.9 & 54.8 & 3.5 & 2.9 & 2.1 \\ 
+  15 & 8 & 128 & 50.4 & 2.2 & 8.5 & 12.0 & 55.0 & 2.7 & 7.7 & 1.5 \\ 
+  100 & 2 & 202 & 134.1 & 15.7 & 3.7 & 0.5 & 144.0 & 11.6 & 3.3 & 0.6 \\ 
+  50 & 4 & 204 & 104.0 & 2.1 & 6.3 & 11.6 & 117.2 & 17.6 & 5.1 & 1.0 \\ 
+  100 & 4 & 404 & 284.3 & 34.3 & 8.0 & 10.9 & 322.5 & 40.8 & 8.2 & 11.1 \\ 
+  50 & 8 & 408 & 248.5 & 48.7 & 10.9 & 1.4 & 271.7 & 43.7 & 11.5 & 1.4 \\ 
+  100 & 8 & 808 & 624.5 & 54.0 & 14.4 & 1.8 & 762.6 & 76.9 & 15.7 & 1.7 \\ 
+  500 & 2 & 1002 & 1758.6 & 86.7 & 10.1 & 10.0 & 1990.0 & 90.6 & 9.3 & 9.8 \\ 
+  500 & 4 & 2004 & 3832.5 & 84.9 & 18.4 & 2.3 & 4460.7 & 82.8 & 19.4 & 2.5 \\ 
+  500 & 8 & 4008 & 8846.5 & 82.1 & 45.2 & 6.2 & 11058.4 & 101.3 & 52.5 & 6.3 \\ 
+   \bottomrule
+
 \end{tabularx}

### 2016-12-24 nobuild/jss_revision/sparseHessianFD.tex
just about done with revision

@@
 
-\usepackage{Sweave}
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 
@@
 %%\newenvironment{knitrout}{\begin{figure}}{\end{figure}}
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 \begin{Soutput}
-[1] 1.1217e-08
+[1] 2.3357e-09
 \end{Soutput}
@@
 \begin{Soutput}
-[1] 6.7543e-18
+[1] 8.0555e-17
 \end{Soutput}

### 2016-12-24 nobuild/jss_revision/response.tex
Revert "just about done with revision"
This reverts commit 7bf2fa8d9479891736b9f3fe6828763c350a30bd.

@@
 explanations and excuses.  Of course, if there are any strong
-objections to any of my responses, I am happy
-to consider further changes in advance of publication.
+objections to how I responded to any of the comments, I am happy
+to make further changes in advance of publication.
 
@@
 
-(error messages related to algorithm.sty removed).
+{\normalfont (error messages related to algorithm.sty removed).}
 
@@
 \begin{response}
-Because \pkg{sparseHessianFD} passes CRAN check, I suspect that the
+Because \pkg{sparseHessianFD} passed CRAN check, I suspect that the
 error is a result of the \pkg{algorithm} package
@@
 \begin{response}
-The help page for \pkg{sparseHessianFD} now includes explicit
-instructions for the initializer.  You are correct that there is
-little guidance in how to document reference classes, and I agree that
-there should be a way for the
-existing mechanism to include documentation for the initializer to
-an RC.  But it's hard to do it using Roxygen2 in way that also passes
-CRAN checks. So I just ``hard coded'' a description of the initializer
-into the details section of the sparseHessianFD-class documentation. I
-could not find a more elegant way of doing it.  
+
+  Note to me:  issue here is that there is no documentation for
+  \func{sparseHessianFD()} as a function.  Good idea to create a
+  documentation page for that.
+
 \end{response}

### 2016-12-24 nobuild/jss_revision/sparseHessianFD.Rnw
Revert "just about done with revision"
This reverts commit 7bf2fa8d9479891736b9f3fe6828763c350a30bd.

@@
   the non-zero elements in the \emph{lower triangle} of the Hessian.\\
+direct& This argument is deprecated, and is included only for
+  backwards compatibility with earlier versions.\\
 delta& The perturbation amount for finite differencing of the

### 2016-12-24 R/sparseHessianFD-class.R

@@ -2,70 +2,33 @@
 ## Copyright (C) 2015-2017 Michael Braun
 ##
 ##' @title sparseHessianFD
-##' @name sparseHessianFD
+##' @name sparseHessianFD-class
 ##' @description A reference class for computing sparse Hessians
 ##' @docType class
 ##' @field fn1 A closure for calling fn(x, ...).
 ##' @field gr1 A closure for calling gr(x, ...).
 ##' @field iRow,jCol Numeric vectors with row and column indices of
-##' the non-zero elements in the lower triangle (including diagonal) of
-##' the Hessian.
+##the non-zero elements in the lower triangle (including diagonal) of
+##the Hessian.
 ##' @field delta The perturbation amount for finite differencing of
-##' the gradient to compute the Hessian. Defaults to 1e-07.
+##the gradient to compute the Hessian. Defaults to sqrt(.Machine$double.eps).
 ##' @field index1 TRUE if rows and cols use 1-based (R format)
-##' indexing (FALSE for 0-based (C format) indexing.
-##' @field complex TRUE if Hessian will be computed using the complex
-##' step method, and FALSE (default) if using finite differences.
+##indexing (FALSE for 0-based (C format) indexing.
+##' @field complex TRUE if Hessian will be computed using the complex step method, and FALSE (default) if using finite differences.
 ##' @field D raw finite differences (internal use only)
 ##' @field nvars Number of variables (length of x)
 ##' @field nnz Number of non-zero elements in the lower triangle of the Hessian.
 ##' @field ready TRUE if object has been initialized, and Hessian has
-##' been partitioned.
+##been partitioned.
 ##' @field idx,pntr Column indices and row pointers for non-zero
-##' elements in lower triangle of the permuted Hessian.  Row-oriented
-##' compressed storage.
+##elements in lower triangle of the permuted Hessian.  Row-oriented
+##compressed storage.
 ##' @field colors A vector representation of the partitioning of the columns.
 ##' There are nvars elements, one for each column of the permuted
-##' Hessian.  The value corresponds to the "color" for that column.
-##' @field perm,invperm Permutation vector and its inverse
-##' @details The sparseHessianFD function calls the initializer for the
-##' sparseHessianFD class, and returns a sparseHessianFD object.
-##' \preformatted{sparseHessianFD(x, fn, gr, rows, cols, delta, index1, complex, ...)}
-##' The function, gradient and sparsity pattern are
[…]

### 2016-12-24 nobuild/jss_revision/sparseHessianFD.Rnw
fixed revision

@@
 gradient: one for the gradient at $x$, and one after perturbing each
-of the $M$ elements of $x$, one at a time.\footnote{More accurate
-  approximations using finite differences require more gradient evaluations.  We will consider
-  only forward finite differences. In Section~\ref{sec:complex} we
-  introduce the complex step method that, under certain conditions,
-  can generate more accurate estimates with the same number of
-  gradient evaluations using forward finite differences.} In a sparse matrix, most of the
+of the $M$ elements of $x$, one at a time. In a sparse matrix, most of the
 elements are constrained to zero.  Depending

### 2016-12-24 nobuild/jss_revision/replication.R

@@ -158,46 +158,46 @@ run_test_tab4 <- function(Nk, reps=50, order.row=TRUE) {
     return(res)
 }
 
-## Replicate Figure 4
+## ## Replicate Figure 4
 
-cases_fig4 <- expand.grid(k=c(8, 5, 2),
-                          N=c(25, 50, 75, seq(100,2500, by=200)),
-                          T=20
-)
-reps_fig4 <- 200
+## cases_fig4 <- expand.grid(k=c(8, 5, 2),
+##                           N=c(25, 50, 75, seq(100,2500, by=200)),
+##                           T=20
+## )
+## reps_fig4 <- 200
 
-runs_fig4 <- plyr::adply(cases_fig4, 1, run_test_fig4, reps=reps_fig4,
-                         order.row=TRUE, .parallel=TRUE)
+## runs_fig4 <- plyr::adply(cases_fig4, 1, run_test_fig4, reps=reps_fig4,
+##                          order.row=TRUE, .parallel=TRUE)
 
-tab_fig4 <- mutate(runs_fig4, ms=bench.time/1000000) %>%
-  dcast(N+k+T+bench.rep+ncolors~bench.expr, value.var="ms")  %>%
-  mutate(nvars=N*k+k) %>%
-  gather(stat, ms, c(f,df,hess,colors,setup)) %>%
-  group_by(N, k, T, stat, nvars) %>%
-  summarize(median=median(ms))
-D2 <- filter(data.frame(tab_fig4), stat %in% c("f", "df", "hess",
-                                               "colors", "setup"))
-D2$stat <- plyr::revalue(D2$stat, c("f"="Function", "df"="Gradient", "hess"="Hessian",
-                              "colors"="Partitioning",
-                              "setup"="Initialization"))
-D2$stat <- factor(D2$stat, levels=c("Function","Gradient","Hessian",
-                                    "Partitioning","Initialization"))
-theme_set(theme_bw())
-fig4 <- ggplot(D2, aes(x=N,y=median, color=as.factor(k), linetype=as.factor(k))) %>%
-    + geom_line(size=.4) %>%
-    + scale_x_continuous("Number of heterogeneous units") %>%
-    + scale_y_continuous("Computation time (milliseconds)") %>%
-    + guides(color=guide_legend("k"), linetype=guide_legend("k")) %>%
-    + facet_wrap(~stat, scales="free",ncol=1) %>%
[…]

### 2016-12-24 nobuild/jss_revision/sparseHessianFD.Rnw
refining the revision

@@
 will use  \pkg{numDeriv} as the baseline against which we can compare the performance of
-\pkg{sparseHessianFD}.  All of the subsequent speed and scalability
-results were generated on an Apple Mac Pro with a 2.7 GHz
-12-core Intel Xeon E5 processor and 64 GB of RAM, running MacOS
-version 10.12.2, and \proglang{R} version 3.3.2.
+\pkg{sparseHessianFD}.
 
@@
 estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
-units ($N$) and within-unit parameters ($k$).  As in the previous
-section, the total number of
+units ($N$) and within-unit parameters ($k$).  The total number of
 variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
-and standard deviations (across 20 replications) for the time (in
-milliseconds) to compute a Hessian using each package.  The difference in run times is dramatic,
-especially because the computation time for \pkg{numDeriv} grows
-quadratically in the number of variables. Times
-were collected using the \pkg{microbenchmark} package
+and standard deviations (across 500 replications) for the time (in
+milliseconds) to compute a Hessian using functions for both the finite
+difference and complex step methods from each package. Times were generated on a
+compute node running Scientific Linux 6 (64-bit) with an 8-core Intel
+Xeon X5560 processor (2.80 GHz) with 24 GB of RAM, and collected using the \pkg{microbenchmark} package
 \citep{R_microbenchmark}. Code to replicate Table~\ref{tab:numDeriv}
 is available as an online supplement to this paper, and in the
-\func{doc/} directory of the installed package.
+\func{doc/} directory of the installed package. In
+Table~\ref{tab:numDeriv} we see that computation times using
+\pkg{sparseHessianFD} and considerably shorter than those using \pkg{numDeriv}.
 
@@
 recorded the run times for different steps in the sparse Hessian
-estimation, across 200 replications.  The steps are summarized in Table~\ref{tab:timeMeasures}.
+estimation, across 200 replications.  The steps are summarized in
+Table~\ref{tab:timeMeasures}.  The times were generated on an
+Apple Mac Pro with a 12-core Intel Xeon E5-2697 processor (2.7 GHz)
+with 64 GB of RAM.
 
@@
 on the x-axis, and median run time, in milliseconds, is on the y-axis.  Each panel shows the
-relationship between $N$ and run time for a different measure of time,
+relationship between $N$ and run time for a different step in the algorithm,
 and each curve in a panel represents a different number of within-unit

### 2016-12-25 nobuild/jss_revision/sparseHessianFD.Rnw
proof of sparsity with complex step method

@@
 \usepackage{amsfonts}
+\usepackage{amsthm}
 \usepackage{subcaption}
@@
 \newcommand{\Real}[1]{\mathbb{R}^{#1}}
+\newcommand{\Complex}[1]{\mathbb{Z}^{#1}}
 \DeclareMathOperator{\Imag}{Im}
@@
 The \pkg{sparseHessianFD} package is a tool for estimating
-sparse Hessians using finite differences of gradients.  Section~\ref{sec:numdiff} will
-cover the specifics, but the basic idea is as follows.  Consider a
+sparse Hessians numerically, using either finite differences or
+complex perturbations of
+gradients.  Section~\ref{sec:numdiff} will
+cover the specifics, but the basic idea is as follows.  Consider a real-valued
 function $f(x)$, its gradient $\nabla{f(x)}$, and its Hessian
@@
 element, and let $\delta$ be a sufficiently small scalar
-constant. A linear approximation to the $m$th column of the Hessian is ${\hess{m}{}\approx\left(\nabla f(x+\delta
+constant. A ``finite difference'' linear approximation to the $m$th column of the Hessian is ${\hess{m}{}\approx\left(\nabla f(x+\delta
   e_m)- \nabla f(x)\right)/\delta}$.  Estimating a dense Hessian in this way involves at least $M+1$ calculations of the
 gradient: one for the gradient at $x$, and one after perturbing each
-of the $M$ elements of $x$, one at a time. In a sparse matrix, most of the
+of the $M$ elements of $x$, one at a time. If the function is complex
+analytic, an alternative approximation that requires $M$ gradient
+evaulations is $\hess{m}{}\approx\Imag\left(\nabla f(x+i\delta
+  e_m)\right)/\delta$, where $i=\sqrt{-1}$ and $\Imag$ returns the imaginary part of a complex
+number \citep{SquireTrapp1998}. This ``complex step'' method can be more accurate than the
+finite difference method because there is no subtraction operation
+that might generate roundoff errors, and thus $\delta$ can be made
+arbitrarily small to avoid truncaton errors \citep{LaiCrassidis2008}. However, it does require that the
+implementation of $f(x)$ accept a complex argument, and return a
+complex result, even though we are interested in differentiating $f(x)$ only along
+the real line.
+
+Regardless of the approximation method used, if the Hessian is sparse, most of the
 elements are constrained to zero.  Depending
@@
 requires the user to provide functions that compute an objective
-function, its gradient, and the sparsity pattern of its Hessian
-matrix. The gradient should be computed exactly (within machine
-precision), as opposed to being numerically estimated (e.g., by finite
-differencing). The sparsity pattern (e.g., location of structural
+function, its gradient (as accurately as possible, to machine precision), and the sparsity pattern of its Hessian
+matrix. The sparsity pattern (e.g., location of structural
 zeros) must be known in advance, and cannot vary across the domain of the
@@
 the space of applications for which \pkg{sparseHessianFD} is the right
-tool for the job. In general, finite differencing is not a ``first
-choice'' method because the result is not exact, so
+tool for the job. In general, numerical approximations are not ``first
+choice'' methods because the result is not exact, so
 \pkg{sparseHessianFD} should not be used when the application cannot
@@
 users might balk at having to provide an exact gradient, even though
-the Hessian will be estimated numerically. However, deriving a vector of first
+the Hessian will be estimated numerically.\footnote{It is possible
+  that a gradient that is approximated using a complex step method
+  would be precise enough to be useful. A further investigation is
+  beyond the scope of this paper.} However, deriving a vector of first
 derivatives, and writing \proglang{R} functions to compute them, is a
@@
 
-The partial derivative of a scalar-valued function $f(x)$ with respect to $x_j$ (the $j$th
+The partial derivative of a real scalar-valued function $f(x)$ with respect to $x_j$ (the $j$th
 component of $x$) is defined as
@@
 
-To estimate the $m$th column of $\hess{}{}$, we again choose a
+\subsubsection[Approximation using finite differences]{Approximation using finite differences}
+
+To estimate the $m$th column of $\hess{}{}$ using finite differences, we choose a
 sufficiently small $\delta$, and compute
@@
   \label{eq:FDhess2}
-  \hess{}{f(x)}&=
+  \hess{}{f(x)}&\approx
   \begin{pmatrix}
@@
   \label{eq:FDhess2sp}
-  \hess{}{f(x)}&=
+  \hess{}{f(x)}&\approx
   \begin{pmatrix}
@@
 the cross-partial derivatives are zero.
+
+\subsubsection[Approximation using complex steps]{Approximation using complex steps}
+
+When $f(x)$ is complex analytic, we can approximate $\parD{}{}{f(x)}$
+and $\hess{}{f(x)}$ using the complex step
+method. To do this, we extend our defintion of a real $x\in\Real{M}$
+to be a complex $x\in\Complex{M}$ for which the imaginary part is
+zero. We then take a Taylor
+series expansion of $f(x)$ in the imaginary direction of the complex
+plane \citep{SquireTrapp1998}, rearrange terms, and take the imaginary
+parts of both sides.
+\begin{align}
+  \label{eq:complexTaylor}
+  f(x+i\delta)&=f(x) + i\delta\parD{}{}{f(x)}+\mathcal{O}(\delta^2)\\
+  \parD{}{}{f(x)}&\approx\frac{\Imag(f(x+i\delta))}{\delta}
+\end{align}
+
+Estimating a first derivative using the complex step method does not
+require a differencing operation, but a second-order approximation of
+the Hessian does \citep{AbreuStich2013}.  However, when the gradient
+is provided, we can estimate the Hessian as if it were a Jacobian of
+the vector-valued gradient function \citep{LaiCrassidis2008}.
+
+\begin{align}
+  \label{eq:CShess2}
+  \hess{}{f(x)}&\approx
+  \begin{pmatrix}
+    \Imag(\parD{}{1}{f(x_1+i\delta, x_2)})& \Imag(\parD{}{1}{f(x_1,x_2+i\delta)})\\
+        \Imag(\parD{}{2}{f(x_1+i\delta, x_2)})&\Imag(\parD{}{2}{f(x_1,x_2+i\delta)})
+    \end{pmatrix}/\delta
+\end{align}
+
+If this matrix were dense, we would need two evaluations of the
+complex $\parD{}{}f(x)$ to estimate it.  But what if it had the same sparsity
+pattern as the Hessian in Equation~\ref{eq:FDhess2sp}? First, we need
+to make an assumption that structural zeros remain zero for all
+complex $x\in\Complex{M}$.  Next, subtract
+$\Imag(\parD{}{}{f(x_1,x_2)})$ from each column of
+$\hess{}{f(x)}$. When $x$ is real, the imaginary part of the gradient
+is zero, so this operation has no effect on the value of the
+Hessian. But for all complex $x$, the sparsity constraints ensure that
+the following identities hold:
+
+\begin{align}
+  \Imag(\parD{}{1}{f(x_1,x_2+i\delta)}) -\Imag(\parD{}{1}{f(x_1,x_2)})&=0\label{eq:I1c}\\
+    \Imag(\parD{}{2}{f(x_1+i\delta,x_2)})-\Imag(\parD{}{2}{f(x_1,x_2)})&=0\label{eq:I2c}
+\end{align}
+
+As with the finite difference method, because Equation~\ref{eq:I1c} holds
+for $x_1$, it must also hold for $x_1+i\delta$, because Equation~\ref{eq:I1d} holds
+for $x_2$, it must also hold for $x_2+i\delta$. Thus, for real $x$,
+
+\begin{align}
+  \label{eq:CShess2sp}
+  \hess{}{f(x)}&\approx
+  \begin{pmatrix}
+    \Imag(\parD{}{1}{f(x_1+i\delta, x_2+i\delta)})&0\\
+        0&\Imag(\parD{}{2}{f(x_1+i\delta,x_2+i\delta)})
+    \end{pmatrix}/\delta
+\end{align}
+
+Only one evaluation of the gradient is required.
+
+\subsubsection[Perturbing groups of variables]{Perturbing groups of variables}
 
@@
 \end{align}
-where $i=\sqrt{-1}$ and $\Imag$ returns the imaginary part of a complex
-number. The advantage of this method is that there is no longer a
+The advantage of this method is that there is no longer a
 subtraction operation to generate truncation error. Thus, $\delta$ can

### 2016-12-25 nobuild/jss_revision/sparseHessianFD.Rnw
proof of sparsity with complex step method

@@
 
-When $f(x)$ is complex analytic, we can approximate $\parD{}{}{f(x)}$
-and $\hess{}{f(x)}$ using the complex step
-method. To do this, we extend our defintion of a real $x\in\Real{M}$
-to be a complex $x\in\Complex{M}$ for which the imaginary part is
-zero. We then take a Taylor
+We are interested in estimating the Hessian only for real
+$x\in\Real{M}$. However, when $f(x)$ is complex-valued over a complex domain $x\in\Complex{M}$, we can approximate $\parD{}{}{f(x)}$
+and $\hess{}{f(x)}$ for real $x$ using the complex step
+method.  This method comes from a Taylor
 series expansion of $f(x)$ in the imaginary direction of the complex
-plane \citep{SquireTrapp1998}, rearrange terms, and take the imaginary
-parts of both sides.
+plane \citep{SquireTrapp1998}.  After rearranging terms, and taking the imaginary
+parts of both sides,
 \begin{align}
@@
 Estimating a first derivative using the complex step method does not
-require a differencing operation, but a second-order approximation of
-the Hessian does \citep{AbreuStich2013}.  However, when the gradient
-is provided, we can estimate the Hessian as if it were a Jacobian of
-the vector-valued gradient function \citep{LaiCrassidis2008}.
+require a differencing operation, so there is no loss of precision
+from rounding error, and the approximation of the derivative can be
+made arbitrarily precise as $\delta\rightarrow 0$.  This is not the
+case for second-order approximations of
+the Hessian \citep{AbreuStich2013}.  However, when the gradient can be
+computed exactly, we can compute a first-order approximation to
+Hessian by treating it as the Jacobian of
+a vector-valued function \citep{LaiCrassidis2008}.
 
@@
 complex $\parD{}{}f(x)$ to estimate it.  But what if it had the same sparsity
-pattern as the Hessian in Equation~\ref{eq:FDhess2sp}? First, we need
-to make an assumption that structural zeros remain zero for all
+pattern as the Hessian in Equation~\ref{eq:FDhess2sp}? First, we
+assume that structural zeros remain zero for all
 complex $x\in\Complex{M}$.  Next, subtract
@@
 is zero, so this operation has no effect on the value of the
-Hessian. But for all complex $x$, the sparsity constraints ensure that
+Hessian. But the sparsity constraints ensure that
 the following identities hold:
@@
 As with the finite difference method, because Equation~\ref{eq:I1c} holds
-for $x_1$, it must also hold for $x_1+i\delta$, because Equation~\ref{eq:I1d} holds
-for $x_2$, it must also hold for $x_2+i\delta$. Thus, for real $x$,
+for $x_1$, it must also hold for $x_1+i\delta$, and because Equation~\ref{eq:I2c} holds
+for $x_2$, it must also hold for $x_2+i\delta$. Thus,
 
@@
 
-Next, define $\Mat{Y}\in\Real{M\times C}$ such that each column is a
-difference in gradients.
+Next, define $\Mat{Y}\in\Real{M\times C}$ such that each column is
+either a difference in gradients, or the imaginary part of a
+complex-valued gradient, depending on the chosen method.
 \begin{align}
   \label{eq:Yg}
-  \Mat{Y}_c&=\nabla f(x+G_c)-\nabla f(x)
+  \Mat{Y}_c&=
+           \begin{cases}  
+             \nabla f(x+G_c)-\nabla f(x)&\text{finite difference method}\\
+             \Imag(\nabla f(x+iG_c))&\text{complex step method}
+             \end{cases}
 \end{align}
@@
 $\hess{im}{}\delta\approx y_{im}$, and we can solve for all elements of $\hess{}{}$
-just by computing $Y$. But if $C<M$,
+just by computing $\Mat{Y}$. But if $C<M$,
 there must be at least one $\Mat{G}_c$ with $\delta$ in at least two
@@
 @
-
-\subsection[Complex step method]{Complex step method}\label{sec:complex}
-
-The \pkg{sparseHessianFD} package also allows for
-estimation of the Hessian using the complex step method
-\citep{SquireTrapp1998} as an alternative to finite differencing. In
-this method, the definition of $\Mat{Y}_c$ in Equation~\ref{eq:Yg} is replaced with
-\begin{align}
-  \label{eq:complexYg}
-  \Mat{Y}_c&=\Imag(\nabla f(x+iG_c))
-\end{align}
-The advantage of this method is that there is no longer a
-subtraction operation to generate truncation error. Thus, $\delta$ can
-be made arbitrarily small, and the resulting estimate of the Hessian
-is more accurate.
 

### 2016-12-25 nobuild/jss_revision/sparseHessianFD.tex
proof of sparsity with complex step method

@@
 \end{Schunk}
-
-\subsection[Complex step method]{Complex step method}\label{sec:complex}
-
-The \pkg{sparseHessianFD} package also allows for
-estimation of the Hessian using the complex step method
-\citep{SquireTrapp1998} as an alternative to finite differencing. In
-this method, the definition of $\Mat{Y}_c$ in Equation~\ref{eq:Yg} is replaced with
-\begin{align}
-  \label{eq:complexYg}
-  \Mat{Y}_c&=\Imag(\nabla f(x+iG_c))
-\end{align}
-The advantage of this method is that there is no longer a
-subtraction operation to generate truncation error. Thus, $\delta$ can
-be made arbitrarily small, and the resulting estimate of the Hessian
-is more accurate.
 

### 2016-12-25 nobuild/jss_revision/sparseHessianFD.Rnw
proof of sparsity with complex step method

@@
 \emph{sparse}, meaning that the proportion of  ``structural''  zeros
-(Hessian matrix elements that are always zero, regardless of the value at which
+(matrix elements that are always zero, regardless of the value at which
 function is estimated) is high.  Consider a log
@@
 gradient: one for the gradient at $x$, and one after perturbing each
-of the $M$ elements of $x$, one at a time. If the function is complex
-analytic, an alternative approximation that requires $M$ gradient
-evaulations is $\hess{m}{}\approx\Imag\left(\nabla f(x+i\delta
-  e_m)\right)/\delta$, where $i=\sqrt{-1}$ and $\Imag$ returns the imaginary part of a complex
-number \citep{SquireTrapp1998}. This ``complex step'' method can be more accurate than the
-finite difference method because there is no subtraction operation
-that might generate roundoff errors, and thus $\delta$ can be made
-arbitrarily small to avoid truncaton errors \citep{LaiCrassidis2008}. However, it does require that the
-implementation of $f(x)$ accept a complex argument, and return a
-complex result, even though we are interested in differentiating $f(x)$ only along
-the real line.
-
-Regardless of the approximation method used, if the Hessian is sparse, most of the
+of the $M$ elements of $x$, one at a time. Under certain conditions, a
+more accurate approximation is the ``complex step'' method:  $\hess{m}{}\approx\Imag\left(\nabla f(x+i\delta
+  e_m)\right)/\delta$, where $i=\sqrt{-1}$ and $\Imag$ returns the
+imaginary part of a complex
+number \citep{SquireTrapp1998}.  Regardless of the approximation method used, if the Hessian is sparse, most of the
 elements are constrained to zero.  Depending
@@
 
-The partial derivative of a real scalar-valued function $f(x)$ with respect to $x_j$ (the $j$th
-component of $x$) is defined as
+The partial derivative of a real scalar-valued function $f(x)$ with
+respect to $x_j$ (the $j$th
+component of $x\in\Real{M}$) is defined as
 \begin{align}
@@
 
-We are interested in estimating the Hessian only for real
-$x\in\Real{M}$. However, when $f(x)$ is complex-valued over a complex domain $x\in\Complex{M}$, we can approximate $\parD{}{}{f(x)}$
-and $\hess{}{f(x)}$ for real $x$ using the complex step
+If $f(x)$ is defined over a complex domain and is holomorphic, then we
+can we can approximate $\parD{}{}{f(x)}$
+and $\hess{}{f(x)}$ at real values of $x$ using the complex step
 method. This method comes from a Taylor
@@
 Estimating a first derivative using the complex step method does not
-require a differencing operation, so there is no loss of precision
-from rounding error, and the approximation of the derivative can be
-made arbitrarily precise as $\delta\rightarrow 0$.  This is not the
+require a differencing operation, so there is no subtraction operation
+that might generate roundoff errors.  Thus, the approximation can be
+made arbitrarily precise as $\delta\rightarrow 0$ \citep{LaiCrassidis2008}.  This is not the
 case for second-order approximations of
@@
 If this matrix were dense, we would need two evaluations of the
-complex $\parD{}{}f(x)$ to estimate it.  But what if it had the same sparsity
-pattern as the Hessian in Equation~\ref{eq:FDhess2sp}? First, we
+$\parD{}{}f(x)$ to estimate it.  If the matrix were sparse, with the same sparsity
+pattern as the Hessian in Equation~\ref{eq:FDhess2sp}, and we
 assume that structural zeros remain zero for all
-complex $x\in\Complex{M}$.  Next, subtract
-$\Imag(\parD{}{}{f(x_1,x_2)})$ from each column of
+complex $x\in\Complex{M}$, then we need only one evaluation.  Suppose
+we were to subtract $\Imag(\parD{}{}{f(x_1,x_2)})$ from each column of
 $\hess{}{f(x)}$. When $x$ is real, the imaginary part of the gradient
@@
 Hessian. But the sparsity constraints ensure that
-the following identities hold:
+the following identities hold all complex $x$.
 
@@
 
-Only one evaluation of the gradient is required.
+for real $x$.  Only one evaluation of the gradient is required.
 
@@
 
-The evaluation of the Hessian is a finite differenced approximation,
+The default choice of method is \funcarg{complex=FALSE}, so the
+evaluation of the Hessian is a finite differenced approximation,
 so it is very close to, but not identical to, the true value, in terms
@@
 
-A call to the \class{sparseHessianFD} \method{hessian} method will
-apply the complex step method if the 
-\funcarg{complex=TRUE} was set in the initializer. The functions
+If \funcarg{complex=TRUE} in the initializer, the call to the \method{hessian} method will
+apply the complex step method. To use this method, the functions
 passed as \funcarg{fn} and \funcarg{gr} must both
@@
 complex implementations. Furthermore, the complex step method is valid
-only if the function is holomorphic
-(complex differentiable). The methods in \class{sparseHessianFD} do
+only if the function is holomorphic. The methods in \class{sparseHessianFD} do
 \emph{not} check that this is the case for the function at hand. We convey the
@@
 \citep{R_numDeriv}, which also implements the complex step method:
-``Avoid this method if you do not know that your function is
+``avoid this method if you do not know that your function is
 suitable. Your mistake may not be caught and the results will be
@@
 
-Fortunately for demonstration purposes, the log posterior density in Equation~\ref{eq:LPD} is holomorphic, so we
+Fortunately for demonstration purposes, the log posterior density in
+Equation~\ref{eq:LPD} is holomorphic, so we
 can estimate its Hessian using the complex step method, and compute
@@
 
-In short, the complex step method can be much more accurate than
+In short, the complex step method can be more accurate than
 finite differencing, but it comes with theoretical and implementation

### 2016-12-25 nobuild/jss_revision/response.tex
fixed revision

--- 2016-12-24 nobuild/jss_revision/response.tex
@@
  \vspace{\baselineskip}
- Response to reviewer
+ Response to editor and reviewer
 \end{center}
@@
 I would like to express my sincere appreciation for the time and effort that
-you invested in the review.  It is obvious to me that you went through
-the paper and code in great detail.  In
-particular, thank you for bringing the complex step
-method to my attention.  I \emph{never} use complex numbers or
-functions in my own work, so I typically ignore anything with the word
-``complex'' in it.  But this review prompted me to learn something
-new, and that is always fun.
-
-I am happy to write that I was able to incorporate a large majority of
-your suggestions into the the revised paper and package. For the
-others, I hope that you, and the editor, will accept my
+was invested in the review.  It is obvious to me that the reviewer went through
+the paper and code in great detail.  I am happy to write that I was
+able to incorporate a large majority of
+the suggestions into the the revised paper and package. For the
+others, I hope that you will accept my
 explanations and excuses.  Of course, if there are any strong
@@
 error is a result of the \pkg{algorithm} package
-not being present in your LaTeX installation.  It is a standard
+not being present in the reviewer's LaTeX installation.  It is a standard
 package in the MacTeX distribution, but I cannot speak with authority about other
@@
 The help page for \pkg{sparseHessianFD} now includes explicit
-instructions for the initializer.  You are correct that there is
+instructions for the initializer.  It is correct that there is
 little guidance in how to document reference classes, and I agree that
@@
   The complex step function is now implemented in the package, and is
-  introduced in Section 2.5 of the paper.  You
-  are correct that ``most R primitives work with complex numbers,''
+  discussed briefly in Section 2.5 of the paper.  It is 
+  correct that ``most R primitives work with complex numbers,''
   but there are many commonly used functions that do not, including 
@@
 \method{simple} and 
-\method{complex} methods. Removing the second-order approximation saves
-a lot of time, and lets me compute Table 4 without resorting to
-parallel computation. That will simplify attempts at replication, and
-removes another variable that might influence results.
-
-Footnote 1 was expanded to acknowledge the complex step method.
+\method{complex} methods.
 
@@
   I added a \func{set.seed} statement near the top of
-  \filename{replication.R}, but you are correct that because only
+  \filename{replication.R}, but it is correct that because only
   times are reported, this does not matter much.
@@
   \begin{response}
-These are all reasonable suggestions.  Let me address them point by point.
+I appreciate the editor's comment that these suggestions might be
+better left for a future paper.  But they are reasonable points that I
+would like to address in turn.
     \begin{itemize}
-    \item It really is all about the sparsity pattern, and
+    \item This is discussed in Section 3.1.  In general, it really is all about the sparsity pattern, and
       specifically, the how few partitions are needed to group the
-      variables.  Of course, it will be hard to keep the number of
-      partitions small if there are too many non-zero elements, so the
-      sparsity itself has some effect.  The arrangement of the non-zeros is
+      variables.  The ``simple sparseness'' comes into play because it
+      is more likely the number of partitions will be small with the
+      the proportion of structural zeros is large. The \emph{arrangement} of the non-zeros is
       related to the ordering of the variables, but the permutation
@@
       arranged in a way that the Hessian has a block-arrow pattern are
-      permuted to get a Hessian with a banded pattern instead.  This
-      is all discussed in Section 3.1.
+      permuted to get a Hessian with a banded pattern instead.
 
-\item I have no idea how to do this in a generalizable way.  I think
+\item I would like to defer this topic for future reserach, mainly
+  because I have no idea how to do this in a generalizable way.  I think
   it has a lot to do with the architecture of the computer
@@
 
-  I do not want to ignore or dismiss this suggestion, but I am simply not clear on
-  what is being asked.  I am open to more specific recommendations.
-
 \item The package now implements the complex step method, and comparisons of
@@
 
-\item Parallel computing on a shared memory architecture really
-  muddles up timing comparisons.  The \emph{only} reason I generated
-  Table 4 in parallel was because of the run time of the \pkg{numDeriv} \func{hessian}
-  function. Now that \func{hessian} was replaced with \func{jacobian},
-  the baseline measures from \pkg{numDeriv} run fast enough that I can
-  generate Table 4 in serial.
-
-Your point that timing comparisons can be affected by
-whether computation is being run on multiple processors is well taken, but the
-``serialization'' of Table 4 makes the issue moot.
-
-  Data for Figure 4 were generated in parallel because of the large
-  number of required simulations.
+\item Again, I would like to defer this topic for future research. In
+  theory, there should be no difference if parallelization is done
+  across distributed notes.  It would be as if each run were completed
+  on a separate computer.  In a shared memory architecture, there
+  could be conflicts in terms of processor demand and memory access,
+  but this is true for any application.
     \end{itemize}
@@
 
+Again, thank you for the helpful comments.  The paper is greatly
+improved by them.
+
 \end{document}

### 2016-12-25 nobuild/jss_revision/response.tex
copied jss article to vignette

@@
  \vspace{\baselineskip}
- Response to editor and reviewer
+ Response to reviewer
 \end{center}
@@
 I would like to express my sincere appreciation for the time and effort that
-was invested in the review.  It is obvious to me that the reviewer went through
-the paper and code in great detail.  I am happy to write that I was
-able to incorporate a large majority of
-the suggestions into the the revised paper and package. For the
-others, I hope that you will accept my
+was invested in the review.  I am happy to write that I was
+able to incorporate a large majority of the suggestions into the the
+revised paper and package. For the
+others, I hope that you and the editor will accept my
 explanations and excuses.  Of course, if there are any strong
@@
 error is a result of the \pkg{algorithm} package
-not being present in the reviewer's LaTeX installation.  It is a standard
+not being present in the LaTeX installation.  It is a standard
 package in the MacTeX distribution, but I cannot speak with authority about other
@@
 existing mechanism to include documentation for the initializer to
-an RC.  But it's hard to do it using Roxygen2 in way that also passes
-CRAN checks. So I just ``hard coded'' a description of the initializer
+an RC.  But it's hard to do it using \pkg{roxygen2} in way that also passes
+CRAN checks. Therefore, I just ``hard coded'' a description of the initializer
 into the details section of the sparseHessianFD-class documentation. I
-could not find a more elegant way of doing it.  
+could not think of a more elegant way of doing it.  
 \end{response}
@@
   \begin{response}
-I included this section primarily as a way to avoid overselling the
-method, but the point that the reader might be scared away by this list of
+In this original submission I included this section primarily as a way to avoid overselling the
+method, but your point that the reader might be scared away by this list of
 restrictions is well-taken.  In the revision, I removed the list, and
-wove the restrictions into the text.  The revision projects a more
+wove the most important restrictions into the text.  The revision projects a more
 positive tone that in the original submission.
@@
   \begin{response}
-Assuming that the R package builder works the way I think it does,
+Assuming that the \proglang{R} package builder works the way I think it does,
 the \filename{replication.R} file, and the data for Table 4 (\filename{vignette\_tab4.Rdata}), will be
@@
 4, I now write that ``code to replicate Table 4 is available as an
-online supplement to this paper.''  I cannot be more specific about
-the name of the replication file, because I believe it depends on the
-volume and issue number of the published version.
+online supplement to this paper.'' In the non-JSS vignette that will be part
+of the package (and posted on CRAN), this sentence will be changed appropriately.
 \end{response}
@@
   I removed the \func{R>} prompt, and now refer to that code as a usage
-  syntax.  Thank you for the suggestion.
+  syntax.
 \end{response}
@@
   The complex step function is now implemented in the package, and is
-  discussed briefly in Section 2.5 of the paper.  It is 
+  described as one of two methods of numerical approximation (the
+  other being finite differences).  It is 
   correct that ``most R primitives work with complex numbers,''
@@
   \func{log1p}, \func{gamma}, and the probability distribution
-  functions. This limitation, combined with the requirement that the function be
-  holomorphic, and my own inexperience with the complex step method, make
-  me hesitant to promote it too heavily in the paper. But it is a nifty
-  trick, and I am happy that it is now included in the package.
+  functions.  This limitation, along
+  with the requirement that the function be
+  holomorphic, are discussed in Section 2.4
 
@@
 \method{complex} methods.
+
+I must admit that I was not familiar with the complex step method
+beforehand.  Incorporating it into the package, and writing about it
+in the paper, was a valuable learning experience for me. That said,
+until now I have \emph{never} come across complex analysis in my own work, and I
+may not have done anything with complex numbers at all since my
+Algebra II class in 10th grade. Thus, it is highly probable that the parts
+of the paper on the complex step method are not precisely correct.
+Any constructive criticism or suggestions in that regard would be
+warmly welcomed.
 
@@
 \begin{response}
-  This information is now included in the first paragraph of Section 4.
+  This information is now included in Section 4.
 \end{response}
@@
 than  \func{numDeriv::hessian}, I was able to increase the size of the
-examples in Table 4. One objective of Table 4 is to compare the relative
-increases in computation time for different $N$ and $k$ between
-\pkg{numDeriv} and \pkg{sparseHessianFD}, and I think that objective
-is met with $N=(15, 50, 100, 500)$.  Please forgive me for not going as high as
-$N=2500$, $k=8$. For the reader who wants to
-replicate the results, a run time of more than a day is a lot to ask.
+examples in Table 4: $N=(15, 50, 100, 500)$ and $k=(2,5,8)$. Please
+excuse me for not going as high as $N=2500$, $k=8$. For the reader
+who wants to replicate the results, a run time of more than a day is a
+lot to ask.
   \end{response}
@@
 \item\begin{revQuote}
-
  p.16  l. -7 "to to compute" -> "to compute"
-
   \end{revQuote}
@@
 \item\begin{revQuote}
-
 While that package is useful, and reasonably demonstrated in the paper,
@@
   \begin{response}
-I appreciate the editor's comment that these suggestions might be
+I appreciate the editor's comment in the decision letter that these suggestions might be
 better left for a future paper.  But they are reasonable points that I
@@
 \end{enumerate}
-
-Again, thank you for the helpful comments.  The paper is greatly
-improved by them.
-
 \end{document}

### 2016-12-25 nobuild/jss_revision/sparseHessianFD.tex
copied jss article to vignette

@@
   \midrule
-15 & 2 & 32 & 12.4 & 1.2 & 2.3 & 0.5 & 12.8 & 1.2 & 2.1 & 0.5 \\ 
-  15 & 4 & 64 & 26.2 & 12.4 & 4.0 & 0.9 & 26.1 & 2.0 & 3.8 & 1.1 \\ 
-  50 & 2 & 102 & 50.5 & 2.4 & 3.0 & 0.9 & 54.8 & 3.5 & 2.9 & 2.1 \\ 
-  15 & 8 & 128 & 50.4 & 2.2 & 8.5 & 12.0 & 55.0 & 2.7 & 7.7 & 1.5 \\ 
-  100 & 2 & 202 & 134.1 & 15.7 & 3.7 & 0.5 & 144.0 & 11.6 & 3.3 & 0.6 \\ 
-  50 & 4 & 204 & 104.0 & 2.1 & 6.3 & 11.6 & 117.2 & 17.6 & 5.1 & 1.0 \\ 
-  100 & 4 & 404 & 284.3 & 34.3 & 8.0 & 10.9 & 322.5 & 40.8 & 8.2 & 11.1 \\ 
-  50 & 8 & 408 & 248.5 & 48.7 & 10.9 & 1.4 & 271.7 & 43.7 & 11.5 & 1.4 \\ 
-  100 & 8 & 808 & 624.5 & 54.0 & 14.4 & 1.8 & 762.6 & 76.9 & 15.7 & 1.7 \\ 
-  500 & 2 & 1002 & 1758.6 & 86.7 & 10.1 & 10.0 & 1990.0 & 90.6 & 9.3 & 9.8 \\ 
-  500 & 4 & 2004 & 3832.5 & 84.9 & 18.4 & 2.3 & 4460.7 & 82.8 & 19.4 & 2.5 \\ 
-  500 & 8 & 4008 & 8846.5 & 82.1 & 45.2 & 6.2 & 11058.4 & 101.3 & 52.5 & 6.3 \\ 
+15 & 2 & 32 & 12.7 & 0.9 & 2.4 & 0.1 & 13.2 & 0.7 & 2.0 & 0.1 \\ 
+  15 & 5 & 80 & 32.2 & 0.9 & 5.0 & 0.5 & 36.1 & 10.1 & 5.2 & 0.8 \\ 
+  50 & 2 & 102 & 51.3 & 9.8 & 3.0 & 0.4 & 54.5 & 1.0 & 2.7 & 0.6 \\ 
+  15 & 8 & 128 & 53.6 & 0.9 & 7.8 & 0.6 & 62.5 & 16.7 & 8.2 & 0.8 \\ 
+  100 & 2 & 202 & 131.7 & 16.9 & 3.7 & 0.3 & 147.6 & 24.1 & 3.4 & 0.5 \\ 
+  50 & 5 & 255 & 144.0 & 27.1 & 6.6 & 0.6 & 162.8 & 29.2 & 6.7 & 0.7 \\ 
+  50 & 8 & 408 & 256.2 & 43.5 & 12.5 & 12.9 & 291.1 & 39.0 & 11.7 & 0.7 \\ 
+  100 & 5 & 505 & 495.7 & 129.5 & 9.8 & 3.1 & 559.8 & 125.9 & 11.8 & 10.4 \\ 
+  100 & 8 & 808 & 850.0 & 171.4 & 20.9 & 20.4 & 999.2 & 217.5 & 23.6 & 17.3 \\ 
+  500 & 2 & 1002 & 1953.4 & 321.7 & 10.8 & 3.7 & 2156.2 & 339.7 & 9.5 & 3.4 \\ 
+  500 & 5 & 2505 & 5444.0 & 951.0 & 27.1 & 10.6 & 6159.3 & 1009.2 & 26.5 & 4.6 \\ 
+  500 & 8 & 4008 & 9533.1 & 190.9 & 54.5 & 12.8 & 14059.4 & 3797.9 & 62.5 & 21.8 \\ 
    \bottomrule
@@
 recorded the run times for different steps in the sparse Hessian
-estimation, across 200 replications.  The steps are summarized in
+estimation, across 500 replications.  The steps are summarized in
 Table~\ref{tab:timeMeasures}.  The times were generated on an

### 2016-12-25 nobuild/jss_revision/replication.R

@@ -10,7 +10,6 @@ library("reshape2")
 library("ggplot2")
 
 set.seed(1234)
-
 registerDoParallel(cores=6)
 
 binary_sim <- function(N, k, T) {
@@ -64,7 +63,6 @@ run_test_fig4 <- function(NkT, reps=50, order.row=FALSE) {
   priors <- priors_sim(k)
   F <- make_funcs(D=data, priors=priors, order.row=order.row)
   nvars <- N*k+k
-
   if (order.row) {
       mm <- Matrix::kronecker(Matrix(1,k,k),Matrix::Diagonal(N))
   } else {
@@ -74,16 +72,13 @@ run_test_fig4 <- function(NkT, reps=50, order.row=FALSE) {
        cBind(Matrix(TRUE, k*(N+1), k)) %>%
        as("sparseMatrix") %>%
        as("nMatrix")
-
   pat <- Matrix.to.Coord(tril(M))
   X <- rnorm(nvars)
-
   obj <- sparseHessianFD(X, F$fn, F$gr, pat$rows, pat$cols, complex=TRUE)
   colors <- obj$partition()
   perm <- obj$get_perm()
   ncolors <- length(unique(colors))
   nvars <- N*k+k
-
   bench <- microbenchmark(
       setup = sparseHessianFD(X, F$fn, F$gr, pat$rows, pat$cols),
       colors = coloring(tril(M[perm,perm])),
@@ -92,20 +87,15 @@ run_test_fig4 <- function(NkT, reps=50, order.row=FALSE) {
       df = obj$gr(X),
       hess = obj$hessian(X),
       times = reps)
-
   vals <- plyr::ddply(data.frame(bench), "expr",
                 function(x) return(data.frame(expr=x$expr,
                                               time=x$time,
                                               rep=1:length(x$expr))))
-
   res <- data.frame(N=N, k=k, T=T,
                     bench=vals,
                     ncolors=ncolors)
-
   cat("Completed N = ",N,"\tk = " , k , "\tT = ",T,"\n")
-
   return(res)
-
 }
 
 
@@ -117,7 +107,6 @@ run_test_tab4 <- function(Nk, reps=50, order.row=TRUE) {
     priors <- priors_sim(k)
     F <- make_funcs(D=data, priors=priors, order.row=order.row)
     nvars <- N*k+k
-
     if (order.row) {
[…]

### 2016-12-25 nobuild/jss_revision/sparseHessianFD.tex
checking with win-builder

@@
   \pkg{numDeriv} and \pkg{sparseHessianFD} packages, and the finite
- difference and complex step methods, across 200 replications. Rows are ordered by the
+ difference and complex step methods, across 500 replications. Rows are ordered by the
   number of variables.}\label{tab:numDeriv}
@@
 recorded the run times for different steps in the sparse Hessian
-estimation, across 500 replications.  The steps are summarized in
+estimation, across 200 replications.  The steps are summarized in
 Table~\ref{tab:timeMeasures}.  The times were generated on an

### 2017-01-24 nobuild/jss_revision/sparseHessianFD.Rnw
copied jss article to vignette
preparing final jss submission

@@
 
-\date{December 22, 2016}
+\date{January 25, 2017}
 
@@
 Define $\parD{2}{x_1,x_2}$ as the product set of cross-partial
-derivatives between all elements in $x_1$ and all elements in $x_2$.
-Furthermore, define an indexing function Ind$(x_i)$ that returns the
-indices in the parameter vector
-$x$ that correspond to the elements in $x_i$, and $\hess{x_1,x_2}{}$
-as the Hessian elements in rows Ind$(x_1)$ and columns Ind$(x_2)$, and
-in rows Ind$(x_2)$ and columns Ind$(x_1)$ (from the symmetry of the
-Hessian).
-
-From the log posterior density in Equation~\ref{eq:LPD}, we can see
+derivatives between all elements in $x_1$ and all elements in
+$x_2$. From the log posterior density in Equation~\ref{eq:LPD}, we can see
 that $\parD{2}{\beta_i, \beta_i}\neq 0$ (one element of $\beta_i$ could be
@@
 greater than $k$, there will be many more zero cross-partial
-derivatives than non-zero.  Regardless of how each $\parD{2}{}$ is
-mapped to $\hess{}{}$ through the indexing function, the Hessian of the log
-posterior density will be sparse.
+derivatives than non-zero.   Each $\parD{2}{}$ is
+mapped to a submatrix of $\hess{}{}$, most of which with be zero.  The
+resulting Hessian of the log posterior density will be sparse.
 
@@
     5&0&0&0&0&1&&\\
-    6&1&0&0&0&1&1&\\
+    6&0&0&0&0&1&1&\\
     7&1&1&1&1&1&1&1

### 2017-01-24 nobuild/jss/sparseHessianFD.Rnw
cleaning up

@@
 %\VignetteIndexEntry{sparseHessianFD}
-%\VignettePackage{sparseHessianFD}
+%\VignetteKeyword{sparseHessianFD}
 
@@
 \newcommand{\Mat}[1]{\mathbf{#1}}
+\newcommand{\Real}[1]{\mathbb{R}^{#1}}
+\newcommand{\Complex}[1]{\mathbb{Z}^{#1}}
+\DeclareMathOperator{\Imag}{Im}
 
@@
 \Keywords{sparse Hessians, sparsity, computation of Hessians,
-    graph coloring, finite differences, differentiation}
+  graph coloring, finite differences, differentiation, complex step}
 \Address{
@@
 
-\date{March 9, 2016}
+\date{January 25, 2017}
 
@@
 knitr::render_sweave()
-knitr::opts_chunk$set(prompt=TRUE, cache=FALSE)
-options(replace.assign=TRUE, width=90,prompt="R> ")
+knitr::opts_chunk$set(prompt=TRUE, cache=FALSE,error=FALSE) #$
+options(replace.assign=TRUE, width=77, prompt="R> ")
 @
@@
 However, for many problems, the Hessian is
-\emph{sparse}, meaning that the proportion of non-zero elements in the
-Hessian is small.  Consider a log
+\emph{sparse}, meaning that the proportion of  ``structural''  zeros
+(matrix elements that are always zero, regardless of the value at which
+function is estimated) is high.  Consider a log
 posterior density in a Bayesian hierarchical
@@
 The \pkg{sparseHessianFD} package is a tool for estimating
-sparse Hessians using finite differences of gradients.  Section \ref{sec:numdiff} will
-cover the specifics, but the basic idea is as follows.  Consider a
+sparse Hessians numerically, using either finite differences or
+complex perturbations of
+gradients.  Section~\ref{sec:numdiff} will
+cover the specifics, but the basic idea is as follows.  Consider a real-valued
 function $f(x)$, its gradient $\nabla{f(x)}$, and its Hessian
@@
 element, and let $\delta$ be a sufficiently small scalar
-constant. A linear approximation to the $m$th column of the Hessian is ${\hess{m}{}\approx\left(\nabla f(x+\delta
+constant. A ``finite difference'' linear approximation to the $m$th column of the Hessian is ${\hess{m}{}\approx\left(\nabla f(x+\delta
   e_m)- \nabla f(x)\right)/\delta}$.  Estimating a dense Hessian in this way involves at least $M+1$ calculations of the
 gradient: one for the gradient at $x$, and one after perturbing each
-of the $M$ elements of $x$, one at a time.\footnote{More accurate
-  approximations require more gradient evaluations.  We will consider
-  only forward finite differences.}  In a sparse matrix, most of the
+of the $M$ elements of $x$, one at a time. Under certain conditions, a
+more accurate approximation is the ``complex step'' method:  $\hess{m}{}\approx\Imag\left(\nabla f(x+i\delta
+  e_m)\right)/\delta$, where $i=\sqrt{-1}$ and $\Imag$ returns the
+imaginary part of a complex
+number \citep{SquireTrapp1998}.  Regardless of the approximation method used, if the Hessian is sparse, most of the
 elements are constrained to zero.  Depending
@@
 The package defines the \class{sparseHessianFD} class, whose initializer
-requires the user to provide functions that compute the objective
-function and the gradient, as well as the sparsity pattern of the
-Hessian.  Some users may find those requirements burdensome, and we
-emphasize that there may be some applications for
-which \pkg{sparseHessianFD} is not an appropriate package to use. To
-extract the maximum benefit from using \pkg{sparseHessianFD}, we need
-to accept some conditions and assumptions.
-
-\begin{enumerate}
-  \item Preferred alternatives to computing the Hessian are not
-  available.  Finite differencing is not generally a ``first choice''
-  method.  Deriving a gradient or Hessian symbolically, and writing a
-  subroutine to compute it, will give an exact answer, but might be
-  tedious or difficult to implement. Algorithmic differentiation (AD) is
-  probably the most efficient method, but requires specialized
-  libraries that, at this moment, are not yet broadly available in \proglang{R}.
- \pkg{sparseHessianFD} makes the most sense when the gradient is easy to compute, but the Hessian is
-  not.
-  \item The application can tolerate the approximation error in the Hessian
-  that comes with finite differencing methods.
-\item The objective function $f(x)$ is twice differentiable, and can be
-  computed ``quickly and easily,'' even for a large number of
-  variables.  We leave the definition of ``quickly and easily''
-  intentionally murky, since no method of differentiation can overcome
-  pathologies in a function that itself is hard to compute.
-\item The gradient can
-  be computed quickly, easily and \emph{exactly} (within machine
-  precision).   We do not recommend using finite differenced gradients
-  when computing finite differenced Hessians.  The
-  approximation errors will be compounded, and the time
-  complexity of computing a gradient grows with the number of
-  variables.
-\item The sparsity pattern is known in advance, and does not depend
-    on the values of the variables.
-\end{enumerate}
-
-Our perspective on the requirements to use \pkg{sparseHessianFD} is
-that deriving a vector of first
+requires the user to provide functions that compute an objective
+function, its gradient (as accurately as possible, to machine precision), and the sparsity pattern of its Hessian
+matrix. The sparsity pattern (e.g., location of structural
+zeros) must be known in advance, and cannot vary across the domain of the
+objective function. The only functions and methods of the class that
+the end user should need to use are the
+initializer, methods that return the Hessian
+in a sparse compressed format, and perhaps some utility functions that
+simplify the construction of the sparsity pattern.  The class also
+defines methods that partition the variables into groups that can be perturbed together in a finite
+differencing step, and recovers the elements of the Hessian via linear
+substitution.  Those methods perform most of the work, but should be
+invisible to the user.
+
+As with any computing method or algorithm, there are boundaries around
+the space of applications for which \pkg{sparseHessianFD} is the right
+tool for the job. In general, numerical approximations are not ``first
+choice'' methods because the result is not exact, so
+\pkg{sparseHessianFD} should not be used when the application cannot
+tolerate any error, no matter how small.  Also, we admit that some
+users might balk at having to provide an exact gradient, even though
+the Hessian will be estimated numerically.\footnote{It is possible
+  that a gradient that is approximated using a complex step method
+  would be precise enough to be useful. A further investigation is
+  beyond the scope of this paper.} However, deriving a vector of first
 derivatives, and writing \proglang{R} functions to compute them, is a
@@
 more accurate than computing second-order approximations from the
-objective function. Even when we have derived the Hessian symbolically, in practice it may still be faster to estimate the Hessian
+objective function. Even when we have derived the Hessian
+symbolically, in practice it may still be faster to estimate the Hessian
 using \pkg{sparseHessianFD} than coding it directly.  These are the
 situations in which \pkg{sparseHessianFD} adds the most value to the
-statistician's toolbox.  If AD software is available to compute the gradient, then
-it is probably available for sparse Hessians as well, and
-\pkg{sparseHessianFD} would not be needed.  Without AD-computed or
-symbolically derived Hessians, \pkg{sparseHessianFD} is the best
-available option.
-
-The only functions and methods the end user should need to use are the
-\class{sparseHessianFD} initializer, methods that return the Hessian
-in a sparse compressed format, and perhaps some utility functions that
-simplify the construction of the sparsity pattern.  The class also
-defines methods that partition
-the variables into groups that can be perturbed together in a finite
-differencing step, and recovers the elements of the Hessian via linear
-substitution.  Those methods perform most of the work, but should be
-invisible to the user.
+statistician's toolbox.
 
@@
 
-The partial derivative of a scalar-valued function $f(x)$ with respect to $x_j$ (the $j$th
-component of $x$) is defined as
+The partial derivative of a real scalar-valued function $f(x)$ with
+respect to $x_j$ (the $j$th
+component of $x\in\Real{M}$) is defined as
 \begin{align}
@@
 
+\subsubsection[Approximation using complex steps]{Approximation using complex steps}
+
+If $f(x)$ is defined over a complex domain and is holomorphic, then we
+can we can approximate $\parD{}{}{f(x)}$
+and $\hess{}{f(x)}$ at real values of $x$ using the complex step
+method. This method comes from a Taylor
+series expansion of $f(x)$ in the imaginary direction of the complex
+plane \citep{SquireTrapp1998}.  After rearranging terms, and taking the imaginary
+parts of both sides,
+\begin{align}
+  \label{eq:complexTaylor}
+  f(x+i\delta)&=f(x) + i\delta\parD{}{}{f(x)}+\mathcal{O}(\delta^2)\\
+  \parD{}{}{f(x)}&\approx\frac{\Imag(f(x+i\delta))}{\delta}
+\end{align}
+
+Estimating a first derivative using the complex step method does not
+require a differencing operation, so there is no subtraction operation
+that might generate roundoff errors.  Thus, the approximation can be
+made arbitrarily precise as $\delta\rightarrow 0$ \citep{LaiCrassidis2008}.  This is not the
+case for second-order approximations of
+the Hessian \citep{AbreuStich2013}.  However, when the gradient can be
+computed exactly, we can compute a first-order approximation to
+Hessian by treating it as the Jacobian of
+a vector-valued function \citep{LaiCrassidis2008}.
+
+\begin{align}
+  \label{eq:CShess2}
+  \hess{}{f(x)}&\approx
+  \begin{pmatrix}
+    \Imag(\parD{}{1}{f(x_1+i\delta, x_2)})& \Imag(\parD{}{1}{f(x_1,x_2+i\delta)})\\
+        \Imag(\parD{}{2}{f(x_1+i\delta, x_2)})&\Imag(\parD{}{2}{f(x_1,x_2+i\delta)})
+    \end{pmatrix}/\delta
+\end{align}
+
+If this matrix were dense, we would need two evaluations of the
+$\parD{}{}f(x)$ to estimate it.  If the matrix were sparse, with the same sparsity
+pattern as the Hessian in Equation~\ref{eq:FDhess2sp}, and we
+assume that structural zeros remain zero for all
+complex $x\in\Complex{M}$, then we need only one evaluation.  Suppose
+we were to subtract $\Imag(\parD{}{}{f(x_1,x_2)})$ from each column of
+$\hess{}{f(x)}$. When $x$ is real, the imaginary part of the gradient
+is zero, so this operation has no effect on the value of the
+Hessian. But the sparsity constraints ensure that
+the following identities hold all complex $x$.
+
+\begin{align}
+  \Imag(\parD{}{1}{f(x_1,x_2+i\delta)}) -\Imag(\parD{}{1}{f(x_1,x_2)})&=0\label{eq:I1c}\\
+    \Imag(\parD{}{2}{f(x_1+i\delta,x_2)})-\Imag(\parD{}{2}{f(x_1,x_2)})&=0\label{eq:I2c}
+\end{align}
+
+As with the finite difference method, because Equation~\ref{eq:I1c} holds
+for $x_1$, it must also hold for $x_1+i\delta$, and because Equation~\ref{eq:I2c} holds
+for $x_2$, it must also hold for $x_2+i\delta$. Thus,
+
+\begin{align}
+  \label{eq:CShess2sp}
+  \hess{}{f(x)}&\approx
+  \begin{pmatrix}
+    \Imag(\parD{}{1}{f(x_1+i\delta, x_2+i\delta)})&0\\
+        0&\Imag(\parD{}{2}{f(x_1+i\delta,x_2+i\delta)})
+    \end{pmatrix}/\delta
+\end{align}
+
+for real $x$.  Only one evaluation of the gradient is required.
+
+\subsubsection[Perturbing groups of variables]{Perturbing groups of variables}
+
 \Citet{CurtisPowellReid1974} describe a method of estimating sparse
@@
 the decision variables into $C$ mutually exclusive groups so that the number of gradient
-evaluations is reduced.  Let $\Mat{G}$ be a $M\times C$
-matrix, where $\Mat{G}_{mc}=\delta$ if variable $m$ belongs to group $c$, and zero
-otherwise.  Define $\Mat{G}_c$ as the $c$th column of $\Mat{G}$.
-
-Next, let $\Mat{Y}$ be a $M\times C$ matrix, where each column is a
-difference in gradients.
+evaluations is reduced.  Define $\Mat{G}\in\Real{M\times C}$ where $\Mat{G}_{mc}=\delta$ if variable $m$ belongs to group $c$, and zero
+otherwise.  Define $\Mat{G}_c\in\Real{M}$ as the $c$th column of $\Mat{G}$.
+
+Next, define $\Mat{Y}\in\Real{M\times C}$ such that each column is
+either a difference in gradients, or the imaginary part of a
+complex-valued gradient, depending on the chosen method.
 \begin{align}
   \label{eq:Yg}
-  \Mat{Y}_c&=\nabla f(x+G_c)-\nabla f(x)
+  \Mat{Y}_c&=
+           \begin{cases}
+             \nabla f(x+G_c)-\nabla f(x)&\text{finite difference method}\\
+             \Imag(\nabla f(x+iG_c))&\text{complex step method}
+             \end{cases}
 \end{align}
@@
  purchase.  The heterogeneous parameter $p_i$ is the same for all $T$
- opportunities, so $y_i$ is a binomial random variable.  Define each $p_i$ such that it depends on both $k$ continuous covariates
- $x_i$, and a heterogeneous coefficient vector $\beta_i$.
-\begin{align}
-  p_i=\frac{\exp(x_i'\beta_i)}{1+\exp(x_i'\beta_i)},~i=1 ... N
-\end{align}
-
-The coefficients are distributed across the population of households
-following a multivariate normal distribution with mean $\mu$ and
-covariance $\Mat{\Sigma}$.   Assume that we know $\Mat{\Sigma}$, but not $\mu$.
-Instead, place a multivariate normal prior on $\mu$, with mean $0$ and
-covariance $\Mat{\Omega}$.  Thus, each $\beta_i$, and $\mu$ are
-$k-$dimensional vectors, and the total number of unknown variables in the model is $(N+1)k$.
+ opportunities, so $y_i$ is a binomial random variable.
+
+ Let $\beta_i\in\Real{k}$ be a heterogeneous coefficient vector that
+ is specific to household $i$, such that
+ $\beta_i=(\beta_{i1},\dotsc,\beta_{ik})$. Similarly,
+ $z_i\in\Real{k}$ is a vector of household-specific covariates. Define each
+ $p_i$ such that the log odds of $p_i$ is a linear function of
+ $\beta_i$ and $z_i$, but does not depend directly on $\beta_j$ and $z_j$ for
+ another household $j\neq i$.
+\begin{align}
+  p_i=\frac{\exp(z_i'\beta_i)}{1+\exp(z_i'\beta_i)},~i=1 ... N
+\end{align}
+
+The coefficient vectors $\beta_i$ are distributed across the population of households
+following a multivariate normal distribution with mean $\mu\in\Real{k}$ and
+covariance $\Mat{\Sigma}\in\Real{k\times k}$.   Assume that we know
+$\Mat{\Sigma}$, but not $\mu$, so we place a multivariate normal prior
+on $\mu$, with mean $0$ and
+covariance $\Mat{\Omega}\in\Real{k\times k}$.  Thus, the parameter
+vector $x\in\Real{(N+1)k}$ consists of the $Nk$ elements in the $N$ $\beta_i$ vectors,
+and the $k$ elements in $\mu$.
 
@@
 
-The sparsity pattern depends on how the variables are
-ordered within the vector. One such ordering is to group all of the
-coefficients for each unit together.
+Let $x_1$ and $x_2$ be two subsets of elements of $x$.
+Define $\parD{2}{x_1,x_2}$ as the product set of cross-partial
+derivatives between all elements in $x_1$ and all elements in
+$x_2$. From the log posterior density in Equation~\ref{eq:LPD}, we can see
+that $\parD{2}{\beta_i, \beta_i}\neq 0$ (one element of $\beta_i$ could be
+correlated with another element of $\beta_i$), and that, for all $i$,
+$\parD{2}{\beta_i, \mu}\neq 0$ (because $\mu$ is the prior mean of each
+$\beta_i$).  However, since the $\beta_i$ and $\beta_j$ are independently distributed, and the $y_i$ are
+conditionally independent, the cross-partial derivatives
+$\parD{2}{\beta_i, \beta_j}=0$ for all $i\neq j$. When $N$ is much
+greater than $k$, there will be many more zero cross-partial
+derivatives than non-zero.   Each $\parD{2}{}$ is
+mapped to a submatrix of $\hess{}{}$, most of which with be zero.  The
+resulting Hessian of the log posterior density will be sparse.
+
+The sparsity pattern depends on the indexing function; that is, on how the variables are
+ordered in $x$. One such ordering is to group all of the
+coefficients in the $\beta_i$ for each unit together.
 \begin{align}
@@
   the non-zero elements in the \emph{lower triangle} of the Hessian.\\
-direct& This argument is deprecated, and is included only for
-  backwards compatibility with earlier versions.\\
 delta& The perturbation amount for finite differencing of the
-  gradient to compute the Hessian (the $\delta$ in Section \ref{sec:numdiff}).  Defaults to
+  gradient to compute the Hessian (the $\delta$ in Section~\ref{sec:numdiff}).  Defaults to
   \code{sqrt(.Machine\$double.eps)}.\\
@@
 <<eval=FALSE>>=
-f <- obj$fn(x)          ## returns numeric
-df <- obj$gr(x)         ## returns numeric vector
-hess <- obj$hessian(x)  ## returns dgCMatrix
-fngr <- obj$fngr(x)     ## returns list
-fngrhs <- obj$fngrhs(x) ## returns list
-@
-
-\subsection[An example]{An example}
+f <- obj$fn(x)
+df <- obj$gr(x)
+hess <- obj$hessian(x)
+fngr <- obj$fngr(x)
+fngrhs <- obj$fngrhs(x)
+@
+
+\subsection[An example]{An example}\label{sec:exHess}
 
 Now we can estimate the Hessian for the
-log posterior density of the model from Section \ref{sec:example}.
+log posterior density of the model from Section~\ref{sec:example}.
 For demonstration purposes, \pkg{sparseHessianFD} includes functions that compute the value
@@
 
-<<binaryInit>>=
+<<binaryInit, tidy=FALSE>>=
 set.seed(123)
@@
 T <- binary[["T"]]
-nvars <- as.integer(N*k + k)
-priors <- list(inv.Sigma = rWishart(1,k+5,diag(k))[,,1],
+nvars <- as.integer(N * k + k)
+priors <- list(inv.Sigma = rWishart(1, k + 5, diag(k))[,,1],
    inv.Omega = diag(k))
@@
 
-Finally, we create an instance of a \class{sparseHessianFD} object,
-evaluate the function, gradient and Hessian, and compare the output to
-the true values.
-
-<<usingSparseHessianFD>>=
+Finally, we create an instance of a \class{sparseHessianFD} object.
+Evaluations of the function and gradient using the \method{fn} and
+\method{gr} methods will always give the same
+results as the true values because they are computed using the same functions.
+
+<<usingSparseHessianFD1, tidy=FALSE>>=
 obj <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
@@
 f <- obj$fn(P)
-all.equal(f, true.f)
+identical(f, true.f)
 gr <- obj$gr(P)
-all.equal(gr, true.grad)
+identical(gr, true.grad)
+@
+
+The default choice of method is \funcarg{complex=FALSE}, so the
+evaluation of the Hessian is a finite differenced approximation,
+so it is very close to, but not identical to, the true value, in terms
+of mean relative difference.
+
+<<echo=FALSE>>=
+options(scipen=0)
+@
+
+<<usingSparseHessianFD2>>=
 hs <- obj$hessian(P)
-all.equal(hs, true.hess)
-@
+mean(abs(hs-true.hess))/mean(abs(hs))
+@
+
+If \funcarg{complex=TRUE} in the initializer, the call to the \method{hessian} method will
+apply the complex step method. To use this method, the functions
+passed as \funcarg{fn} and \funcarg{gr} must both
+accept a complex argument, and return a complex
+result, even though we are differentiating a real-valued function. Although \pkg{base} \proglang{R} supports complex arguments
+for most basic mathematical functions, many common functions (e.g.,  \func{gamma}, \func{log1p},
+\func{expm1}, and the probability distribution functions) do not have
+complex implementations. Furthermore, the complex step method is valid
+only if the function is holomorphic. The methods in \class{sparseHessianFD} do
+\emph{not} check that this is the case for the function at hand. We convey the
+following warning from the documentation of the \pkg{numDeriv} package
+\citep{R_numDeriv}, which also implements the complex step method:
+``avoid this method if you do not know that your function is
+suitable. Your mistake may not be caught and the results will be
+spurious.''
+
+Fortunately for demonstration purposes, the log posterior density in
+Equation~\ref{eq:LPD} is holomorphic, so we
+can estimate its Hessian using the complex step method, and compute
+the mean relative difference from the true Hessian.
+
+<<tidy=FALSE>>=
+obj2 <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
+   rows=pattern[["rows"]], cols=pattern[["cols"]],
+   complex=TRUE,
+   data=binary, priors=priors, order.row=order.row)
+hs2 <- obj2$hessian(P)
+mean(abs(hs2-true.hess))/mean(abs(hs2))
+@
+
+<<echo=FALSE>>=
+options(scipen=999)
+@
+
+In short, the complex step method can be more accurate than
+finite differencing, but it comes with theoretical and implementation
+restrictions that may limit its universality.
 
@@
 algorithms are adapted from \citet{ColemanGarbow1985}, who provided
-Fortran implementations as \citet{ColemanGarbow1985b}.  Earlier
+\proglang{Fortran} implementations as \citet{ColemanGarbow1985b}.  Earlier
 versions of \pkg{sparseHessianFD} included licensed copies of the
@@
 As far as we know, \pkg{numDeriv} \citep{R_numDeriv} is the only other \proglang{R} package
-that computes numerical approximations to derivatives.  It differs from
-\pkg{sparseHessianFD} in some important ways.
-
-\begin{enumerate}
-\item It treats all Hessians as dense;
-  \item It computes each element of the Hessian using a second-order
-    finite differencing approximation that does not require the user
-    to supply the gradient; and
-    \item It implements iterative algorithms to improve accuracy, at
-      the expense of speed.
-\end{enumerate}
-
-Nevertheless, \pkg{numDeriv} is an easy-to-use tool for numerical differentiation,
-so it is worthwhile to compare its performance to that of
-\pkg{sparseHessianFD}.  To prepare Table~\ref{tab:numDeriv}, we
+that computes numerical approximations to derivatives.  Like
+\pkg{sparseHessianFD} it includes functions to compute Hessians from
+user-supplied gradients (through the \func{jacobian} function), and
+implements both the finite differencing and complex step methods.  Its
+most important distinction from \pkg{sparseHessianFD} is that it treats all Hessians as dense.  Thus, we
+will use  \pkg{numDeriv} as the baseline against which we can compare the performance of
+\pkg{sparseHessianFD}.
+
+To prepare Table~\ref{tab:numDeriv}, we
 estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
-units ($N$) and within-unit parameters ($k$).  As in the previous
-section, the total number of
+units ($N$) and within-unit parameters ($k$).  The total number of
 variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
-and standard deviations (across 20 replications) for the time (in milliseconds) to
-to compute a Hessian using each package.  The difference in run times is dramatic,
-especially because the computation time for \pkg{numDeriv} grows
-quadratically in the number of variables.  The setup time for
-\pkg{sparseHessianFD} was about 7 milliseconds for all cases.  Times
-were collected using the \pkg{microbenchmark} package \citep{R_microbenchmark}.
+and standard deviations (across 500 replications) for the time (in
+milliseconds) to compute a Hessian using functions for both the finite
+difference and complex step methods from each package. Times were generated on a
+compute node running Scientific Linux 6 (64-bit) with an 8-core Intel
+Xeon X5560 processor (2.80 GHz) with 24 GB of RAM, and collected using the \pkg{microbenchmark} package
+\citep{R_microbenchmark}. Code to replicate Table~\ref{tab:numDeriv}
+is available as an online supplement to this paper, and in the
+\func{doc/} directory of the installed package. In
+Table~\ref{tab:numDeriv} we see that computation times using
+\pkg{sparseHessianFD} and considerably shorter than those using \pkg{numDeriv}.
+
+<<echo=FALSE>>=
+suppressMessages(library(xtable))
+suppressMessages(library(dplyr))
+load("vignette_tab4.Rdata")
+@
 
 \begin{table}[ptb]\centering
-  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }cccp{2pt}rrrrp{2pt}rrrr}
+  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }rrr|rrrr|rrrr}
   \toprule
-&&&&\multicolumn{4}{c}{\bfseries{Hessian
-    time}}&&\multicolumn{4}{c}{\bfseries{Hessian/Gradient ratio}}\\
-&&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
-$N$&$k$&$M$&&mean&sd&mean&sd&&mean&sd&mean&sd\\
-  \cmidrule{1-3}\cmidrule{5-8}\cmidrule{10-13}
-   9 &    2 &   20 &&  363 & 6.4 &  2.8 & 2.2 &&  871 & 63 & 6.5 & 4.2 \\
-    12 &    2 &   26 &&  629 &  8.9 &  2.9 & 2.1 && 1469 &  124 & 6.5 & 3.3 \\
-     9 &    3 &   30 &&  819 & 27.7 &  3.6 & 2.2 && 1974 &  161 & 8.4 & 3.7 \\
-    15 &    2 &   32 &&  973 & 25.8 &    3.0 & 2.2 && 2191 &  209 &  6.5 & 3.4 \\
-    12 &    3 &   39 && 1407 & 39.3 & 3.5 & 2.0 && 3291 &  285 &    8.0 & 3.5 \\
-     9 &    4 &   40 && 1445 & 28.8 & 4.3 & 2.1 && 3453 &  292 & 9.9 & 3.1 \\
-    15 &    3 &   48 && 2138 & 92.7 & 3.5 & 2.0 && 4861 &  341 & 7.8 & 3.4 \\
-    12 &    4 &   52 && 2400 & 98.6 & 4.3 & 2.1 && 5679 &  503 & 10.1 & 3.7 \\
-    15 &    4 &   64 && 3643 &  209.0 & 4.2 & 2.2 && 8945 &  828 & 10.0 &  2.8 \\
-   \bottomrule
+&&&\multicolumn{4}{c|}{\bfseries{Finite differencing}}&\multicolumn{4}{c}{\bfseries{Complex step}}\\
+&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c|}{\pkg{sparseHessianFD}}&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
+$N$&$k$&$M$&mean&sd&mean&sd&mean&sd&mean&sd\\
+%%\midrule
+<<results='asis',echo=FALSE>>=
+arrange(tab4, M) %>%
+xtable(digits=c(rep(0,4),rep(1,8))) %>%
+    print(only.contents=TRUE, comment=FALSE, booktabs=TRUE,
+      include.rownames=FALSE, include.colnames=FALSE)
+@
 \end{tabularx}
 \caption{Computation times (milliseconds) for computing Hessians using the
-  \pkg{numDeriv} and \pkg{sparseHessianFD} packages, across 200 replications. Rows are ordered by the
+  \pkg{numDeriv} and \pkg{sparseHessianFD} packages, and the finite
+ difference and complex step methods, across 500 replications. Rows are ordered by the
   number of variables.}\label{tab:numDeriv}
@@
 In the plots in Figure~\ref{fig:timing}, the number of heterogeneous units ($N$) is
-on the x-axis, and mean run time, in milliseconds, is on the y-axis.  Each panel shows the
-relationship between $N$ and run time for a different measure of time,
+on the x-axis, and median run time, in milliseconds, is on the y-axis.  Each panel shows the
+relationship between $N$ and run time for a different step in the algorithm,
 and each curve in a panel represents a different number of within-unit

### 2017-01-24 nobuild/jss/sparseHessianFD.tex
cleaning up

@@
-\documentclass[article]{jss}\usepackage[]{graphicx}\usepackage[]{color}
+ \documentclass[jss]{jss}\usepackage[]{graphicx}\usepackage[]{color}
 %% maxwidth is the original width if it is less than linewidth
@@
   symbolic or algorithmic methods are impractical, inefficient or
-    unavailable. }
-
-\date{March 9, 2016}
+  unavailable.}
+
+\date{January 25, 2017}
 
@@
 
-All of the $\beta_i$ are correlated with
-$\mu$.  Thus, $\hess{\beta_{ik},\mu_k}{}\neq 0$ for all $i$. Since the $\beta_i$ are independently distributed, and the $y_i$ are
+\subsection[Sparsity patterns]{Sparsity patterns}\label{sec:sparsity}
+
+Let $x_1$ and $x_2$ be two subsets of elements of $x$.
+Define $\parD{2}{x_1,x_2}$ as the product set of cross-partial
+derivatives between all elements in $x_1$ and all elements in
+$x_2$. From the log posterior density in Equation~\ref{eq:LPD}, we can see
+that $\parD{2}{\beta_i, \beta_i}\neq 0$ (one element of $\beta_i$ could be
+correlated with another element of $\beta_i$), and that, for all $i$,
+$\parD{2}{\beta_i, \mu}\neq 0$ (because $\mu$ is the prior mean of each
+$\beta_i$).  However, since the $\beta_i$ and $\beta_j$ are independently distributed, and the $y_i$ are
 conditionally independent, the cross-partial derivatives
-$\hess{\beta_i,\beta_j}{}=\parD{2}{\beta_i, \beta_j}{}=0$ for all $i\neq
-j$.  When $N$ is much greater than $k$, the Hessian of the log
-posterior density will be sparse.
-
-\subsection[Sparsity patterns]{Sparsity patterns}\label{sec:sparsity}
-
-The sparsity pattern depends on how the variables are
-ordered within the vector. One such ordering is to group all of the
-coefficients for each unit together.
+$\parD{2}{\beta_i, \beta_j}=0$ for all $i\neq j$. When $N$ is much
+greater than $k$, there will be many more zero cross-partial
+derivatives than non-zero.   Each $\parD{2}{}$ is
+mapped to a submatrix of $\hess{}{}$, most of which with be zero.  The
+resulting Hessian of the log posterior density will be sparse.
+
+The sparsity pattern depends on the indexing function; that is, on how the variables are
+ordered in $x$. One such ordering is to group all of the
+coefficients in the $\beta_i$ for each unit together.
 \begin{align}
@@
 
+If there is uncertainty about whether an element is a structural
+   zero or not, one should err on the side of it being non-zero, and
+   include that element in the sparsity pattern.
+   There might be a loss of efficiency if the element really is a
+   structural zero, but the result will still be correct.  All that
+   would happen is that the numerical estimate for that element would be zero
+   (within machine precision). On the other hand, excluding a non-zero
+   element from the sparsity pattern will likely lead to an incorrect
+   estimate of the Hessian.
+
 \subsection[The sparseHessianFD class]{The \class{sparseHessianFD} class}
@@
 
-\subsection[An example]{An example}
+\subsection[An example]{An example}\label{sec:exHess}
 
 Now we can estimate the Hessian for the
-log posterior density of the model from Section \ref{sec:example}.
+log posterior density of the model from Section~\ref{sec:example}.
 For demonstration purposes, \pkg{sparseHessianFD} includes functions that compute the value
@@
 R> T <- binary[["T"]]
-R> nvars <- as.integer(N*k + k)
-R> priors <- list(inv.Sigma = rWishart(1,k+5,diag(k))[,,1],
+R> nvars <- as.integer(N * k + k)
+R> priors <- list(inv.Sigma = rWishart(1, k + 5, diag(k))[,,1],
 +    inv.Omega = diag(k))
@@
 
-Finally, we create an instance of a \class{sparseHessianFD} object,
-evaluate the function, gradient and Hessian, and compare the output to
-the true values.
+Finally, we create an instance of a \class{sparseHessianFD} object.
+Evaluations of the function and gradient using the \method{fn} and
+\method{gr} methods will always give the same
+results as the true values because they are computed using the same functions.
 
@@
 
+The default choice of method is \funcarg{complex=FALSE}, so the
+evaluation of the Hessian is a finite differenced approximation,
+so it is very close to, but not identical to, the true value, in terms
+of mean relative difference.
+
+\begin{Schunk}
+\begin{Sinput}
+R> hs <- obj$hessian(P)
+R> mean(abs(hs-true.hess))/mean(abs(hs))
+\end{Sinput}
+\begin{Soutput}
+[1] 2.3357e-09
+\end{Soutput}
+\end{Schunk}
+
+If \funcarg{complex=TRUE} in the initializer, the call to the \method{hessian} method will
+apply the complex step method. To use this method, the functions
+passed as \funcarg{fn} and \funcarg{gr} must both
+accept a complex argument, and return a complex
+result, even though we are differentiating a real-valued function. Although \pkg{base} \proglang{R} supports complex arguments
+for most basic mathematical functions, many common functions (e.g.,  \func{gamma}, \func{log1p},
+\func{expm1}, and the probability distribution functions) do not have
+complex implementations. Furthermore, the complex step method is valid
+only if the function is holomorphic. The methods in \class{sparseHessianFD} do
+\emph{not} check that this is the case for the function at hand. We convey the
+following warning from the documentation of the \pkg{numDeriv} package
+\citep{R_numDeriv}, which also implements the complex step method:
+``avoid this method if you do not know that your function is
+suitable. Your mistake may not be caught and the results will be
+spurious.''
+
+Fortunately for demonstration purposes, the log posterior density in
+Equation~\ref{eq:LPD} is holomorphic, so we
+can estimate its Hessian using the complex step method, and compute
+the mean relative difference from the true Hessian.
+
+\begin{Schunk}
+\begin{Sinput}
+R> obj2 <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
++    rows=pattern[["rows"]], cols=pattern[["cols"]],
++    complex=TRUE,
++    data=binary, priors=priors, order.row=order.row)
+R> hs2 <- obj2$hessian(P)
+R> mean(abs(hs2-true.hess))/mean(abs(hs2))
+\end{Sinput}
+\begin{Soutput}
+[1] 8.0555e-17
+\end{Soutput}
+\end{Schunk}
+
+In short, the complex step method can be more accurate than
+finite differencing, but it comes with theoretical and implementation
+restrictions that may limit its universality.
+
 \section[Algorithms]{Algorithms}\label{sec:algorithms}
@@
 As far as we know, \pkg{numDeriv} \citep{R_numDeriv} is the only other \proglang{R} package
-that computes numerical approximations to derivatives.  It differs from
-\pkg{sparseHessianFD} in some important ways.
-
-\begin{enumerate}
-\item It treats all Hessians as dense;
-  \item It computes each element of the Hessian using a second-order
-    finite differencing approximation that does not require the user
-    to supply the gradient; and
-    \item It implements iterative algorithms to improve accuracy, at
-      the expense of speed.
-\end{enumerate}
-
-Nevertheless, \pkg{numDeriv} is an easy-to-use tool for numerical differentiation,
-so it is worthwhile to compare its performance to that of
-\pkg{sparseHessianFD}.  To prepare Table~\ref{tab:numDeriv}, we
+that computes numerical approximations to derivatives.  Like
+\pkg{sparseHessianFD} it includes functions to compute Hessians from
+user-supplied gradients (through the \func{jacobian} function), and
+implements both the finite differencing and complex step methods.  Its
+most important distinction from \pkg{sparseHessianFD} is that it treats all Hessians as dense.  Thus, we
+will use  \pkg{numDeriv} as the baseline against which we can compare the performance of
+\pkg{sparseHessianFD}.
+
+To prepare Table~\ref{tab:numDeriv}, we
 estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
-units ($N$) and within-unit parameters ($k$).  As in the previous
-section, the total number of
+units ($N$) and within-unit parameters ($k$).  The total number of
 variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
-and standard deviations (across 20 replications) for the time (in milliseconds) to
-to compute a Hessian using each package.  The difference in run times is dramatic,
-especially because the computation time for \pkg{numDeriv} grows
-quadratically in the number of variables.  The setup time for
-\pkg{sparseHessianFD} was about 7 milliseconds for all cases.  Times
-were collected using the \pkg{microbenchmark} package
-\citep{R_microbenchmark}, using a 2013-edition Mac Pro with 12
-processing cores and 64GB of RAM.
+and standard deviations (across 500 replications) for the time (in
+milliseconds) to compute a Hessian using functions for both the finite
+difference and complex step methods from each package. Times were generated on a
+compute node running Scientific Linux 6 (64-bit) with an 8-core Intel
+Xeon X5560 processor (2.80 GHz) with 24 GB of RAM, and collected using the \pkg{microbenchmark} package
+\citep{R_microbenchmark}. Code to replicate Table~\ref{tab:numDeriv}
+is available as an online supplement to this paper, and in the
+\func{doc/} directory of the installed package. In
+Table~\ref{tab:numDeriv} we see that computation times using
+\pkg{sparseHessianFD} and considerably shorter than those using \pkg{numDeriv}.
 
 \begin{table}[ptb]\centering
-  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }cccp{2pt}rrrrp{2pt}rrrr}
+  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }rrr|rrrr|rrrr}
   \toprule
-&&&&\multicolumn{4}{c}{\bfseries{Hessian
-    time}}&&\multicolumn{4}{c}{\bfseries{Hessian/Gradient ratio}}\\
-&&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
-$N$&$k$&$M$&&mean&sd&mean&sd&&mean&sd&mean&sd\\
-  \cmidrule{1-3}\cmidrule{5-8}\cmidrule{10-13}
-   9 &    2 &   20 &&  363 & 6.4 &  2.8 & 2.2 &&  871 & 63 & 6.5 & 4.2 \\
-    12 &    2 &   26 &&  629 &  8.9 &  2.9 & 2.1 && 1469 &  124 & 6.5 & 3.3 \\
-     9 &    3 &   30 &&  819 & 27.7 &  3.6 & 2.2 && 1974 &  161 & 8.4 & 3.7 \\
-    15 &    2 &   32 &&  973 & 25.8 &    3.0 & 2.2 && 2191 &  209 &  6.5 & 3.4 \\
-    12 &    3 &   39 && 1407 & 39.3 & 3.5 & 2.0 && 3291 &  285 &    8.0 & 3.5 \\
-     9 &    4 &   40 && 1445 & 28.8 & 4.3 & 2.1 && 3453 &  292 & 9.9 & 3.1 \\
-    15 &    3 &   48 && 2138 & 92.7 & 3.5 & 2.0 && 4861 &  341 & 7.8 & 3.4 \\
-    12 &    4 &   52 && 2400 & 98.6 & 4.3 & 2.1 && 5679 &  503 & 10.1 & 3.7 \\
-    15 &    4 &   64 && 3643 &  209.0 & 4.2 & 2.2 && 8945 &  828 & 10.0 &  2.8 \\
+&&&\multicolumn{4}{c|}{\bfseries{Finite differencing}}&\multicolumn{4}{c}{\bfseries{Complex step}}\\
+&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c|}{\pkg{sparseHessianFD}}&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
+$N$&$k$&$M$&mean&sd&mean&sd&mean&sd&mean&sd\\
+%%\midrule
+  \midrule
+15 & 2 & 32 & 12.7 & 0.9 & 2.4 & 0.1 & 13.2 & 0.7 & 2.0 & 0.1 \\ 
+  15 & 5 & 80 & 32.2 & 0.9 & 5.0 & 0.5 & 36.1 & 10.1 & 5.2 & 0.8 \\ 
+  50 & 2 & 102 & 51.3 & 9.8 & 3.0 & 0.4 & 54.5 & 1.0 & 2.7 & 0.6 \\ 
+  15 & 8 & 128 & 53.6 & 0.9 & 7.8 & 0.6 & 62.5 & 16.7 & 8.2 & 0.8 \\ 
+  100 & 2 & 202 & 131.7 & 16.9 & 3.7 & 0.3 & 147.6 & 24.1 & 3.4 & 0.5 \\ 
+  50 & 5 & 255 & 144.0 & 27.1 & 6.6 & 0.6 & 162.8 & 29.2 & 6.7 & 0.7 \\ 
+  50 & 8 & 408 & 256.2 & 43.5 & 12.5 & 12.9 & 291.1 & 39.0 & 11.7 & 0.7 \\ 
+  100 & 5 & 505 & 495.7 & 129.5 & 9.8 & 3.1 & 559.8 & 125.9 & 11.8 & 10.4 \\ 
+  100 & 8 & 808 & 850.0 & 171.4 & 20.9 & 20.4 & 999.2 & 217.5 & 23.6 & 17.3 \\ 
+  500 & 2 & 1002 & 1953.4 & 321.7 & 10.8 & 3.7 & 2156.2 & 339.7 & 9.5 & 3.4 \\ 
+  500 & 5 & 2505 & 5444.0 & 951.0 & 27.1 & 10.6 & 6159.3 & 1009.2 & 26.5 & 4.6 \\ 
+  500 & 8 & 4008 & 9533.1 & 190.9 & 54.5 & 12.8 & 14059.4 & 3797.9 & 62.5 & 21.8 \\ 
    \bottomrule
+
 \end{tabularx}
 \caption{Computation times (milliseconds) for computing Hessians using the
-  \pkg{numDeriv} and \pkg{sparseHessianFD} packages, across 200 replications. Rows are ordered by the
+  \pkg{numDeriv} and \pkg{sparseHessianFD} packages, and the finite
+ difference and complex step methods, across 500 replications. Rows are ordered by the
   number of variables.}\label{tab:numDeriv}
@@
 
-It should not be surprising that run times
+Computation times
 for the function and gradient, as well as the setup and
 partitioning times for the \class{sparseHessianFD} object, grow
-linearly with the number of heterogenous units.  The run time
+linearly with the number of heterogenous units.  The time
 for the Hessian grows linearly as well, and that might be partially
-surprising.  We saw in Section \ref{sec:coloring} that adding
+surprising.  We saw in Section~\ref{sec:coloring} that adding
 additional heterogeneous units in a hierarchical model does not

### 2017-01-24 nobuild/jss/numDeriv.R

@@ -1,122 +1,0 @@
-rm(list=ls())
-gc()
-
-library(sparseHessianFD)
-library(mvtnorm)
-library(plyr)
-library(microbenchmark)
-library(doParallel)
-library(numDeriv)
-
-run.par <- TRUE
-if (run.par) registerDoParallel(cores=12) else registerDoParallel(cores=1)
-
-binary_sim <- function(N, k, T) {
-
-    x.mean <- rep(0,k)
-    x.cov <- diag(k)
-    x.cov[1,1] <- .02
-    x.cov[k,k] <- x.cov[1,1]
-    mu <- seq(-2,2,length=k)
-    Omega <- diag(k)
-
-    X <- t(rmvnorm(N, mean=x.mean, sigma=x.cov)) ## k x N
-    B <- t(rmvnorm(N, mean=mu, sigma=Omega)) ## k x N
-    XB <- colSums(X * B)
-    log.p <- XB - log1p(exp(XB))
-    Y <- sapply(log.p, function(q) return(rbinom(1,T,exp(q))))
-
-    list(Y=Y, X=X, T=T)
-}
-
-priors_sim <- function(k) {
-
-    list(inv.Omega = solve(diag(k)),
-         inv.Sigma = rWishart(1,k+5,diag(k))[,,1])
-}
-
-
-make_funcs <- function(D, priors) {
-    res <- vector("list", length=3)
-    names(res) <- c("fn", "gr", "hessian")
-    res$fn <-  function(pars) {
-        binary.f(pars, data=D, priors=priors)
-    }
-    res$gr <-  function(pars) {
-        binary.grad(pars, data=D, priors=priors)
-    }
-    res$hessian <-  function(pars) {
-        binary.hess(pars, data=D, priors=priors)
-    }
-    return(res)
-}
-
-make_perm <- function(iRow, jCol) {
-    tmp <- sparseMatrix(i=iRow, j=jCol, index1=TRUE, symmetric=TRUE)
-    perm <- order(Matrix::rowSums(tmp), decreasing=TRUE)
-    invperm <- invPerm(perm)
-    L <- tril(tmp[perm,perm])
-    ptr <- Matrix.to.Pointers(L, order="row", index1=TRUE)
-    idx <- ptr[[1]]
-    pntr <- ptr[[2]]
-}
-
-get_id <- function(x) 1:length(x)
-
-run_test <- function(Nk, reps=50) {
-
-    N <- as.numeric(Nk["N"])
-    k <- as.numeric(Nk["k"])
-
[…]

### 2017-01-24 nobuild/jss/summarize_timings.R

@@ -1,52 +1,0 @@
-library(ggplot2)
-library(plyr)
-library(dplyr)
-library(tidyr)
-library(reshape2)
-
-theme_set(theme_bw())
-
-data.file <- "inst/examples/timings.Rdata"
-load(data.file)
-
-tab <- mutate(res, ms=bench.time/1000000) %>%
-  dcast(N+k+T+ord+bench.rep+ncolors~bench.expr, value.var="ms")  %>%
-  mutate(nvars=N*k+k, hess_f=hess/f, hess_df=hess/df,
-         hess_k=hess/k, hess_ncol = hess/ncolors, total=setup+hess,
-         total_f=total/f, total_df = total/df, df/nvars)
-
-tab2 <- gather(tab, stat, ms, c(setup:hess,hess_f:total_df)) %>%
-  group_by(N, k, T, ord, stat, nvars) %>%
-  summarize(mean=mean(ms), sd=sd(ms), lower=mean-1.96*sd, upper=mean+1.96*sd)
-
-
-D2 <- subset(data.frame(tab2), stat %in% c("f", "df", "hess",
-                                           "colors", "setup","hess_df"))
-D2$stat <- revalue(D2$stat, c("f"="Function", "df"="Gradient", "hess"="Hessian",
-                              "colors"="Partitioning",
-                              "setup"="Initialization",
-                              "hess_df"="Hessian/Gradient"))
-D2$stat <- relevel(D2$stat, "Hessian/Gradient") %>%
-  relevel("Initialization") %>%
-  relevel("Partitioning") %>%
-  relevel("Hessian") %>%
-  relevel("Gradient") %>%
-  relevel("Function")
-
-
-
-P2 <- ggplot(D2, aes(x=N,y=mean, color=as.factor(k), linetype=as.factor(k))) %>%
-  + geom_line(size=.4) %>%
-  + scale_x_continuous("Number of heterogeneous units") %>%
-  + scale_y_continuous("Computation time (milliseconds)") %>%
-  + guides(color=guide_legend("k"), linetype=guide_legend("k")) %>%
-  + facet_wrap(~stat, scales="free") %>%
-  + theme(text=element_text(size=8), legend.position="right")
-
-
-pdf(file="vignettes/timings.pdf", width=6, height=4)
-print(P2)
-dev.off()
-
-
-

### 2017-01-24 nobuild/jss_revision/replication.R

@@ -1,200 +1,0 @@
-library("sparseHessianFD")
-library("Matrix")
-library("mvtnorm")
-library("microbenchmark")
-library("dplyr")
-library("tidyr")
-library("numDeriv")
-library("reshape2")
-library("ggplot2")
-
-set.seed(1234)
-
-binary_sim <- function(N, k, T) {
-  x.mean <- rep(0,k)
-  x.cov <- diag(k)
-  x.cov[1,1] <- .02
-  x.cov[k,k] <- x.cov[1,1]
-  mu <- seq(-2,2,length=k)
-  Omega <- diag(k)
-  X <- t(rmvnorm(N, mean=x.mean, sigma=x.cov)) ## k x N
-  B <- t(rmvnorm(N, mean=mu, sigma=Omega)) ## k x N
-  XB <- colSums(X * B)
-  log.p <- XB - log1p(exp(XB))
-  Y <- sapply(log.p, function(q) return(rbinom(1,T,exp(q))))
-  list(Y=Y, X=X, T=T)
-}
-
-priors_sim <- function(k) {
-  list(inv.Omega = solve(diag(k)),
-       inv.Sigma = rWishart(1,k+5,diag(k))[,,1])
-}
-
-
-make_funcs <- function(D, priors, order.row=FALSE) {
-  res <- vector("list", length=3)
-  names(res) <- c("fn", "gr", "hessian")
-  res$fn <-  function(pars) binary.f(pars, data=D, priors=priors,
-                                     order.row=order.row)
-  res$gr <-  function(pars) binary.grad(pars, data=D, priors=priors,
-                                        order.row=order.row)
-  res$hessian <-  function(pars) binary.hess(pars, data=D, priors=priors,
-                                             order.row=order.row)
-  return(res)
-}
-
-make_perm <- function(iRow, jCol) {
-  tmp <- sparseMatrix(i=iRow, j=jCol, index1=TRUE, symmetric=TRUE)
-  perm <- order(Matrix::rowSums(tmp), decreasing=TRUE)
-  invperm <- invPerm(perm)
-  L <- tril(tmp[perm,perm])
-  ptr <- Matrix.to.Pointers(L, order="row", index1=TRUE)
-  idx <- ptr[[1]]
-  pntr <- ptr[[2]]
-}
-
-get_id <- function(x) 1:length(x)
-
-run_test_fig4 <- function(NkT, reps=50, order.row=FALSE) {
-  ## Replication function for Figure 4
-  N <- as.numeric(NkT["N"])
[…]

### 2017-01-25 vignettes/replication.R
merging

@@ -80,7 +80,7 @@ run_test_fig4 <- function(NkT, reps=50, order.row=FALSE) {
 
   obj <- sparseHessianFD(X, F$fn, F$gr, pat$rows, pat$cols, complex=TRUE)
   colors <- obj$partition()
-  perm <- obj$get_perm()
+  perm <- obj$get_perm()S
   ncolors <- length(unique(colors))
   nvars <- N*k+k
 

### 2018-03-26 nobuild/jss/2717/paper.tex

--- 2017-01-24 nobuild/jss/sparseHessianFD.tex
@@
- \documentclass[jss]{jss}\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\usepackage{Sweave}
-
-%\VignetteEngine{knitr::knitr}
-%\VignetteIndexEntry{sparseHessianFD}
-%\VignettePackage{sparseHessianFD}
+\documentclass[article]{jss}
+\usepackage{thumbpdf} 
+\graphicspath{{Figures/}}
 
@@
 \usepackage{amsfonts}
-\usepackage{amsthm}
 \usepackage{subcaption}
@@
 \usepackage{tabularx}
-\usepackage{dcolumn}
-\usepackage{booktabs}
-\usepackage{etoolbox}
-\usepackage{placeins}
-\newtoggle{tikz}
-\toggletrue{tikz}
-
-\iftoggle{tikz}{
 \usepackage{tikz}
-}{}
-
-   \newcolumntype{d}[1]{D{.}{.}{#1}}
-%%\newcommand{\pkg}[1]{\emph{#1}}
-%%\newcommand{\proglang}[1]{\textsf{#1}}
-%%\newcommand{\filename}[1]{\textit{#1}}
-%%\newcommand{\code}[1]{\texttt{#1}}
-\newcommand{\func}[1]{\code{#1}}
-\newcommand{\class}[1]{\textsl{#1}}
+
+\newcommand{\func}[1]{\code{#1()}}
+\newcommand{\class}[1]{`\code{#1}'}
 \newcommand{\funcarg}[1]{\code{#1}}
@@
 
- \newcommand{\df}[3]{\mathsf{d}^{#1}f(#2;#3)}
 \newcommand{\parD}[3]{\mathsf{D}^{#1}_{#2}#3}
@@
 
-\author{Michael Braun\\Edwin L. Cox School of Business\\Southern Methodist University}
+\author{Michael Braun\\Southern Methodist University}
 \Plainauthor{Michael Braun}
-\title{\pkg{sparseHessianFD}: An R Package for Estimating Sparse Hessian Matrices}
+
+\title{\pkg{sparseHessianFD}: An \proglang{R} Package for Estimating Sparse Hessian Matrices}
+
 \Plaintitle{sparseHessianFD: An R Package for Estimating Sparse Hessian Matrices}
-\Shorttitle{sparseHessianFD:  Sparse Hessians}
+
 \Keywords{sparse Hessians, sparsity, computation of Hessians,
   graph coloring, finite differences, differentiation, complex step}
+
 \Address{
@@
   algorithms. By exploiting the known sparsity pattern of a Hessian, methods in the
-  sparseHessianFD package require many fewer function or gradient
+  \pkg{sparseHessianFD} package require many fewer function or gradient
   evaluations than would be required if the Hessian were treated as
@@
   linear substitution algorithms that were previously unavailable to
-  R users, and is most useful when other numerical,
+  \proglang{R} users, and is most useful when other numerical,
   symbolic or algorithmic methods are impractical, inefficient or
@@
 
-\date{January 25, 2017}
-
-%% need no \usepackage{Sweave.sty}
-
-%%\newenvironment{knitrout}{\begin{figure}}{\end{figure}}
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 The Hessian matrix of a log likelihood function or log posterior
-density function plays
-an important role in statistics.  From a frequentist point of view,
-the inverse of the negative Hessian is the asymptotic covariance of
-the sampling distribution of a maximum
+density function plays an important role in statistics.  From a
+frequentist point of view, the inverse of the negative Hessian is the
+asymptotic covariance of the sampling distribution of a maximum
 likelihood estimator.  In Bayesian analysis, when evaluated at the
@@
 applications with hundreds of thousands of variables, computing the
-Hessian even once might not be practical under time, storage or
+Hessian even once might not be practical under the given time, storage or
 processor constraints.  Hierarchical models, in
@@
 
-However, for many problems, the Hessian is
-\emph{sparse}, meaning that the proportion of  ``structural''  zeros
-(matrix elements that are always zero, regardless of the value at which
-function is estimated) is high.  Consider a log
-posterior density in a Bayesian hierarchical
-model.  If the outcomes across units are conditionally
+However, for many problems, the Hessian is \emph{sparse}, meaning that
+the proportion of ``structural'' zeros (matrix elements that are
+always zero, regardless of the value at which the function is
+estimated) is high.  Consider a log posterior density in a Bayesian
+hierarchical model.  If the outcomes across units are conditionally
 independent, the cross-partial derivatives of heterogeneous variables
-across units are zero.  As the number of units
-increases, the size of the Hessian still grows quadratically, but the number
-of \emph{non-zero} elements grows only linearly; the Hessian
-becomes increasingly sparse.  The row and column indices of the
-non-zero elements comprise the \emph{sparsity pattern} of the Hessian,
-and are typically known in advance, before computing the values of
-those elements.  \proglang{R} packages such as \pkg{trustOptim}
+across units are zero.  As the number of units increases, the size of
+the Hessian still grows quadratically, but the number of
+\emph{non-zero} elements grows only linearly; the Hessian becomes
+increasingly sparse.  The row and column indices of the non-zero
+elements comprise the \emph{sparsity pattern} of the Hessian, and are
+typically known in advance, before computing the values of those
+elements.  \proglang{R} \citep{R} packages such as \pkg{trustOptim}
 \citep{R_trustOptim}, \pkg{sparseMVN} \citep{R_sparseMVN} and
@@
 
-The \pkg{sparseHessianFD} package is a tool for estimating
-sparse Hessians numerically, using either finite differences or
-complex perturbations of
-gradients.  Section~\ref{sec:numdiff} will
-cover the specifics, but the basic idea is as follows.  Consider a real-valued
-function $f(x)$, its gradient $\nabla{f(x)}$, and its Hessian
-$\hess{}{f(x)}$, for $x\in\mathbb{R}^M$.  Define the derivative vector
-as the transpose of the gradient, and a vector of partial derivatives, so
-${\parD{}{}{f(x)}=\nabla f(x)^\top=\left(\parD{}{1}{},\mathellipsis,\parD{}{M}{}\right)}$.   (Throughout the paper, we will try
-to reduce notational clutter by referring to the derivative and
-Hessian as $\parD{}{}{}$ and $\hess{}{}$, respectively, without the
-$f(x)$ symbol).  Let $e_m$ be a vector of zeros, except with a 1 in the $m$th
-element, and let $\delta$ be a sufficiently small scalar
-constant. A ``finite difference'' linear approximation to the $m$th column of the Hessian is ${\hess{m}{}\approx\left(\nabla f(x+\delta
-  e_m)- \nabla f(x)\right)/\delta}$.  Estimating a dense Hessian in this way involves at least $M+1$ calculations of the
-gradient: one for the gradient at $x$, and one after perturbing each
-of the $M$ elements of $x$, one at a time. Under certain conditions, a
-more accurate approximation is the ``complex step'' method:  $\hess{m}{}\approx\Imag\left(\nabla f(x+i\delta
-  e_m)\right)/\delta$, where $i=\sqrt{-1}$ and $\Imag$ returns the
-imaginary part of a complex
-number \citep{SquireTrapp1998}.  Regardless of the approximation method used, if the Hessian is sparse, most of the
-elements are constrained to zero.  Depending
-on the sparsity pattern of the Hessian, those constraints may let us recover
-the Hessian with fewer gradient evaluations by perturbing multiple
-elements of $x$ together.  For some sparsity patterns, estimating a Hessian in
-this way can be profoundly efficient.  In fact, for the hierarchical models that we consider in
-this paper, the number of gradient evaluations does not increase with
-the number of additional heterogeneous units.
+The \pkg{sparseHessianFD} package \citep{sparseHessianFD} is a tool
+for estimating sparse Hessians numerically, using either finite
+differences or complex perturbations of gradients. The package is
+available from the Comprehensive \proglang{R} Archive Network (CRAN)
+at \url{https://CRAN.R-project.org/package=sparseHessianFD}.
+Section~\ref{sec:numdiff} will cover the specifics, but the basic idea
+is as follows.  Consider a real-valued function $f(x)$, its gradient
+$\nabla{f(x)}$, and its Hessian $\hess{}{f(x)}$, for
+$x\in\mathbb{R}^M$.  Define the derivative vector as the transpose of
+the gradient, and a vector of partial derivatives, so
+${\parD{}{}{f(x)}=\nabla
+  f(x)^\top=\left(\parD{}{1}{},\mathellipsis,\parD{}{M}{}\right)}$.
+(Throughout the paper, we will try to reduce notational clutter by
+referring to the derivative and Hessian as $\parD{}{}{}$ and
+$\hess{}{}$, respectively, without the $f(x)$ symbol).  Let $e_m$ be a
+vector of zeros, except with a 1 in the $m$th element, and let
+$\delta$ be a sufficiently small scalar constant. A ``finite
+difference'' linear approximation to the $m$th column of the Hessian
+is
+${\hess{m}{}\approx\left(\nabla f(x+\delta e_m)- \nabla
+    f(x)\right)/\delta}$.  Estimating a dense Hessian in this way
+involves at least $M+1$ calculations of the gradient: one for the
+gradient at $x$, and one after perturbing each of the $M$ elements of
+$x$, one at a time. Under certain conditions, a more accurate
+approximation is the ``complex step'' method:
+$\hess{m}{}\approx\Imag\left(\nabla f(x+i\delta e_m)\right)/\delta$,
+where $i=\sqrt{-1}$ and $\Imag()$ returns the imaginary part of a
+complex number \citep{SquireTrapp1998}.  Regardless of the
+approximation method used, if the Hessian is sparse, most of the
+elements are constrained to zero.  Depending on the sparsity pattern
+of the Hessian, those constraints may let us recover the Hessian with
+fewer gradient evaluations by perturbing multiple elements of $x$
+together.  For some sparsity patterns, estimating a Hessian in this
+way can be profoundly efficient.  In fact, for the hierarchical models
+that we consider in this paper, the number of gradient evaluations
+does not increase with the number of additional heterogeneous units.
 
@@
 As with any computing method or algorithm, there are boundaries around
-the space of applications for which \pkg{sparseHessianFD} is the right
-tool for the job. In general, numerical approximations are not ``first
-choice'' methods because the result is not exact, so
-\pkg{sparseHessianFD} should not be used when the application cannot
-tolerate any error, no matter how small.  Also, we admit that some
-users might balk at having to provide an exact gradient, even though
-the Hessian will be estimated numerically.\footnote{It is possible
-  that a gradient that is approximated using a complex step method
-  would be precise enough to be useful. A further investigation is
-  beyond the scope of this paper.} However, deriving a vector of first
-derivatives, and writing \proglang{R} functions to compute them, is a
-lot easier than doing the same for a matrix of second derivatives, and
-more accurate than computing second-order approximations from the
-objective function. Even when we have derived the Hessian
-symbolically, in practice it may still be faster to estimate the Hessian
-using \pkg{sparseHessianFD} than coding it directly.  These are the
-situations in which \pkg{sparseHessianFD} adds the most value to the
-statistician's toolbox.
+the space of applications for which package \pkg{sparseHessianFD} is
+the right tool for the job. In general, numerical approximations are
+not ``first choice'' methods because the result is not exact, so
+package \pkg{sparseHessianFD} should not be used when the application
+cannot tolerate any error, no matter how small.  Also, we admit that
+some users might balk at having to provide an exact gradient, even
+though the Hessian will be estimated numerically.\footnote{It is
+  possible that a gradient that is approximated using a complex step
+  method would be precise enough to be useful. A further investigation
+  is beyond the scope of this paper.} However, deriving a vector of
+first derivatives, and writing \proglang{R} functions to compute them,
+is a lot easier than doing the same for a matrix of second
+derivatives, and more accurate than computing second-order
+approximations from the objective function. Even when we have derived
+the Hessian symbolically, in practice it may still be faster to
+estimate the Hessian using package \pkg{sparseHessianFD} than coding
+it directly.  These are the situations in which package
+\pkg{sparseHessianFD} adds the most value to the statistician's
+toolbox.
 
@@
 defined in the \pkg{Matrix} package \citep{R_Matrix}, which are used
-extensively in \pkg{sparseHessianFD}.
+extensively in package \pkg{sparseHessianFD}.
 
@@
 component of $x\in\Real{M}$) is defined as
-\begin{align}
-  \label{eq:defParD}
-\parD{}{j}{f(x)}=\lim\limits_{\delta\to 0}\frac{f(x+\delta e_j)-f(x)}{\delta}
-\end{align}
+\begin{align*}
+\parD{}{j}{f(x)}=\lim\limits_{\delta\to 0}\frac{f(x+\delta e_j)-f(x)}{\delta}.
+\end{align*}
 For a sufficiently small $\delta$, this definition allows for a
 linear approximation to $\parD{}{j}{f(x)}$.  The derivative of $f(x)$
-is the vector of all $M$ partial derivatives.
-\begin{align}
-  \parD{}{}{f(x)}=\left(\parD{}{1}{f(x)},\mathellipsis,\parD{}{M}{f(x)}\right)
-  \end{align}
+is the vector of all $M$ partial derivatives:
+\begin{align*}
+  \parD{}{}{f(x)}=\left(\parD{}{1}{f(x)},\mathellipsis,\parD{}{M}{f(x)}\right).
+\end{align*}
 The \emph{gradient} is defined as $\nabla f(x)=\parD{}{}{f(x)}^\top$.
@@
 We define the second-order partial derivative as
-\begin{align}
-  \label{eq:14}
+\begin{align*}
   \parD{2}{jk}{}=\lim\limits_{\delta\to 0}\frac{\parD{}{j}{f(x+\delta e_k)}-\parD{}{j}{f(x)}}{\delta}
-\end{align}
+\end{align*}
 and the Hessian as
-\begin{align}
-  \label{eq:15}
+\begin{align*}
   \hess{}{f(x)}=
@@
     \parD{2}{M1}{}&  \parD{2}{M2}{}&  \mathellipsis &  \parD{2}{MM}{}
-    \end{pmatrix}
-\end{align}
-
+    \end{pmatrix}.
+\end{align*}
 The Hessian is symmetric, so $\parD{2}{ij}{}=\parD{2}{ji}{}$.
@@
 sufficiently small $\delta$, and compute
-\begin{align}
-  \label{eq:1}
-  \hess{m}{f(x)}&\approx\frac{\parD{}{}{f(x+\delta e_m)}-\parD{}{}{f(x)}}{\delta}
-\end{align}
-
+\begin{align*}
+  \hess{m}{f(x)}&\approx\frac{\parD{}{}{f(x+\delta e_m)}-\parD{}{}{f(x)}}{\delta}.
+\end{align*}
 For $M=2$, our estimate of a general $\hess{}{f(x)}$ would be
-\begin{align}
-  \label{eq:FDhess2}
+\begin{align*}
   \hess{}{f(x)}&\approx
@@
         \parD{}{2}{f(x_1+\delta, x_2)}-\parD{}{2}{f(x_1,x_2)}&  \parD{}{2}{f(x_1,x_2+\delta)}-\parD{}{2}{f(x_1,x_2)}
-    \end{pmatrix}/\delta
-\end{align}
-
+    \end{pmatrix}/\delta.
+\end{align*}
 This estimate requires three evaluations of the gradient to get
@@
 Now suppose that the Hessian is sparse, and that the off-diagonal
-elements are zero.  That means that
+elements are zero.  That means that:
 \begin{align}
-  \parD{}{1}{f(x_1,x_2+\delta)}-\parD{}{1}{f(x_1,x_2)}&=0\label{eq:I1}\\
-    \parD{}{2}{f(x_1+\delta,x_2)}-\parD{}{2}{f(x_1,x_2)}&=0\label{eq:I2}
+  \parD{}{1}{f(x_1,x_2+\delta)}-\parD{}{1}{f(x_1,x_2)}&=0,\label{eq:I1}\\
+    \parD{}{2}{f(x_1+\delta,x_2)}-\parD{}{2}{f(x_1,x_2)}&=0.\label{eq:I2}
 \end{align}
@@
         0&  \parD{}{2}{f(x_1+\delta,x_2+\delta)}-\parD{}{2}{f(x_1,x_2)}
-    \end{pmatrix}/\delta
+    \end{pmatrix}/\delta.
 \end{align}
-
 Only two gradients,
@@
 If $f(x)$ is defined over a complex domain and is holomorphic, then we
-can we can approximate $\parD{}{}{f(x)}$
-and $\hess{}{f(x)}$ at real values of $x$ using the complex step
-method. This method comes from a Taylor
+can approximate $\parD{}{}{f(x)}$ and $\hess{}{f(x)}$ at real values
+of $x$ using the complex step method. This method comes from a Taylor
 series expansion of $f(x)$ in the imaginary direction of the complex
-plane \citep{SquireTrapp1998}.  After rearranging terms, and taking the imaginary
-parts of both sides,
-\begin{align}
-  \label{eq:complexTaylor}
-  f(x+i\delta)&=f(x) + i\delta\parD{}{}{f(x)}+\mathcal{O}(\delta^2)\\
-  \parD{}{}{f(x)}&\approx\frac{\Imag(f(x+i\delta))}{\delta}
-\end{align}
-
+plane \citep{SquireTrapp1998}.  After rearranging terms, and taking
+the imaginary parts of both sides,
+\begin{align*}
+  f(x+i\delta)&=f(x) + i\delta\parD{}{}{f(x)}+\mathcal{O}(\delta^2),\\
+  \parD{}{}{f(x)}&\approx\frac{\Imag(f(x+i\delta))}{\delta}.
+\end{align*}
 Estimating a first derivative using the complex step method does not
@@
 the Hessian \citep{AbreuStich2013}.  However, when the gradient can be
-computed exactly, we can compute a first-order approximation to
+computed exactly, we can compute a first-order approximation to the
 Hessian by treating it as the Jacobian of
-a vector-valued function \citep{LaiCrassidis2008}.
-
-\begin{align}
-  \label{eq:CShess2}
+a vector-valued function \citep{LaiCrassidis2008}:
+\begin{align*}
   \hess{}{f(x)}&\approx
@@
         \Imag(\parD{}{2}{f(x_1+i\delta, x_2)})&\Imag(\parD{}{2}{f(x_1,x_2+i\delta)})
-    \end{pmatrix}/\delta
-\end{align}
-
+    \end{pmatrix}/\delta.
+\end{align*}
 If this matrix were dense, we would need two evaluations of the
@@
 Hessian. But the sparsity constraints ensure that
-the following identities hold all complex $x$.
-
+the following identities hold for all complex $x$:
 \begin{align}
-  \Imag(\parD{}{1}{f(x_1,x_2+i\delta)}) -\Imag(\parD{}{1}{f(x_1,x_2)})&=0\label{eq:I1c}\\
-    \Imag(\parD{}{2}{f(x_1+i\delta,x_2)})-\Imag(\parD{}{2}{f(x_1,x_2)})&=0\label{eq:I2c}
+  \Imag(\parD{}{1}{f(x_1,x_2+i\delta)}) -\Imag(\parD{}{1}{f(x_1,x_2)})&=0,\label{eq:I1c}\\
+    \Imag(\parD{}{2}{f(x_1+i\delta,x_2)})-\Imag(\parD{}{2}{f(x_1,x_2)})&=0.\label{eq:I2c}
 \end{align}
-
 As with the finite difference method, because Equation~\ref{eq:I1c} holds
@@
 for $x_2$, it must also hold for $x_2+i\delta$. Thus,
-
-\begin{align}
-  \label{eq:CShess2sp}
+\begin{align*}
   \hess{}{f(x)}&\approx
@@
     \end{pmatrix}/\delta
-\end{align}
-
+\end{align*}
 for real $x$.  Only one evaluation of the gradient is required.
@@
 either a difference in gradients, or the imaginary part of a
-complex-valued gradient, depending on the chosen method.
+complex-valued gradient, depending on the chosen method:
 \begin{align}
@@
            \begin{cases}
-             \nabla f(x+G_c)-\nabla f(x)&\text{finite difference method}\\
-             \Imag(\nabla f(x+iG_c))&\text{complex step method}
+             \nabla f(x+G_c)-\nabla f(x)&\text{finite difference method},\\
+             \Imag(\nabla f(x+iG_c))&\text{complex step method}.
              \end{cases}
 \end{align}
-
 If $C=M$, then $\Mat{G}$ is a diagonal matrix with $\delta$ in each
@@
 of the Hessian. Consider an example with the following values and
-sparsity pattern.
-\begin{align}
-  \label{eq:7}
+sparsity pattern:
+\begin{align*}
  \hess{}{f(x)}= \begin{pmatrix}
@@
     0&0&h_{53}&0&h_{55}
-  \end{pmatrix}
-\end{align}
-
-Suppose $C=2$, and define group membership of the five variables through the following $\Mat{G}$ matrix.
-\begin{align}
-  \label{eq:7}
+  \end{pmatrix}.
+\end{align*}
+Suppose $C=2$, and define group membership of the five variables through the following $\Mat{G}$ matrix:
+\begin{align*}
  \Mat{G}^\top= \begin{pmatrix}
@@
  0&0&\delta&\delta&0
- \end{pmatrix}
-\end{align}
+ \end{pmatrix}.
+\end{align*}
 Variables 1, 2 and 5 are in group 1, while variables 3 and 4 are in
@@
 Next, compute the columns of $\Mat{Y}$ using Equation~\ref{eq:Yg}.  We now
-have the following system of linear equations from $\hess{}{}\Mat{G}=\Mat{Y}$.
-\begin{align}
-  \label{eq:11}
+have the following system of linear equations from $\hess{}{}\Mat{G}=\Mat{Y}$:
+\begin{align*}
   \begin{aligned}[c]
@@
   \end{aligned}
-\end{align}
+\end{align*}
 Note that this system is overdetermined.  Both $h_{31}=y_{12}$ and $h_{53}=y_{52}$
@@
 algorithms for assigning decision variables to groups, and for
-extracting nonzero Hessian elements via substitution, are described in
+extracting non-zero Hessian elements via substitution, are described in
 Section~\ref{sec:algorithms}.
 
-\subsection[Sparse matrices in R]{Sparse matrices and the \pkg{Matrix}
+\subsection[Sparse matrices and the Matrix package]{Sparse matrices and the \pkg{Matrix}
   package}\label{sec:Rsparse}
@@
 precision) data, stored in
-column-compressed format, has a class \class{dtCMatrix}.
-
-\begin{table}[tbh]\centering
-  \begin{tabular}{ll|ccc}
-      \toprule
+column-compressed format, has the class \class{dtCMatrix}.
+
+\begin{table}[t!]
+  \centering
+  \begin{tabular}{llccc}
+      \hline
    Storage&Layout&\multicolumn{3}{c}{Data type}\\
     &&numeric&logical&pattern\\
-    \midrule
+    \hline
     Triplet&general&\class{dgTMatrix}&\class{lgTMatrix}&\class{ngTMatrix}\\
@@
       &symmetric&\class{dsTMatrix}&\class{lsTMatrix}&\class{nsTMatrix}\\
-    \midrule
         Row-compressed&general&\class{dgRMatrix}&\class{lgRMatrix}&\class{ngRMatrix}\\
@@
       &symmetric&\class{dsRMatrix}&\class{lsRMatrix}&\class{nsRMatrix}\\
-    \midrule
             Column-compressed&general&\class{dgCMatrix}&\class{lgCMatrix}&\class{ngCMatrix}\\
@@
       &symmetric&\class{dsCMatrix}&\class{lsCMatrix}&\class{nsCMatrix}\\
-\bottomrule
+\hline
   \end{tabular}
     \caption{Class names for sparse matrices, as defined in the
-    Matrix package.}\label{tab:MatrixClasses}
+      \pkg{Matrix} package.}\label{tab:MatrixClasses}
 \end{table}
 
-\pkg{Matrix} also defines some other classes of
+Package \pkg{Matrix} also defines some other classes of
 sparse and dense matrices that we will not discuss here. The
@@
 The distinctions among sparse matrix classes is important because
-\pkg{sparseHessianFD}'s \method{hessian} method returns a
+package \pkg{sparseHessianFD}'s \method{hessian} method returns a
 \class{dgCMatrix}, even though the Hessian is symmetric. Depending on how the
@@
 
-In this section, we demonstrate how to use the \pkg{sparseHessianFD},
+In this section, we demonstrate how to use the \pkg{sparseHessianFD} package,
 using a hierarchical binary choice model as an example.  Then, we
@@
 
-\subsection[Example: hierarchical binary choice]{Example model: hierarchical binary choice}\label{sec:example}
+\subsection[Example model: Hierarchical binary choice]{Example model: Hierarchical binary choice}\label{sec:example}
 
@@
  $\beta_i$ and $z_i$, but does not depend directly on $\beta_j$ and $z_j$ for
- another household $j\neq i$.
+ another household $j\neq i$:
 \begin{align}
-  p_i=\frac{\exp(z_i'\beta_i)}{1+\exp(z_i'\beta_i)},~i=1 ... N
+  p_i=\frac{\exp(z_i^{\top}\beta_i)}{1+\exp(z_i^{\top}\beta_i)},~i=1, \ldots, N.
 \end{align}
-
+%
 The coefficient vectors $\beta_i$ are distributed across the population of households
@@
   -\frac{1}{2}\left(\beta_i-\mu\right)^\top\Mat{\Sigma}^{-1}\left(\beta_i-\mu\right)\right)
--\frac{1}{2}\mu^\top\Mat{\Omega}^{-1}\mu
+-\frac{1}{2}\mu^\top\Mat{\Omega}^{-1}\mu.
 \end{align}
@@
 ordered in $x$. One such ordering is to group all of the
-coefficients in the $\beta_i$ for each unit together.
+coefficients in the $\beta_i$ for each unit together:
 \begin{align}
-\beta_{11},\dotsc,\beta_{1k},\beta_{21},\dotsc,\beta_{2k},~\dotsc~,\beta_{N1},\dotsc,\beta_{Nk},\mu_1,\dotsc,\mu_k
+\beta_{11},\dotsc,\beta_{1k},\beta_{21},\dotsc,\beta_{2k},~\dotsc~,\beta_{N1},\dotsc,\beta_{Nk},\mu_1,\dotsc,\mu_k.
 \end{align}
@@
 Another possibility is to group coefficients for each covariate
-together.
+together:
 \begin{align}
-\beta_{11},\dotsc,\beta_{N1},\beta_{12},\dotsc,\beta_{N2},~\dotsc~,\beta_{1k},\dotsc,\beta_{Nk},\mu_1,\dotsc,\mu_k
+\beta_{11},\dotsc,\beta_{N1},\beta_{12},\dotsc,\beta_{N2},~\dotsc~,\beta_{1k},\dotsc,\beta_{Nk},\mu_1,\dotsc,\mu_k.
 \end{align}
-Now the Hessian has an "banded" sparsity pattern, as in Figure~\ref{fig:banded}.
-\begin{figure}[tbp]
+Now the Hessian has a ``banded'' sparsity pattern, as in Figure~\ref{fig:banded}.
+\begin{figure}[t!]
   \begin{subfigure}[b]{.5\textwidth}
@@
 
-In both cases, the number of non-zeros is the same.   There are
-144 elements in this symmetric matrix, but only 64 are
-non-zero, and only 38 values are unique.  Although the reduction in
-RAM from using a sparse matrix structure for the Hessian may be
-modest, consider what would happen if $N=1,000$ instead.  In that case,
-there are 2,002 variables in the problem, and more than $4$ million
-elements in the Hessian.  However, only $12,004$ of those elements are
-non-zero.  If we work with only the lower triangle of the Hessian, then we need to work with
-only 7,003 values.
-
-The sparsity pattern required by \pkg{sparseHessianFD} consists of the row and column
-indices of the non-zero elements in the \emph{lower triangle} the
-Hessian, and it is the responsibility of the user to ensure that
-the pattern is correct.  In practice, rather than trying to keep track of the row and column indices
-directly, it might be easier to construct a pattern matrix first,
-check visually that the matrix has the right pattern, and then extract
-the indices.  The package defines utility functions (Table~\ref{tab:helpers}) to convert
+In both cases, the number of non-zeros is the same.  There are 144
+elements in this symmetric matrix, but only 64 are non-zero, and only
+38 values are unique.  Although the reduction in RAM from using a
+sparse matrix structure for the Hessian may be modest, consider what
+would happen if $N=1,000$ instead.  In that case, there are $2,002$
+variables in the problem, and more than $4$ million elements in the
+Hessian.  However, only $12,004$ of those elements are non-zero.  If
+we work only with the lower triangle of the Hessian, then we need to
+work with only $7,003$ values.
+
+The sparsity pattern required by package \pkg{sparseHessianFD}
+consists of the row and column indices of the non-zero elements in the
+\emph{lower triangle} of the Hessian, and it is the responsibility of the
+user to ensure that the pattern is correct.  In practice, rather than
+trying to keep track of the row and column indices directly, it might
+be easier to construct a pattern matrix first, check visually that the
+matrix has the right pattern, and then extract the indices.  The
+package defines utility functions (Table~\ref{tab:helpers}) to convert
 between sparse matrices, and the vectors of row and column indices
@@
 
-\begin{table}[ptb]
+\begin{table}[t!]
   \centering
@@
   \begin{tabularx}{1.0\linewidth}{lX}
-    \toprule
+    \hline
    \func{Matrix.to.Coord} & Returns a list of vectors containing row
@@
                                sparse matrix.\\
-    \func{Coord.to.Pointers}&Converts list of row and column indices
+    \func{Coord.to.Pointers}&Converts a list of row and column indices
                               (triplet format) to a list of indices
                               and pointers (compressed format).\\
-  \bottomrule
+  \hline
   \end{tabularx}
-  \caption{sparseHessianFD matrix conversion functions.}
+  \caption{`\code{sparseHessianFD}' matrix conversion functions.}
   \label{tab:helpers}
@@
 matrix, and prints the sparsity pattern of its lower triangle.
-
-\begin{minipage}{\columnwidth}
+%
 \begin{Schunk}
@@
 \end{Schunk}
-\end{minipage}
-
-\vspace{1em}
+%
 To check that a proposed sparsity pattern represents the
 intended matrix visually, use the \pkg{Matrix} \func{sparseMatrix} constructor.
-
+%
 \begin{Schunk}
 \begin{Sinput}
-R> pattern <- sparseMatrix(i=mc$rows, j=mc$cols)
+R> pattern <- sparseMatrix(i = mc$rows, j = mc$cols)
 R> printSpMatrix(pattern)
@@
 \end{Schunk}
-
-If there is uncertainty about whether an element is a structural
-   zero or not, one should err on the side of it being non-zero, and
-   include that element in the sparsity pattern.
-   There might be a loss of efficiency if the element really is a
-   structural zero, but the result will still be correct.  All that
-   would happen is that the numerical estimate for that element would be zero
-   (within machine precision). On the other hand, excluding a non-zero
-   element from the sparsity pattern will likely lead to an incorrect
-   estimate of the Hessian.
-
-\subsection[The sparseHessianFD class]{The \class{sparseHessianFD} class}
-
-The function \func{sparseHessianFD} is an initializer that returns a reference to a
-\class{sparseHessianFD} object.  The initializer determines an
-appropriate permutation and partitioning
-of the variables, and performs some additional validation tests.  The arguments to the initializer are
+%
+If there is uncertainty about whether an element is a structural zero
+or not, one should err on the side of it being non-zero, and include
+that element in the sparsity pattern.  There might be a loss of
+efficiency if the element really is a structural zero, but the result
+will still be correct.  All that would happen is that the numerical
+estimate for that element would be zero (within machine precision). On
+the other hand, excluding a non-zero element from the sparsity pattern
+will likely lead to an incorrect estimate of the Hessian.
+
+\subsection[The `sparseHessianFD' class]{The \class{sparseHessianFD}
+  class}
+
+The function \func{sparseHessianFD} in package \pkg{sparseHessianFD}
+is an initializer that returns an object of class
+\class{sparseHessianFD}.  The initializer determines an appropriate
+permutation and partitioning of the variables, and performs some
+additional validation tests.  The arguments to the initializer are
 described in Table~\ref{tab:init}.
@@
 {\setlength{\extrarowheight}{.25em}
-\begin{table}[ptb]
-\begin{tabularx}{\linewidth}{>{\bfseries}r X}
-  \toprule
+\begin{table}[t!]
+\begin{tabularx}{\linewidth}{>{\ttfamily}r X}
+  \hline
 x& A numeric vector, with length $M$ at which the object will be
   initialized and tested.\\
-fn,gr& \proglang{R} Functions that return the value of the
+fn, gr& \proglang{R} functions that return the value of the
   objective function, and its gradient. The first argument is the numeric
   variable vector.  Other named arguments can be passed to \func{fn}
-  and {gr} as well (see the \funcarg{...} argument below).\\
+  and \func{gr} as well (see the \funcarg{...} argument below).\\
 rows, cols& Sparsity pattern: integer vectors of the row and column indices of
@@
 ...& Additional arguments to be passed to \func{fn} and \func{gr}.\\
- \bottomrule
+ \hline
 \end{tabularx}
-\caption{Arguments to the sparseHessianFD initializer.}\label{tab:init}
+\caption{Arguments to the \class{sparseHessianFD} initializer.}\label{tab:init}
 \end{table}
@@
 arguments, the usage syntax to create a \class{sparseHessianFD} object is
-\begin{Schunk}
-\begin{Sinput}
+%
+\begin{Code}
 obj <- sparseHessianFD(x, fn, gr, rows, cols, ...)
-\end{Sinput}
-\end{Schunk}
+\end{Code}
+%
 where \funcarg{...} represents all other arguments that are
@@
 \func{fngrhs} method includes the Hessian as well.
-
-\begin{Schunk}
-\begin{Sinput}
-R> f <- obj$fn(x)
-R> df <- obj$gr(x)
-R> hess <- obj$hessian(x)
-R> fngr <- obj$fngr(x)
-R> fngrhs <- obj$fngrhs(x)
-\end{Sinput}
-\end{Schunk}
-
+%
+\begin{Code}
+f <- obj$fn(x)
+df <- obj$gr(x)
+hess <- obj$hessian(x)
+fngr <- obj$fngr(x)
+fngrhs <- obj$fngrhs(x)
+\end{Code}
+%
 \subsection[An example]{An example}\label{sec:exHess}
 
-Now we can estimate the Hessian for the
-log posterior density of the model from Section~\ref{sec:example}.
-For demonstration purposes, \pkg{sparseHessianFD} includes functions that compute the value
-(\func{binary.f}), the gradient (\func{binary.grad}) and the Hessian
-(\func{binary.hess}) of this model. We will treat the result from \func{binary.hess} as a
-  ``true'' value against which we will compare the numerical
-  estimates.
+Now we can estimate the Hessian for the log posterior density of the
+model from Section~\ref{sec:example}.  For demonstration purposes,
+package \pkg{sparseHessianFD} includes functions that compute the
+value (\func{binary.f}), the gradient (\func{binary.grad}) and the
+Hessian (\func{binary.hess}) of this model. We will treat the result
+from \func{binary.hess} as a ``true'' value against which we will
+compare the numerical estimates.
 
@@
 values for $\Mat{\Sigma}^{-1}$ and $\Mat{\Omega}^{-1}$, and simulate a
-vector of variables at which to evaluate the function. The \func{binary.f} and \func{binary.grad} functions take the data and
-priors as lists.  The \func{data(binary)} call adds the appropriate data
-list to the environment, but we need to construct the prior list
-ourselves.
-
+vector of variables at which to evaluate the function. The
+\func{binary.f} and \func{binary.grad} functions take the data and
+priors as lists.  The \code{data("binary", package =
+  "sparseHessianFD")} call adds the appropriate data list to the
+environment, but we need to construct the prior list ourselves.
+%
 \begin{Schunk}
@@
 R> set.seed(123)
-R> data("binary")
+R> data("binary", package = "sparseHessianFD")
 R> str(binary)
@@
 R> nvars <- as.integer(N * k + k)
-R> priors <- list(inv.Sigma = rWishart(1, k + 5, diag(k))[,,1],
+R> priors <- list(inv.Sigma = rWishart(1, k + 5, diag(k))[, , 1],
 +    inv.Omega = diag(k))
@@
 \end{Schunk}
-
+%
 This dataset represents the simulated choices for $N= 50$ customers
-over $T= 20$ purchase opportunties, where the probability of purchase
+over $T= 20$ purchase opportunities, where the probability of purchase
 is influenced by $k= 4$ covariates.
@@
 Hessian will have a block-arrow pattern.
-
+%
 \begin{Schunk}
@@
 R> order.row <- FALSE
-R> true.f <- binary.f(P, binary, priors, order.row=order.row)
-R> true.grad <- binary.grad(P, binary, priors, order.row=order.row)
-R> true.hess <- binary.hess(P, binary, priors, order.row=order.row)
+R> true.f <- binary.f(P, binary, priors, order.row = order.row)
+R> true.grad <- binary.grad(P, binary, priors, order.row = order.row)
+R> true.hess <- binary.hess(P, binary, priors, order.row = order.row)
 \end{Sinput}
 \end{Schunk}
-
+%
 The sparsity pattern of the Hessian is specified by two integer
-vectors:  one each for the row and column indices of the non-zero
-elements of the lower triangule of the Hessian.  For this example, we happen to have a
+vectors:  one for the row and and one for the column indices of the non-zero
+elements of the lower triangle of the Hessian.  For this example, we happen to have a
 matrix with the same sparsity pattern
@@
 create a block diagonal matrix, and concatenate dense rows and columns to the margins.
-
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
+%
 Finally, we create an instance of a \class{sparseHessianFD} object.
@@
 results as the true values because they are computed using the same functions.
-
+%
 \begin{Schunk}
 \begin{Sinput}
-R> obj <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
-+    rows=pattern[["rows"]], cols=pattern[["cols"]],
-+    data=binary, priors=priors, order.row=order.row)
+R> obj <- sparseHessianFD(P, fn = binary.f, gr = binary.grad,
++    rows = pattern[["rows"]], cols = pattern[["cols"]],
++    data = binary, priors = priors, order.row = order.row)
 R> f <- obj$fn(P)
@@
 \end{Schunk}
-
-The default choice of method is \funcarg{complex=FALSE}, so the
+%
+The default choice of method is \funcarg{complex = FALSE}, so the
 evaluation of the Hessian is a finite differenced approximation,
@@
 of mean relative difference.
-
+%
 \begin{Schunk}
@@
 R> hs <- obj$hessian(P)
-R> mean(abs(hs-true.hess))/mean(abs(hs))
+R> mean(abs(hs - true.hess)) / mean(abs(hs))
 \end{Sinput}
@@
 \end{Schunk}
-
-If \funcarg{complex=TRUE} in the initializer, the call to the \method{hessian} method will
-apply the complex step method. To use this method, the functions
-passed as \funcarg{fn} and \funcarg{gr} must both
-accept a complex argument, and return a complex
-result, even though we are differentiating a real-valued function. Although \pkg{base} \proglang{R} supports complex arguments
-for most basic mathematical functions, many common functions (e.g.,  \func{gamma}, \func{log1p},
-\func{expm1}, and the probability distribution functions) do not have
-complex implementations. Furthermore, the complex step method is valid
-only if the function is holomorphic. The methods in \class{sparseHessianFD} do
-\emph{not} check that this is the case for the function at hand. We convey the
-following warning from the documentation of the \pkg{numDeriv} package
+%
+If \funcarg{complex = TRUE} in the initializer, the call to the
+\method{hessian} method will apply the complex step method. To use
+this method, the functions passed as \funcarg{fn} and \funcarg{gr}
+must both accept a complex argument, and return a complex result, even
+though we are differentiating a real-valued function. Although
+\pkg{base} \proglang{R} supports complex arguments for most basic
+mathematical functions, many common functions (e.g., \func{gamma},
+\func{log1p}, \func{expm1}, and the probability distribution
+functions) do not have complex implementations. Furthermore, the
+complex step method is valid only if the function is holomorphic. The
+functions in package \pkg{sparseHessianFD} do \emph{not} check that
+this is the case for the function at hand. We convey the following
+warning from the documentation of the \pkg{numDeriv} package
 \citep{R_numDeriv}, which also implements the complex step method:
@@
 the mean relative difference from the true Hessian.
-
+%
 \begin{Schunk}
 \begin{Sinput}
-R> obj2 <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
-+    rows=pattern[["rows"]], cols=pattern[["cols"]],
-+    complex=TRUE,
-+    data=binary, priors=priors, order.row=order.row)
+R> obj2 <- sparseHessianFD(P, fn = binary.f, gr = binary.grad,
++    rows = pattern[["rows"]], cols = pattern[["cols"]], complex = TRUE,
++    data = binary, priors = priors, order.row = order.row)
 R> hs2 <- obj2$hessian(P)
-R> mean(abs(hs2-true.hess))/mean(abs(hs2))
+R> mean(abs(hs2 - true.hess)) / mean(abs(hs2))
 \end{Sinput}
 \begin{Soutput}
-[1] 8.0555e-17
+[1] 7.9673e-17
 \end{Soutput}
 \end{Schunk}
-
+%
 In short, the complex step method can be more accurate than
@@
 
-In this section, we explain how \pkg{sparseHessianFD} works. The
+In this section, we explain how package \pkg{sparseHessianFD} works. The
 algorithms are adapted from \citet{ColemanGarbow1985}, who provided
 \proglang{Fortran} implementations as \citet{ColemanGarbow1985b}.  Earlier
-versions of \pkg{sparseHessianFD} included licensed copies of the
+versions of package \pkg{sparseHessianFD} included licensed copies of the
 \citet{ColemanGarbow1985b} code, on which the current version no
@@
 
-This intersection graph, and the number of colors needed to
-color it, are not invariant to
-permutation of the rows and columns of $\hess{}{}$.  Let $\pi$
-represent such a permutation, and let $\hessLT{\pi}{}$ be the lower
-triangle of $\pi\hess{}{}{\pi^\top}$.    \Citet[Theorem
-6.1]{ColemanMore1984} show that a coloring is triangular if and only
-if it is also a proper coloring of the intersection
-graph of $\hessLT{\pi}{}$.  Furthermore,
+This intersection graph, and the number of colors needed to color it,
+are not invariant to permutation of the rows and columns of
+$\hess{}{}$.  Let $\pi$ represent such a permutation, and let
+$\hessLT{\pi}{}$ be the lower triangle of $\pi\hess{}{}\pi^{\top}$.
+\citet[][Theorem 6.1]{ColemanMore1984} show that a coloring is
+triangular if and only if it is also a proper coloring of the
+intersection graph of $\hessLT{\pi}{}$.  Furthermore,
 \citet{ColemanCai1986} prove that a partitioning is consistent with a
 substitution method if and only if it is an acyclic coloring of the
-graph of the sparsity pattern of the Hessian.   Therefore,
-finding an optimal partitioning of the variables involves finding an
-optimal combination of a permutation $\pi$, and coloring algorithm
-for the intersection graph of $\hessLT{\pi}{}$.
+graph of the sparsity pattern of the Hessian.  Therefore, finding an
+optimal partitioning of the variables involves finding an optimal
+combination of a permutation $\pi$, and coloring algorithm for the
+intersection graph of $\hessLT{\pi}{}$.
 
@@
 efficient than treating the Hessian as if it were dense.
-\begin{figure}[ptb]
+\begin{figure}[t!]
   \begin{subfigure}[b]{.32\textwidth}\centering
@@
 \begin{subfigure}[b]{.32\textwidth}\centering
-\iftoggle{tikz}{
 \begin{tikzpicture}
@@
 \end{tikzpicture}
-}{} %% end toggle
 \caption{Sparsity graph.}\label{fig:graph1pic}
@@
 \end{tikzpicture}
-}{} %% end toggle
 \caption{Intersection graph.}\label{fig:graph1int}
@@
 
-Now suppose we were to rearrange $\hess{}{}$ so the last row and
-and column were moved to the front.  In Figure~\ref{fig:graph2adj}, all columns share at least one non-zero row with
-the column for variable 7, but variable groups $\{2,4,6\}$ and
-$\{1,3,5\}$ have empty intersection sets.  The intersection graph in
-Figure~\ref{fig:graph2int} has fewer edges than Figure~\ref{fig:graph1int}, and can be colored
-with only three colors.
-\begin{figure}[ptb]
+Now suppose we were to rearrange $\hess{}{}$ so that the last row and
+column were moved to the front.  In Figure~\ref{fig:graph2adj}, all
+columns share at least one non-zero row with the column for variable
+7, but variable groups $\{2,4,6\}$ and $\{1,3,5\}$ have empty
+intersection sets.  The intersection graph in
+Figure~\ref{fig:graph2int} has fewer edges than
+Figure~\ref{fig:graph1int}, and can be colored with only three colors.
+\begin{figure}[t!]
   \begin{subfigure}[m]{.32\textwidth}\centering
@@
 \begin{subfigure}[m]{.32\textwidth}\centering
-\iftoggle{tikz}{
 \begin{tikzpicture}
@@
 \end{tikzpicture}
-}{} %% end toggle
 \caption{Sparsity graph.}\label{fig:graph2pic}
@@
 \end{tikzpicture}
-}{} %% end toggle
 \caption{Intersection graph.}\label{fig:graph2int}
@@
 
-The \class{sparseHessianFD}
-class finds a permutation, and partitions the variables, when it is
-initialized. The problem of finding a cyclic
+The \class{sparseHessianFD} class finds a permutation, and partitions
+the variables, when it is initialized. The problem of finding a cyclic
 coloring of the graph of the sparsity pattern is NP-complete
@@
 Fortunately, we just need the partitioning to be reasonably good, to
-make the effort worth our
-while.  A plethora of vertex coloring heuristics have been proposed,
-and we make no claims that any of the algorithms in \pkg{sparseHessianFD} are even
-``best available'' for all situations.
+make the effort worth our while.  A plethora of vertex coloring
+heuristics have been proposed, and we make no claims that any of the
+algorithms in package \pkg{sparseHessianFD} are even ``best
+available'' for all situations.
 
@@
 
-\begin{algorithm}
+\begin{algorithm}[t!]
   \begin{algorithmic}
@@
     \REQUIRE{$F[i],i=1,\mathellipsis,M$:  sets of ``forbidden'' colors for
-         vertex $i$ (initially empty)}
-    \REQUIRE{$U$:  set of used colors (initially empty)}
+         vertex $i$ (initially empty).}
+    \REQUIRE{$U$:  set of used colors (initially empty).}
     \REQUIRE{$C[i], i=1,\mathellipsis,M$: vector to store output of
-      assigned colors (initially all zero)}.
-    \STATE{$k\leftarrow 0$} \COMMENT{Largest color index used}
-    \STATE{Insert $0$ in $U$}
+      assigned colors (initially all zero).}
+    \STATE{$k\leftarrow 0$} \COMMENT{Largest color index used.}
+    \STATE{Insert $0$ in $U$.}
     \FOR{$i=1$ \TO $M$}
@@
         \ELSE
-           \STATE{$V\leftarrow U - F[i]$}\COMMENT{Used colors that are
-             not forbidden}
+           \STATE{$V\leftarrow U - F[i]$} \COMMENT{Used colors that are
+             not forbidden.}
            \IF{$V$ is empty}
               \STATE{$k\leftarrow k+1$}
-              \STATE{Insert $k$ into $U$}
+              \STATE{Insert $k$ into $U$.}
               \STATE{$C[i]\leftarrow k$}
@@
              \STATE{$C[i]\leftarrow\min(V)$}
-\COMMENT{Assign smallest existing non-forbidden color to $i$}
+\COMMENT{Assign smallest existing non-forbidden color to $i$.}
 \ENDIF
@@
     \FOR{$j$ in $P[i]$}
-    \STATE{Insert $C[i]$ into $F[j]$}
+    \STATE{Insert $C[i]$ into $F[j]$.}
 \COMMENT{Make $i$'s color forbidden to all of
-      its uncolored neighbors}
+      its uncolored neighbors.}
     \ENDFOR
@@
 variable $m$.  The substitution method is defined in
-\citet[Equation 6.1]{ColemanMore1984}.
-\begin{align}
-  \label{eq:subst}
-  \hess{ij}{f(x)}&=\Mat{Y}_{i,C_j}/\delta - \sum_{l>i,l\in C_j}\hess{li}{f(x)}
-\end{align}
-
+\citet[Equation 6.1]{ColemanMore1984}:
+\begin{align*}
+  \hess{ij}{f(x)}&=\Mat{Y}_{i,C_j}/\delta - \sum_{l>i,l\in C_j}\hess{li}{f(x)}.
+\end{align*}
 We implement the substitution method using Algorithm \ref{alg:subst}.
@@
     \REQUIRE{$C[i], i=1,\mathellipsis,M$: vector of
-      assigned colors}
-    \REQUIRE{$\Mat{H}$, an $M\times M$ Hessian (initialized to zero)}
-    \REQUIRE{$\Mat{B}$, a $\max(C)\times M$ matrix (initialized to zero)}
-    \REQUIRE{$\Mat{Y}$, a matrix of finite differences}
-    \REQUIRE{$\delta$, the small constant used to estimate $\Mat{Y}$}
+      assigned colors.}
+    \REQUIRE{$\Mat{H}$, an $M\times M$ Hessian (initialized to zero).}
+    \REQUIRE{$\Mat{B}$, a $\max(C)\times M$ matrix (initialized to zero).}
+    \REQUIRE{$\Mat{Y}$, a matrix of finite differences.}
+    \REQUIRE{$\delta$, the small constant used to estimate $\Mat{Y}$.}
     \FOR{$i$ = $M$ \TO $1$}
@@
 
-\subsection[Software libraries]{Software libraries}
+\subsection[Software libraries and R packages]{Software libraries and \proglang{R} packages}
 The coloring and substitution algorithms use the \pkg{Eigen} numerical
@@
 \pkg{testthat} \citep{R_testthat}, \pkg{scales} \citep{R_scales} and
-\pkg{knitr} \citep{R_knitr} packages were used for testing, and to prepare this article.
+\pkg{knitr} \citep{R_knitr} packages were used for testing, and to
+prepare this article.
 
@@
 
-As far as we know, \pkg{numDeriv} \citep{R_numDeriv} is the only other \proglang{R} package
-that computes numerical approximations to derivatives.  Like
-\pkg{sparseHessianFD} it includes functions to compute Hessians from
-user-supplied gradients (through the \func{jacobian} function), and
-implements both the finite differencing and complex step methods.  Its
-most important distinction from \pkg{sparseHessianFD} is that it treats all Hessians as dense.  Thus, we
-will use  \pkg{numDeriv} as the baseline against which we can compare the performance of
-\pkg{sparseHessianFD}.
-
-To prepare Table~\ref{tab:numDeriv}, we
-estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
-units ($N$) and within-unit parameters ($k$).  The total number of
-variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
-and standard deviations (across 500 replications) for the time (in
-milliseconds) to compute a Hessian using functions for both the finite
-difference and complex step methods from each package. Times were generated on a
-compute node running Scientific Linux 6 (64-bit) with an 8-core Intel
-Xeon X5560 processor (2.80 GHz) with 24 GB of RAM, and collected using the \pkg{microbenchmark} package
+As far as we know, package \pkg{numDeriv} \citep{R_numDeriv} is the
+only other \proglang{R} package that computes numerical approximations
+to derivatives.  Like package \pkg{sparseHessianFD} it includes
+functions to compute Hessians from user-supplied gradients (through
+the \func{jacobian} function), and implements both the finite
+differencing and complex step methods.  Its most important distinction
+from package \pkg{sparseHessianFD} is that it treats all Hessians as
+dense.  Thus, we will use package \pkg{numDeriv} as the baseline
+against which we can compare the performance of package \pkg{sparseHessianFD}.
+
+To prepare Table~\ref{tab:numDeriv}, we estimated Hessians of the log
+posterior density in Equation~\ref{eq:LPD} with different numbers of
+heterogeneous units ($N$) and within-unit parameters ($k$).  The total
+number of variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the
+mean and standard deviations (across 500 replications) for the time
+(in milliseconds) to compute a Hessian using functions for both the
+finite difference and complex step methods from each package. Times
+were generated on a compute node running Scientific Linux 6 (64-bit)
+with an 8-core Intel Xeon X5560 processor (2.80 GHz) with 24 GB of
+RAM, and collected using the \pkg{microbenchmark} package
 \citep{R_microbenchmark}. Code to replicate Table~\ref{tab:numDeriv}
 is available as an online supplement to this paper, and in the
-\func{doc/} directory of the installed package. In
+\code{doc/} directory of the installed package. In
 Table~\ref{tab:numDeriv} we see that computation times using
-\pkg{sparseHessianFD} and considerably shorter than those using \pkg{numDeriv}.
-
-\begin{table}[ptb]\centering
+package \pkg{sparseHessianFD} and considerably shorter than those using
+package \pkg{numDeriv}.
+
+\begin{table}[t!]
+  \centering
   \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }rrr|rrrr|rrrr}
-  \toprule
-&&&\multicolumn{4}{c|}{\bfseries{Finite differencing}}&\multicolumn{4}{c}{\bfseries{Complex step}}\\
+  \hline
+&&&\multicolumn{4}{c|}{{Finite differencing}}&\multicolumn{4}{c}{{Complex step}}\\
 &&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c|}{\pkg{sparseHessianFD}}&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
 $N$&$k$&$M$&mean&sd&mean&sd&mean&sd&mean&sd\\
-%%\midrule
-  \midrule
+%%\hline
+  \hline
 15 & 2 & 32 & 12.7 & 0.9 & 2.4 & 0.1 & 13.2 & 0.7 & 2.0 & 0.1 \\ 
@@
   500 & 8 & 4008 & 9533.1 & 190.9 & 54.5 & 12.8 & 14059.4 & 3797.9 & 62.5 & 21.8 \\ 
-   \bottomrule
+   \hline
 
@@
 
-To help us understand just how scalable
+To help us understand just how scalable package
 \pkg{sparseHessianFD} is, we ran another set of simulations, for the
@@
 
-\begin{table}[ptb]\centering
-  \begin{tabularx}{1.0\linewidth}{>{\bfseries}l X}
-    \toprule
+\begin{table}[t!]
+  \centering
+  \begin{tabularx}{1.0\linewidth}{l X}
+    \hline
     Measure&Description\\
-    \midrule
-Function&estimating the objective function\\
-Gradient&estimating the gradient\\
-Hessian&computing the Hessian (not including initialization or
-  partitioning time)\\
-Partitioning&finding a consistent partitioning of the
-  variables (the vertex coloring problem)\\
-Initialization&total setup time (including the partitioning time)\\
-    \bottomrule
+    \hline
+Function&Estimating the objective function.\\
+Gradient&Estimating the gradient.\\
+Hessian&Computing the Hessian (not including initialization or
+  partitioning time).\\
+Partitioning&Finding a consistent partitioning of the
+  variables (the vertex coloring problem).\\
+Initialization&Total setup time (including the partitioning time).\\
+    \hline
   \end{tabularx}
@@
 
-In the plots in Figure~\ref{fig:timing}, the number of heterogeneous units ($N$) is
-on the x-axis, and median run time, in milliseconds, is on the y-axis.  Each panel shows the
-relationship between $N$ and run time for a different step in the algorithm,
-and each curve in a panel represents a different number of within-unit
-parameters ($k$).
-
-\begin{figure}[tbp]
+In the plots in Figure~\ref{fig:timing}, the number of heterogeneous
+units ($N$) is on the $x$-axis, and median run time, in milliseconds,
+is on the $y$-axis.  Each panel shows the relationship between $N$ and
+run time for a different step in the algorithm, and each curve in a
+panel represents a different number of within-unit parameters ($k$).
+
+\begin{figure}[t!]
   \centering
@@
 
-\FloatBarrier
-\bibliography{sparseHessianFD}
+\bibliography{ref}
 

### 2021-09-23 README.Rmd
set up pkgdown

@@
   md_document:
-    variant: markdown_github
+    variant: gfm
 ---
@@
 
-# Why use the sparseHessianFD package?
+# sparseHessianFD: Because gradients are easy, but Hessians are hard
 
@@
 
-should suffice.  The package is being developed on github at
-github.com/braunm/sparseHessian, and you can get development branches
-there.
+should suffice.  The package is being developed on [Github](www.github.com/braunm/sparseHessianFD), so you can also install from
+
+```
+devtools::install_github("braunm/sparseHessianFD")
+```
 
